<!-- Methods child of paper.Rmd. Every reported value is read inline from results/; see paper/load_precomputed.R. -->

```{r methods-repro, echo=FALSE, message=FALSE, warning=FALSE}
.bo     <- read_result("desurv_bo_results_tcgacptac")
.bor    <- .bo$runs[[1]]; .bof <- .bor$fixed
.bost   <- table(.bor$history$stage)
.boh    <- .bor$history[!is.na(.bor$history$mean_cindex), ]
.ks     <- read_result("tar_k_selection_tcgacptac")
.k3     <- max(.boh$mean_cindex[round(.boh$k_grid) == .ks$k_selected])
.sf     <- read_result("desurv_seed_fits_tcgacptac")
.sfh    <- .sf$fits[[1]]$hyper
.sfn    <- vapply(.sf$fits, function(f) length(f$lossit), numeric(1))
.sftol  <- sprintf("10^{%d}", as.integer(round(log10(.sfh$tol))))
.botol  <- sprintf("10^{%d}", as.integer(round(log10(.bof$tol))))
.pe     <- read_result("consensus_endpoint_stats")
.rng2   <- function(v) sprintf("%.2f--%.2f", min(v), max(v))
.hci    <- read_result("validation_heterogeneity_ci")
.brm    <- read_result("desurv_brier_stats")$meta
.pbst   <- read_result("tar_params_best_tcgacptac")
.p0     <- read_result("tar_params_best_alpha0_tcgacptac")
.a0k3   <- read_result("tar_params_best_a0k3_tcgacptac")
.svt    <- read_result("desurv_vs_supervised_tuned")
.nword  <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten")
# Rank-by-supervision sensitivity grid at the selected non-rank hyperparameters,
# restricted to the ntop the final model uses. Two one-SE thresholds exist and are
# distinguished in the prose: this grid's own, and the Bayesian-optimization one.
.cvg    <- read.csv("results/cv_grid/cv_grid_summary.csv")
.cvg    <- .cvg[!is.na(.cvg$ntop) & .cvg$ntop == as.integer(round(.pbst$ntop)), ]
.cvgb   <- do.call(rbind, lapply(split(.cvg, .cvg$k), function(d) d[which.max(d$mean_cindex), ]))
.cvgi   <- which.max(.cvgb$mean_cindex)
.cvgthr <- .cvgb$mean_cindex[.cvgi] - .cvgb$se_cindex[.cvgi]
.cvgn   <- .cvg[.cvg$k == .ks$k_selected, ]
.cvgn   <- .cvgn[which.min(abs(.cvgn$alpha - .pbst$alpha)), ]
stopifnot(min(.cvgb$k[.cvgb$mean_cindex >= .cvgthr]) == .ks$k_selected,
          .cvgn$mean_cindex > .ks$lcb_threshold, .cvgn$mean_cindex < .cvgthr)
# Own-ntop sensitivity for the matched-rank unsupervised control (scored, like every
# other method, at the supervised model's ntop).
.ownc   <- merge(read_result("val_cindex_desurv_a0k3_tcgacptac"),
                 read_result("val_cindex_desurv_a0k3_ownntop_tcgacptac"), by = "dataset")
.ownd   <- max(abs(.ownc$cindex.x - .ownc$cindex.y))
# Single-owner guard for the one reproducibility parameter that is a cached string.
stopifnot(identical(.brm$horizon, "per-cohort 80th percentile of event times"))
```

## Problem formulation and notation
Let $X \in \mathbb{R}_{\ge 0}^{p \times n}$ denote the nonnegative gene expression matrix. DeSurv approximates $X \approx WH$, where $W \in \mathbb{R}_{\ge 0}^{p\times k}$ contains nonnegative gene programs and $H \in \mathbb{R}_{\ge 0}^{k\times n}$ contains sample loadings. Additionally, let $y \in \mathbb{R}^{n}_{>0}$ denote patient survival times, $\delta \in \{0,1\}^{n}$ the event indicators ($\delta_i = 1$ for an observed event, $\delta_i = 0$ for censoring), and $\beta \in \mathbb{R}^{k}$ the Cox regression coefficients. Survival outcomes are modeled through a Cox proportional hazards model with factor scores $Z = W^\top X \in \mathbb{R}^{k \times n}$: for sample $i$ with projected score $z_i = W^\top x_i$, the hazard is $h(t \mid x_i) = h_0(t)\exp(z_i^\top \beta)$, where $h_0$ is the baseline hazard. Defining factor scores by projection onto the learned gene programs ($Z = W^\top X$) rather than by the inferred loadings $H$ lets the same gene weights score new cohorts without refitting, the property we exploit for external validation. We use *factor* for a mathematical component and *program* for its biological interpretation, including for comparator methods. "Unsupervised" describes a factorization objective without a survival term; outcome-based tuning is stated separately.

## The DeSurv model
DeSurv integrates NMF with penalized Cox regression to identify gene programs associated with survival outcomes. The method operates in a survival-supervised framework: the reconstruction loss penalizes deviation from the observed expression data, while the survival term directs gene programs toward outcome-relevant structure. Five quantities are tuned. The rank $k$ is the number of programs; $\alpha \in [0,1)$ controls the relative contribution of the two objectives; $\lambda$ and $\xi$ are the strength and elastic-net mixing of the Cox penalty; and $n_{\text{top}}$ determines each program's top-gene set, whose union defines the scoring support. A sixth, the optional ridge penalty $\lambda_H$ on the loadings, was held at zero throughout.

The joint objective is
\begin{equation}
\label{eqn:desurv}
\mathcal{L}(W,H,\beta) =
(1-\alpha)\,\mathcal{L}_{\mathrm{NMF}}(W,H)
- \alpha\,\mathcal{L}_{\mathrm{Cox}}(W,\beta),
\end{equation}
with components
\begin{align}
\mathcal{L}_{\mathrm{NMF}}(W,H) &=
  \frac{1}{2\lVert X \rVert_F^2}\lVert X - WH \rVert_F^2
  + \frac{\lambda_H}{2nk}\lVert H \rVert_F^2, \nonumber \\
\mathcal{L}_{\mathrm{Cox}}(W,\beta) &=
  \frac{2}{n_{\mathrm{event}}}\,\ell(W,\beta)
  - \lambda\left(\xi\lVert b \rVert_1 + \tfrac{1-\xi}{2}\lVert b \rVert_2^2\right), \nonumber
\end{align}
where $\lVert \cdot \rVert_F$ is the Frobenius norm, $\ell(W,\beta)$ is the Cox log-partial likelihood, the fit criterion of the Cox model, and $n_{\mathrm{event}}$ is the number of observed deaths. The elastic-net penalty acts on the standardized-score coefficients $b = D_Z\beta$. Here $D_Z = \diag(\mathrm{sd}(Z_{1\cdot}),\ldots,\mathrm{sd}(Z_{k\cdot}))$ contains the standard deviations across samples of the factor-score rows at the current $W$. The inverse map $\beta = D_Z^{-1} b$ is applied only on the retained, nondegenerate score rows; a row whose score standard deviation falls below the numerical floor given in Supplementary Information, Optimization Algorithm, is dropped and its coefficient set to zero. This is the objective the implementation evaluates and records. The constants $1/(2\lVert X \rVert_F^2)$, $\lambda_H/(2nk)$ and $2/n_{\mathrm{event}}$ normalize the component losses before $\alpha$ sets their relative contribution (Supplementary Information, Model details).

A key design choice is where survival supervision enters the factorization. Because factor scores are defined as $Z = W^\top X$, the Cox partial likelihood $\mathcal{L}_{\mathrm{Cox}}$ is a direct function of $W$ and $\beta$, and the survival gradient acts explicitly on the gene program matrix $W$. The sample-level loadings $H$ enter only through the reconstruction term and receive no survival gradient, preserving their interpretation as mixture coefficients [@lee1999learning; @gaujoux2010flexible]; the entries of $H$ are sample loadings on the learned programs, not estimated cell-type fractions. This construction makes clinical outcome part of program estimation rather than a post hoc filter applied to an unsupervised decomposition.

When $\alpha = 0$, DeSurv reduces to its unsupervised limit. The Cox coefficients $\beta$ are then fit post hoc by penalized Coxnet on the projected scores $Z = W^\top X$ and are used only for scoring and cross-validation; they do not affect the factorization. At $\alpha = 0$, $\lambda$ and $\xi$ therefore govern the post hoc Cox fit rather than entering the factorization objective. Note that the survival-tuned $n_{\text{top}}$ still enters the consensus initialization of the final $\alpha = 0$ model (Supplementary Information, Bayesian Optimization). The Cox component also supports optional adjustment for sample-level covariates (for example, stage or grade); the analyses reported here use factor scores alone.

## Optimization
Optimization alternates updates for $H$, $W$, and $\beta$ (Supplementary Information, Optimization Algorithm). The reconstruction term is minimized using multiplicative updates that preserve nonnegativity, and the supervision term enters the proposal for $W$ as a Cox log-partial-likelihood derivative. The proposal rescales the weighted reconstruction and survival gradients using the norm-based factor defined in the Supplementary Information, subject to its numerical stabilizer and cap. It does not differentiate the $W$-dependent penalty on the standardized-score coefficients and is therefore not the gradient of the full objective; each candidate is instead evaluated under the full objective by a backtracking line search. Every candidate $W$ is first normalized to unit-Euclidean-norm columns, with $H$ and $\beta$ rescaled by the same constants, so that $WH$, the linear predictor and $b = D_Z\beta$ are all preserved; with the optional ridge penalties on $W$ and $H$ both zero ($\lambda_W = \lambda_H = 0$), as in the reported fits, the normalization alone leaves the displayed objective unchanged.

$\beta$ is then updated by Coxnet-style cyclic coordinate descent on the current factor scores $Z = W^\top X$ with elastic-net penalty parameters $(\lambda, \xi)$, starting from the current raw coefficients after the normalization step and using the current $D_Z$. The scores are centered as well as standardized, which shifts the linear predictor by a common additive constant that cancels from the Cox partial likelihood. The coordinate descent weights its quadratic approximation of the partial likelihood by $1/n_{\mathrm{event}}$ rather than the $2/n_{\mathrm{event}}$ of the objective while penalizing $b$, so within that parameterization the effective penalty is $2\lambda$. Its proposal is accepted only if the evaluated objective does not increase (for $\alpha > 0$ this is equivalent, with $W$ and $H$ fixed, to not increasing the penalized survival loss; at $\alpha = 0$ the full objective is independent of $\beta$), which makes the $\beta$ update non-increasing in that objective without establishing convergence to a conditional or joint optimum, and $\beta$ is refit on the converged $W$ at the end. The $\beta$ block is therefore an approximate rather than exact conditional minimizer. Convergence of the outer loop is assessed by the change in $W$ and $H$ between successive iterations, not by the change in the objective. Iteration stops when the cosine change of both $W$ and $H$ falls below $`r .sftol`$ for the restart and final fits, and below $`r .botol`$ for the cross-validation fits inside the hyperparameter search, or after `r format(.sfh$maxit, big.mark = ",")` iterations; cosine change is defined in Supplementary Information, Optimization Algorithm. The reported fits were produced by this implementation; the nominal tuning parameters are reported together with the implemented scaling and standardization conventions.

Because the implemented algorithm combines normalization, gradient rescaling and a Coxnet approximation, we assessed optimization behavior empirically rather than relying on a formal convergence guarantee. Across `r length(.sf$fits)` independent random initializations of the reported PDAC model, every run terminated on the stopping criterion rather than the iteration cap, and none failed numerically (median `r median(.sfn)` iterations, range `r min(.sfn)`--`r max(.sfn)`; Supplementary Information, Empirical optimization stability). The recorded outer-loop objective decreased monotonically. Monotone decrease together with the cosine stopping rule is what these runs establish, and they are checks on the optimizer rather than on which programs it recovers. Closed-form update equations are given in Supplementary Information, Model details and Optimization Algorithm, and Bayesian-optimization control parameters in Bayesian Optimization.

## Hyperparameter selection and cross-validation
Hyperparameters $(k,\alpha,\lambda,\xi,n_{\text{top}})$ were selected by maximizing the cross-validated Harrell C-index using Bayesian optimization (BO) [@snoek2012practical; @shahriari2016taking]. The objective was the mean `r .bof$nfolds`-fold cross-validated C-index, with folds constructed separately within each training cohort and stratified by event status under a fixed seed (`r .bor$seed`). Each configuration was evaluated from `r .bof$n_starts` random initializations at the cross-validation stopping tolerance $`r .botol`$, and its C-index averaged over initializations and folds. The search used an initial design of `r .bost[["initial"]]` evaluations followed by `r .bost[["bo"]]` Gaussian-process-guided evaluations; search bounds, acquisition settings and parameter transformations are given in Supplementary Information, Bayesian Optimization. The final rank $k$ was chosen by a one-standard-error rule: for each rank, the Bayesian-optimization configuration with the highest observed mean cross-validated C-index was retained, the threshold was set one fold-level standard error below the best observed configuration's mean, and the smallest rank whose retained configuration met the threshold was selected; the remaining hyperparameters were those of the best observed configuration (Supplementary Information, Bayesian Optimization). The best observed configuration had mean cross-validated C-index `r sprintf("%.3f", .ks$best_mean)` (fold-level SE `r sprintf("%.3f", .ks$best_se)`) at $k = `r .ks$k_best`$, giving a threshold of `r sprintf("%.3f", .ks$lcb_threshold)`, and the best configuration observed at $k = `r .ks$k_selected`$ reached `r sprintf("%.3f", .k3)` and was therefore eligible. All model selection was performed entirely within the training data using cross-validation; no validation cohort data were used during any stage of tuning, and out-of-sample generalization was assessed only on the external validation cohorts. The consensus initialization described next was used only for the final model and is not reproduced within each fold.

Because this rule combined $k = `r .ks$k_selected`$ with the non-rank hyperparameters of the best observed configuration, a configuration not itself evaluated in the search, we ran a sensitivity check on rank selection using the same training data and selected non-rank hyperparameters. That check is an exhaustive grid over ranks $k = 2$ to $k = 12$ and supervision strengths $\alpha$, holding $\lambda$, $\xi$ and $n_{\text{top}}$ at the retained values, and is restricted throughout to $n_{\text{top}} = `r as.integer(round(.pbst$ntop))`$. Applying the same one-standard-error rule to that grid also selects $k = `r .ks$k_selected`$. Within its $k = `r .ks$k_selected`$ slice, the nearest evaluated supervision level, $\alpha = `r .cvgn$alpha`$, gave a mean cross-validated C-index of `r sprintf("%.3f (SE %.3f)", .cvgn$mean_cindex, .cvgn$se_cindex)`. That value clears the Bayesian-optimization threshold of `r sprintf("%.3f", .ks$lcb_threshold)` but not the grid's own one-standard-error threshold of `r sprintf("%.3f", .cvgthr)`, which is set by the grid's best rank, $k = `r .cvgb$k[.cvgi]`$ (Supplementary Information, Bayesian Optimization and Cross-validated C-index across factorization rank). That grid point is a sensitivity check at a nearby $\alpha$, not the cross-validated performance of the final configuration itself.

Because the goal is prognostic factorization, DeSurv selects $k$ using the criterion that directly measures prognostic performance (cross-validated C-index) rather than the unsupervised diagnostics commonly used for NMF rank selection [@frigyesi2008nmf; @brunet2004metagenes]. In the PDAC training data, standard NMF diagnostics (reconstruction residuals, cophenetic correlation, mean silhouette width) yielded inconsistent guidance across candidate ranks, a known limitation of unsupervised rank-selection criteria that can disagree on the same dataset (Supplementary Fig.\ \ref{fig:fig-nmf-diagnostics}). On the sensitivity grid, cross-validated concordance varied little across candidate ranks, with no clear optimum anywhere between $k = 2$ and $k = 12$ and a total spread across that range of `r sprintf("%.1f", (max(.cvgb$mean_cindex) - min(.cvgb$mean_cindex)) / .cvgb$se_cindex[.cvgi])` standard errors (Supplementary Fig.\ \ref{fig:fig-cindex-by-k}). The one-standard-error rule applied to the rank-wise BO results accordingly selected the parsimonious $k = `r .ks$k_selected`$; the final model retained $\alpha = `r round(.pbst$alpha, 3)`$, $\lambda = `r sprintf("%.3f", .pbst$lambda)`$, $\xi = `r sprintf("%.3f", .pbst$nu)`$ and $n_{\text{top}} = `r as.integer(round(.pbst$ntop))`$ from the best observed configuration. Reference-program correspondence and the GATA6 comparison support biological interpretation of the selected programs; they do not establish a uniquely correct rank.

## Consensus initialization and program stability
At the selected hyperparameters, `r length(.sf$fits)` fits from independent random initializations were aggregated into a gene-by-gene co-occurrence matrix, recording how often each pair of genes appeared together among the top genes of the same program, ranked by factor specificity. Genes reaching a top set in fewer than `r ceiling(0.3 * length(.sf$fits))` of the runs were treated as inactive and excluded; the remaining genes were partitioned into $k$ groups by average-linkage hierarchical clustering, each cluster defining one column of the initialization $W_0$, with inactive and otherwise empty entries set to a small positive floor. The loadings were initialized by nonnegative least squares given $W_0$, and the initial Cox coefficients were set to zero; the final model was then fit once from that initialization. Because the consensus is built from gene-by-gene co-occurrence, it is invariant to factor-label permutation and requires no alignment of labels across runs (Supplementary Information, Consensus initialization for the final model).

We assessed program stability by partitioning the cached restart pool into disjoint halves and constructing a consensus initialization and final fit from each half. Across `r .pe$meta$n_splits` random splits, median agreement with the reported basis was `r .rng2(.pe$vs_reported$loading[, "median"])` for gene-loading correlations and `r .rng2(.pe$vs_reported$top_jaccard[, "median"])` for top-ranked-gene Jaccard overlap across the three programs. These comparisons assess sensitivity to restart-pool composition within the training data, rather than independent repetition of the full procedure (Supplementary Information, Stability of the recovered gene programs). That section also reports stability under patient subsampling, a separate and weaker property.

## Frozen gene-program and risk scoring
Scoring uses a truncated copy of the fitted basis. Within each program the genes were ranked by factor specificity (Supplementary Information, Gene Program Correlation Analysis) and the top $n_{\text{top}} = `r as.integer(round(.pbst$ntop))`$ retained; the gene support for scoring is the union of the $k$ programs' top sets, giving the truncated matrix $\widetilde{W}$ that was held fixed for every validation and treated cohort. All columns retain their weights on the same union of gene rows, and truncated columns are not renormalized. This union construction defines the primary external-validation scores, including the headline, leave-one-dataset-out and classifier-adjusted risk-score analyses, and it is the construction that cross-validation evaluated when selecting $n_{\text{top}}$. Genes outside the union still shape the factorization but do not enter the linear predictor. Other analyses use explicitly identified scoring constructions, namely full-support scores for the training-data contribution analysis and for the frozen-axis comparator analysis, and per-program masks for the tuned supervised-comparator analysis and for the Supplementary Information analyses that identify them. Both supervised-comparator constructions, the one on the full shared support and the one on per-program masks, also refit unpenalized Cox coefficients on the training scores. Treated-cohort preprocessing and missing-gene handling are described separately below.

Factor-level scores retain all $k$ columns of the truncated basis, $Z = \widetilde{W}^\top X_{\text{new}}$, and the risk score is $\eta = Z^\top \beta$. Where a single risk score is required, as in cross-validated tuning, the columns are first collapsed to $\theta = \widetilde{W}\beta$ and $\theta$ is scaled to unit $\ell_2$ norm before projection, which changes the resulting scalar risk score only by a positive multiplicative constant. The normalization applies to the collapsed score rather than to individual columns of $W$. Here, projection denotes multiplication by the fixed truncated gene-weight matrix, $\widetilde{W}^\top X_{\text{new}}$; because truncation does not renormalize the columns, these scores are not orthogonal geometric projections. During model fitting, $\beta$ was estimated from the full-support scores $W^\top X$ over all analyzed genes. For external scoring, the fitted $\beta$ was carried forward unchanged and applied to scores computed on the union support $\widetilde{W}$; $\beta$ was not refit after truncation. The elastic-net penalty set the D2 coefficient to exactly zero, so the frozen risk score is a weighted combination of D1 and D3.

## Unsupervised and supervised comparators
The matched-rank comparator shown in the figure is Lee multiplicative NMF (`NMF::nmf` with `nrun = 30`, which returns the single run of the thirty with the lowest reconstruction residual, followed by a post hoc Cox fit on its factor scores), which does not receive DeSurv's consensus initialization or hyperparameter tuning. We therefore also fit a matched-budget unsupervised control, an unsupervised factorization control with outcome-based hyperparameter tuning. At $\alpha = 0$ survival does not enter the factorization objective, but cross-validated survival performance still tunes the remaining hyperparameters and $n_{\text{top}}$. Bayesian optimization was run with $\alpha$ pinned to 0 and the rank pinned to DeSurv's selected $k$, matched to the supervised search on optimizer, search budget, gene filter, cross-validation folds, consensus procedure and stopping settings. The selected model was then fit by the same `r length(.sf$fits)`-restart consensus procedure. This control isolates removal of survival supervision at the pipeline level, with the remaining hyperparameters re-optimized rather than held numerically fixed. A second, rank-optimized control let Bayesian optimization choose the rank at $\alpha = 0$ as well, and the same one-standard-error rule applied to those results selected $k = `r .p0$k`$ (Supplementary Information, Factor structure of the rank-optimized unsupervised control). A further sensitivity analysis instead held DeSurv's own $\lambda$, $\xi$ and $n_{\text{top}}$ fixed and varied only $\alpha$. All three controls are reported in Supplementary Information, Matched-budget unsupervised control at the same rank.

The primary factorization comparison used a common number of top genes per program when constructing each model's scoring union. Total union size can still differ with rank and overlap among programs. Both unsupervised controls were therefore scored externally at the supervised model's $n_{\text{top}} = `r as.integer(round(.pbst$ntop))`$ rather than at their own tuning-selected values, which were `r as.integer(round(.a0k3$ntop))` for the matched-rank control and `r as.integer(round(.p0$ntop))` for the rank-optimized control. Scoring the matched-rank control at its own selected value instead changes no per-dataset C-index by more than `r sprintf("%.3f", .ownd)` (Supplementary Information, Matched-budget unsupervised control at the same rank).

To distinguish prognostic discrimination from biological representation, we independently fit supervised principal components (superpc; cross-validated screening threshold and component number) and lasso-penalized Cox regression (glmnet; $\lambda_{\min}$) using the same 1,970-gene training universe as DeSurv. Each method performed feature selection and tuning using the training data only, after which the fitted model was frozen and applied to rank-transformed validation cohorts without refitting.

The two comparators select features by different mechanisms, and neither matches DeSurv's. The penalized Cox comparator uses a pure lasso penalty (\texttt{glmnet} with mixing parameter 1, distinct from DeSurv's supervision parameter $\alpha$), whereas supervised principal components applies no penalty at all, instead screening genes by a cross-validated feature-score threshold and extracting a cross-validated number of components. DeSurv's internal Cox step uses an elastic net whose Bayesian-optimization-selected mixing parameter ($\xi = `r sprintf("%.3f", .pbst$nu)`$) sits close to ridge. Both comparators retain far fewer genes than DeSurv's `r .n_support`-gene support (across `r .svt$meta$nseed` resampling seeds, sparse Cox selected a median of `r .svt$genes_per_seed[["Sparse Cox"]][["median"]]` genes, range `r .svt$genes_per_seed[["Sparse Cox"]][["min"]]`--`r .svt$genes_per_seed[["Sparse Cox"]][["max"]]`, and supervised principal components `r .svt$genes_per_seed[["Supervised PCA"]][["median"]]`, range `r .svt$genes_per_seed[["Supervised PCA"]][["min"]]`--`r .svt$genes_per_seed[["Supervised PCA"]][["max"]]`), and both were tuned by their standard method-specific cross-validation rather than a budget matched to DeSurv. For this comparison DeSurv's program scores used per-program top-$n_{\text{top}}$ masks rather than the union support, and the Cox weights combining the three program scores were re-estimated without penalty on the training scores, so that all three methods enter as training-fitted linear scores. Because tuning budgets were not matched, the comparison characterizes the representations each approach recovers rather than benchmarking tuned predictive performance. A separate exploratory comparison used restricted supervised principal components, Cox partial least squares and a single sparse Cox fit (Supplementary Information, The D1 axis is not specific to DeSurv).

No matched-budget control was run for these supervised comparators, as was done for the unsupervised arm. Correspondence between each supervised risk score and the DeSurv programs D1--D3 was quantified in the training cohort using absolute Pearson correlation. Additional analyses of selected-gene enrichment and compartment attribution are described in the Supplementary Information.

## Biological correspondence and factor contributions
To characterize the biological identity of the learned factors, we quantified how strongly each factor's gene loadings aligned with independently defined PDAC tumor and microenvironmental gene programs. For each factor, we correlated its gene loadings with binary membership indicators for each reference program (Spearman correlation computed over the union of the top 50 genes of all factors of that method, intersected with the reference-program gene universe; this differs from the factor-to-factor loading correlations, which use all analyzed genes), annotating Benjamini--Hochberg-adjusted $P$ values within each factor as descriptive marks, because the gene universe is itself selected by loading. The same analysis was applied to DeSurv and to standard NMF fit at the same rank. For display, reference programs whose Spearman correlation exceeded 0.2 with at least one factor of either method are shown, as the same row set in both panels of the main figure; the same threshold defines correspondence in the coverage analysis (Supplementary Information, Reference-program coverage, edges and multiplicity).

Two per-factor contribution measures are reported alongside this correspondence. Reconstruction shares are exact Shapley values over subsets of the fixed rank-one factor contributions, computed without refitting the factorization, and each subset's value is its reconstruction gain relative to the zero-factor reconstruction; the shares are normalized to sum to 100\%. The separately reported centered $R^2$ instead takes the grand mean of the analysis matrix as its reference. Reconstruction gain and centered $R^2$ describe different aspects of reconstruction and should not be interpreted as interchangeable measures. Survival contributions are the loss in maximized unpenalized Cox partial log-likelihood between a model on all factor scores and a model refitted after omitting one factor, computed from full-support training scores; these conditional increments are nonadditive and are not changes in the trained penalized DeSurv predictor. Because the programs were estimated from the same training outcomes, both measures are descriptive rather than independent tests of prognosis (Supplementary Information, Reconstruction and Survival Contribution per Factor).

To corroborate the classical component of D1 with an independent assay, we evaluated the association between projected D1 scores and GATA6 RNA in situ hybridization measurements in the COMPASS cohort among the 33 tumors with both measurements, using Spearman correlation. Because *GATA6* is itself in the `r .n_support`-gene projection support (`r read_result("treated_cohort_stats")$gata6$n_support` of these genes were measured in the COMPASS expression data and contribute to the projected scores), part of this correlation could reflect the score being built from the assayed gene. We therefore repeated the comparison with *GATA6* removed from the projection, and the correlation was essentially unchanged (Spearman $\rho = `r sprintf("%.2f", read_result("treated_cohort_stats")$gata6$rho_excl)`$ versus $`r sprintf("%.2f", read_result("treated_cohort_stats")$gata6$rho)`$), indicating the association is not an artifact of self-correlation.

## External validation and survival analysis
External validation cohorts were scored by projecting new datasets onto the learned programs via $Z = \widetilde{W}^\top X_{\text{new}}$, and survival associations were evaluated using Harrell's C-index and hazard ratios. Analyses combining validation datasets represented each patient once and used Cox models stratified by dataset, so that each dataset carries its own baseline hazard. PurIST and DeCAF calls entered only as fixed adjustment covariates and were never inputs to the factorization or to tuning. For a fixed fitted predictor and scoring support, the manuscript uses the following reporting scalings. Primary risk-score hazard ratios, comprising the headline estimate, the covariate-adjusted estimates and the leave-one-dataset-out estimates, are expressed per standard deviation of the linear predictor over the pooled de-duplicated validation set. The per-factor forest estimates and the D1 classifier-adjustment analysis instead standardize each score within validation dataset, whereas the between-cohort slope analysis standardizes within patient cohort, so hazard ratios under these scalings are not interchangeable. Within a given cohort or dataset, centering and positive rescaling of a fixed score leave its C-index unchanged. Discrimination was summarized as the event-weighted mean of the `r .nword[nrow(.hci$per_cohort)]` cohort C-indices, so no concordance pair spans two cohorts, with percentile confidence intervals from `r format(.hci$meta$B, big.mark = ",")` bootstrap resamples of patients drawn within cohort (seed `r .hci$meta$seed`); resampling preserves cohort membership. Cohort-specific slopes were compared with a common slope by likelihood-ratio test, and proportional hazards were assessed from scaled Schoenfeld residuals. Because the association attenuated over follow-up, the pooled hazard ratio is a time-averaged summary.

Harrell's C-index is rank-based and assesses discrimination only, not calibration or time-dependent prediction error, so we additionally report a proper score as a sensitivity check. This is the integrated Brier score and its index of prediction accuracy relative to a Kaplan--Meier reference, computed by recalibrating only the frozen score's scalar slope and baseline hazard within each validation dataset and integrating to the 80th percentile of that dataset's event times. These are apparent, in-sample recalibrated values rather than out-of-sample prediction error, and because each validation dataset is integrated over its own horizon the pooled figure is a descriptive event-weighted summary (Supplementary Information, Proper-scoring sensitivity of the external validation).

Because the survival-supervised information lives in $W$ rather than $H$, DeSurv scores new samples by multiplication with the fixed gene-weight matrix, $\widetilde{W}^\top X_{\text{new}}$, without using their survival outcomes. Related supervised-NMF methods instead place the survival term on the sample side: CoxNMF maximizes a Cox partial likelihood whose covariates are the columns of $H$ [@huang2020low], and SurvNMF jointly learns the factorization and a Cox model on the sample-side latent matrix, which its own transposed notation writes as $W$ [@le2025survnmf]. Because the risk score under either formulation is a function of the sample loadings rather than of the gene-program basis, a new sample needs loadings of its own. SurvNMF documents such a step, obtaining them for held-out samples by a constrained least-squares fit; that step holds the gene basis fixed and needs neither relearning of the basis nor validation survival outcomes, so it is an additional fitting step at transfer rather than a leak. CoxNMF's published description does not specify a new-sample scoring procedure (Supplementary Table \ref{tab:architecture}). Sample-side supervision can nonetheless influence the basis indirectly, through the alternating reconstruction updates. External survival outcomes were used to estimate associations and evaluate performance, including the separately described recalibration analysis, rather than to learn or tune the primary programs or risk score. Further training and validation details appear in Supplementary Information, Part I.

## Simulation studies
```{r sim-defaults, echo=FALSE, message=FALSE, warning=FALSE}
.sim_env <- new.env(); sys.source("R/simulation_functions/scenario_defaults.R", envir = .sim_env)
.sim_def <- .sim_env$get_desurv_defaults("R0")
# The null and mixed scenarios share the prognostic scenario's baseline hazard and
# censoring rate; the prose says so, so the equality is asserted rather than typed.
stopifnot(vapply(c("R00", "R_mixed"), function(s) {
  d <- .sim_env$get_desurv_defaults(s)
  identical(d$baseline_hazard, .sim_def$baseline_hazard) && identical(d$censor_rate, .sim_def$censor_rate)
}, logical(1)))
.sim_sf <- read_result("sim_figs_by_scenario"); .sim_sid <- sapply(.sim_sf, `[[`, "scenario_id")
.sim_pd <- .sim_sf[[which(.sim_sid == "R0_easy")[1]]]$k_hist$data; .sim_n_rep <- sum(.sim_pd$method == "DeSurv")
```

Simulation studies were conducted to assess recovery of prognostic latent structure and survival prediction under controlled conditions in three scenarios: a prognostic scenario in which the factor-specific survival signal contributes little transcriptomic variation relative to the shared background that is excluded from the survival-generating gene set, a null scenario with no survival signal, and a mixed scenario where prognostic and variance-dominant signals partially overlap. Gene expression data were generated from a nonnegative factor model $X=WH$, where gene loadings $W$ comprised three gene classes: marker genes, background genes, and noise genes. Marker genes were simulated to load strongly on a single factor and weakly on others, background genes to load strongly across all factors, and noise genes to have uniformly low loadings. Specifically, marker gene primary loadings were drawn from $\text{Gamma}(\text{shape}=3, \text{rate}=0.8)$ and cross-loadings onto other factors from $\text{Gamma}(1, 20)$; background gene loadings from $\text{Gamma}(2, 1)$ across all factors; and noise gene loadings from $\text{Gamma}(1, 20)$. Sample-level factor activities $H$ were drawn from $\text{Gamma}(2, 2)$. Each dataset comprised $p = 3{,}000$ genes, $n = 200$ samples, and $k = 3$ factors, with 150 marker genes per factor and 500 shared background genes. The three factors are exchangeable in this design: the prognostic factor is not generated with lower variance than the other two. The variance-dominant signal excluded from the survival-generating gene set is the shared background block, which loads on every factor, so the factor-specific marker genes that carry survival contribute less of the total expression variation than that block. Because the background genes load on the same latent activities as the markers, their expression remains correlated with the linear predictor; "background" here denotes exclusion from the survival-generating gene set, not statistical independence of outcome. The expression matrix actually analyzed was not the noise-free product $WH$: a per-gene baseline drawn as $\mathcal{N}(3,1)$ was added to every row, independent Gaussian noise with standard deviation $`r .sim_def$noise_sd`$ was added to every entry, and negative values were clipped to zero. Cox regression coefficients were $\beta = (2, 0, 0)^\top$ in the prognostic and mixed scenarios and $\beta = 0$ in the null scenario, which otherwise used the same baseline hazard parameter and censoring rate as the prognostic scenario. The mixed scenario differs from the prognostic scenario only in its survival gene set, 300 genes per column of the survival-generating weight matrix $W^{\ast}$, comprising the 150 markers of that factor and 150 randomly sampled background genes, which creates partial overlap between the prognostic signal and the variance-dominant background.

Survival times were generated from an exponential distribution, from the noise-free expression only. Let $X_0 = WH$ denote that noise-free matrix and $W^{\ast}$ the survival-generating weight matrix, a masked copy of the column-normalized $W$ that retains, in each column, only the loadings of that factor's survival-generating genes and is zero elsewhere: the 150 markers of that factor in the prognostic and null scenarios, and 150 markers plus 150 background genes in the mixed scenario. The raw generating scores were $Z_{\text{raw}} = X_0^\top W^{\ast} \in \mathbb{R}^{n \times k}$, written here as the transpose of the model-section convention $Z = W^\top X$ so that rows index samples, and $Z_{\text{true}}$ denotes their column-wise standardization across all $n = 200$ samples, applied before the 70/30 train/test split; the linear predictor was $\eta = Z_{\text{true}} \beta$, so with $\beta = (2,0,0)^\top$ only factor 1 contributes; event times were drawn as $T \sim \text{Exponential}(\lambda_0 e^{\eta})$ with $\lambda_0 = `r .sim_def$baseline_hazard`$, and censoring times independently as $\text{Exponential}(`r .sim_def$censor_rate`)$. Each scenario comprised `r .sim_n_rep` replicates, each with a held-out test split. DeSurv incorporated survival during factorization, whereas its unsupervised limit used a post hoc Cox model on projected scores; both pipelines were tuned by Bayesian optimization using cross-validated concordance. Performance was summarized across replicates using test-set concordance and matched-factor precision of prognostic-gene recovery.

Matched-factor precision was quantified separately for each survival-driving factor. We first matched each true program to the learned factor whose top-ranked genes had the highest precision against it. Precision was then the proportion of that learned factor's top-$n_{\text{top}}$ genes belonging to the ground-truth prognostic gene set, with genes ranked by factor-specificity score and $n_{\text{top}}$ set to the Bayesian-optimization-selected signature size. Because the matching selects the factor by the precision that is then reported, matched-factor precision is optimistic by construction. We also report global marker recall, the fraction of the true survival-gene set recovered anywhere in the union of all learned factors' top-$n_{\text{top}}$ gene sets; it is insensitive to factor matching, although, like precision, it depends on each method's selected $n_{\text{top}}$ (Supplementary Information). In the primary scenario the target comprised the 150 factor-specific marker genes driving survival ($\beta = (2,0,0)$); in the mixed scenario, marker-only and full survival-gene definitions of the target were evaluated separately.

## PDAC datasets and preprocessing
We analyzed publicly available and controlled-access RNA sequencing (RNA-seq) and microarray cohorts of PDAC with corresponding overall survival outcomes. Eligible samples were primary PDAC tumors with a measured bulk expression profile and valid overall-survival annotation, subject to cohort-specific quality-control and exclusion criteria detailed in Supplementary Information, PDAC Datasets; accrual and follow-up periods are as defined in each cohort's original publication. Gene expression matrices were analyzed on a log scale: RNA-seq cohorts as log\textsubscript{2}(TPM + 1); microarray cohorts in platform-normalized log-scale intensities. Within each training cohort, genes were ranked separately in descending order of mean expression and of variance, and the 3,000 genes with the smallest sum of the two ranks were retained, selecting genes that are both highly expressed and highly variable. The intersection across training cohorts yielded 1,970 genes used for all analyses. Validation datasets were restricted to this same gene set. Survival times and censoring indicators were taken from the associated clinical annotations. No missing-data imputation was performed: all analyses used complete cases, and analyses adjusted for the PurIST and DeCAF classifiers were restricted to samples with available classifier calls. Of the seven PDAC expression datasets we considered, two were used for training (TCGA and CPTAC) and five for external validation (Dijk, Moffitt, PACA-AU array, PACA-AU RNA-seq, and Puleo); because the two PACA-AU platforms profile one patient cohort, these five validation datasets span four independent patient cohorts. To harmonize differences in scale across cohorts, filtered gene expression data were within-sample rank transformed before model training (ranks are inherently nonnegative, preserving NMF compatibility). Within-sample ranking reduces platform- and library-size differences in expression scale that would otherwise dominate cross-cohort projection. The training Cox objective was fit to the pooled TCGA and CPTAC samples without a cohort term; cross-validation folds were constructed within cohort with stratification by event status. Both training cohorts are RNA-seq, were rank transformed within sample, and had similar event rates (`r local({si <- read_result("tar_data_filtered_tcgacptac")$sampInfo; sprintf("%.0f\\%% in TCGA-PAAD and %.0f\\%% in CPTAC", 100*mean(si$event[si$dataset == "TCGA_PAAD"]), 100*mean(si$event[si$dataset == "CPTAC"]))})`; Supplementary Table \ref{tab:cohorts}); all validation analyses stratify by dataset, and a descriptive check against cohort-driven association is given in the Supplementary Information (Cohort handling in the training objective). More details about the datasets can be found in Supplementary Information, PDAC Datasets.

Per-cohort participant flow is summarized in Supplementary Table \ref{tab:cohorts}. Of 321 pooled training-cohort samples (TCGA-PAAD, $n = 181$; CPTAC-3, $n = 140$), 48 were excluded for missing survival time, missing event indicator, missing tumor grade (an inclusion criterion of the curated training set, not a model input), or quality-control failure, yielding `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)` analytic training samples (TCGA-PAAD `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$dataset=="TCGA_PAAD")`, CPTAC-3 `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$dataset=="CPTAC")`; `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)` events). Of 979 pooled validation-cohort samples, 363 were excluded for non-PDAC histology, non-tumor tissue, or missing survival annotation, yielding 616 analytic validation profiles across five platform datasets: Dijk ($n = 90$, 81 events), Moffitt GEO array ($n = 123$, 83), PACA-AU array ($n = 63$, 38), PACA-AU RNA-seq ($n = 52$, 31), and Puleo array ($n = 288$, 181).

PACA-AU included partially overlapping array and RNA-seq datasets (46 shared patients). Platform-specific analyses retained each dataset separately; for analyses combining validation datasets, duplicate patient accessions were represented once, preferentially retaining the RNA-seq profile and excluding the corresponding array profile, consistent with the preprocessing rule used in the previously published DeCAF analysis [@peng2024determination]. This yielded a combined validation population of `r n_val_total` unique patients (`r n_val_events` deaths) spanning four independent patient cohorts.

## Ethics, sex and gender, and ancestry

Each dataset was collected under the ethical approvals and consent procedures of the original study. This secondary analysis used only de-identified, previously collected data and was not submitted for separate Institutional Review Board (IRB) review at the University of North Carolina at Chapel Hill. The patient cohorts analyzed in this study were curated by the original investigators or consortia (TCGA, CPTAC, ICGC, and the original Moffitt, Dijk, and Puleo studies) and our access was limited to the de-identified expression and survival data they provide. Sex was recorded in every cohort and was included as a covariate in the clinical-adjustment analyses (Supplementary Table \ref{tab:clinical-adjustment}); cohort-specific estimates and complete-case sample sizes are reported there. We did not conduct sex-stratified analyses, because the study was not powered for subgroup comparisons within individual cohorts, and we did not stratify by race or ethnicity because those annotations were not consistently available across cohorts. Whether the prognostic relevance of these gene programs differs by sex, race, or ethnicity remains an important question for adequately powered, more comprehensively annotated cohorts.

## Treated-cohort analyses

Two independently treated PDAC cohorts were used only for projection of the frozen programs (Supplementary Information, Translational transportability). Projection used the frozen union support without any retraining: expression was rank transformed within sample over the genes each cohort shares with the trained gene-program namespace, support genes absent from a cohort were omitted from that cohort's projection rather than imputed (`r read_result("treated_cohort_stats")$gata6$n_support` of the support genes were measured in the COMPASS data), and the resulting program scores were standardized within each treated cohort before analysis. In the O'Kane/COMPASS cohort of treated metastatic PDAC ($n = `r read_result("treated_cohort_stats")$okane$n`$ patients, `r read_result("treated_cohort_stats")$okane$events` deaths; laser-capture microdissected, epithelial-enriched expression), the association of the projected D1 score with survival was reported both as arm-specific estimates and as a common slope from a Cox model stratified by treatment arm, which allows arm-specific baseline hazards, with a treatment-by-D1 interaction test reported descriptively; no estimate is pooled across cohorts. In the Linehan cohort of borderline-resectable or locally advanced disease, `r read_result("treated_cohort_stats")$paired$n` patients with paired pre- and post-treatment biopsies were compared on the projected D2 score by a paired Wilcoxon signed-rank test; this analysis concerns program level, not survival. Both analyses were exploratory and not preregistered.

## Use of large language models
During preparation of this manuscript, the authors used large language models (Claude, Anthropic; ChatGPT, OpenAI) for language editing, for assistance with code used to generate and format the display items, for verification of the written algorithm description against the released code, and for critical review of the manuscript's presentation and evidentiary framing. All model output was directed, reviewed, and verified by the authors, who take full responsibility for the work. The models were not used to design the study or to develop the DeSurv methodology, and they do not meet the criteria for authorship.
