<!-- REVISED RESULTS — prediction-validation structure (originally for PNAS; retained for Nature submission) -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- Supplementary Information figure order (si_appendix.Rmd, post-reorganization):
     S1  fig-converge          empirical optimization-stability diagnostic
     S2  fig-sim-null-mixed    null and mixed simulation scenarios
     S3  fig-cindex-by-k       training/CV C-index vs k sensitivity
     S4  fig-nmf-diagnostics   standard NMF rank selection heuristics
     S5  fig-cutpoint-km       cutpoint selection and KM curves
     S6  fig-nmf-k7-heatmap    NMF gene overlap at k=7 (fragmentation)
     S7+ fig-treated           treated-cohort transportability
     Removed in this revision: subtype-overlap, DeSurv/NMF k=5 heatmaps,
     k=5 variance-survival figure, and the per-cohort BIC table.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(magick)
library(gt)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)
desurv_maxvar_label <- paste0("D", desurv_maxvar$factor)
nmf_maxvar_label    <- paste0("N", nmf_maxvar$factor)
# Exocrine NMF factor (N2) — referenced explicitly because of its biological
# role as the high-variance exocrine-associated factor
nmf_n2              <- nmf_vs[nmf_vs$factor == 2, ]
nmf_n2_pct          <- round(nmf_n2$variance_explained * 100, 1)
nmf_n2_dl           <- round(nmf_n2$delta_loglik, 2)
# Per-factor survival contribution for the variance-versus-prognosis sentence:
# D1 carries essentially all of it despite the smallest reconstruction share.
.dl <- setNames(desurv_vs$delta_loglik, paste0("D", desurv_vs$factor))
.vp <- setNames(round(100 * desurv_vs$variance_explained, 1), paste0("D", desurv_vs$factor))

# ── Clinical / purity adjustment, per cohort ─────────────────────────────────
.capur <- local({
  ca <- read_result("desurv_clinical_adjustment")
  t(vapply(ca, function(r) c(unadj = r$unadjusted$hr,
                             pur   = if (is.null(r$purity_adjusted)) NA_real_ else r$purity_adjusted$hr),
           numeric(2)))
})

# ── Whole-matrix reconstruction R^2, centered denominator ────────────────────
# The cached helper reports an UNCENTERED R^2 on a rank matrix, where the grand
# mean alone carries ~75% of the denominator. Quote the conventional centered
# value in prose (see R/reconstruction_helpers.R and SI Table S3).
source("R/reconstruction_helpers.R")
.Xr2      <- as.matrix(read_result("tar_data_filtered_tcgacptac")$ex)
.fit_d_r2 <- read_result("tar_fit_desurv_tcgacptac")
.fit_n_r2 <- read_result("fit_std_desurvk_tcgacptac")
.r2c_desurv <- overall_recon_r2_centered(.fit_d_r2$W, .fit_d_r2$H, .Xr2)
.r2c_nmf    <- overall_recon_r2_centered(.fit_n_r2$W, .fit_n_r2$H, .Xr2)
.r2c_gap    <- 100 * (.r2c_nmf - .r2c_desurv)

# Size of the DeSurv projection support: union across factors of the
# BO-selected top genes. Used in Methods when contrasting comparator sparsity.
source("R/get_top_genes.R")
.n_support <- length(unique(unlist(get_top_genes(.fit_d_r2$W,
                    as.integer(read_result("tar_params_best_tcgacptac")$ntop))$top_genes)))

# ── Cross-method correspondence and projection R^2 ───────────────────────────
# Previously typed as literals in prose (0.84 / 0.87 / 0.36 / 0.09). Values are
# unchanged; they now come from the fits so they cannot drift out of date.
.Wd <- .fit_d_r2$W; .Wn <- .fit_n_r2$W
.corrC   <- cor(.Wd, .Wn, method = "spearman")
.corr_D2 <- max(abs(.corrC[2, ])); .corr_D3 <- max(abs(.corrC[3, ]))
.corr_D1 <- max(abs(.corrC[1, ]))
.r2_projD1 <- projection_r2(.Wd, .Wn)[1]

# Stromal-resolution accessors were removed with the count-based sentence:
# the main text now describes the pattern shown in Fig. 2a,b rather than
# quoting marker-gene counts, which read as weak in isolation. The numbers
# remain in code/23 and the SI (Stromal resolution).

# ── Tuning-budget control (code/03 sec. 5, code/04 sec. 1d) ─────────────────
# The matched-rank comparator above is NMF::nmf, which gets neither DeSurv's
# consensus initialization nor its tuning budget. To show the reorganization is
# not an artifact of that asymmetry, an unsupervised model (alpha = 0) was fit
# at the same rank using DeSurv's own optimizer, consensus initialization and
# an identical Bayesian-optimization budget. These are its correspondences.
.Wf <- read_result("tar_fit_desurv_a0k3_tcgacptac")$W
.corrF   <- cor(.Wd, .Wf, method = "spearman")
.corrF_D1 <- max(abs(.corrF[1, ]))
.corrF_D2 <- max(abs(.corrF[2, ]))
.corrF_D3 <- max(abs(.corrF[3, ]))

# Supervised risk scores versus the D2 stromal program (tuned comparators)
.ax <- read_result("desurv_vs_supervised_tuned")$axis_decomposition

# ── Orthogonal GATA6 validation (COMPASS) ────────────────────────────────────
# Values come from the cached treated-cohort statistics rather than being typed
# in. rho_excl is the self-correlation sensitivity with GATA6 removed from its
# own projection support (see code/11).

# Treated-cohort quantitative results reported in the Fig. 4 caption: the
# O'Kane/COMPASS arm-stratified common association (panel a), the paired
# post-treatment D2 shift (panel b), and the between-cohort heterogeneity of D1
# and the paired post-treatment D2 shift (panel b).
.tr    <- read_result("treated_cohort_stats")
.g6    <- .tr$gata6
.trok  <- .tr$okane$stratified
.trint <- .tr$okane$interaction_p
.trpr  <- .tr$paired

# Only two treated resources are used: the O'Kane/COMPASS arm-stratified D1
# transportability (panel a) and the Linehan paired pre/post D2 shift (panel b).
# No program-by-survival association is reported in bulk treated tissue.

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(id = rownames(surv), time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score,
               stringsAsFactors = FALSE)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df <- df[is.finite(df$time) & is.finite(df$risk_score), ]
  # Combined analysis: one record per unique patient (drop PACA-AU array
  # duplicates, keep RNA-seq), per the shared DeCAF-consistent rule.
  dedup_combined(df, dataset_col = "dataset", id_col = "id")
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))

# ── Combined per-factor associations (the diamonds in Fig. 3a) ───────────────
# The Results assert D1 is the only individual program associated with survival
# in the combined analysis; these are the numbers behind that claim. Computed
# here rather than typed, on the same de-duplicated, cohort-stratified footing
# as the pooled linear predictor above.
.pf <- local({
  rows <- lapply(val_latent_desurv_tcgacptac, function(e) {
    Z <- as.matrix(e$latent); sv <- e$survival
    data.frame(sample_id = rownames(sv), dataset = e$dataset,
               D1 = Z[, 1], D2 = Z[, 2], D3 = Z[, 3],
               time = sv$time, event = sv$event, stringsAsFactors = FALSE)
  })
  d <- do.call(rbind, rows); d <- d[is.finite(d$time) & d$time > 0, ]
  d <- dedup_combined(d, dataset_col = "dataset", id_col = "sample_id")
  out <- lapply(c("D1", "D2", "D3"), function(p) {
    d$z <- ave(d[[p]], d$dataset, FUN = function(x) as.numeric(scale(x)))
    m <- summary(coxph(Surv(time, event) ~ z + strata(dataset), data = d))
    data.frame(prog = p, hr = m$conf.int["z", 1], lo = m$conf.int["z", 3],
               hi = m$conf.int["z", 4], p = m$coefficients["z", 5])
  })
  do.call(rbind, out)
})
.pff <- function(p) { r <- .pf[.pf$prog == p, ]
  sprintf("$%.2f$ ($%.2f$--$%.2f$)", r$hr, r$lo, r$hi) }
.pfp <- function(p) { r <- .pf[.pf$prog == p, ]
  if (r$p < 0.001) "<0.001" else sprintf("%.2f", r$p) }

# ── Leave-one-cohort-out stability of the pooled hazard ────────────────
# Puleo contributes roughly half the combined validation set, so the pooled
# estimate should be shown not to rest on it. Computed here rather than typed.
# The concentration itself is stated in the Fig. 3 legend; the LOCO range below
# is what demonstrates the estimate does not depend on it.
.n_puleo   <- sum(grepl("^Puleo", pool_df$dataset))
.pct_puleo <- 100 * .n_puleo / n_val_total
.loco <- local({
  d <- pool_df
  fitone <- function(x) {
    x$rz <- x$risk_score / sd(x$risk_score)
    summary(coxph(Surv(time, event) ~ rz + strata(dataset), data = x))$conf.int["rz", "exp(coef)"]
  }
  vapply(sort(unique(d$dataset)), function(co) fitone(d[d$dataset != co, ]), numeric(1))
})
.loco_lo <- min(.loco); .loco_hi <- max(.loco)

# Build-time safeguard against pipeline drift: report the combined-cohort counts
# and flag any deviation from the expected de-duplicated totals. dedup_combined()
# already hard-asserts one record per unique patient.
message(sprintf("Unique validation patients: %d\nOS events: %d", n_val_total, n_val_events))
if (n_val_total != DESURV_EXPECTED_COMBINED_PATIENTS || n_val_events != DESURV_EXPECTED_COMBINED_EVENTS)
  warning(sprintf("Combined validation counts (%d patients, %d events) differ from expected (%d, %d): check PACA-AU de-duplication / cohort inputs.",
                  n_val_total, n_val_events, DESURV_EXPECTED_COMBINED_PATIENTS, DESURV_EXPECTED_COMBINED_EVENTS))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Pooled adjusted (PurIST + DeCAF) statistics for DeSurv ───────────────────
load_result("data_val_filtered_tcgacptac")

extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(sample_id = rownames(surv), time = surv$time, event = surv$event,
                     dataset = entry$dataset, risk_score = entry$risk_score,
                     stringsAsFactors = FALSE)
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo; si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df <- df[is.finite(df$time) & is.finite(df$risk_score), ]
  # Combined analysis: one record per unique patient, per the shared rule.
  dedup_combined(df, dataset_col = "dataset", id_col = "sample_id")
}

pool_df_adj_complete <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
pool_df_adj_complete <- pool_df_adj_complete[
  complete.cases(pool_df_adj_complete[, c("PurIST", "DeCAF")]), ]
rs_sd_adj <- sd(pool_df_adj_complete$risk_score, na.rm = TRUE)
if (is.na(rs_sd_adj) || rs_sd_adj < 1e-12) rs_sd_adj <- 1
pool_df_adj_complete$risk_z <- pool_df_adj_complete$risk_score / rs_sd_adj
cox_adj_sum <- summary(coxph(
  Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
  data = pool_df_adj_complete))
hr_adj      <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "exp(coef)"])
ci_adj_lo   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "lower .95"])
ci_adj_hi   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "upper .95"])
p_adj_val   <- cox_adj_sum$coefficients["risk_z", "Pr(>|z|)"]
p_adj_text  <- if (p_adj_val < 0.001) "< 0.001" else paste0("= ", signif(p_adj_val, 2))

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

```

## DeSurv incorporates survival information into gene-program estimation {.unnumbered}

DeSurv learns gene programs by balancing reconstruction of the tumor transcriptome with association to patient survival. Unlike the conventional sequence of unsupervised factorization followed by post hoc clinical screening, survival directly shapes the gene programs during discovery. After training, the gene weights are fixed, allowing the same programs to be scored and tested in new tumors without refitting the factorization (Fig.\ \ref{fig:schema}A; Methods). We trained DeSurv in pooled TCGA and CPTAC PDAC cohorts [@raphael2017integrated; @cao2021proteogenomic] ($n = `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)`$, `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)` events). Settings were chosen by Bayesian optimization [@snoek2012practical; @shahriari2016taking], a sequential search algorithm that proposes each new setting from the results of previous ones to optimize a pre-specified objective (Methods). Cross-validated concordance (the C-index, the probability that the model ranks a pair of patients in the correct survival order) was high across a broad region of intermediate supervision and moderate rank, declining toward the unsupervised $\alpha = 0$ boundary (Supplementary Fig.\ \ref{fig:fig-si-bo-tuning}). Concordance changed little across the ranks examined, so we applied the one-standard-error rule, which takes the most parsimonious model within one standard error of the best, giving a three-program model ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$, $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$; peak C-index $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$; Fig.\ \ref{fig:schema}B and Supplementary Fig.\ \ref{fig:fig-cindex-by-k}). By contrast, standard unsupervised NMF rank criteria (reconstruction residuals, cophenetic correlation and silhouette width [@brunet2004metagenes; @frigyesi2008nmf]) gave conflicting choices in the same data (Supplementary Fig.\ \ref{fig:fig-nmf-diagnostics}). All model and hyperparameter selection was completed within the training cohorts before the programs were applied to external validation datasets.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. Survival supervision acts on the gene-program matrix $W$ during discovery, rather than being applied only after an unsupervised factorization has been completed; following training-only model selection, the resulting gene weights are fixed for direct evaluation in external cohorts. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "model_schematic_final.pdf"))
```


## Survival supervision resolves a distinct tumor--stroma program structure {.unnumbered}

To determine how survival supervision altered the biological representation, we compared DeSurv with standard NMF at the same rank ($k = 3$), thereby holding model complexity constant. DeSurv resolved three distinct programs (D1--D3): D1, enriched for classical tumor and restCAF-like stromal signatures; D2, enriched for proCAF-like stroma; and D3, enriched for basal-like tumor programs (Fig.\ \ref{fig:pdac}a). Together, these programs define a tumor--stroma prognostic architecture spanning classical/restCAF-like and basal-like/proCAF-like states across patients. The classical identity of D1 was independently supported in the COMPASS cohort, where D1 scores correlated with GATA6 RNA in situ hybridization (Spearman $\rho = `r sprintf("%.2f", .g6$rho)`$, $P `r if (.g6$p < 0.001) "< 0.001" else sprintf("= %.3f", .g6$p)`$, $n = `r .g6$n`$; Fig.\ \ref{fig:pdac}e).

Standard NMF produced a different representation at the same rank (Fig.\ \ref{fig:pdac}b). Its malignant factor (N1) was enriched for both classical and basal-like tumor programs, its microenvironmental factor (N3) for multiple stromal and immune programs [@leary2023s; @Maurer2019], and a third factor (N2) for exocrine-associated variation. DeSurv instead separated basal-like tumor (D3) and proCAF-like stroma (D2) while recovering the distinct classical/restCAF-like survival-aligned D1 program. Both corresponded comparably to the activated-stroma and proCAF-like reference programs, the stromal distinctions bulk deconvolution conventionally resolves, but differed at the restraining end: the restCAF and iCAF reference programs correlated with a DeSurv program, and with no standard NMF factor at the same rank (Fig.\ \ref{fig:pdac}a,b).

This difference reflected a selective reorganization rather than wholesale loss of expression structure. D1 accounted for the smallest share of the full model's reconstruction gain, $`r sprintf("%.1f", .vp["D1"])`$% (Shapley shares, which sum to 100\% across the three programs [@shapley1953value]), yet carried essentially all of the survival signal (Fig.\ \ref{fig:pdac}c): omitting it lowered the Cox partial log-likelihood, the fit criterion of the Cox model, by $`r sprintf("%.1f", .dl["D1"])`$, against $`r sprintf("%.1f", .dl["D3"])`$ for D3 and $`r sprintf("%.2f", .dl["D2"])`$ for D2, even though D3 accounted for $`r sprintf("%.0f", .vp["D3"])`$% of it. Because D1 was estimated using these outcomes, its increment describes where the model placed the survival signal rather than testing whether that signal is real. The unsupervised factors are not subject to the same concern: they were fit without survival, and the largest reconstructor among them, at $`r sprintf("%.0f", max(100*nmf_vs$variance_explained))`$% of the reconstruction gain, lowered the partial log-likelihood by only $`r sprintf("%.2f", max(nmf_vs$delta_loglik))`$. The programs that best reconstruct the transcriptome need not be those most relevant to survival [@aran2015systematic].

Correlating each NMF factor's gene loadings with each DeSurv factor's, across all analysed genes, showed that the principal NMF tumor and microenvironmental factors were largely retained as D3 (Spearman $\rho = `r sprintf("%.2f", .corr_D3)`$) and D2 ($\rho = `r sprintf("%.2f", .corr_D2)`$), whereas D1 correlated only weakly with every matched-rank NMF factor ($\rho \leq `r sprintf("%.2f", .corr_D1)`$; Fig.\ \ref{fig:pdac}d). Consistently, D1 was poorly reconstructable from the full set of NMF factors ($R^2 = `r sprintf("%.2f", .r2_projD1)`$; Supplementary Table \ref{tab:projection-r2}), while reconstructing the analysis matrix somewhat less completely than unsupervised NMF ($R^2$ `r sprintf("%.2f", .r2c_desurv)` versus `r sprintf("%.2f", .r2c_nmf)`, a difference of `r sprintf("%.1f", .r2c_gap)` percentage points; Supplementary Table \ref{tab:overall-r2}). Survival supervision therefore preserved most of the variance-dominant tumor and stromal structure while resolving a distinct survival-aligned program, at a measurable cost in reconstruction. Tuning effort does not appear to explain the difference. We repeated the comparison with an unsupervised model at the same rank, given DeSurv's optimizer, its tuning budget and its consensus initialization (starting values pooled over many restarts). It recovered D2 ($\rho = `r sprintf("%.2f", .corrF_D2)`$) and D3 ($\rho = `r sprintf("%.2f", .corrF_D3)`$) as before, but recovered D1 no better than the standard implementation did ($\rho \leq `r sprintf("%.2f", .corrF_D1)`$).
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (a,b) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (a) and standard NMF (b) at $k = 3$; reference programs with $r > 0.2$ for at least one factor are shown. (c) Per-factor reconstruction contribution (normalized Shapley share, summing to 100\\%) versus survival contribution ($\\Delta$ Cox partial log-likelihood on factor removal). (d) Pairwise correspondence between NMF and DeSurv factors (Spearman correlation of W-matrix gene loadings over all genes). (e) DeSurv D1 score versus GATA6 RNA in situ hybridization level in the COMPASS cohort (Spearman correlation; $n = 33$). \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig3_tcgacptac.pdf"))
```


## Frozen survival-aligned programs retain prognostic relevance across independent untreated and treated PDAC cohorts {.unnumbered}

We next applied the frozen programs to five external validation datasets spanning four independent patient cohorts (Dijk, Moffitt, PACA-AU [array and RNA-seq], and Puleo; non-metastatic, treatment-naive; Supplementary Table \ref{tab:cohorts}). Each tumor was scored by projecting its expression onto the trained programs, restricted to their top-ranked genes ($Z = \widetilde{W}^\top X_{\text{new}}$; Methods). Combined analyses represented each patient once, giving $`r n_val_total`$ unique patients (`r n_val_events` deaths; Methods). The survival-aligned D1 program was protective in every validation dataset and was the only individual program associated with survival in the combined, cohort-stratified analysis (HR per SD `r .pff("D1")`, $P$ `r .pfp("D1")`). D3 pointed toward higher risk in four of the five datasets but was not significant in any of them, nor combined (`r .pff("D3")`, $P = `r .pfp("D3")`$), and the proCAF-like D2 program was near-null (`r .pff("D2")`, $P = `r .pfp("D2")`$) and more variable (Fig.\ \ref{fig:val}a). The frozen multivariable DeSurv linear predictor, the weighted sum of the three program scores (oriented so that higher values indicate higher risk), was associated with overall survival in a cohort-stratified analysis (pooled HR per SD $= `r hr_pool`$, 95\% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$, $P = `r p_pool`$; Table \ref{tab:hr-adjusted}). The estimate did not depend on any single dataset (Fig.\ \ref{fig:val}a). At the same complexity, unsupervised predictors were not associated with survival, whether tuned by DeSurv's own budget or not (Table \ref{tab:hr-adjusted}).

We then asked whether comparable prognostic performance could be achieved without survival-supervised deconvolution. The two methods differ most in how their performance depends on rank (Supplementary Fig. \ref{fig:fig-cindex-by-k}). DeSurv reaches essentially its full cross-validated concordance at the smallest ranks examined and is flat thereafter, so added factors buy it nothing. Unsupervised NMF starts well below DeSurv at low rank and climbs steadily as factors are added, closing most but not all of the gap by the largest rank examined; across this range its cross-validated concordance does not reach the value DeSurv attains with three factors. The two curves meet at a single rank, and there because the DeSurv curve dips rather than because NMF gains. Unsupervised factorization therefore requires many more factors to approach the same prognostic information, and at the rank where it does best in external validation those factors redundantly overlap the same established signatures rather than resolving additional biology (Supplementary Fig. \ref{fig:fig-nmf-k7-heatmap}). Survival supervision is thus not required to recover prognostic signal, but it concentrates that signal into a decomposition less than half the size (Table \ref{tab:hr-adjusted}; Supplementary Table \ref{tab:tab-cindex-comparison}).

```{r suprec-bench, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_stats")
load_result("desurv_vs_supervised_tuned")
.svb  <- function(m) sprintf("%.2f", desurv_vs_supervised_stats$pooled[m, "pooled_C"])
.svbt <- function(m) sprintf("%.2f", desurv_vs_supervised_tuned$prediction[m, "POOLED"])
```



```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC patient cohorts (five external validation datasets: Dijk, Moffitt GEO array, PACA-AU array, PACA-AU RNA-seq and Puleo; combined $n=", n_val_total, "$ unique patients, ", n_val_events, " deaths, with PACA-AU patients profiled on both platforms represented once via RNA-seq). (a) Forest plot of per-dataset univariate hazard ratios (95\\% CIs) for each factor; the platform-specific PACA-AU array and RNA-seq estimates are shown separately, and the black diamonds denote the combined estimate over unique patients (per-factor cohort-stratified Cox on the de-duplicated patient-level data; PACA-AU array duplicates excluded). Puleo is the largest contributor, supplying $", .n_puleo, "$ of the $", n_val_total, "$ unique patients (", sprintf("%.0f", .pct_puleo), "\\%), which is why the combined estimate is also reported with each dataset omitted in turn; those leave-one-dataset-out estimates range from $", sprintf("%.2f", .loco_lo), "$ to $", sprintf("%.2f", .loco_hi), "$. (b) Correspondence ($|$Spearman $r|$) of tuned supervised risk scores (penalized Cox and supervised PCA) with the three DeSurv programs (D1--D3); supervised scores concentrate on D1 and are only weakly correlated with the D2 stromal program. No validation-cohort survival outcomes were used in training, scoring or hyperparameter selection. \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig4_tcgacptac.pdf"))
```

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(knitr); library(kableExtra)
# extract_pool_df_with_covariates() and data_val_filtered_tcgacptac
# are defined in the setup chunks above.
load_result("val_latent_desurv_alpha0_tcgacptac")

pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(hr = round(ci_row[, "exp(coef)"], 2),
         ci_lo = round(ci_row[, "lower .95"], 2),
         ci_hi = round(ci_row[, "upper .95"], 2),
         p = coef_row[, "Pr(>|z|)"])
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

pool_desurv_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv_t1 <- pooled_cox_unadj_adj(pool_desurv_t1)
pool_nmf7_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7_t1 <- pooled_cox_unadj_adj(pool_nmf7_t1)
# Tuning-budget control: unsupervised at DeSurv's rank, DeSurv's optimizer,
# consensus initialization and an identical BO budget (code/03 sec. 5).
load_result("val_latent_desurv_a0k3_tcgacptac")
pool_a0k3_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_a0k3_tcgacptac, data_val_filtered_tcgacptac)
stats_a0k3_t1 <- pooled_cox_unadj_adj(pool_a0k3_t1)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)",
             "Unsupervised, matched rank ($k = 3$, $\\alpha = 0$)",
             "Unsupervised NMF ($k = 7$, $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv_t1$unadj), fmt_hr(stats_a0k3_t1$unadj),
                 fmt_hr(stats_nmf7_t1$unadj)),
  Adjusted = c(fmt_hr(stats_desurv_t1$adj), fmt_hr(stats_a0k3_t1$adj),
               fmt_hr(stats_nmf7_t1$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per SD (95\\% CI), $P$",
                       "Adjusted HR per SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = paste0("Pooled external validation hazard ratios (per SD of the linear predictor) for DeSurv at $k = 3$ and standard NMF at BO-selected $k = 7$ ($\\alpha = 0$), across five validation datasets spanning four independent patient cohorts (combined $n = ", n_val_total, "$ unique patients, ", n_val_events, " deaths). Adjusted models include the PurIST tumor-intrinsic and DeCAF stromal molecular classifiers as covariates, with stratification by dataset. Both methods retain significant prognostic value after adjustment. Both unsupervised comparators shown here are the unsupervised limit of the same optimizer, run at $\\alpha = 0$ with the identical consensus initialization and Bayesian-optimization budget given to the supervised model: one at its own BO-selected rank ($k = 7$) and one held at DeSurv's rank ($k = 3$), the latter serving as the tuning-budget control for the representational comparison in Fig.\\ \\ref{fig:pdac}. The matched-rank comparator used for the factor-structure panels of Fig.\\ \\ref{fig:pdac} is Lee multiplicative NMF, the field-standard implementation (Supplementary Table \\ref{tab:tab-cindex-comparison}). \\label{tab:hr-adjusted}")) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

```{r d1-partition-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("d1_variance_partition_stats")
.d1p <- d1_variance_partition_stats
.d1pf <- function(p) if (p < 1e-4) "< 0.0001" else sprintf("= %.4f", p)
load_result("supervised_recovery_stats")
.d1p_sr_mean  <- supervised_recovery_stats$sup_mean_score_cor
.d1p_sr_unsup <- abs(supervised_recovery_stats$table$score_cor_D1[!supervised_recovery_stats$table$supervised])
.srt2 <- supervised_recovery_stats$table
.d1p_srcor <- function(m) sprintf("%.2f", abs(.srt2$score_cor_D1[.srt2$method == m]))
```

The programs above were learned in treatment-naive, non-metastatic disease. Projected without retraining into independently treated PDAC cohorts, the survival-aligned D1 program remained associated with overall survival in treated metastatic disease under an arm-stratified analysis, and the proCAF-like D2 program increased after treatment (Fig.\ \ref{fig:treated}; Supplementary Information, Translational transportability).

```{r fig-treated, fig.cap = paste0("Transportability of the frozen DeSurv programs into independently treated PDAC cohorts. (a) Prognostic transportability of D1 in treated metastatic PDAC. Forest plot of hazard ratios per SD of the frozen D1 score within each O'Kane/COMPASS treatment arm (FOLFIRINOX, gemcitabine/nab-paclitaxel [GA], and GA/experimental); the diamond is the common association from an arm-stratified Cox model with treatment-specific baseline hazards (not a pooled-arm estimate), HR $", sprintf("%.2f", .trok["hr"]), "$ (95\\% CI $", sprintf("%.2f", .trok["lo"]), "$--$", sprintf("%.2f", .trok["hi"]), "$, $P=", sprintf("%.3f", .trok["p"]), "$). D1 was protective in all three arms with no evidence of heterogeneity by treatment arm (interaction $P=", sprintf("%.2f", .trint), "$). (b) The proCAF program D2 before and after treatment in paired Linehan biopsies ($n=", .trpr$n, "$, paired Wilcoxon $P=", sprintf("%.3f", .trpr$wilcox_p), "$); grey lines are individual patients, the red line the group mean ($", sprintf("%+.2f", .trpr$mean_delta), "$ SD). Error bars denote 95\\% confidence intervals. Because treatment, stage and cohort were not independently varied, these analyses establish prognostic and biological transportability rather than treatment-predictive effects. \\label{fig:treated}"), fig.env='figure*', fig.pos='t', fig.width=7.2, fig.height=3.4, out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(magick::image_read("figures/fig_treated_context.pdf", density = 300))
```

## The survival-aligned structure is not method-specific and is complementary to established PDAC classifiers {.unnumbered}

We next asked whether the dominant D1 direction simply recapitulated established PDAC subtype classifications. Across the pooled validation cohorts ($n = `r .d1p$meta$n`$, `r .d1p$meta$n_events` events), binary PurIST and DeCAF classifications together explained `r sprintf("%.0f", 100*.d1p$variance$explained_total)`\% of the variance in the continuous D1 score (ordinary $R^2$ from a linear model with cohort fixed effects, fitted to D1 scores already standardized within cohort), indicating substantial alignment with known tumor-intrinsic and stromal subtypes but incomplete overlap. D1 remained associated with survival after simultaneous adjustment for both classifiers (HR per SD $= `r sprintf("%.2f", .d1p$prognosis$d1_adj_hr)`$, 95\% CI $`r sprintf("%.2f", .d1p$prognosis$d1_adj_lo)`$--$`r sprintf("%.2f", .d1p$prognosis$d1_adj_hi)`$, $P `r .d1pf(.d1p$prognosis$d1_adj_p)`$); adding D1 to a stratified Cox model containing PurIST and DeCAF improved model fit (partial likelihood-ratio $\chi^2 = `r sprintf("%.1f", .d1p$prognosis$lr_chisq)`$ on 1 df, $P `r .d1pf(.d1p$prognosis$lr_p)`$) and concordance from `r sprintf("%.3f", .d1p$prognosis$c_base)` to `r sprintf("%.3f", .d1p$prognosis$c_with_d1)`. The DeSurv linear predictor also remained prognostic after adjustment for standard clinical covariates. Because D1 spans tumor and stromal programs, we asked whether it tracks tumor purity. Where purity was measured on a full cohort the association was essentially unchanged (PACA-AU array, `r sprintf("%.2f", .capur["PACA_AU_array","unadj"])` unadjusted versus `r sprintf("%.2f", .capur["PACA_AU_array","pur"])` adjusted; RNA-seq, `r sprintf("%.2f", .capur["PACA_AU_seq","unadj"])` versus `r sprintf("%.2f", .capur["PACA_AU_seq","pur"])`), arguing against a purity artifact (Supplementary Table \ref{tab:clinical-adjustment}). Thus, the survival-aligned representation overlaps established tumor and stromal classifications without being reducible to them.

The dominant D1 direction was also not specific to the DeSurv optimizer. Three independent survival-supervised approaches, in simple single-component form (supervised principal components, Cox partial least squares and sparse Cox), recovered patient scores strongly correlated with D1 in the training cohort ($|r| = `r .d1p_srcor("Supervised PCA")`$, $`r .d1p_srcor("Cox-PLS")`$ and $`r .d1p_srcor("Sparse Cox")`$, respectively), whereas the leading unsupervised principal component was only weakly correlated with D1 ($|r| = `r sprintf("%.2f", .d1p_sr_unsup)`$) and instead tracked higher-variance D2/D3 structure (Supplementary Information, The D1 axis is recovered by survival supervision, not by DeSurv specifically). Thus, D1 reflects a survival-associated expression direction shared across supervised approaches rather than an idiosyncrasy of DeSurv.

Having established that D1 is not specific to DeSurv, we asked what DeSurv adds beyond recovering it. Independently tuned supervised principal components and penalized Cox achieved pooled external concordance comparable to DeSurv (C-index $`r .svbt("Supervised PCA")`$, $`r .svbt("Sparse Cox")`$ and $`r .svbt("DeSurv full LP")`$, respectively) and captured the dominant D1-associated prognostic direction. Their selected genes, however, were consistently enriched for basal-like tumor programs but not proCAF, restCAF, immune or exocrine programs, and their risk scores were weakly correlated with the separately resolved proCAF-like D2 program (penalized Cox, $|r| = `r sprintf("%.2f", .ax["Sparse Cox", "D2"])`$; supervised PCA, $|r| = `r sprintf("%.2f", .ax["Supervised PCA", "D2"])`$; Fig. \ref{fig:val}b). Although these scores retained modest stromal signal, they did not resolve it as a distinct stromal program (Supplementary Information). DeSurv instead places that direction within a fixed set of tumor and stromal programs that can be measured and interpreted separately, and applied unchanged to new cohorts.


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

We used simulations with known latent structure to characterize when survival supervision alters program recovery. In the primary scenario, the gene program driving survival contributed less transcriptomic variation than outcome-neutral programs ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies), so that the strongest sources of expression variation were not the most prognostically relevant. DeSurv and standard NMF ($\alpha = 0$) were tuned using the same cross-validation procedure, isolating the effect of survival supervision during factorization.

Under this setting, DeSurv more accurately recovered the genes defining the true prognostic program, whereas standard NMF largely followed variance-dominant structure (Fig.\ \ref{fig:sim}a--b). This distinction is important because the genes assigned to a recovered program, not only its ability to rank patients by risk, determine its subsequent biological interpretation and its potential use in candidate signatures or biomarker development. Across 100 replicates, median test-set C-index was $`r desurv_sim_median`$ for DeSurv versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$), and DeSurv recovered the true factorization rank ($k = 3$) more frequently (Fig.\ \ref{fig:sim}c).

The advantage depended on the relationship between transcriptomic variance and prognosis. It attenuated when prognostic and variance-dominant structure partially overlapped and disappeared under the null, where neither method identified prognostic signal (Supplementary Fig.\ \ref{fig:fig-sim-null-mixed}). In the mixed setting, the advantage remained strongest for recovery of the factor-specific marker genes rather than shared background survival genes, supporting a specific effect of supervision on identifying the genes that define the prognostic program.

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Survival supervision improves recovery of prognostic gene programs when variance and prognosis diverge. (a) Test-set concordance across 100 simulation replicates comparing DeSurv with standard NMF ($\\alpha = 0$; $p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; paired Wilcoxon $P ", sim_wilcox_p, "$). (b) Recovery of genes defining the true prognostic program, quantified as the proportion of selected genes in the best-matching learned program that belonged to the ground-truth prognostic gene set. In (a) and (b), boxes show median and interquartile range; whiskers extend to $1.5\\times$IQR; points are individual replicates. (c) Distribution of selected ranks across replicates for DeSurv versus standard NMF. (d) The same gene-program recovery metric as in (b), computed for two signal regimes: a prognostic program that explains low transcriptomic variance (separated from the dominant variance) versus one that partially overlaps the dominant variance. DeSurv recovers the prognostic program in both regimes, whereas standard NMF, which follows variance, does not; the null scenario (no prognostic program) is shown in the Supplementary Information. These scenarios are constructed to isolate the regime in which the prognostic program carries low transcriptomic variance, and therefore quantify recovery under that construction; the concordance gap in (a) is not an expected discrimination gain in real data, where the two approaches perform comparably (Table \\ref{tab:hr-adjusted}). \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

knitr::include_graphics(file.path("..", "figures", "fig2_tcgacptac.pdf"))

```
