<!-- REVISED RESULTS — prediction-validation structure for PNAS -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- SI Appendix figure numbering (supplement.Rmd order, post-reorganization):
     S1  fig-converge        convergence trajectories
     S2  fig-sim-null-mixed  null and mixed simulation scenarios
     S3  fig-cindex-by-k     C-index vs K sensitivity
     S4  fig-nmf-diagnostics standard NMF rank selection heuristics (demoted from old Fig 2A-C)
     S5  fig-k3-k7-nesting   K=3 vs K=7 factor correspondence
     S6  fig-cutpoint-km     cutpoint selection and KM curves
     S7  fig-subtype-overlap subtype composition of risk groups
     S8  fig-nmf-k5-heatmap  NMF gene overlap at k=5
     S9  fig-nmf-k7-heatmap  NMF gene overlap at k=7
     S10 fig-varsurvival-k5  variance-survival at k=5
     Note: Former S7 (variance-survival), S8 (W-corr), S9 (forest) are now
           main text Fig 3C-D and Fig 4A respectively.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
# library(enrichplot)  # not used in chunks; figures come from cached .rds
library(magick)
# library(Seurat)      # not used in chunks; figures come from cached .rds
library(gt)
# library(VAM)         # not used in chunks; figures come from cached .rds
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

# ── KM dichotomized risk-group statistics ─────────────────────────────────────
# Extract HR/CI/p from the annotation layer embedded in the KM plot objects
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# Extract exact risk-group counts from KM risk tables (time = 0)
km_n_high_desurv <- fig_median_survival_desurv_tcgacptac$data.survtable$n.risk[
  fig_median_survival_desurv_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_desurv_tcgacptac$data.survtable$strata == "factor=1"]
km_n_high_nmf <- fig_median_survival_std_desurvk_tcgacptac$data.survtable$n.risk[
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$strata == "factor=1"]
km_reclass_diff <- km_n_high_desurv - km_n_high_nmf
```

## Model Overview {.unnumbered}

We developed DeSurv, a survival-supervised deconvolution framework (Fig. \ref{fig:schema}; Methods), and evaluated it in controlled simulations and PDAC cohort analysis with external validation.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..","figures", "model_schematic_final.pdf"))
```


## Survival supervision clarifies NMF rank selection {.unnumbered}

We first asked whether incorporating survival outcomes resolves the instability of unsupervised rank selection heuristics [@frigyesi2008nmf; @brunet2004metagenes]. Using gene expression data from the PDAC training cohorts (The Cancer Genome Atlas (TCGA) [@raphael2017integrated] and the Clinical Proteomic Tumor Analysis Consortium (CPTAC) [@ellis2013clinical]; Materials and Methods), which consist of non-metastatic, treatment-naive specimens (SI Appendix, Table S1), we evaluated commonly used unsupervised criteria across a range of candidate ranks.

Standard NMF diagnostics (reconstruction residuals, cophenetic correlation, mean silhouette width) yielded inconsistent guidance across candidate ranks --- a known limitation of unsupervised rank-selection criteria, which can disagree on the same dataset [@brunet2004metagenes; @frigyesi2008nmf] (SI Appendix, Fig. S4).

Because the goal is prognostic factorization, DeSurv selects $k$ using the criterion that directly measures prognostic performance: cross-validated concordance index (C-index), optimized jointly with supervision strength $\alpha$ via Bayesian Optimization (BO) [@snoek2012practical; @shahriari2016taking]. The resulting C-index surface identified a coherent region of high performance --- a broad plateau spanning $k = 3$ through $k = 12$ along DeSurv's supervised dimension (SI Appendix, Fig. S3), indicating that three supervised factors capture nearly as much prognostic information as twelve. Model selection followed the one-standard-error rule (Methods): we selected the smallest $k$ whose predicted performance lay within one standard error of the maximum, yielding a parsimonious choice. Applying the 1-SE rule to the C-index surface selected $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ and $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$ (peak C-index $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$; $n = `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)`$ patients, $`r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)`$ events). Survival-supervised model selection thus provides a more informed criterion for rank choice than unsupervised heuristics, yielding a single parsimonious factorization for downstream evaluation.

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Simulation studies and model selection. (A) Test-set concordance index across 100 simulation replicates ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$) comparing DeSurv to standard NMF ($\\alpha = 0$); paired Wilcoxon $P ", sim_wilcox_p, "$. (B) Precision (fraction of a factor's top-weighted genes --- defined in main text --- overlapping a true prognostic program) across replicates. In (A) and (B), boxes show median and interquartile range (IQR); whiskers extend to $1.5\\times$IQR; points beyond are individual replicates. The y-axis in (A) is truncated at 0.5 because C-index values below 0.5 indicate worse-than-random concordance. (C) Distribution of selected $k$ values across the same 100 replicates for DeSurv versus standard NMF. (D) Gaussian-process--predicted mean cross-validated C-index from Bayesian optimization over factorization rank ($k$) and supervision strength ($\\alpha$) in PDAC training data (TCGA + CPTAC). \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

load_result("fig_bo_heat_tcgacptac")
fig_bo_heat_tcgacptac = fig_bo_heat_tcgacptac +
  guides(fill = guide_colorbar(barheight = unit(2, "cm"), barwidth = unit(0.3, "cm"))) +
  theme(legend.title = element_text(size = 7), legend.text = element_text(size = 7))
leg_grob = ggplotGrob(fig_bo_heat_tcgacptac)
guide_idx <- which(leg_grob$layout$name == "guide-box-right")
heat_legend <- leg_grob$grobs[[guide_idx]]

fig_bo_heat_tcgacptac = fig_bo_heat_tcgacptac + theme(legend.position = "none")
upper = plot_grid(
  alt_plots$cindex_box + labs(title = NULL) + scale_y_continuous(limits = c(.5, 1)),
  alt_plots$precision_box + labs(title = NULL),
  NULL,
  ncol = 3, labels = c("A", "B",""),
  rel_widths = c(4,4,.8)
)

lower = plot_grid(
  alt_plots$k_hist,
  fig_bo_heat_tcgacptac,
  heat_legend,
  ncol = 3, labels = c("C", "D",""),
  rel_widths = c(4,4,.8)
)

plot_grid(upper, lower, nrow = 2)

```


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

To test whether survival supervision improves gene program recovery under nonnegative constraints, we designed simulations with known ground-truth latent structure. Simulated expression matrices were generated from a nonnegative factor model in which prognostic gene programs explained low variance relative to outcome-neutral background signals, with survival times generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies). To test whether survival information should enter during factorization rather than only afterward, we compared DeSurv to standard NMF ($\alpha=0$), where factors are learned by reconstruction error alone and survival associations are estimated by a separately fit Cox model. Both methods were tuned via BO over cross-validated C-index, so that differences in performance reflect the role of supervision during factorization, not during model selection.

In the primary scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index (Fig. \ref{fig:sim}A) and substantially higher precision for recovering the true prognostic gene programs (Fig. \ref{fig:sim}B). Here precision denotes the fraction of a factor's top-weighted genes --- those with the largest factor specificity score $s_{ij}$, with $n_{\text{top}}$ --- the number of factor-specific genes entering the survival linear predictor --- chosen by BO (SI Appendix, Sections 6 and 9) --- that overlap with a true prognostic program's gene set. The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This confirms that outcome supervision during factorization is necessary to recover prognostic programs that do not dominate expression variance. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$). The smaller C-index gap reflects what each metric measures: NMF factor scores aggregate many genes, so partial marker recovery at moderate loadings preserves moderate within-cohort C-index even when top-weighted positions are dominated by outcome-neutral background --- a gene-attribution failure that precision exposes. Cross-cohort transportability depends on the biological stability of the factors themselves and is assessed in the PDAC external validation analyses.

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under a null scenario ($\beta = 0$) and a mixed scenario with partial overlap between variance-dominant and prognostic structure (SI Appendix, Fig. S2). DeSurv's benefit scaled with the degree to which prognostic programs were masked by outcome-neutral variation: largest in the primary scenario ($\Delta = `r delta_sim_cindex`$), attenuated when variance and prognosis partially overlapped ($\Delta = `r delta_mixed_cindex`$, paired Wilcoxon $P `r mixed_wilcox_p`$), and absent when no survival signal existed, where both methods yielded C-index near 0.5. DeSurv also recovered the true rank ($k = 3$) more frequently than standard NMF, which systematically under-selected near $k = 2$ (Fig. \ref{fig:sim}C).




## Survival supervision and standard NMF produce different factor structures in PDAC {.unnumbered}

BO selected $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ and $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$ for DeSurv in the PDAC training cohorts (TCGA and CPTAC). To compare the factor structures produced with and without survival supervision at the same dimensionality, we fit standard NMF at the same rank ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$) and examined the overlap between each method's factor-specific gene rankings and established PDAC gene programs (Fig. \ref{fig:pdac}A--B). Matching $k$ across methods isolates the effect of survival supervision on factor content from differences in model complexity; standard NMF's own rank-selection diagnostics did not converge on a clear alternative (SI Appendix, Fig.\ S4), and a sensitivity analysis (SI Appendix, Table S2) evaluates NMF performance across a range of ranks and supervision strengths. We first describe the biological identity of each method's factors, then ask whether these structural differences correspond to differences in survival contribution.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. In the DeSurv factorization, D1 correlated with both Classical tumor programs and proCAF-associated stromal signatures, D2 with restCAF-associated programs, and D3 with Basal-like tumor programs (Fig. \ref{fig:pdac}A). The co-occurrence of Classical and proCAF signatures within a single factor is consistent with evidence that survival stratification in PDAC improves when tumor-intrinsic and CAF subtypes are considered jointly [@peng2024determination; @elyada2019cross]. No DeSurv factor showed a significant positive correlation with exocrine-associated gene programs (DECODER Exocrine, Collisson Exocrine, Bailey aberrantly differentiated endocrine--exocrine (ADEX); Fig. \ref{fig:pdac}A). The survival contributions of these factors and their relationship to expression variance are quantified below.

Standard NMF produced a different factor organization (Fig. \ref{fig:pdac}B). One factor (N2) was strongly correlated with exocrine-associated gene programs and negatively correlated with immune and stromal signatures (Fig. \ref{fig:pdac}B). A second factor (N1) correlated with Classical tumor identity (Fig. \ref{fig:pdac}B). A third factor (N3) correlated with immune, fibroblast, and extracellular matrix (ECM)-related programs (including Puleo Desmoplastic, SCISSORS proCAF and restCAF [@leary2023s], Maurer Immune and ECM [@Maurer2019], and DECODER Activated Stroma), aggregating these into a single factor rather than separating them as DeSurv does (D1 and D2; Fig. \ref{fig:pdac}A). At $k = 3$, NMF thus devotes one of its three factors to exocrine-associated variation, leaving two factors to capture all tumor-intrinsic and microenvironmental programs. Whether these structural differences correspond to differences in prognostic information is quantified next.

We quantified per-factor contribution using two metrics (Fig. \ref{fig:pdac}C; SI Appendix, Sections 10--11). For each factor, we computed the conditional variance explained (semi-partial $R^2$: the incremental fraction of total variance accounted for by that factor given all other factors), and its survival contribution, measured as the change in Cox partial log-likelihood ($\Delta\ell$) when that factor is dropped from a $k$-factor Cox model. All three NMF factors had negligible $\Delta\ell$ despite spanning a range of variance, while DeSurv concentrated nearly all survival contribution into D1. Specifically, D1 explained $`r desurv_maxsurv_pct`$% of variance with $\Delta\ell = `r desurv_maxsurv_dl`$, while the highest-variance NMF factor explained $`r nmf_maxvar_pct`$% of expression variance but contributed only $\Delta\ell = `r nmf_maxvar_dl`$ to survival. The highest-variance DeSurv factor (D3, $`r desurv_maxvar_pct`$% of variance) also contributed minimally to survival after accounting for D1. Thus, in PDAC the highest-variance factors are not the most prognostic, consistent with tumor purity analyses across cancer types [@aran2015systematic].

To further characterize how the two factorizations relate, we examined pairwise W-matrix gene loading correlations restricted to the top 50 genes per factor (Fig. \ref{fig:pdac}D). NMF's tumor-identity factor (N1, top-correlated with Classical programs) was anti-correlated with DeSurv's tumor factor (D3, top-correlated with Basal-like programs), reflecting genuine subtype discordance between Classical and Basal-like gene programs in the top-weighted genes of each method. NMF's microenvironmental factor (N3) mapped primarily to DeSurv's D2. Notably, DeSurv's most prognostic factor (D1) showed only weak, roughly uniform correlations with all three NMF factors, indicating that the gene program structure underlying D1 does not have a direct counterpart in the unsupervised factorization. Because D1 couples Classical tumor and proCAF-associated stromal signatures into a single survival-aligned program, its signal is distributed across multiple NMF factors, none of which individually captures the same tumor--stroma coupling. NMF's exocrine-associated factor (N2) similarly lacked a strong DeSurv counterpart, with its strongest (though weak) correlation to D3, consistent with D3's high variance and low survival contribution. Together, the two methods recover overlapping tumor and microenvironmental structure but differ in two key respects: NMF dedicates a factor to exocrine-compositional variation that DeSurv suppresses, while DeSurv constructs a prognostic tumor--stroma factor that NMF does not isolate. The net result is that DeSurv concentrates survival contribution into a single factor ($\Delta\ell = `r desurv_maxsurv_dl`$), whereas no NMF factor achieves comparable survival contribution at the same rank. Whether this translates to improved generalization across independent cohorts is tested below.
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (A--B) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$. Only $|r| > 0.2$ shown. DeSurv concentrates survival signal into D1 (Classical tumor + proCAF stroma) while suppressing exocrine variation; standard NMF devotes one factor to exocrine signals. (C) Fraction of expression variance explained versus survival contribution ($\\Delta$ Cox partial log-likelihood upon factor removal) for each factor. DeSurv concentrates survival signal into D1 while the highest-variance NMF factor (N2, exocrine-associated) contributes negligibly to survival. (D) Pairwise correspondence between NMF and DeSurv factors, measured by Spearman rank correlation of W-matrix gene loadings restricted to the top 50 genes per factor. \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.height= "7in", out.width='\\textwidth'}

library(cowplot)
library(ggplot2)

# Panels A-B: gene program heatmaps
load_result("fig_gene_overlap_heatmap_desurv_tcgacptac")
load_result("fig_gene_overlap_heatmap_std_desurvk_tcgacptac")

# Panel C: variance vs survival (promoted from supplement)
load_result("fig_variation_explained_tcgacptac")

# Panel D: W-matrix correlation (promoted from supplement)
load_result("fig_desurv_std_correlation_tcgacptac_top50")

# ── Panels A-B: heatmaps with shared legend ──
# Add top margin so panel labels (A, B) don't overlap heatmap titles
plot_3a <- fig_gene_overlap_heatmap_desurv_tcgacptac$plot +
  theme(plot.margin = margin(t = 14, r = 2, b = 2, l = 2))
plot_3b <- fig_gene_overlap_heatmap_std_desurvk_tcgacptac$plot +
  theme(plot.margin = margin(t = 14, r = 2, b = 2, l = 2))
legend_ab <- fig_gene_overlap_heatmap_desurv_tcgacptac$legend
legend_ab <- gtable::gtable_add_padding(
  legend_ab, padding = unit(c(0, 10, 0, 0), "pt")
)

top_row <- plot_grid(
  plot_3a, plot_3b, cowplot::ggdraw(legend_ab),
  ncol = 3, labels = c("A", "B", ""), align = "hv",
  label_size = 12, rel_widths = c(3.5, 3.5, 0.3)
)

# ── Panel C: variance vs survival scatterplot ──
plot_3c <- fig_variation_explained_tcgacptac +
  theme(
    legend.position    = c(1, 0.5),
    legend.justification = c(1, 0),
    legend.background  = element_rect(color = "black"),
    axis.title         = element_text(size = 8),
    plot.margin        = margin(2, 30, 2, 2)
  )

# ── Panel D: W-matrix correlation ──
legend_d <- ggplot(
  data.frame(x = 0, y = seq(-0.5, 1, length.out = 100)),
  aes(x = x, y = y, fill = y)
) +
  geom_tile() +
  scale_fill_gradientn(
    colors = grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100),
    limits = c(-0.5, 1),
    breaks = c(-0.4, 0, 0.4, 0.8),
    name = "Spearman\ncorrelation"
  ) +
  guides(fill = guide_colorbar(
    barwidth  = unit(0.3, "cm"),
    barheight = unit(2, "cm"),
    title.position = "top",
    title.hjust = 0.5
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 6),
    legend.text  = element_text(size = 6)
  )
legend_d_grob <- cowplot::get_legend(legend_d)

plot_3d <- plot_grid(
  fig_desurv_std_correlation_tcgacptac_top50$plot +
    theme(plot.margin = margin(2, 2, 2, 20)),
  plot_grid(
    NULL,
    cowplot::ggdraw(legend_d_grob),
    nrow = 2, rel_heights = c(0.08, 0.92)
  ),
  ncol = 2, rel_widths = c(4, 1)
)

bottom_row <- plot_grid(
  plot_3c, plot_3d,
  ncol = 2, labels = c("C", "D"),
  label_size = 12, rel_widths = c(0.55, 0.45)
)

plot_grid(top_row, bottom_row, nrow = 2, rel_heights = c(1.3, 0.7))
```


## Survival-aligned programs generalize across independent PDAC cohorts with lower complexity {.unnumbered}

A prognostic factorization is useful only if it generalizes beyond the training cohort. Having established that survival supervision and standard NMF produce different factor structures in the training data, we next tested whether each method's structure transfers to independent cohorts without retraining. The five external validation cohorts (Dijk, Moffitt, PACA-AU array and RNA-seq, and Puleo) likewise consist of non-metastatic, treatment-naive specimens (SI Appendix, Table S1). We computed factor scores in each validation dataset by projecting the gene programs learned in the training cohort ($Z = \widetilde{W}^\top X_{\text{new}}$), yielding a continuous score for each factor per sample.

External validation C-index for NMF at independently selected ranks is summarized in SI Appendix, Table S2. Standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ substantially improved over NMF at $k = 3$, with $k = 7$ matching or exceeding DeSurv's per-cohort C-index in most cohorts --- though only when DeSurv's tuning framework is applied ($\alpha = 0$ in the joint BO over rank and hyperparameters), since NMF's own rank-selection heuristics (cophenetic, silhouette, residual; SI Appendix, Fig.\ S4) do not converge on this rank. However, this concordance gain still requires more than twice as many factors. DeSurv at $k = 3$ retains a significant adjusted hazard ratio after adjustment for PurIST and DeCAF (SI Appendix, Table S3), achieving independent prognostic value more parsimoniously. Across `r n_val_cohorts` independent PDAC cohorts ($n = `r n_val_total`$), the DeSurv linear predictor showed a consistent survival association (stratified Cox model: pooled HR per SD $`r hr_pool`$; 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$; $P = `r p_pool`$; per-cohort forest plot in Fig. \ref{fig:val}A; KM curves in SI Appendix, Fig. S5C--F). At the matched rank ($k = 3$), the analogous NMF linear predictor showed a weaker pooled effect (HR per SD $`r hr_pool_nmf`$; 95% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$; $P = `r p_pool_nmf`$); a direct comparison using dichotomized risk groups is presented below.

To assess clinical interpretability of these continuous associations, we dichotomized validation samples into high- and low-risk groups using method-specific cutpoints, each independently optimized via cross-validation on the training data. The DeSurv linear predictor separated survival trajectories in the pooled validation cohort (HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:val}B). The same procedure applied to NMF at $k = 3$ yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:val}C). The DeSurv cutpoint identified a substantially larger high-risk group than NMF (`r km_n_high_desurv` vs `r km_n_high_nmf` high-risk patients; `r km_reclass_diff` reclassified), thereby providing finer prognostic discrimination. The DeSurv high-risk group was enriched for both established poor-prognosis subtypes (Basal-like by PurIST, proCAF by DeCAF; Fisher's exact, SI Appendix, Fig. S6), capturing established consensus biology while adding independent prognostic information. The same analysis applied to NMF risk groups at $k = 3$ showed Basal-like enrichment but no significant proCAF enrichment (SI Appendix, Fig. S6), consistent with NMF's failure to isolate the tumor--stroma coupling that underlies D1. Together, these results indicate that incorporating survival information during factorization yields gene programs with more transportable prognostic associations than the standard discover-then-evaluate approach.

```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC cohorts (pooled $n=", n_val_total, "$ patients across ", n_val_cohorts, " cohorts: ", paste(sprintf("%s $n=%d$", gsub("AU seq", "AU RNA-seq", gsub("PACA AU", "PACA-AU", gsub("_", " ", names(table(pool_df$dataset))))), as.integer(table(pool_df$dataset))), collapse = "; "), "). (A) Forest plot of univariate hazard ratios (95\\% CIs) from separate Cox models for each factor, showing each factor's marginal survival association in each validation cohort. DeSurv factor D1 shows a consistent protective association across all five cohorts (pooled HR = 0.76), while no standard NMF factor achieves a comparable pooled effect. Error bars are 95\\% confidence intervals; black diamonds denote inverse-variance-weighted pooled estimates. (B--C) Kaplan--Meier curves for pooled external validation samples stratified by the combined multivariate linear predictor ($\\hat{\\beta}_1 Z_1 + \\hat{\\beta}_2 Z_2 + \\hat{\\beta}_3 Z_3$, where coefficients are learned from training data), dichotomized at a cross-validated optimal cutpoint: DeSurv (B) and standard NMF (C). \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

library(cowplot)
library(ggplot2)
library(dplyr)

# ── Panel A: forest plot (rebuilt from stored data with pooled HRs) ──
load_result("fig_hr_forest_tcgacptac")
d <- fig_hr_forest_tcgacptac$data

# Clean dataset labels
d$dataset <- dplyr::recode(d$dataset,
  "Puleo_array" = "Puleo", "Moffitt_GEO_array" = "Moffitt",
  "PACA_AU_seq" = "PACA seq", "PACA_AU_array" = "PACA array")

# Compute inverse-variance-weighted pooled HR per method × factor
pooled <- d %>%
  mutate(logHR = log(HR),
         se = (log(upper) - log(lower)) / (2 * 1.96),
         w = 1 / se^2) %>%
  group_by(method, factor_name) %>%
  summarise(pooled_logHR = sum(logHR * w) / sum(w),
            pooled_se = sqrt(1 / sum(w)), .groups = "drop") %>%
  mutate(HR = exp(pooled_logHR),
         lower = exp(pooled_logHR - 1.96 * pooled_se),
         upper = exp(pooled_logHR + 1.96 * pooled_se),
         dataset = "Pooled")

keep_cols <- c("factor_name", "HR", "lower", "upper", "dataset", "method", "row_type")
d$row_type <- "cohort"
pooled$row_type <- "pooled"
all_data <- rbind(d[, keep_cols], pooled[, keep_cols])

# Method label for facet panels
all_data$method <- factor(all_data$method, levels = c("DeSurv", "NMF"),
                          labels = c("DeSurv (D1-D3)", "Standard NMF (N1-N3)"))
all_data$dataset <- factor(all_data$dataset,
  levels = c("Puleo", "PACA seq", "PACA array", "Moffitt", "Dijk", "Pooled"))
# Okabe-Ito palette (color-blind safe; Okabe & Ito 2008) plus distinct shapes
# so cohorts remain distinguishable under deuteranopia/protanopia/tritanopia.
cohort_cols <- c("Dijk" = "#E69F00", "Moffitt" = "#56B4E9", "PACA array" = "#009E73",
                 "PACA seq" = "#0072B2", "Puleo" = "#CC79A7", "Pooled" = "#000000")
cohort_shapes <- c("Dijk" = 16, "Moffitt" = 17, "PACA array" = 15,
                   "PACA seq" = 25, "Puleo" = 8,  "Pooled" = 18)
cohort_sizes <- c("Dijk" = 2.2, "Moffitt" = 2.2, "PACA array" = 2.2,
                  "PACA seq" = 2.2, "Puleo" = 2.2, "Pooled" = 4)

base_size <- 8
dodge_w <- 0.6

# Split into DeSurv / NMF using relabeled factor levels
desurv_dat <- all_data[all_data$method == "DeSurv (D1-D3)", ]
desurv_dat$factor_name <- factor(desurv_dat$factor_name,
  levels = c("D1", "D2", "D3"))
nmf_dat <- all_data[all_data$method == "Standard NMF (N1-N3)", ]
nmf_dat$factor_name <- factor(nmf_dat$factor_name,
  levels = c("N1", "N2", "N3"))

# Helper to build one forest panel
make_forest_panel <- function(dat, title, highlight_level) {
  hl <- data.frame(
    factor_name = factor(highlight_level, levels = levels(dat$factor_name)),
    x = 1)
  ggplot(dat, aes(x = HR, y = factor_name)) +
    geom_tile(data = hl, aes(x = x, y = factor_name),
              width = 100, height = 0.8,
              fill = "steelblue", alpha = 0.08, inherit.aes = FALSE) +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60",
               linewidth = 0.4) +
    geom_errorbar(aes(xmin = lower, xmax = upper, colour = dataset,
                      linewidth = row_type),
                  orientation = "y", width = 0,
                  position = position_dodge(width = dodge_w)) +
    geom_point(aes(colour = dataset, shape = dataset, size = dataset),
               position = position_dodge(width = dodge_w)) +
    scale_linewidth_manual(values = c("cohort" = 0.6, "pooled" = 1.2),
                           guide = "none") +
    scale_colour_manual(values = cohort_cols, name = NULL, drop = FALSE) +
    scale_shape_manual(values = cohort_shapes, name = NULL, drop = FALSE) +
    scale_size_manual(values = cohort_sizes, name = NULL, drop = FALSE) +
    scale_x_log10(limits = c(0.35, 3)) +
    labs(x = "Hazard ratio (95% CI)", y = NULL, title = title) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", size = 9, hjust = 0.5),
      axis.text.y = element_text(size = 8, face = "bold"),
      axis.text.x = element_text(size = 8, face = "bold"),
      axis.title.x = element_text(size = 9),
      plot.margin = margin(2, 6, 2, 4)
    )
}

p_desurv <- make_forest_panel(desurv_dat, "DeSurv (D1-D3)", "D1")
p_nmf    <- make_forest_panel(nmf_dat, "Standard NMF (N1-N3)", "N1")

# Extract shared legend
p_leg <- ggplot(all_data, aes(x = HR, y = factor_name, colour = dataset,
                               shape = dataset, size = dataset)) +
  geom_point() +
  scale_colour_manual(values = cohort_cols, name = NULL) +
  scale_shape_manual(values = cohort_shapes, name = NULL) +
  scale_size_manual(values = cohort_sizes, name = NULL) +
  guides(colour = guide_legend(nrow = 1, override.aes = list(size = 2.5)),
         shape = guide_legend(nrow = 1), size = guide_legend(nrow = 1)) +
  theme_void(base_size = base_size) +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))
leg_grob = ggplotGrob(p_leg)
forest_legend <- gtable::gtable_filter(leg_grob, "guide-box")

plot_forest <- plot_grid(
  plot_grid(p_desurv, p_nmf, ncol = 2, align = "hv", axis = "tb"),
  forest_legend,
  nrow = 2, rel_heights = c(15, 1)
)

# ── Panels B-C: KM curves ──
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")

theme_pnas <- theme_classic(base_size = base_size) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.margin = margin(6, 10, 6, 15),
    legend.box.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(6, "pt"),
    legend.key.width = unit(12, "pt")
  )

km_text_size    <- base_size
risk_title_size <- base_size

stack_surv <- function(surv_obj, title, keep_legend = FALSE) {
  p <- surv_obj$plot +
    theme_pnas +
    theme(
      legend.position = if (keep_legend) "bottom" else "none",
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = km_text_size),
      axis.text.x  = element_text(size = km_text_size),
      axis.text.y  = element_text(size = km_text_size),
      plot.margin  = margin(4, 10, 0, 10),
      plot.title   = element_text(size = 9)
    ) +
    labs(title = title)

  t <- surv_obj$table +
    theme_pnas +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.x  = element_text(size = km_text_size),
      axis.text.y  = element_text(size = km_text_size),
      text = element_text(size = km_text_size),
      axis.title.x = element_text(size = km_text_size),
      plot.title   = element_text(size = risk_title_size),
      plot.margin  = margin(4, 10, 2, 10)
    ) +
    labs(x = "Time (months)")
  t$layers[[1]]$aes_params$size <- km_text_size / ggplot2::.pt
  plot_grid(p, t, ncol = 1, rel_heights = c(2.5, 1.5), align = "v", axis = "lr")
}

km_b <- stack_surv(fig_median_survival_desurv_tcgacptac, "DeSurv")
km_c <- stack_surv(fig_median_survival_std_desurvk_tcgacptac, "NMF")

# Shared KM legend
legend_plot <- ggplot(
  data.frame(x = 1:2, y = 1:2,
             group = factor(c("Low", "High"), levels = c("Low", "High"))),
  aes(x = x, y = y, colour = group)
) +
  geom_line() +
  scale_colour_manual(values = c("Low" = "violetred2", "High" = "turquoise4"),
                      name = "Risk group") +
  theme_pnas +
  theme(legend.position = "bottom",
        legend.text = element_text(size = km_text_size),
        legend.title = element_text(size = km_text_size)) +
  guides(colour = guide_legend(nrow = 1))

legend_grob <- ggplotGrob(legend_plot)
km_legend <- legend_grob$grobs[
  sapply(legend_grob$grobs, function(x) x$name) == "guide-box"
][[1]]

km_block <- plot_grid(
  km_b, km_c, ggdraw(km_legend),
  nrow = 3, labels = c("B", "C", ""),
  label_size = 12, rel_heights = c(5, 5, 0.6)
)

# ── Full figure: forest (left) + KM curves (right) ──
plot_grid(plot_forest, km_block, ncol = 2, rel_widths = c(1.4, 1),
          labels = c("A", ""), label_size = 12)
```

