<!-- REVISED RESULTS — prediction-validation structure (originally for PNAS; retained for Nature submission) -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- Supplementary Information figure numbering (supplement.Rmd order, post-reorganization):
     S1  fig-converge          convergence trajectories
     S2  fig-sim-null-mixed    null and mixed simulation scenarios
     S3  fig-cindex-by-k       C-index vs K sensitivity
     S4  fig-nmf-diagnostics   standard NMF rank selection heuristics (demoted from old Fig 2A-C)
     S5  fig-cutpoint-km       cutpoint selection and KM curves
     S6  fig-subtype-overlap   subtype composition of risk groups
     S7  fig-desurv-k5-heatmap DeSurv gene overlap at k=5
     S8  fig-nmf-k5-heatmap    NMF gene overlap at k=5
     S9  fig-nmf-k7-heatmap    NMF gene overlap at k=7
     S10 fig-varsurvival-k5    variance-survival at k=5
     Note: Former S7 (variance-survival), S8 (W-corr), S9 (forest) are now
           main text Fig 3C-D and Fig 4A respectively.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(magick)
library(gt)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)
desurv_maxvar_label <- paste0("D", desurv_maxvar$factor)
nmf_maxvar_label    <- paste0("N", nmf_maxvar$factor)
# Exocrine NMF factor (N2) — referenced explicitly because of its biological
# role as the high-variance exocrine-associated factor
nmf_n2              <- nmf_vs[nmf_vs$factor == 2, ]
nmf_n2_pct          <- round(nmf_n2$variance_explained * 100, 1)
nmf_n2_dl           <- round(nmf_n2$delta_loglik, 2)

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Pooled adjusted (PurIST + DeCAF) statistics for DeSurv ───────────────────
load_result("data_val_filtered_tcgacptac")

extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(sample_id = rownames(surv), time = surv$time, event = surv$event,
                     dataset = entry$dataset, risk_score = entry$risk_score,
                     stringsAsFactors = FALSE)
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo; si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

pool_df_adj_complete <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
pool_df_adj_complete <- pool_df_adj_complete[
  complete.cases(pool_df_adj_complete[, c("PurIST", "DeCAF")]), ]
rs_sd_adj <- sd(pool_df_adj_complete$risk_score, na.rm = TRUE)
if (is.na(rs_sd_adj) || rs_sd_adj < 1e-12) rs_sd_adj <- 1
pool_df_adj_complete$risk_z <- pool_df_adj_complete$risk_score / rs_sd_adj
cox_adj_sum <- summary(coxph(
  Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
  data = pool_df_adj_complete))
hr_adj      <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "exp(coef)"])
ci_adj_lo   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "lower .95"])
ci_adj_hi   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "upper .95"])
p_adj_val   <- cox_adj_sum$coefficients["risk_z", "Pr(>|z|)"]
p_adj_text  <- if (p_adj_val < 0.001) "< 0.001" else paste0("= ", signif(p_adj_val, 2))

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

# ── KM dichotomized risk-group statistics ─────────────────────────────────────
# Extract HR/CI/p from the annotation layer embedded in the KM plot objects
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# Extract exact risk-group counts from KM risk tables (time = 0)
km_n_high_desurv <- fig_median_survival_desurv_tcgacptac$data.survtable$n.risk[
  fig_median_survival_desurv_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_desurv_tcgacptac$data.survtable$strata == "factor=1"]
km_n_high_nmf <- fig_median_survival_std_desurvk_tcgacptac$data.survtable$n.risk[
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$strata == "factor=1"]
km_reclass_diff <- km_n_high_desurv - km_n_high_nmf
```

## DeSurv learns survival-associated gene programs from bulk transcriptomes {.unnumbered}

To identify prognostic yet interpretable gene programs, we applied DeSurv (Fig.\ \ref{fig:schema}A, Methods), which shapes programs by both expression reconstruction and survival association acting through the gene-program matrix, preserving the interpretation of sample loadings as mixture coefficients, so that trained programs can be projected into new cohorts without retraining.

We applied DeSurv to the PDAC training cohorts from TCGA and CPTAC [@raphael2017integrated; @cao2021proteogenomic] (Methods). Cross-validated concordance plateaued from $k = 3$ through $k = 12$ (Supplementary Fig.\ \ref{fig:fig-cindex-by-k}), and Bayesian optimization over rank and supervision strength with a one-standard-error rule selected a parsimonious three-factor solution, $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ at $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$ (Fig.\ \ref{fig:schema}B). The peak cross-validated C-index was $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$ across `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)` patients with `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)` events. By contrast, standard unsupervised NMF diagnostics, including reconstruction residuals, cophenetic correlation and silhouette width [@brunet2004metagenes; @frigyesi2008nmf], did not identify a consistent rank in PDAC (Supplementary Fig.\ \ref{fig:fig-nmf-diagnostics}). All rank and hyperparameter selection was performed within the training set before projection into validation cohorts.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "model_schematic_final.pdf"))
```


## Survival supervision reshapes PDAC factor structure {.unnumbered}

To determine how survival supervision shaped the recovered factor structure, we fit standard NMF at the same rank selected for DeSurv ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$) and compared each method's factor-specific gene rankings with established PDAC gene programs (Fig.\ \ref{fig:pdac}A--B). Matching $k$ across methods controls for differences in model complexity when comparing the effect of survival supervision on factor content; a sensitivity analysis using NMF's independently selected ranks (elbow $k = 5$, BO $k = 7$) is provided in Supplementary Table \ref{tab:tab-cindex-comparison}. We compared the biological content of each factor and then quantified how much each contributed to reconstruction and survival.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. In the DeSurv factorization, D1 correlated with classical tumor programs and restCAF-like stromal signatures, D2 with proCAF-like programs, and D3 with basal-like tumor programs (Fig. \ref{fig:pdac}A). In the DeSurv solution, the dominant survival-associated signal is aligned with a tumor--stroma axis: higher D1 scores were associated with a classical/restCAF-like, lower-risk state, whereas higher basal-like tumor and proCAF-like stromal scores were associated with higher risk. This organization is consistent with prior evidence that joint tumor-intrinsic and CAF state stratification improves prognostic resolution in PDAC [@peng2024determination]. No DeSurv factor showed positive correlation with exocrine-associated programs above the display threshold, including DECODER Exocrine, Collisson Exocrine and Bailey aberrantly differentiated endocrine--exocrine (ADEX) programs (Fig. \ref{fig:pdac}A).

Standard NMF produced a different factor organization (Fig. \ref{fig:pdac}B). One factor, N2, was strongly correlated with exocrine-associated programs and negatively correlated with immune and stromal signatures. A second factor, N1, correlated with malignant-epithelial identity and carried both classical and basal-like markers. A third factor, N3, correlated with immune, fibroblast and extracellular matrix (ECM)-related programs, including Puleo Desmoplastic, SCISSORS proCAF and restCAF [@leary2023s], Maurer Immune and ECM [@Maurer2019], and DECODER Activated Stroma. Thus, at $k = 3$, standard NMF devoted one factor to exocrine-associated variation and aggregated tumor-intrinsic and microenvironmental programs into two broad factors, whereas DeSurv separated basal-like tumor, proCAF-like stroma and a classical/restCAF-like survival-aligned program.

NMF seeks a low-rank \emph{reconstruction} of the transcriptome; many factorizations can reconstruct comparably well, and the unsupervised objective has no incentive to prefer prognostically informative directions. We therefore quantified each factor's contribution using two metrics (Fig. \ref{fig:pdac}C; Supplementary Information, Sections 10--11): a normalized reconstruction Shapley value, which partitions reconstruction contribution across factors and sums to $100\%$, and survival contribution, defined as the change in Cox partial log-likelihood ($\Delta\ell$) when that factor is dropped from a $k$-factor Cox model. Reconstruction and survival contribution were decoupled: all three NMF factors had small $\Delta\ell$ despite spanning a wide range of reconstruction contribution, whereas in the training set (pooled TCGA + CPTAC) DeSurv concentrated nearly all survival contribution into D1 ($\Delta\ell = `r desurv_maxsurv_dl`$ from only $`r desurv_maxsurv_pct`$% of reconstruction; per-factor values in Fig. \ref{fig:pdac}C). Thus, in PDAC, factors that best reconstruct expression are not necessarily the factors that best explain survival, consistent with tumor-purity analyses across cancer types [@aran2015systematic].

To further characterize the relationship between the two factorizations, we examined pairwise W-matrix gene-loading correlations across all 1,970 genes (Fig. \ref{fig:pdac}D; Spearman correlations over full loading profiles, since top-gene subsets can overstate preservation). At $k = 3$, NMF captured the malignant-epithelial compartment largely in a single factor, N1, which carried both classical and basal-like markers and therefore is labeled \emph{tumor} rather than classical. N1 mapped strongly onto DeSurv's basal-like tumor factor, D3 (Spearman $\rho = 0.84$), while NMF's microenvironmental factor, N3, mapped onto DeSurv's proCAF-like stromal factor, D2 ($\rho = 0.87$). In contrast, DeSurv's most prognostic factor, D1, was only weakly correlated with all three NMF factors (all $\rho \le 0.36$). Because D1 couples classical tumor and restCAF-like stromal signatures into a single survival-aligned program, it is not simply a reweighting of an NMF factor: little of its gene-loading profile is linearly reconstructable from the matched-rank NMF basis (Supplementary Table \ref{tab:projection-r2}). NMF's exocrine factor, N2, had no comparably strong DeSurv counterpart (maximum $\rho = 0.61$), consistent with reduced emphasis on exocrine-associated variation under survival supervision. Together, these analyses indicate that, at $k = 3$, standard NMF separated broad tumor, stromal/immune and exocrine-associated structure, whereas DeSurv produced a survival-aligned tumor--stroma program not closely matched by any single NMF factor.
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (A,B) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$; only $|r| > 0.2$ shown. (C) Per-factor reconstruction contribution (normalized Shapley share, summing to 100\\%) versus survival contribution ($\\Delta$ Cox partial log-likelihood on factor removal). (D) Pairwise correspondence between NMF and DeSurv factors (Spearman correlation of W-matrix gene loadings over all genes). \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.height= "7in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig3_tcgacptac.pdf"))
```


## Survival-aligned programs generalize across independent PDAC cohorts {.unnumbered}

We next tested whether the programs learned in the TCGA and CPTAC training cohort transferred to five independent PDAC cohorts without retraining (Dijk, Moffitt, PACA-AU array, PACA-AU RNA-seq and Puleo; non-metastatic, treatment-naive specimens; Supplementary Table \ref{tab:cohorts}), projecting the training-derived programs into each validation expression matrix ($Z = \widetilde{W}^\top X_{\text{new}}$) to obtain a continuous per-factor score for each sample.

Across the `r n_val_cohorts` validation cohorts ($n = `r n_val_total`$), the DeSurv linear predictor (oriented so that higher values indicate higher risk) was consistently associated with survival in a stratified Cox model (pooled HR per SD $= `r hr_pool`$, 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$, $P = `r p_pool`$; Fig. \ref{fig:val}A). At the matched rank of $k = 3$, the analogous standard NMF predictor showed a weaker pooled association (HR per SD $= `r hr_pool_nmf`$, 95% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$, $P = `r p_pool_nmf`$). The combined DeSurv linear predictor remained prognostic after simultaneous adjustment for PurIST and DeCAF classifications (per-SD HR $= `r hr_adj`$, 95% CI $`r ci_adj_lo`$--$`r ci_adj_hi`$, $P `r p_adj_text`$; Table \ref{tab:hr-adjusted}), indicating that the tumor--stroma axis captures information beyond established malignant and stromal subtype labels.

```{r suprec-bench, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_stats")
.svb <- function(m) sprintf("%.2f", desurv_vs_supervised_stats$pooled[m, "pooled_C"])
```

The continuous, stratified-Cox associations above are our primary analysis; to illustrate them clinically, we dichotomized validation samples into high- and low-risk groups using method-specific cutpoints independently optimized by cross-validation in the training data. The DeSurv linear predictor separated survival trajectories in the pooled validation cohort (HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:val}B), whereas standard NMF at $k = 3$ yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:val}C). The DeSurv high-risk group was enriched for both established poor-prognosis subtypes, basal-like by PurIST and proCAF by DeCAF (Fisher's exact test; Supplementary Fig. \ref{fig:fig-subtype-overlap}), whereas NMF risk groups at $k = 3$ showed basal-like enrichment but no significant proCAF enrichment (Supplementary Fig. \ref{fig:fig-subtype-overlap}), consistent with NMF not isolating the tumor--stroma axis.

Higher-rank NMF models achieved similar validation concordance in some cohorts but required more than twice as many factors and did not recover the same compact tumor--stroma axis: NMF at elbow-selected $k = 5$ and Bayesian-optimization-selected $k = 7$ improved over $k = 3$ (Supplementary Table \ref{tab:tab-cindex-comparison}), yet NMF's own rank-selection diagnostics did not converge on these ranks (Supplementary Fig. \ref{fig:fig-nmf-diagnostics}), and DeSurv's three-factor model was preferred by BIC over NMF's seven-factor model in three of five cohorts (Supplementary Table \ref{tab:tab-bic-percohort}). Prediction alone likewise did not distinguish DeSurv from simpler survival-supervised models: its pooled transfer concordance (C-index $= `r .svb("DeSurv (full LP)")`$) was on par with penalized (lasso) Cox ($`r .svb("Sparse Cox")`$), Cox partial least squares ($`r .svb("Cox-PLS")`$) and supervised principal components ($`r .svb("Supervised PCA")`$), whereas unsupervised directions transferred poorly (standard NMF at $\alpha = 0$, $`r .svb("Unsup. NMF (a=0)")`$; the leading principal component, $`r .svb("Unsup. PCA (PC1)")`$), consistent with evidence that learned representations do not reliably out-predict penalized regression for out-of-sample prognosis [@smith2020standard]. DeSurv's contribution is therefore not higher predictive accuracy but the decomposition of a transferable prognostic signal into a compact, interpretable, source-attributable tumor--stroma program.

```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC cohorts (pooled $n=", n_val_total, "$ patients across ", n_val_cohorts, " cohorts: ", paste(sprintf("%s $n=%d$", gsub("AU seq", "AU RNA-seq", gsub("PACA AU", "PACA-AU", gsub("_", " ", names(table(pool_df$dataset))))), as.integer(table(pool_df$dataset))), collapse = "; "), "). (A) Forest plot of per-cohort univariate hazard ratios (95\\% CIs) for each factor; black diamonds denote inverse-variance-weighted pooled estimates. (B,C) Kaplan--Meier curves for pooled validation samples stratified by the training-derived multivariate linear predictor, dichotomized at a cross-validated optimal cutpoint: DeSurv (B) and standard NMF (C). No validation-cohort survival outcomes were used in training, scoring or hyperparameter selection. \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig4_tcgacptac.pdf"))
```

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(knitr); library(kableExtra)
# extract_pool_df_with_covariates() and data_val_filtered_tcgacptac
# are defined in the setup chunks above.
load_result("val_latent_desurv_alpha0_tcgacptac")

pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(hr = round(ci_row[, "exp(coef)"], 2),
         ci_lo = round(ci_row[, "lower .95"], 2),
         ci_hi = round(ci_row[, "upper .95"], 2),
         p = coef_row[, "Pr(>|z|)"])
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

pool_desurv_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv_t1 <- pooled_cox_unadj_adj(pool_desurv_t1)
pool_nmf7_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7_t1 <- pooled_cox_unadj_adj(pool_nmf7_t1)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)", "Std NMF ($k = 7$, BO $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv_t1$unadj), fmt_hr(stats_nmf7_t1$unadj)),
  Adjusted = c(fmt_hr(stats_desurv_t1$adj), fmt_hr(stats_nmf7_t1$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per SD (95\\% CI), $P$",
                       "Adjusted HR per SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = paste0("Pooled external validation hazard ratios (per SD of the linear predictor) for DeSurv at $k = 3$ and standard NMF at BO-selected $k = 7$ ($\\alpha = 0$), across the five validation cohorts ($n = ", n_val_total, "$ patients, ", n_val_events, " events). Adjusted models include the PurIST tumor-intrinsic and DeCAF stromal molecular classifiers as covariates, with stratification by dataset. Both methods retain significant prognostic value after adjustment. \\label{tab:hr-adjusted}")) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

```{r d1-partition-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("d1_variance_partition_stats")
.d1p <- d1_variance_partition_stats
.d1pf <- function(p) if (p < 1e-4) "< 0.0001" else sprintf("= %.4f", p)
load_result("supervised_recovery_stats")
.d1p_sr_mean  <- supervised_recovery_stats$sup_mean_score_cor
.d1p_sr_unsup <- abs(supervised_recovery_stats$table$score_cor_D1[!supervised_recovery_stats$table$supervised])
```

## The tumor--stroma axis adds prognostic information beyond established PDAC classifiers {.unnumbered}

We next asked whether the D1 axis provided information beyond established discrete subtype calls. Across the pooled validation cohorts ($n = `r .d1p$meta$n`$, `r .d1p$meta$n_events` events), binary PurIST and DeCAF classifications together explained `r sprintf("%.0f", 100*.d1p$variance$explained_total)`\% of the within-cohort variance in the continuous D1 score, so D1 aligned with known tumor-intrinsic and stromal subtypes while capturing a finer-grained continuum of tumor--stroma state.

D1 remained independently associated with survival after simultaneous adjustment for PurIST and DeCAF classifications (adjusted HR per SD = `r sprintf("%.2f", .d1p$prognosis$d1_adj_hr)`, 95\% CI `r sprintf("%.2f", .d1p$prognosis$d1_adj_lo)`--`r sprintf("%.2f", .d1p$prognosis$d1_adj_hi)`, $P `r .d1pf(.d1p$prognosis$d1_adj_p)`$). Adding D1 to a stratified Cox model containing PurIST and DeCAF improved model fit (partial likelihood-ratio $\chi^2 = `r sprintf("%.1f", .d1p$prognosis$lr_chisq)`$ on 1 df, $P `r .d1pf(.d1p$prognosis$lr_p)`$) and increased concordance from `r sprintf("%.3f", .d1p$prognosis$c_base)` to `r sprintf("%.3f", .d1p$prognosis$c_with_d1)`. The tumor--stroma axis therefore captures prognostic variation not fully represented by either classifier alone.

This axis was not specific to the DeSurv optimizer. Three independent survival-supervised approaches (supervised principal components, Cox partial least squares and sparse Cox) recovered scores highly correlated with D1 in the training cohort (mean $|r| = `r sprintf("%.2f", .d1p_sr_mean)`$), whereas the leading unsupervised direction did not ($|r| = `r sprintf("%.2f", .d1p_sr_unsup)`$; Supplementary Information, supervised-method recovery). These results indicate that D1 reflects a reproducible survival-associated expression direction rather than an idiosyncrasy of a single factorization algorithm.

Finally, in the fully annotated external PACA-AU cohorts, the DeSurv score remained prognostic after adjustment for stage, grade, age and sex, and further adjustment for tumor purity did not eliminate the association in the two PACA-AU cohorts with purity annotations (Supplementary Table \ref{tab:clinical-adjustment}). In the smaller Dijk and Moffitt cohorts, the DeSurv score was not significantly associated with survival before adjustment, and the Moffitt purity-adjusted analysis was underpowered ($n = 29$). These analyses support the tumor--stroma axis as a prognostic continuum that is not reducible to established subtype labels, standard clinical covariates or tumor purity.

In an exploratory, hypothesis-generating analysis, we also projected the trained programs into three bulk treated PDAC cohorts (the metastatic Rash-Accept series and the borderline-resectable/locally advanced Linehan cohort); because treatment, stage and cohort were confounded, these associations are prognostic, not predictive. The proCAF program $D_2$ showed a consistent protective direction on overall survival (reaching significance in the less-advanced Linehan cohort), whereas the basal--classical axis was setting-dependent, attenuating in metastatic disease as it does for established classifiers (Supplementary Information, Section \ref{si:treated}).


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

We asked whether survival supervision improves recovery of prognostic gene programs when prognostic variation is not the dominant source of transcriptomic variance, a regime that probes weak prognostic signal rather than all transcriptomic settings. We designed simulations with known ground-truth latent structure, in which prognostic gene programs explained low variance relative to outcome-neutral background signals and survival times were generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies). We compared DeSurv to standard NMF ($\alpha=0$), where factors are learned by reconstruction error alone and survival associations are estimated by a separately fit Cox model; both were tuned via BO over cross-validated C-index, so differences reflect the role of supervision during factorization, not during model selection.

In the prognostic scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index (Fig.\ \ref{fig:sim}A) and substantially higher precision for recovering the true prognostic gene programs (Fig.\ \ref{fig:sim}B). Here precision measures how well the learned factors recover the true prognostic programs, computed per program against its best-matching factor and averaged (Supplementary Information, Sections 6 and 9). The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This indicates that outcome supervision during factorization improves recovery of prognostic programs that do not dominate expression variance. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$). The gain in gene-program precision exceeded the gain in C-index, indicating that unsupervised factorization can achieve similar predictive performance while recovering different gene programs. Survival supervision therefore improves not only prediction but also recovery of the underlying prognostic biology.

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under a null scenario ($\beta = 0$) and a mixed scenario with partial overlap between variance-dominant and prognostic structure (Supplementary Fig.\ \ref{fig:fig-sim-null-mixed}). The benefit of survival supervision scaled with the separation between prognostic and variance-dominant structure: largest in the prognostic scenario ($\Delta = `r delta_sim_cindex`$), attenuated when variance and prognosis partially overlapped ($\Delta = `r delta_mixed_cindex`$, paired Wilcoxon $P `r mixed_wilcox_p`$), and absent when no survival signal existed, where both methods yielded C-index near 0.5. DeSurv also recovered the true rank ($k = 3$) more frequently than standard NMF, which systematically under-selected near $k = 2$ (Fig.\ \ref{fig:sim}C).

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Survival supervision improves recovery of prognostic gene programs in simulations and PDAC training data. (A) Test-set concordance across 100 simulation replicates comparing DeSurv with standard NMF ($\\alpha = 0$; $p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; paired Wilcoxon $P ", sim_wilcox_p, "$). (B) Precision of recovered prognostic gene programs (overlap between each true program and the top-weighted genes of its best-matching learned factor). In (A) and (B), boxes show median and interquartile range; whiskers extend to $1.5\\times$IQR; points are individual replicates. (C) Distribution of selected ranks across replicates for DeSurv versus standard NMF. (D) Gaussian-process--predicted mean cross-validated C-index over factorization rank and supervision strength in PDAC training data (TCGA + CPTAC); this panel shows empirical tuning, not a simulation result. \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

knitr::include_graphics(file.path("..", "figures", "fig2_tcgacptac.pdf"))

```

