<!-- REVISED RESULTS — prediction-validation structure (originally for PNAS; retained for Nature submission) -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- Supplementary Information figure numbering (supplement.Rmd order, post-reorganization):
     S1  fig-converge          convergence trajectories
     S2  fig-sim-null-mixed    null and mixed simulation scenarios
     S3  fig-cindex-by-k       C-index vs K sensitivity
     S4  fig-nmf-diagnostics   standard NMF rank selection heuristics (demoted from old Fig 2A-C)
     S5  fig-cutpoint-km       cutpoint selection and KM curves
     S6  fig-subtype-overlap   subtype composition of risk groups
     S7  fig-desurv-k5-heatmap DeSurv gene overlap at k=5
     S8  fig-nmf-k5-heatmap    NMF gene overlap at k=5
     S9  fig-nmf-k7-heatmap    NMF gene overlap at k=7
     S10 fig-varsurvival-k5    variance-survival at k=5
     Note: Former S7 (variance-survival), S8 (W-corr), S9 (forest) are now
           main text Fig 3C-D and Fig 4A respectively.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(magick)
library(gt)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)
desurv_maxvar_label <- paste0("D", desurv_maxvar$factor)
nmf_maxvar_label    <- paste0("N", nmf_maxvar$factor)
# Exocrine NMF factor (N2) — referenced explicitly because of its biological
# role as the high-variance exocrine-associated factor
nmf_n2              <- nmf_vs[nmf_vs$factor == 2, ]
nmf_n2_pct          <- round(nmf_n2$variance_explained * 100, 1)
nmf_n2_dl           <- round(nmf_n2$delta_loglik, 2)

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Pooled adjusted (PurIST + DeCAF) statistics for DeSurv ───────────────────
load_result("data_val_filtered_tcgacptac")

extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(sample_id = rownames(surv), time = surv$time, event = surv$event,
                     dataset = entry$dataset, risk_score = entry$risk_score,
                     stringsAsFactors = FALSE)
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo; si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

pool_df_adj_complete <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
pool_df_adj_complete <- pool_df_adj_complete[
  complete.cases(pool_df_adj_complete[, c("PurIST", "DeCAF")]), ]
rs_sd_adj <- sd(pool_df_adj_complete$risk_score, na.rm = TRUE)
if (is.na(rs_sd_adj) || rs_sd_adj < 1e-12) rs_sd_adj <- 1
pool_df_adj_complete$risk_z <- pool_df_adj_complete$risk_score / rs_sd_adj
cox_adj_sum <- summary(coxph(
  Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
  data = pool_df_adj_complete))
hr_adj      <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "exp(coef)"])
ci_adj_lo   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "lower .95"])
ci_adj_hi   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "upper .95"])
p_adj_val   <- cox_adj_sum$coefficients["risk_z", "Pr(>|z|)"]
p_adj_text  <- if (p_adj_val < 0.001) "< 0.001" else paste0("= ", signif(p_adj_val, 2))

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

# ── KM dichotomized risk-group statistics ─────────────────────────────────────
# Extract HR/CI/p from the annotation layer embedded in the KM plot objects
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# Extract exact risk-group counts from KM risk tables (time = 0)
km_n_high_desurv <- fig_median_survival_desurv_tcgacptac$data.survtable$n.risk[
  fig_median_survival_desurv_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_desurv_tcgacptac$data.survtable$strata == "factor=1"]
km_n_high_nmf <- fig_median_survival_std_desurvk_tcgacptac$data.survtable$n.risk[
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$strata == "factor=1"]
km_reclass_diff <- km_n_high_desurv - km_n_high_nmf
```

## DeSurv learns survival-associated gene programs from bulk transcriptomes {.unnumbered}

DeSurv learns gene programs that balance reconstruction of the tumor transcriptome with association to survival. After training, the learned gene-program basis ($W$) is fixed, allowing the same programs to be scored directly in new tumors without refitting the factorization (Fig.\ \ref{fig:schema}A; Methods). We trained DeSurv in pooled TCGA and CPTAC PDAC cohorts [@raphael2017integrated; @cao2021proteogenomic] ($n = `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)`$, `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)` events). Cross-validated concordance changed little beyond three programs, and Bayesian optimization with the one-standard-error rule selected a parsimonious three-program model ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$, $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$; peak cross-validated C-index, $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$; Fig.\ \ref{fig:schema}B and Supplementary Fig.\ \ref{fig:fig-cindex-by-k}). By contrast, standard unsupervised NMF rank criteria (reconstruction residuals, cophenetic correlation and silhouette width [@brunet2004metagenes; @frigyesi2008nmf]) gave conflicting choices in the same data (Supplementary Fig.\ \ref{fig:fig-nmf-diagnostics}), underscoring rank selection as an additional source of ambiguity in unsupervised decompositions. All model and hyperparameter selection was completed within the training cohorts before the programs were applied to external validation datasets.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "model_schematic_final.pdf"))
```


## Survival supervision resolves a distinct tumor--stroma program structure {.unnumbered}

To determine how survival supervision altered the biological representation, we compared DeSurv with standard NMF at the same rank ($k = 3$), thereby holding model complexity constant (a sensitivity analysis using NMF's independently selected ranks, elbow $k = 5$ and BO $k = 7$, is in Supplementary Table \ref{tab:tab-cindex-comparison}). DeSurv resolved three distinct programs (D1--D3): D1, enriched for classical tumor and restCAF-like stromal signatures; D2, enriched for proCAF-like stroma; and D3, enriched for basal-like tumor programs (Fig.\ \ref{fig:pdac}A). Together, these programs define a tumor--stroma prognostic architecture spanning classical/restCAF-like and basal-like/proCAF-like states across patients. The classical identity of D1 was independently supported in the COMPASS cohort, where D1 scores correlated with GATA6 RNA in situ hybridization (Spearman $\rho = 0.68$, $P < 0.001$, $n = 33$; Fig.\ \ref{fig:pdac}E).

Standard NMF produced a different organization at the same rank (Fig.\ \ref{fig:pdac}B). Its malignant factor (N1) combined classical and basal-like tumor programs, its microenvironmental factor (N3) combined multiple stromal and immune programs [@leary2023s; @Maurer2019], and a third factor (N2) captured exocrine-associated variation. DeSurv instead separated basal-like tumor (D3) and proCAF-like stroma (D2) while recovering the distinct classical/restCAF-like survival-aligned D1 program.

This difference reflected a selective reorganization rather than wholesale loss of expression structure. D1 accounted for only $`r desurv_maxsurv_pct`$% of reconstruction contribution (normalized Shapley value; Fig.\ \ref{fig:pdac}C) but nearly all of the factor-specific survival contribution in the training data ($\Delta\ell = `r desurv_maxsurv_dl`$, the change in Cox partial log-likelihood on factor removal), illustrating that the programs that best reconstruct the transcriptome need not be those most relevant to survival [@aran2015systematic]. Pairwise gene-loading correlations across all 1{,}970 genes showed that the principal NMF tumor and microenvironmental factors were largely retained as D3 (Spearman $\rho = 0.84$) and D2 ($\rho = 0.87$), whereas D1 correlated only weakly with every matched-rank NMF factor ($\rho \leq 0.36$; Fig.\ \ref{fig:pdac}D). Consistently, D1 was poorly reconstructable from the full NMF basis ($R^2 = 0.09$), while DeSurv sacrificed only about 1.5 percentage points of whole-transcriptome reconstruction $R^2$ (Supplementary Table \ref{tab:projection-r2}). Thus, survival supervision largely preserved variance-dominant tumor and stromal structure while reorganizing the representation to expose a distinct survival-aligned program.
```{r fig-pdac, fig.width=7, fig.height=9, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (A,B) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$; only $|r| > 0.2$ shown. (C) Per-factor reconstruction contribution (normalized Shapley share, summing to 100\\%) versus survival contribution ($\\Delta$ Cox partial log-likelihood on factor removal). (D) Pairwise correspondence between NMF and DeSurv factors (Spearman correlation of W-matrix gene loadings over all genes). (E) DeSurv D1 score versus GATA6 RNA in situ hybridization level in the COMPASS cohort (Spearman correlation; $n = 33$). \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.height= "9in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig3_tcgacptac.pdf"))
```


## Survival-aligned programs generalize across independent PDAC cohorts {.unnumbered}

We next tested whether the programs learned in TCGA and CPTAC generalized without retraining to five independent PDAC cohorts (Dijk, Moffitt, PACA-AU array, PACA-AU RNA-seq and Puleo; non-metastatic, treatment-naive; Supplementary Table \ref{tab:cohorts}; $n = `r n_val_total`$, `r n_val_events` events), projecting the trained programs into each validation expression matrix ($Z = \widetilde{W}^\top X_{\text{new}}$). The frozen DeSurv linear predictor (oriented so that higher values indicate higher risk) was associated with overall survival in a cohort-stratified analysis (pooled HR per SD $= `r hr_pool`$, 95\% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$, $P = `r p_pool`$; Fig. \ref{fig:val}A). At the same three-program complexity, the corresponding standard NMF predictor showed a substantially weaker association (HR per SD $= `r hr_pool_nmf`$, 95\% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$, $P = `r p_pool_nmf`$).

We then asked whether comparable prognostic performance could be achieved without survival-supervised deconvolution. Increasing the rank of standard NMF substantially improved external discrimination: NMF at $k = 7$ matched or exceeded DeSurv's concordance in most validation cohorts and, like DeSurv, remained prognostic after simultaneous adjustment for PurIST and DeCAF (Table \ref{tab:hr-adjusted}; Supplementary Table \ref{tab:tab-cindex-comparison}). Thus, survival supervision was not required to recover prognostic information from the transcriptome. Increasing NMF rank, however, distributed this information across a rank-sensitive set of overlapping factors: at $k = 7$, multiple factors shared the same established PDAC signatures and the additional factors contributed little individually to survival despite strong performance of the combined predictor (Supplementary Fig. \ref{fig:fig-nmf-k7-heatmap}).

```{r suprec-bench, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_stats")
load_result("desurv_vs_supervised_tuned")
.svb  <- function(m) sprintf("%.2f", desurv_vs_supervised_stats$pooled[m, "pooled_C"])
.svbt <- function(m) sprintf("%.2f", desurv_vs_supervised_tuned$prediction[m, "POOLED"])
```

As a secondary, clinical visualization, a risk cutpoint selected entirely within the training data separated survival in the pooled external validation cohort (HR $= `r km_desurv$hr`$, 95\% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:val}B); per-cohort Kaplan--Meier analyses are provided in the Supplementary Information. Prediction alone therefore did not distinguish DeSurv from higher-rank NMF, and, as we show next, comparable prediction is also achieved by conventional survival-supervised models. We therefore asked whether DeSurv instead provided a different biological representation of the prognostic signal, beyond that available from established classifiers and other survival-supervised approaches.

```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC cohorts (pooled $n=", n_val_total, "$ patients across ", n_val_cohorts, " cohorts: ", paste(sprintf("%s $n=%d$", gsub("AU seq", "AU RNA-seq", gsub("PACA AU", "PACA-AU", gsub("_", " ", names(table(pool_df$dataset))))), as.integer(table(pool_df$dataset))), collapse = "; "), "). (A) Forest plot of per-cohort univariate hazard ratios (95\\% CIs) for each factor; black diamonds denote inverse-variance-weighted pooled estimates. (B) Kaplan--Meier curves for pooled validation samples stratified by the training-derived multivariate DeSurv linear predictor, dichotomized at a cross-validated optimal cutpoint. (C) Correspondence ($|$Spearman $r|$) of tuned supervised risk scores (penalized Cox and supervised PCA) with the three DeSurv programs (D1--D3); supervised scores concentrate on D1 and are near-orthogonal to the D2 stromal program. No validation-cohort survival outcomes were used in training, scoring or hyperparameter selection. \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig4_tcgacptac.pdf"))
```

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(knitr); library(kableExtra)
# extract_pool_df_with_covariates() and data_val_filtered_tcgacptac
# are defined in the setup chunks above.
load_result("val_latent_desurv_alpha0_tcgacptac")

pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(hr = round(ci_row[, "exp(coef)"], 2),
         ci_lo = round(ci_row[, "lower .95"], 2),
         ci_hi = round(ci_row[, "upper .95"], 2),
         p = coef_row[, "Pr(>|z|)"])
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

pool_desurv_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv_t1 <- pooled_cox_unadj_adj(pool_desurv_t1)
pool_nmf7_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7_t1 <- pooled_cox_unadj_adj(pool_nmf7_t1)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)", "Std NMF ($k = 7$, BO $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv_t1$unadj), fmt_hr(stats_nmf7_t1$unadj)),
  Adjusted = c(fmt_hr(stats_desurv_t1$adj), fmt_hr(stats_nmf7_t1$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per SD (95\\% CI), $P$",
                       "Adjusted HR per SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = paste0("Pooled external validation hazard ratios (per SD of the linear predictor) for DeSurv at $k = 3$ and standard NMF at BO-selected $k = 7$ ($\\alpha = 0$), across the five validation cohorts ($n = ", n_val_total, "$ patients, ", n_val_events, " events). Adjusted models include the PurIST tumor-intrinsic and DeCAF stromal molecular classifiers as covariates, with stratification by dataset. Both methods retain significant prognostic value after adjustment. \\label{tab:hr-adjusted}")) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

```{r d1-partition-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("d1_variance_partition_stats")
.d1p <- d1_variance_partition_stats
.d1pf <- function(p) if (p < 1e-4) "< 0.0001" else sprintf("= %.4f", p)
load_result("supervised_recovery_stats")
.d1p_sr_mean  <- supervised_recovery_stats$sup_mean_score_cor
.d1p_sr_unsup <- abs(supervised_recovery_stats$table$score_cor_D1[!supervised_recovery_stats$table$supervised])
```

## The survival-aligned structure is reproducible and complementary to established PDAC classifiers {.unnumbered}

We next asked whether the dominant D1 direction simply recapitulated established PDAC subtype classifications. Across the pooled validation cohorts ($n = `r .d1p$meta$n`$, `r .d1p$meta$n_events` events), binary PurIST and DeCAF classifications together explained `r sprintf("%.0f", 100*.d1p$variance$explained_total)`\% of the within-cohort variance in the continuous D1 score, indicating substantial alignment with known tumor-intrinsic and stromal subtypes but incomplete overlap. D1 remained associated with survival after simultaneous adjustment for both classifiers (HR per SD $= `r sprintf("%.2f", .d1p$prognosis$d1_adj_hr)`$, 95\% CI $`r sprintf("%.2f", .d1p$prognosis$d1_adj_lo)`$--$`r sprintf("%.2f", .d1p$prognosis$d1_adj_hi)`$, $P `r .d1pf(.d1p$prognosis$d1_adj_p)`$); adding D1 to a stratified Cox model containing PurIST and DeCAF improved model fit (partial likelihood-ratio $\chi^2 = `r sprintf("%.1f", .d1p$prognosis$lr_chisq)`$ on 1 df, $P `r .d1pf(.d1p$prognosis$lr_p)`$) and concordance from `r sprintf("%.3f", .d1p$prognosis$c_base)` to `r sprintf("%.3f", .d1p$prognosis$c_with_d1)`. The DeSurv linear predictor also remained prognostic after adjustment for standard clinical covariates and, where adequately powered, for tumor purity (Supplementary Table \ref{tab:clinical-adjustment}). Thus, the survival-aligned representation overlaps established tumor and stromal classifications without being reducible to them.

The dominant D1 direction was also not specific to the DeSurv optimizer. Three independent survival-supervised approaches---supervised principal components, Cox partial least squares and sparse Cox---recovered patient scores strongly correlated with D1 in the training cohort ($|r| = 0.81$, $0.83$ and $0.85$, respectively), whereas the leading unsupervised principal component was nearly orthogonal to D1 ($|r| = `r sprintf("%.2f", .d1p_sr_unsup)`$) and instead tracked higher-variance D2/D3 structure (Supplementary Information, supervised-method recovery). Thus, D1 reflects a reproducible survival-associated expression direction shared across supervised approaches rather than an idiosyncrasy of DeSurv.

Having established D1 as a reproducible survival-associated direction, we asked what DeSurv adds beyond recovering it. Independently tuned supervised principal components and penalized Cox achieved pooled external concordance comparable to DeSurv (C-index $`r .svbt("Supervised PCA")`$, $`r .svbt("Sparse Cox")`$ and $`r .svbt("DeSurv full LP")`$, respectively) and captured the dominant D1-associated prognostic direction, but were nearly orthogonal to the proCAF-associated D2 program ($|r| = 0.06$ and $0.14$; Fig. \ref{fig:val}C). Their selected genes consistently enriched basal-like tumor but not proCAF programs; although their risk scores retained modest stromal signal, they did not resolve it as a distinct stromal program (Supplementary Information). By jointly preserving transcriptomic reconstruction while allowing survival to reshape the learned basis, DeSurv instead organized the transcriptomic context around the dominant prognostic direction into distinct classical/restCAF-like, proCAF-like and basal-like programs. This organization is consistent with DeSurv retaining major expression structure while reorganizing it around prognosis, rather than optimizing prediction or reconstruction alone. Importantly, it converts a single prognostic direction into biologically attributable tumor and stromal programs that can be independently validated and interrogated.


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

We used simulations with known latent structure to define when survival supervision alters program recovery. In the primary scenario, the gene program driving survival contributed less transcriptomic variation than outcome-neutral programs ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies), so that the strongest sources of expression variation were not the most prognostically relevant. DeSurv and standard NMF ($\alpha = 0$) were tuned using the same cross-validation procedure, isolating the effect of survival supervision during factorization.

Under this setting, DeSurv more accurately recovered the genes defining the true prognostic program, whereas standard NMF largely followed variance-dominant structure (Fig.\ \ref{fig:sim}A--B). This distinction is important because the genes assigned to a recovered program---not only its ability to rank patients by risk---determine its subsequent biological interpretation and its potential use in candidate signatures or biomarker development. Across 100 replicates, median test-set C-index was $`r desurv_sim_median`$ for DeSurv versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$), and DeSurv recovered the true factorization rank ($k = 3$) more frequently (Fig.\ \ref{fig:sim}C).

The advantage depended on the relationship between transcriptomic variance and prognosis. It attenuated when prognostic and variance-dominant structure partially overlapped and disappeared under the null, where neither method identified prognostic signal (Supplementary Fig.\ \ref{fig:fig-sim-null-mixed}). In the mixed setting, the advantage remained strongest for recovery of the factor-specific marker genes rather than shared background survival genes, supporting a specific effect of supervision on identifying the genes that define the prognostic program. Thus, survival supervision is most useful for biological discovery when clinically relevant programs are not themselves the dominant sources of transcriptomic variation.

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Survival supervision improves recovery of prognostic gene programs in simulations and PDAC training data. (A) Test-set concordance across 100 simulation replicates comparing DeSurv with standard NMF ($\\alpha = 0$; $p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; paired Wilcoxon $P ", sim_wilcox_p, "$). (B) Precision of recovered prognostic gene programs (overlap between each true program and the top-weighted genes of its best-matching learned factor). In (A) and (B), boxes show median and interquartile range; whiskers extend to $1.5\\times$IQR; points are individual replicates. (C) Distribution of selected ranks across replicates for DeSurv versus standard NMF. (D) The same gene-program recovery metric as in (B), computed for two signal regimes: a prognostic program that explains low transcriptomic variance (separated from the dominant variance) versus one that partially overlaps the dominant variance. DeSurv recovers the prognostic program in both regimes, whereas standard NMF, which follows variance, does not; the null scenario (no prognostic program) is shown in the Supplementary Information. \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

knitr::include_graphics(file.path("..", "figures", "fig2_tcgacptac.pdf"))

```

