<!-- REVISED RESULTS — prediction-validation structure (originally for PNAS; retained for Nature submission) -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- SI Appendix figure numbering (supplement.Rmd order, post-reorganization):
     S1  fig-converge          convergence trajectories
     S2  fig-sim-null-mixed    null and mixed simulation scenarios
     S3  fig-cindex-by-k       C-index vs K sensitivity
     S4  fig-nmf-diagnostics   standard NMF rank selection heuristics (demoted from old Fig 2A-C)
     S5  fig-cutpoint-km       cutpoint selection and KM curves
     S6  fig-subtype-overlap   subtype composition of risk groups
     S7  fig-desurv-k5-heatmap DeSurv gene overlap at k=5
     S8  fig-nmf-k5-heatmap    NMF gene overlap at k=5
     S9  fig-nmf-k7-heatmap    NMF gene overlap at k=7
     S10 fig-varsurvival-k5    variance-survival at k=5
     Note: Former S7 (variance-survival), S8 (W-corr), S9 (forest) are now
           main text Fig 3C-D and Fig 4A respectively.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(magick)
library(gt)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)
desurv_maxvar_label <- paste0("D", desurv_maxvar$factor)
nmf_maxvar_label    <- paste0("N", nmf_maxvar$factor)
# Exocrine NMF factor (N2) — referenced explicitly because of its biological
# role as the high-variance exocrine-associated factor
nmf_n2              <- nmf_vs[nmf_vs$factor == 2, ]
nmf_n2_pct          <- round(nmf_n2$variance_explained * 100, 1)
nmf_n2_dl           <- round(nmf_n2$delta_loglik, 2)

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Pooled adjusted (PurIST + DeCAF) statistics for DeSurv ───────────────────
load_result("data_val_filtered_tcgacptac")

extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(sample_id = rownames(surv), time = surv$time, event = surv$event,
                     dataset = entry$dataset, risk_score = entry$risk_score,
                     stringsAsFactors = FALSE)
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo; si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

pool_df_adj_complete <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
pool_df_adj_complete <- pool_df_adj_complete[
  complete.cases(pool_df_adj_complete[, c("PurIST", "DeCAF")]), ]
rs_sd_adj <- sd(pool_df_adj_complete$risk_score, na.rm = TRUE)
if (is.na(rs_sd_adj) || rs_sd_adj < 1e-12) rs_sd_adj <- 1
pool_df_adj_complete$risk_z <- pool_df_adj_complete$risk_score / rs_sd_adj
cox_adj_sum <- summary(coxph(
  Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
  data = pool_df_adj_complete))
hr_adj      <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "exp(coef)"])
ci_adj_lo   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "lower .95"])
ci_adj_hi   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "upper .95"])
p_adj_val   <- cox_adj_sum$coefficients["risk_z", "Pr(>|z|)"]
p_adj_text  <- if (p_adj_val < 0.001) "< 0.001" else paste0("= ", signif(p_adj_val, 2))

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

# ── KM dichotomized risk-group statistics ─────────────────────────────────────
# Extract HR/CI/p from the annotation layer embedded in the KM plot objects
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# Extract exact risk-group counts from KM risk tables (time = 0)
km_n_high_desurv <- fig_median_survival_desurv_tcgacptac$data.survtable$n.risk[
  fig_median_survival_desurv_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_desurv_tcgacptac$data.survtable$strata == "factor=1"]
km_n_high_nmf <- fig_median_survival_std_desurvk_tcgacptac$data.survtable$n.risk[
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$strata == "factor=1"]
km_reclass_diff <- km_n_high_desurv - km_n_high_nmf
```

## Model Overview {.unnumbered}

We developed DeSurv, a survival-supervised deconvolution framework that learns two linked quantities from bulk transcriptomes paired with patient outcomes: gene programs and their survival associations (Fig.\ \ref{fig:schema}A, Methods). DeSurv combines nonnegative matrix factorization with Cox proportional hazards modeling so that each gene program is shaped by both reconstruction of the expression matrix and association with survival. The survival objective acts on the gene programs but not on the sample loadings, allowing the learned programs to be applied to new samples by projection without retraining. A supervision parameter controls the balance between reconstruction and survival, and both this parameter and the factorization rank are selected by Bayesian optimization on cross-validated concordance (Fig.\ \ref{fig:schema}B, Methods).

Applied to the PDAC training cohorts from TCGA [@raphael2017integrated] and CPTAC [@ellis2013clinical] (Materials and Methods), cross-validated concordance varied across candidate ranks but plateaued from $k = 3$ through $k = 12$ (SI Appendix, Fig.\ S3), indicating that three supervised factors capture nearly as much prognostic information as twelve. Within this plateau, joint Bayesian optimization over rank and supervision strength with a one-standard-error rule (Methods) accordingly selected the parsimonious $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ at $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$. Peak cross-validated C-index was $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$ across $n = `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)`$ patients with $`r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)`$ events. Standard NMF rank-selection diagnostics (reconstruction residuals, cophenetic correlation, silhouette width) [@frigyesi2008nmf; @brunet2004metagenes] did not converge on a consistent rank in PDAC (SI Appendix, Fig.\ S4), motivating the supervised approach.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "model_schematic_final.pdf"))
```


## Survival supervision and standard NMF produce different factor structures in PDAC {.unnumbered}

To examine how survival supervision shapes the recovered factor structure, we fit standard NMF at the same rank as DeSurv ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$, selected above) and examined the overlap between each method's factor-specific gene rankings and established PDAC gene programs (Fig.\ \ref{fig:pdac}A--B). Matching $k$ across methods isolates the effect of survival supervision on factor content from differences in model complexity; a sensitivity analysis at NMF's independently selected ranks (elbow $k = 5$, BO $k = 7$) is provided in SI Appendix, Table S2. We first asked which malignant, stromal, immune and other microenvironmental programs were represented in each factorization, and then tested whether these structural differences corresponded to differences in survival contribution.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. In the DeSurv factorization, D1 correlated with Classical tumor programs and restCAF-like stromal signatures, D2 with proCAF-like programs, and D3 with Basal-like tumor programs (Fig. \ref{fig:pdac}A). Among the programs detected, DeSurv organized the dominant survival-associated variation along a tumor--stroma axis, with D1 representing the Classical/restCAF-like low-risk pole and increasing risk reflecting movement toward Basal-like and proCAF-like structure, consistent with evidence that survival stratification in PDAC improves when tumor-intrinsic and CAF subtypes are considered jointly [@peng2024determination; @elyada2019cross]. No DeSurv factor showed a significant positive correlation with exocrine-associated gene programs (DECODER Exocrine, Collisson Exocrine, Bailey aberrantly differentiated endocrine--exocrine (ADEX); Fig. \ref{fig:pdac}A). The survival contributions of these factors and their relationship to reconstruction of the expression matrix are quantified below.

Standard NMF produced a different factor organization (Fig. \ref{fig:pdac}B). One factor (N2) was strongly correlated with exocrine-associated gene programs and negatively correlated with immune and stromal signatures (Fig. \ref{fig:pdac}B). A second factor (N1) correlated with malignant-epithelial (tumor) identity, carrying both Classical and Basal-like markers (Fig. \ref{fig:pdac}B). A third factor (N3) correlated with immune, fibroblast, and extracellular matrix (ECM)-related programs (including Puleo Desmoplastic, SCISSORS proCAF and restCAF [@leary2023s], Maurer Immune and ECM [@Maurer2019], and DECODER Activated Stroma), aggregating these into a single factor rather than separating them as DeSurv does (D1 and D2; Fig. \ref{fig:pdac}A). At $k = 3$, NMF thus devotes one of its three factors to exocrine-associated variation, leaving two factors to capture all tumor-intrinsic and microenvironmental programs. Whether these structural differences correspond to differences in prognostic information is quantified next.

NMF seeks a low-rank \emph{reconstruction} of the transcriptome; many factorizations reconstruct comparably well, and the unsupervised objective gives no incentive to prefer a prognostically informative direction. We therefore quantified per-factor contribution using two metrics (Fig. \ref{fig:pdac}C; SI Appendix, Sections 10--11): each factor's contribution to reconstruction (the normalized reconstruction Shapley value, which partitions the model's reconstruction and sums to $100\%$) and its survival contribution (the change in Cox partial log-likelihood $\Delta\ell$ when that factor is dropped from a $k$-factor Cox model). NMF distributed reconstruction across its factors in proportions unrelated to prognosis: all three NMF factors had negligible $\Delta\ell$ despite spanning a wide range of reconstruction share, while DeSurv concentrated nearly all survival contribution into D1. Specifically, D1 contributed $`r desurv_maxsurv_pct`$% of reconstruction with $\Delta\ell = `r desurv_maxsurv_dl`$, while the largest-reconstruction NMF factor (`r nmf_maxvar_label`) contributed $`r nmf_maxvar_pct`$% of reconstruction but only $\Delta\ell = `r nmf_maxvar_dl`$ to survival. The exocrine-associated NMF factor (N2) contributed $`r nmf_n2_pct`$% of reconstruction yet only $\Delta\ell = `r nmf_n2_dl`$ to survival. The largest-reconstruction DeSurv factor (`r desurv_maxvar_label`, $`r desurv_maxvar_pct`$% of reconstruction) likewise contributed minimally to survival after accounting for D1. Thus, in PDAC the factors that contribute most to reconstruction are not the most prognostic, consistent with tumor purity analyses across cancer types [@aran2015systematic].

To further characterize how the two factorizations relate, we examined pairwise W-matrix gene loading correlations over all 1,970 genes (Fig. \ref{fig:pdac}D; Spearman over the full loading profile, since a top-gene subset would overstate preservation). At $k = 3$, NMF captures the entire malignant-epithelial compartment in a single factor (N1), which we therefore label ``tumor'' rather than Classical; N1 carries both Classical and Basal markers and maps strongly onto DeSurv's Basal-like tumor factor (N1$\leftrightarrow$D3 Spearman $0.84$), a tumor$\leftrightarrow$tumor correspondence, while NMF's microenvironmental factor maps onto DeSurv's proCAF stroma (N3$\leftrightarrow$D2 $0.87$). DeSurv's most prognostic factor (D1) is essentially uncorrelated with all three NMF factors (entire D1 column $\le 0.36$): because D1 couples Classical tumor and restCAF-like stromal signatures into a single survival-aligned program, its structure is novel relative to NMF rather than a reweighting of an existing factor (SI Appendix, Table~\ref{tab:projection-r2}). NMF's exocrine factor (N2) has no strong DeSurv counterpart (row capped at $0.61$), consistent with its dissolution under supervision. In short, NMF lumps the tumor compartment into one factor, whereas DeSurv resolves it into Classical- (D1) and Basal-aligned (D3) programs and constructs the survival-aligned tumor--stroma coupling (D1) that NMF does not isolate. Whether this translates to improved generalization across independent cohorts is tested below.
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (A--B) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$. Only $|r| > 0.2$ shown. DeSurv concentrates survival signal into D1 (Classical tumor + restCAF-like stroma) while suppressing exocrine variation; standard NMF devotes one factor to exocrine signals. (C) Per-factor contribution to reconstruction of the expression matrix (normalized reconstruction Shapley share, which partitions and sums to 100\\%) versus survival contribution ($\\Delta$ Cox partial log-likelihood upon factor removal). DeSurv concentrates survival signal into D1 while NMF's largest-reconstruction factors --- including the exocrine-associated factor N2 --- contribute negligibly to survival. (D) Pairwise correspondence between NMF and DeSurv factors, measured by Spearman rank correlation of W-matrix gene loadings over all genes. \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.height= "7in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig3_tcgacptac.pdf"))
```


## Survival-aligned programs generalize across independent PDAC cohorts with lower complexity {.unnumbered}

A prognostic factorization is useful only if it generalizes beyond the training cohort. Having established that survival supervision and standard NMF produce different factor structures in the training data, we next tested whether each method's structure transfers to independent cohorts without retraining. The five external validation cohorts (Dijk, Moffitt, PACA-AU array and RNA-seq, and Puleo) likewise consist of non-metastatic, treatment-naive specimens (SI Appendix, Table S1). We computed factor scores in each validation dataset by projecting the gene programs learned in the training cohort ($Z = \widetilde{W}^\top X_{\text{new}}$), yielding a continuous score for each factor per sample.

External validation C-index for NMF at independently selected ranks is summarized in SI Appendix, Table S2. Standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ substantially improved over NMF at $k = 3$, with $k = 7$ matching or exceeding DeSurv's per-cohort C-index in most cohorts, though only when DeSurv's tuning framework is applied ($\alpha = 0$ in the joint BO over rank and hyperparameters), since NMF's own rank-selection heuristics (cophenetic, silhouette, residual; SI Appendix, Fig.\ S4) do not converge on this rank. However, this concordance gain still requires more than twice as many factors. DeSurv at $k = 3$ retains a significant adjusted hazard ratio after adjustment for PurIST and DeCAF (per-SD HR = `r hr_adj`, 95\% CI `r ci_adj_lo`--`r ci_adj_hi`, $P `r p_adj_text`$; Table \ref{tab:hr-adjusted}), achieving independent prognostic value more parsimoniously. When $\beta$ coefficients are re-fit on each validation cohort, DeSurv's three-factor decomposition is BIC-preferred over NMF's seven-factor decomposition in three of five cohorts, with the remaining two essentially tied (SI Appendix, Table S4). Across `r n_val_cohorts` independent PDAC cohorts ($n = `r n_val_total`$), the DeSurv linear predictor showed a consistent survival association (stratified Cox model: pooled HR per SD $`r hr_pool`$; 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$; $P = `r p_pool`$; per-cohort forest plot in Fig. \ref{fig:val}A; KM curves in SI Appendix, Fig. S5C--F). At the matched rank ($k = 3$), the analogous NMF linear predictor showed a weaker pooled effect (HR per SD $`r hr_pool_nmf`$; 95% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$; $P = `r p_pool_nmf`$); a direct comparison using dichotomized risk groups is presented below.

To assess clinical interpretability of these continuous associations, we dichotomized validation samples into high- and low-risk groups using method-specific cutpoints, each independently optimized via cross-validation on the training data. The DeSurv linear predictor separated survival trajectories in the pooled validation cohort (HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:val}B). The same procedure applied to NMF at $k = 3$ yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:val}C). The DeSurv cutpoint classified a larger validation subset as high risk (`r km_n_high_desurv` vs `r km_n_high_nmf` high-risk patients; `r km_reclass_diff` reclassified) while preserving strong survival separation. The DeSurv high-risk group was enriched for both established poor-prognosis subtypes (Basal-like by PurIST, proCAF by DeCAF; Fisher's exact, SI Appendix, Fig. S6), capturing established consensus biology while adding independent prognostic information. The same analysis applied to NMF risk groups at $k = 3$ showed Basal-like enrichment but no significant proCAF enrichment (SI Appendix, Fig. S6), consistent with NMF's failure to isolate the tumor--stroma coupling that underlies D1. Together, these results indicate that survival supervision yields a compact, biologically interpretable prognostic structure that generalizes across independent PDAC cohorts; DeSurv identifies a lower-dimensional tumor--stroma coupling program with independent prognostic value beyond PurIST and DeCAF, whereas higher-rank NMF models can achieve similar per-cohort concordance only at substantially greater complexity without recovering the same coupling axis.

```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC cohorts (pooled $n=", n_val_total, "$ patients across ", n_val_cohorts, " cohorts: ", paste(sprintf("%s $n=%d$", gsub("AU seq", "AU RNA-seq", gsub("PACA AU", "PACA-AU", gsub("_", " ", names(table(pool_df$dataset))))), as.integer(table(pool_df$dataset))), collapse = "; "), "). (A) Forest plot of univariate hazard ratios (95\\% CIs) from separate Cox models for each factor, showing each factor's marginal survival association in each validation cohort. DeSurv factor D1 shows a consistent protective association across all five cohorts (pooled HR = 0.76), while no standard NMF factor achieves a comparable pooled effect. Error bars are 95\\% confidence intervals; black diamonds denote inverse-variance-weighted pooled estimates. (B--C) Kaplan--Meier curves for pooled external validation samples stratified by the combined multivariate linear predictor ($\\hat{\\beta}_1 Z_1 + \\hat{\\beta}_2 Z_2 + \\hat{\\beta}_3 Z_3$, where coefficients are learned from training data), dichotomized at a cross-validated optimal cutpoint: DeSurv (B) and standard NMF (C). \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig4_tcgacptac.pdf"))
```

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(knitr); library(kableExtra)
# extract_pool_df_with_covariates() and data_val_filtered_tcgacptac
# are defined in the setup chunks above.
load_result("val_latent_desurv_alpha0_tcgacptac")

pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(hr = round(ci_row[, "exp(coef)"], 2),
         ci_lo = round(ci_row[, "lower .95"], 2),
         ci_hi = round(ci_row[, "upper .95"], 2),
         p = coef_row[, "Pr(>|z|)"])
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

pool_desurv_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv_t1 <- pooled_cox_unadj_adj(pool_desurv_t1)
pool_nmf7_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7_t1 <- pooled_cox_unadj_adj(pool_nmf7_t1)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)", "Std NMF ($k = 7$, BO $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv_t1$unadj), fmt_hr(stats_nmf7_t1$unadj)),
  Adjusted = c(fmt_hr(stats_desurv_t1$adj), fmt_hr(stats_nmf7_t1$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per SD (95\\% CI), $P$",
                       "Adjusted HR per SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = paste0("Pooled external validation hazard ratios (per SD of the linear predictor) for DeSurv at $k = 3$ and standard NMF at BO-selected $k = 7$ ($\\alpha = 0$), across the five validation cohorts ($n = ", n_val_total, "$ patients, ", n_val_events, " events). Adjusted models include the PurIST tumor-intrinsic and DeCAF stromal molecular classifiers as covariates, with stratification by dataset. Both methods retain significant prognostic value after adjustment, with DeSurv's larger HR drop reflecting biological concordance with established subtypes and NMF $k = 7$'s smaller drop reflecting orthogonal residual content from its additional, less interpretable factors. \\label{tab:hr-adjusted}")) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

To test whether the recovered prognostic structure extends to treated disease, we projected the trained gene-program matrix onto two independent bulk-tissue treated PDAC cohorts with overall survival (RASH-ACCEPT, metastatic, $n = 155$; Linehan, borderline-resectable/locally advanced, $n = 66$), both of which --- being bulk tissue --- permit the full PurIST + DeCAF adjustment used in treatment-naive validation (Table~\ref{tab:hr-adjusted}). Results were cohort-dependent. In RASH-ACCEPT, the DeSurv linear predictor remained independently prognostic after PurIST + DeCAF adjustment (HR per SD $= 1.25$, $95\%$ CI $1.02$--$1.52$, $P = 0.033$), as did the D1 program (HR $= 0.81$, $P = 0.038$), although their marginal effects were not significant --- a suppression pattern in which the established classifiers partially mask the DeSurv signal. In Linehan, the D1 program was prognostic marginally (HR $= 0.59$, $P = 0.03$) but its effect was absorbed by joint adjustment. The proCAF stromal program D2 was independent of both classifiers in both cohorts (RASH-ACCEPT HR $= 0.74$, $P = 0.010$; Linehan HR $= 0.51$, $P = 0.002$). These treated bulk cohorts thus provide partial, cohort-dependent evidence that the DeSurv programs carry prognostic information beyond PurIST and DeCAF in treated disease (SI Appendix, treated-cohort translational analysis).


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

The cross-cohort generalization above is corroborated by simulations with known ground-truth latent structure. To test whether survival supervision improves gene program recovery under nonnegative constraints, we designed simulations in which prognostic gene programs explained low variance relative to outcome-neutral background signals, with survival times generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies). To test whether survival information should enter during factorization rather than only afterward, we compared DeSurv to standard NMF ($\alpha=0$), where factors are learned by reconstruction error alone and survival associations are estimated by a separately fit Cox model. Both methods were tuned via BO over cross-validated C-index, so that differences in performance reflect the role of supervision during factorization, not during model selection.

In the primary scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index (Fig.\ \ref{fig:sim}A) and substantially higher precision for recovering the true prognostic gene programs (Fig.\ \ref{fig:sim}B). Here precision denotes the fraction of a factor's top-weighted genes (those with the largest factor specificity score $s_{ij}$, where $n_{\text{top}}$, the number of factor-specific genes entering the survival linear predictor, is chosen by BO; SI Appendix, Sections 6 and 9) that overlap with a true prognostic program's gene set. The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This confirms that outcome supervision during factorization is necessary to recover prognostic programs that do not dominate expression variance. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$). The precision advantage exceeded the C-index advantage, reflecting that unsupervised factorization can match predictive performance while assigning different genes to each factor; survival supervision constrains gene-program identity itself. A similar decoupling appears empirically in PDAC at $k = 5$ (NMF's elbow-selected rank), where supervised (DeSurv) and unsupervised (NMF) factorizations yield substantially different factor-specific top genes (SI Appendix, Fig.\ S7 [DeSurv $k = 5$] vs Fig.\ S8 [NMF $k = 5$]), with similar C-indices in external cohorts.

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under a null scenario ($\beta = 0$) and a mixed scenario with partial overlap between variance-dominant and prognostic structure (SI Appendix, Fig.\ S2). DeSurv's benefit scaled with the degree to which prognostic programs were masked by outcome-neutral variation: largest in the primary scenario ($\Delta = `r delta_sim_cindex`$), attenuated when variance and prognosis partially overlapped ($\Delta = `r delta_mixed_cindex`$, paired Wilcoxon $P `r mixed_wilcox_p`$), and absent when no survival signal existed, where both methods yielded C-index near 0.5. DeSurv also recovered the true rank ($k = 3$) more frequently than standard NMF, which systematically under-selected near $k = 2$ (Fig.\ \ref{fig:sim}C).

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Survival supervision improves recovery of true prognostic gene programs in controlled simulations. (A) Test-set concordance index across 100 simulation replicates ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$) comparing DeSurv to standard NMF ($\\alpha = 0$); paired Wilcoxon $P ", sim_wilcox_p, "$. (B) Precision (fraction of a factor's top-weighted genes, defined in main text, overlapping a true prognostic program) across replicates. In (A) and (B), boxes show median and interquartile range (IQR); whiskers extend to $1.5\\times$IQR; points beyond are individual replicates. The y-axis in (A) is truncated at 0.5 because C-index values below 0.5 indicate worse-than-random concordance. (C) Distribution of selected $k$ values across the same 100 replicates for DeSurv versus standard NMF. (D) Gaussian-process--predicted mean cross-validated C-index from Bayesian optimization over factorization rank ($k$) and supervision strength ($\\alpha$) in PDAC training data (TCGA + CPTAC), showing the broad plateau spanning $k = 3$--$12$ from which the one-standard-error rule selects $k = 3$ (Methods). \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

knitr::include_graphics(file.path("..", "figures", "fig2_tcgacptac.pdf"))

```

