<!-- REVISED RESULTS — prediction-validation structure for PNAS -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- SI Appendix figure numbering (supplement.Rmd order, post-reorganization):
     S1  fig-converge        convergence trajectories
     S2  fig-sim-null-mixed  null and mixed simulation scenarios
     S3  fig-cindex-by-k     C-index vs K sensitivity
     S4  fig-nmf-diagnostics standard NMF rank selection heuristics (demoted from old Fig 2A-C)
     S5  fig-k3-k7-nesting   K=3 vs K=7 factor correspondence
     S6  fig-cutpoint-km     cutpoint selection and KM curves
     S7  fig-subtype-overlap subtype composition of risk groups
     S8  fig-nmf-k5-heatmap  NMF gene overlap at k=5
     S9  fig-nmf-k7-heatmap  NMF gene overlap at k=7
     S10 fig-varsurvival-k5  variance-survival at k=5
     Note: Former S7 (variance-survival), S8 (W-corr), S9 (forest) are now
           main text Fig 3C-D and Fig 4A respectively.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(enrichplot)
library(magick)
library(Seurat)
library(gt)
library(VAM)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

alt_plots = lapply(alt_plots, set_fig_font, size = 10)
null_plots = lapply(null_plots, set_fig_font, size = 10)

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
mixed_plots = lapply(mixed_plots, set_fig_font, size = 10)
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
sim_wilcox_p <- signif(wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                    paired = TRUE)$p.value, 2)

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
mixed_wilcox_p <- signif(wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                      paired = TRUE)$p.value, 2)

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- desurv_stats$hr
ci_pool_lo   <- desurv_stats$ci_lo
ci_pool_hi   <- desurv_stats$ci_hi
p_pool       <- desurv_stats$p

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- nmf_stats$hr
ci_pool_lo_nmf <- nmf_stats$ci_lo
ci_pool_hi_nmf <- nmf_stats$ci_hi
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

# ── KM dichotomized risk-group statistics ─────────────────────────────────────
# Extract HR/CI/p from the annotation layer embedded in the KM plot objects
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# Extract exact risk-group counts from KM risk tables (time = 0)
km_n_high_desurv <- fig_median_survival_desurv_tcgacptac$data.survtable$n.risk[
  fig_median_survival_desurv_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_desurv_tcgacptac$data.survtable$strata == "factor=1"]
km_n_high_nmf <- fig_median_survival_std_desurvk_tcgacptac$data.survtable$n.risk[
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$strata == "factor=1"]
km_reclass_diff <- km_n_high_desurv - km_n_high_nmf
```

## Model Overview {.unnumbered}

We developed DeSurv, a survival-supervised deconvolution framework (Fig. \ref{fig:schema}; Methods), and evaluated it in controlled simulations and PDAC cohort analysis with external validation.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..","figures", "model_schematic_final.pdf"))
```


## Survival supervision clarifies NMF rank selection {.unnumbered}

We first asked whether incorporating survival outcomes resolves the instability of unsupervised rank selection heuristics [@frigyesi2008nmf; @brunet2004metagenes]. Using gene expression data from the PDAC training cohorts (TCGA [@raphael2017integrated] and CPTAC [@ellis2013clinical]; Methods), we evaluated commonly used unsupervised criteria across a range of candidate ranks.

Standard NMF diagnostics (reconstruction residuals, cophenetic correlation, mean silhouette width) yielded inconsistent guidance across candidate ranks (SI Appendix, Fig. S4).

Because the goal is prognostic factorization, DeSurv selects $k$ using the criterion that directly measures prognostic performance: cross-validated concordance index (C-index), optimized jointly with supervision strength $\alpha$. The resulting C-index surface identified a coherent region of high performance (Fig. \ref{fig:sim}D). Model selection followed the one-standard-error rule (Methods): we selected the smallest $k$ whose predicted performance lay within one standard error of the maximum, yielding a parsimonious choice. Applying the 1-SE rule to the Bayesian optimization (BO) surface selected $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ and $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$ (peak C-index $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$; $n = `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)`$ patients, $`r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)`$ events). Survival-supervised model selection thus resolves the rank ambiguity that unsupervised heuristics leave open, yielding a single parsimonious factorization for downstream evaluation.

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = "Simulation studies and model selection. (A) Test-set concordance index across 100 simulation replicates ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$) comparing DeSurv to standard NMF ($\\alpha = 0$). (B) Precision (fraction of top-weighted genes in a learned factor overlapping a true prognostic program) across replicates. (C) Distribution of selected $k$ values for DeSurv versus standard NMF. (D) Gaussian process predicted mean cross-validated C-index from Bayesian optimization over factorization rank ($k$) and supervision strength ($\\alpha$) in PDAC training data (TCGA + CPTAC). \\label{fig:sim}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

load_result("fig_bo_heat_tcgacptac")
fig_bo_heat_tcgacptac = set_fig_font(fig_bo_heat_tcgacptac, size = 10) +
  guides(fill = guide_colorbar(barheight = unit(2, "cm"), barwidth = unit(0.3, "cm"))) +
  theme(legend.title = element_text(size = 7), legend.text = element_text(size = 7))
leg_grob = ggplotGrob(fig_bo_heat_tcgacptac)
guide_idx <- which(leg_grob$layout$name == "guide-box-right")
heat_legend <- leg_grob$grobs[[guide_idx]]

fig_bo_heat_tcgacptac = fig_bo_heat_tcgacptac + theme(legend.position = "none")
upper = plot_grid(
  alt_plots$cindex_box + labs(title = NULL) + scale_y_continuous(limits = c(.5, 1)),
  alt_plots$precision_box + labs(title = NULL),
  NULL,
  ncol = 3, labels = c("A", "B",""),
  rel_widths = c(4,4,.8)
)

lower = plot_grid(
  alt_plots$k_hist,
  fig_bo_heat_tcgacptac,
  heat_legend,
  ncol = 3, labels = c("C", "D",""),
  rel_widths = c(4,4,.8)
)

plot_grid(upper, lower, nrow = 2)

```


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

To test whether survival supervision improves gene program recovery under nonnegative constraints, we designed simulations with known ground-truth latent structure. Simulated expression matrices were generated from a nonnegative factor model in which prognostic gene programs explained low variance relative to outcome-neutral background signals, with survival times generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies). To test whether survival information should enter during factorization rather than only afterward, we compared DeSurv to standard NMF ($\alpha=0$), where factors are learned by reconstruction error alone and survival associations are estimated by a separately fit Cox model. Both methods were tuned via BO over cross-validated C-index, so that differences in performance reflect the role of supervision during factorization, not during model selection.

In the primary scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index (Fig. \ref{fig:sim}A) and substantially higher precision for recovering the true prognostic gene programs (Fig. \ref{fig:sim}B). Here precision denotes the fraction of top-weighted genes in a learned factor that overlap with a true prognostic program's gene set. The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This confirms that outcome supervision during factorization is necessary to recover prognostic programs that do not dominate expression variance. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P = `r sim_wilcox_p`$).

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under a null scenario ($\beta = 0$) and a mixed scenario with partial overlap between variance-dominant and prognostic structure (SI Appendix, Fig. S2). DeSurv's benefit scaled with the degree to which prognostic programs were masked by outcome-neutral variation: largest in the primary scenario ($\Delta = `r delta_sim_cindex`$), attenuated when variance and prognosis partially overlapped ($\Delta = `r delta_mixed_cindex`$, paired Wilcoxon $P = `r mixed_wilcox_p`$), and absent when no survival signal existed, where both methods yielded C-index near 0.5. DeSurv also recovered the true rank ($k = 3$) more frequently than standard NMF, which systematically under-selected near $k = 2$ (Fig. \ref{fig:sim}C).




## Survival supervision and standard NMF produce different factor structures in PDAC {.unnumbered}

BO selected $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ and $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$ for DeSurv in the PDAC training cohorts (TCGA and CPTAC). To compare the factor structures produced with and without survival supervision at the same dimensionality, we fit standard NMF at DeSurv's BO-selected rank ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$) and examined the overlap between each method's factor-specific gene rankings and established PDAC gene programs (Fig. \ref{fig:pdac}A--B). Matching ranks ensures that differences in factor content reflect the presence or absence of survival supervision rather than differences in model complexity; whether NMF's performance improves at higher ranks is evaluated in the sensitivity analysis below and in SI Appendix, Table S5. We first describe the biological identity of each method's factors, then ask whether these structural differences correspond to differences in survival contribution.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. In the DeSurv factorization, D1 correlated with both classical tumor programs and restCAF-associated (iCAF) stromal signatures, D2 with proCAF-associated (activated immune-stromal) programs, and D3 with basal-like tumor programs (Fig. \ref{fig:pdac}A). The co-occurrence of classical and restCAF signatures within a single factor is consistent with evidence that survival stratification in PDAC improves when tumor-intrinsic and CAF subtypes are considered jointly [@peng2024determination; @elyada2019cross]. No DeSurv factor showed a significant positive correlation with exocrine-associated gene programs (DECODER Exocrine, Collisson Exocrine, Bailey ADEX; Fig. \ref{fig:pdac}A). The survival contributions of these factors and their relationship to expression variance are quantified below (Fig. \ref{fig:pdac}C).

Standard NMF produced a different factor organization (Fig. \ref{fig:pdac}B). One factor (N2) was strongly correlated with exocrine-associated gene programs and negatively correlated with immune and stromal signatures (Fig. \ref{fig:pdac}B). A second factor (N1) correlated with classical tumor identity (Fig. \ref{fig:pdac}B). A third factor (N3) correlated with immune, fibroblast, and extracellular matrix-related programs (including Puleo Desmoplastic, SCISSORS panCAF and myCAF [@leary2023s], Maurer Immune and ECM [@Maurer2019], and DECODER Activated Stroma), aggregating these into a single factor rather than separating them as DeSurv does (D1 and D2; Fig. \ref{fig:pdac}A). At $k = 3$, NMF thus devotes one of its three factors to exocrine-associated variation, leaving two factors to capture all tumor-intrinsic and microenvironmental programs. Whether these structural differences correspond to differences in prognostic information is quantified next.

We quantified per-factor contribution using two metrics (Fig. \ref{fig:pdac}C; SI Appendix, Sections 10--11). For each factor, we computed the conditional variance explained (semi-partial $R^2$: the incremental fraction of total variance accounted for by that factor given all other factors), and its survival contribution, measured as the change in Cox partial log-likelihood ($\Delta\ell$) when that factor is dropped from a $k$-factor Cox model. All three NMF factors had negligible $\Delta\ell$ despite spanning a range of variance, while DeSurv concentrated nearly all survival contribution into D1. Specifically, D1 explained $`r desurv_maxsurv_pct`$% of variance with $\Delta\ell = `r desurv_maxsurv_dl`$, while the highest-variance NMF factor explained $`r nmf_maxvar_pct`$% of expression variance but contributed only $\Delta\ell = `r nmf_maxvar_dl`$ to survival. The highest-variance DeSurv factor (D3, $`r desurv_maxvar_pct`$% of variance) also contributed minimally to survival after accounting for D1. Thus, in PDAC the highest-variance factors are not the most prognostic, consistent with tumor purity analyses across cancer types [@aran2015systematic].

To further characterize how the two factorizations relate, we examined pairwise W-matrix gene loading correlations (Fig. \ref{fig:pdac}D). Two clear correspondences emerged: NMF's classical-associated factor (N1) mapped strongly to DeSurv's basal-like factor (D3), and NMF's microenvironmental factor (N3) mapped primarily to DeSurv's D2. Notably, DeSurv's most prognostic factor (D1) showed only weak, roughly uniform correlations with all three NMF factors, indicating that the gene program structure underlying D1 does not have a direct counterpart in the unsupervised factorization. Because D1 couples classical tumor and restCAF-associated stromal signatures into a single survival-aligned program, its signal is distributed across multiple NMF factors, none of which individually captures the same tumor--stroma coupling. NMF's exocrine-associated factor (N2) similarly lacked a strong DeSurv counterpart, with its strongest (though weak) correlation to D3, consistent with D3's high variance and low survival contribution. Together, the two methods recover overlapping tumor and microenvironmental structure but differ in two key respects: NMF dedicates a factor to exocrine-compositional variation that DeSurv suppresses, while DeSurv constructs a prognostic tumor--stroma factor that NMF does not isolate. The net result is that DeSurv concentrates survival contribution into a single factor ($\Delta\ell = `r desurv_maxsurv_dl`$), whereas no NMF factor achieves comparable survival contribution at the same rank. Whether this translates to improved generalization across independent cohorts is tested below.
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap="Survival supervision produces different factor structures in PDAC. (A--B) Spearman correlations between factor gene rankings and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$. Only $|r| > 0.2$ shown. DeSurv concentrates survival signal into D1 (classical tumor + restCAF stroma) while suppressing exocrine variation; standard NMF devotes one factor to exocrine signals. (C) Fraction of expression variance explained versus survival contribution ($\\Delta$ Cox partial log-likelihood upon factor removal) for each factor. DeSurv concentrates survival signal into D1 while the highest-variance NMF factor (N2, exocrine-associated) contributes negligibly to survival. (D) Pairwise correspondence between NMF and DeSurv factors, measured by Spearman rank correlation of W-matrix gene loadings across all genes. \\label{fig:pdac}", fig.env='figure*', fig.pos='t', out.height= "7in", out.width='\\textwidth'}

library(cowplot)
library(ggplot2)

# Panels A-B: gene program heatmaps
load_result("fig_gene_overlap_heatmap_desurv_tcgacptac")
load_result("fig_gene_overlap_heatmap_std_desurvk_tcgacptac")

# Panel C: variance vs survival (promoted from supplement)
load_result("fig_variation_explained_tcgacptac")

# Panel D: W-matrix correlation (promoted from supplement)
load_result("fig_desurv_std_correlation_tcgacptac")

# ── Panels A-B: heatmaps with shared legend ──
# Add top margin so panel labels (A, B) don't overlap heatmap titles
plot_3a <- fig_gene_overlap_heatmap_desurv_tcgacptac$plot +
  theme(plot.margin = margin(t = 14, r = 2, b = 2, l = 2))
plot_3b <- fig_gene_overlap_heatmap_std_desurvk_tcgacptac$plot +
  theme(plot.margin = margin(t = 14, r = 2, b = 2, l = 2))
legend_ab <- fig_gene_overlap_heatmap_desurv_tcgacptac$legend
legend_ab <- gtable::gtable_add_padding(
  legend_ab, padding = unit(c(0, 10, 0, 0), "pt")
)

top_row <- plot_grid(
  plot_3a, plot_3b, cowplot::ggdraw(legend_ab),
  ncol = 3, labels = c("A", "B", ""), align = "hv",
  label_size = 12, rel_widths = c(3.5, 3.5, 0.3)
)

# ── Panel C: variance vs survival scatterplot ──
plot_3c <- set_fig_font(fig_variation_explained_tcgacptac, size = 10) +
  theme(
    legend.position    = c(1, 0.5),
    legend.justification = c(1, 0),
    legend.background  = element_rect(color = "black"),
    axis.title         = element_text(size = 8),
    plot.margin        = margin(2, 30, 2, 2)
  )

# ── Panel D: W-matrix correlation ──
legend_d <- ggplot(
  data.frame(x = 0, y = seq(-0.5, 1, length.out = 100)),
  aes(x = x, y = y, fill = y)
) +
  geom_tile() +
  scale_fill_gradientn(
    colors = grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100),
    limits = c(-0.5, 1),
    breaks = c(-0.4, 0, 0.4, 0.8),
    name = "Spearman\ncorrelation"
  ) +
  guides(fill = guide_colorbar(
    barwidth  = unit(0.3, "cm"),
    barheight = unit(2, "cm"),
    title.position = "top",
    title.hjust = 0.5
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 6),
    legend.text  = element_text(size = 6)
  )
legend_d_grob <- cowplot::get_legend(legend_d)

plot_3d <- plot_grid(
  fig_desurv_std_correlation_tcgacptac$plot +
    theme(plot.margin = margin(2, 2, 2, 20)),
  plot_grid(
    NULL,
    cowplot::ggdraw(legend_d_grob),
    nrow = 2, rel_heights = c(0.08, 0.92)
  ),
  ncol = 2, rel_widths = c(4, 1)
)

bottom_row <- plot_grid(
  plot_3c, plot_3d,
  ncol = 2, labels = c("C", "D"),
  label_size = 12, rel_widths = c(0.55, 0.45)
)

plot_grid(top_row, bottom_row, nrow = 2, rel_heights = c(1.3, 0.7))
```


## Survival-aligned programs generalize across independent PDAC cohorts {.unnumbered}

A prognostic factorization is useful only if it generalizes beyond the training cohort. Having established that survival supervision and standard NMF produce different factor structures in the training data, we next tested whether DeSurv's structure transfers to independent cohorts without retraining. We computed factor scores in each validation dataset by projecting the gene programs learned in the training cohort ($Z = \widetilde{W}^\top X_{\text{new}}$), yielding a continuous score for each factor per sample.

For DeSurv, we evaluated D1 (the classical--restCAF co-program with the largest training-data $\Delta\ell$); for standard NMF at the same rank, we selected the factor with the largest $\Delta\ell$ (N1, the classical-associated factor). External validation C-index for NMF at independently selected ranks is summarized in SI Appendix, Table S5. Standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ substantially improved over NMF at $k = 3$, with $k = 7$ matching or exceeding DeSurv's per-cohort C-index in most cohorts. However, this concordance gain requires more than twice as many factors and produces a fragmented structure whose prognostic signal largely overlaps with existing molecular classifiers (SI Appendix, Tables S2--S4), whereas DeSurv at $k = 3$ retains significance after classifier adjustment. Across `r n_val_cohorts` independent PDAC cohorts ($n = `r n_val_total`$), the DeSurv linear predictor showed a consistent survival association (stratified Cox model: pooled HR per SD $`r hr_pool`$; 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$; $P = `r p_pool`$; per-cohort forest plot in Fig. \ref{fig:val}A; KM curves in SI Appendix, Fig. S6C--F). The analogous NMF linear predictor showed a weaker pooled effect (HR per SD $`r hr_pool_nmf`$; 95% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$; $P = `r p_pool_nmf`$); a direct comparison using dichotomized risk groups is presented below.

To assess clinical interpretability of these continuous associations, we dichotomized validation samples into high- and low-risk groups using method-specific cutpoints, each independently optimized via cross-validation on the training data. The DeSurv linear predictor separated survival trajectories in the pooled validation cohort (HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:val}B). The same procedure applied to NMF yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:val}C). The DeSurv cutpoint identified a substantially larger high-risk group than NMF (`r km_n_high_desurv` vs `r km_n_high_nmf` high-risk patients; `r km_reclass_diff` reclassified), thereby providing finer prognostic discrimination. Together, these results indicate that incorporating survival information during factorization yields gene programs with more transportable prognostic associations than the standard discover-then-evaluate approach.

```{r fig-val, fig.cap="Survival-aligned programs generalize across independent PDAC cohorts. (A) Forest plot of univariate hazard ratios (95\\% CIs) from separate Cox models for each factor, showing each factor's marginal survival association in each validation cohort. DeSurv factor D1 shows a consistent protective association across all five cohorts (pooled HR = 0.76), while no standard NMF factor achieves a comparable pooled effect. Black diamonds denote inverse-variance-weighted pooled estimates. (B--C) Kaplan--Meier curves for pooled external validation samples stratified by the combined multivariate linear predictor ($\\hat{\\beta}_1 Z_1 + \\hat{\\beta}_2 Z_2 + \\hat{\\beta}_3 Z_3$, where coefficients are learned from training data), dichotomized at a cross-validated optimal cutpoint: DeSurv (B) and standard NMF (C). \\label{fig:val}", fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

library(cowplot)
library(ggplot2)
library(dplyr)

# ── Panel A: forest plot (rebuilt from stored data with pooled HRs) ──
load_result("fig_hr_forest_tcgacptac")
d <- fig_hr_forest_tcgacptac$data

# Clean dataset labels
d$dataset <- dplyr::recode(d$dataset,
  "Puleo_array" = "Puleo", "Moffitt_GEO_array" = "Moffitt",
  "PACA_AU_seq" = "PACA seq", "PACA_AU_array" = "PACA array")

# Compute inverse-variance-weighted pooled HR per method × factor
pooled <- d %>%
  mutate(logHR = log(HR),
         se = (log(upper) - log(lower)) / (2 * 1.96),
         w = 1 / se^2) %>%
  group_by(method, factor_name) %>%
  summarise(pooled_logHR = sum(logHR * w) / sum(w),
            pooled_se = sqrt(1 / sum(w)), .groups = "drop") %>%
  mutate(HR = exp(pooled_logHR),
         lower = exp(pooled_logHR - 1.96 * pooled_se),
         upper = exp(pooled_logHR + 1.96 * pooled_se),
         dataset = "Pooled")

keep_cols <- c("factor_name", "HR", "lower", "upper", "dataset", "method", "row_type")
d$row_type <- "cohort"
pooled$row_type <- "pooled"
all_data <- rbind(d[, keep_cols], pooled[, keep_cols])

# Method label for facet panels
all_data$method <- factor(all_data$method, levels = c("DeSurv", "NMF"),
                          labels = c("DeSurv (D1-D3)", "Standard NMF (N1-N3)"))
all_data$dataset <- factor(all_data$dataset,
  levels = c("Puleo", "PACA seq", "PACA array", "Moffitt", "Dijk", "Pooled"))
cohort_cols <- c("Dijk" = "#F8766D", "Moffitt" = "#B79F00", "PACA array" = "#00BA38",
                 "PACA seq" = "#619CFF", "Puleo" = "#F564E3", "Pooled" = "black")
cohort_shapes <- c("Dijk" = 16, "Moffitt" = 16, "PACA array" = 16,
                   "PACA seq" = 16, "Puleo" = 16, "Pooled" = 18)
cohort_sizes <- c("Dijk" = 2.2, "Moffitt" = 2.2, "PACA array" = 2.2,
                  "PACA seq" = 2.2, "Puleo" = 2.2, "Pooled" = 4)

base_size <- 8
dodge_w <- 0.6

# Split into DeSurv / NMF using relabeled factor levels
desurv_dat <- all_data[all_data$method == "DeSurv (D1-D3)", ]
desurv_dat$factor_name <- factor(desurv_dat$factor_name,
  levels = c("D1", "D2", "D3"))
nmf_dat <- all_data[all_data$method == "Standard NMF (N1-N3)", ]
nmf_dat$factor_name <- factor(nmf_dat$factor_name,
  levels = c("N1", "N2", "N3"))

# Helper to build one forest panel
make_forest_panel <- function(dat, title, highlight_level) {
  hl <- data.frame(
    factor_name = factor(highlight_level, levels = levels(dat$factor_name)),
    x = 1)
  ggplot(dat, aes(x = HR, y = factor_name)) +
    geom_tile(data = hl, aes(x = x, y = factor_name),
              width = 100, height = 0.8,
              fill = "steelblue", alpha = 0.08, inherit.aes = FALSE) +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60",
               linewidth = 0.4) +
    geom_errorbar(aes(xmin = lower, xmax = upper, colour = dataset,
                      linewidth = row_type),
                  orientation = "y", width = 0,
                  position = position_dodge(width = dodge_w)) +
    geom_point(aes(colour = dataset, shape = dataset, size = dataset),
               position = position_dodge(width = dodge_w)) +
    scale_linewidth_manual(values = c("cohort" = 0.6, "pooled" = 1.2),
                           guide = "none") +
    scale_colour_manual(values = cohort_cols, name = NULL, drop = FALSE) +
    scale_shape_manual(values = cohort_shapes, name = NULL, drop = FALSE) +
    scale_size_manual(values = cohort_sizes, name = NULL, drop = FALSE) +
    scale_x_log10(limits = c(0.35, 3)) +
    labs(x = "Hazard ratio (95% CI)", y = NULL, title = title) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", size = 9, hjust = 0.5),
      axis.text.y = element_text(size = 8, face = "bold"),
      axis.text.x = element_text(size = 8, face = "bold"),
      axis.title.x = element_text(size = 9),
      plot.margin = margin(2, 6, 2, 4)
    )
}

p_desurv <- make_forest_panel(desurv_dat, "DeSurv (D1-D3)", "D1")
p_nmf    <- make_forest_panel(nmf_dat, "Standard NMF (N1-N3)", "N1")

# Extract shared legend
p_leg <- ggplot(all_data, aes(x = HR, y = factor_name, colour = dataset,
                               shape = dataset, size = dataset)) +
  geom_point() +
  scale_colour_manual(values = cohort_cols, name = NULL) +
  scale_shape_manual(values = cohort_shapes, name = NULL) +
  scale_size_manual(values = cohort_sizes, name = NULL) +
  guides(colour = guide_legend(nrow = 1, override.aes = list(size = 2.5)),
         shape = guide_legend(nrow = 1), size = guide_legend(nrow = 1)) +
  theme_void(base_size = base_size) +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))
leg_grob = ggplotGrob(p_leg)
forest_legend <- gtable::gtable_filter(leg_grob, "guide-box")

plot_forest <- plot_grid(
  plot_grid(p_desurv, p_nmf, ncol = 2, align = "hv", axis = "tb"),
  forest_legend,
  nrow = 2, rel_heights = c(15, 1)
)

# ── Panels B-C: KM curves ──
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")

theme_pnas <- theme_classic(base_size = base_size) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.margin = margin(6, 10, 6, 15),
    legend.box.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(6, "pt"),
    legend.key.width = unit(12, "pt")
  )

km_text_size    <- base_size
risk_title_size <- base_size

stack_surv <- function(surv_obj, title, keep_legend = FALSE) {
  p <- surv_obj$plot +
    theme_pnas +
    theme(
      legend.position = if (keep_legend) "bottom" else "none",
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = km_text_size),
      axis.text.x  = element_text(size = km_text_size),
      axis.text.y  = element_text(size = km_text_size),
      plot.margin  = margin(4, 10, 0, 10),
      plot.title   = element_text(size = 9)
    ) +
    labs(title = title)

  t <- surv_obj$table +
    theme_pnas +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.x  = element_text(size = km_text_size),
      axis.text.y  = element_text(size = km_text_size),
      text = element_text(size = km_text_size),
      axis.title.x = element_text(size = km_text_size),
      plot.title   = element_text(size = risk_title_size),
      plot.margin  = margin(4, 10, 2, 10)
    ) +
    labs(x = "Time (months)")
  t$layers[[1]]$aes_params$size <- km_text_size / ggplot2::.pt
  plot_grid(p, t, ncol = 1, rel_heights = c(2.5, 1.5), align = "v", axis = "lr")
}

km_b <- stack_surv(fig_median_survival_desurv_tcgacptac, "DeSurv")
km_c <- stack_surv(fig_median_survival_std_desurvk_tcgacptac, "NMF")

# Shared KM legend
legend_plot <- ggplot(
  data.frame(x = 1:2, y = 1:2,
             group = factor(c("Low", "High"), levels = c("Low", "High"))),
  aes(x = x, y = y, colour = group)
) +
  geom_line() +
  scale_colour_manual(values = c("Low" = "violetred2", "High" = "turquoise4"),
                      name = "Risk group") +
  theme_pnas +
  theme(legend.position = "bottom",
        legend.text = element_text(size = km_text_size),
        legend.title = element_text(size = km_text_size)) +
  guides(colour = guide_legend(nrow = 1))

legend_grob <- ggplotGrob(legend_plot)
km_legend <- legend_grob$grobs[
  sapply(legend_grob$grobs, function(x) x$name) == "guide-box"
][[1]]

km_block <- plot_grid(
  km_b, km_c, ggdraw(km_legend),
  nrow = 3, labels = c("B", "C", ""),
  label_size = 12, rel_heights = c(5, 5, 0.6)
)

# ── Full figure: forest (left) + KM curves (right) ──
plot_grid(plot_forest, km_block, ncol = 2, rel_widths = c(1.4, 1),
          labels = c("A", ""), label_size = 12)
```


## Factorization rank k=3 is robust across supervision settings {.unnumbered}

```{r ksens-counts, echo=FALSE}
# ── Load K-sensitivity RDS outputs ────────────────────────────────────────────
# Generated by:
#   Rscript inst/cv_grid_validation_analysis_amber_optimal.R  → adj_p_270, hcor_270, master_rows_270
#   Rscript inst/cv_grid_validation_analysis_allgenes.R       → adj_p_all, hcor_all, master_rows_all
#   Rscript inst/k3_k7_nesting_heatmap.R                     → k3_k7_summary
#   Rscript inst/create_ksens_rds.R                          → lam300_summary, production_summary (bootstrap)
RDS_DIR <- "results/cv_grid"

adj_p_270    <- readRDS(file.path(RDS_DIR, "adj_p_270_matrix.rds"))
adj_p_all    <- readRDS(file.path(RDS_DIR, "adj_p_all_matrix.rds"))
unadj_p_270  <- readRDS(file.path(RDS_DIR, "unadj_p_270_matrix.rds"))
hcor_270     <- readRDS(file.path(RDS_DIR, "hcor_270_matrix.rds"))
hcor_all     <- readRDS(file.path(RDS_DIR, "hcor_all_matrix.rds"))
master_270   <- readRDS(file.path(RDS_DIR, "master_rows_270.rds"))
master_all   <- readRDS(file.path(RDS_DIR, "master_rows_all.rds"))
lam300       <- readRDS(file.path(RDS_DIR, "lam300_summary.rds"))
prod_sum     <- readRDS(file.path(RDS_DIR, "production_summary.rds"))

# ── Derived scalars from unadjusted p-value matrix ────────────────────────────
n_sig_270            <- sum(unadj_p_270 < 0.05)
n_k3_sig_270         <- sum(unadj_p_270["K=3", ] < 0.05)
n_k7_sig_270         <- sum(unadj_p_270["K=7", ] < 0.05)
k3_alpha55_p_270     <- unadj_p_270["K=3", "a=0.55"]
k7_alpha75_p_270     <- unadj_p_270["K=7", "a=0.75"]

# ── Derived scalars from adjusted p-value matrix ─────────────────────────────
n_k3_adj_sig_270     <- sum(adj_p_270["K=3", ] < 0.05)
n_k7_adj_sig_270     <- sum(adj_p_270["K=7", ] < 0.05)
n_sig_all            <- sum(adj_p_all < 0.05)
k3_alpha55_p_all     <- adj_p_all["K=3", "a=0.55"]
k5_alpha55_p_270     <- adj_p_270["K=5", "a=0.55"]
k5_alpha25_p_all     <- adj_p_all["K=5", "a=0.25"]

# ── H-cor values from hcor matrices ───────────────────────────────────────────
hcor_k3_a55_270  <- hcor_270["K=3", "a=0.55"]
hcor_k5_a55_270  <- hcor_270["K=5", "a=0.55"]
hcor_k5_a25_all  <- hcor_all["K=5", "a=0.25"]

# ── λ=0.300 analysis scalars ───────────────────────────────────────────────────
n_sig_all_prior  <- lam300$n_sig
production_adj_p <- prod_sum$adj_p

# ── Grid dimensions (for "of N combinations" and "of M alpha values" prose) ──
GRID_K      <- c(2, 3, 5, 7, 9)
GRID_ALPHA  <- c(0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95)
n_alpha     <- length(GRID_ALPHA)
n_grid      <- length(GRID_K) * length(GRID_ALPHA)
```

The 1-SE model selection rule chose $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ over the global BO maximum ($k = `r read_result("tar_k_selection_tcgacptac")$k_best`$; CV C-index $`r k_best_cindex`$ versus $`r k3_best_cindex`$; margin $`r k_margin`$). To assess whether this reflects a robust solution or an artifact of a specific tuning configuration, we conducted a sensitivity analysis across $k \in \{2, 3, 5, 7, 9\}$ and $\alpha \in \{0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95\}$ (`r n_grid` combinations), holding all other hyperparameters at their BO-selected values and evaluating each combination in independent external validation cohorts via stratified Cox regression (SI Appendix, Tables S2--S4). $k = 3$ achieved significance ($P < 0.05$) across `r n_k3_sig_270` of `r n_alpha` supervision strengths, and $k = `r read_result("tar_k_selection_tcgacptac")$k_best`$ at a comparable `r n_k7_sig_270` of `r n_alpha`. After additional adjustment for PurIST and DeCAF molecular classifiers, $k = 3$ retained significance at `r n_k3_adj_sig_270` of `r n_alpha` supervision strengths, while $k = `r read_result("tar_k_selection_tcgacptac")$k_best`$ dropped to `r n_k7_adj_sig_270` of `r n_alpha` (SI Appendix, Table S3). This indicates that $k = `r read_result("tar_k_selection_tcgacptac")$k_best`$'s prognostic content is largely recoverable from existing classifiers applied to the same data, whereas DeSurv's D1 factor provides independent prognostic information not captured by PurIST or DeCAF. The parsimonious $k = 3$ solution thus validates robustly across supervision strengths and retains prognostic value beyond known classifiers, providing empirical support for the 1-SE rule.
