<!-- REVISED RESULTS — prediction-validation structure (originally for PNAS; retained for Nature submission) -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- SI Appendix figure numbering (supplement.Rmd order, post-reorganization):
     S1  fig-converge          convergence trajectories
     S2  fig-sim-null-mixed    null and mixed simulation scenarios
     S3  fig-cindex-by-k       C-index vs K sensitivity
     S4  fig-nmf-diagnostics   standard NMF rank selection heuristics (demoted from old Fig 2A-C)
     S5  fig-cutpoint-km       cutpoint selection and KM curves
     S6  fig-subtype-overlap   subtype composition of risk groups
     S7  fig-desurv-k5-heatmap DeSurv gene overlap at k=5
     S8  fig-nmf-k5-heatmap    NMF gene overlap at k=5
     S9  fig-nmf-k7-heatmap    NMF gene overlap at k=7
     S10 fig-varsurvival-k5    variance-survival at k=5
     Note: Former S7 (variance-survival), S8 (W-corr), S9 (forest) are now
           main text Fig 3C-D and Fig 4A respectively.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(magick)
library(gt)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)


load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)
desurv_maxvar_label <- paste0("D", desurv_maxvar$factor)
nmf_maxvar_label    <- paste0("N", nmf_maxvar$factor)
# Exocrine NMF factor (N2) — referenced explicitly because of its biological
# role as the high-variance exocrine-associated factor
nmf_n2              <- nmf_vs[nmf_vs$factor == 2, ]
nmf_n2_pct          <- round(nmf_n2$variance_explained * 100, 1)
nmf_n2_dl           <- round(nmf_n2$delta_loglik, 2)

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
desurv_stats <- pooled_cox_stats(pool_df)
hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Pooled adjusted (PurIST + DeCAF) statistics for DeSurv ───────────────────
load_result("data_val_filtered_tcgacptac")

extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(sample_id = rownames(surv), time = surv$time, event = surv$event,
                     dataset = entry$dataset, risk_score = entry$risk_score,
                     stringsAsFactors = FALSE)
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo; si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

pool_df_adj_complete <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
pool_df_adj_complete <- pool_df_adj_complete[
  complete.cases(pool_df_adj_complete[, c("PurIST", "DeCAF")]), ]
rs_sd_adj <- sd(pool_df_adj_complete$risk_score, na.rm = TRUE)
if (is.na(rs_sd_adj) || rs_sd_adj < 1e-12) rs_sd_adj <- 1
pool_df_adj_complete$risk_z <- pool_df_adj_complete$risk_score / rs_sd_adj
cox_adj_sum <- summary(coxph(
  Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
  data = pool_df_adj_complete))
hr_adj      <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "exp(coef)"])
ci_adj_lo   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "lower .95"])
ci_adj_hi   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "upper .95"])
p_adj_val   <- cox_adj_sum$coefficients["risk_z", "Pr(>|z|)"]
p_adj_text  <- if (p_adj_val < 0.001) "< 0.001" else paste0("= ", signif(p_adj_val, 2))

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

# ── KM dichotomized risk-group statistics ─────────────────────────────────────
# Extract HR/CI/p from the annotation layer embedded in the KM plot objects
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
load_result("fig_median_survival_desurv_tcgacptac")
load_result("fig_median_survival_std_desurvk_tcgacptac")
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# Extract exact risk-group counts from KM risk tables (time = 0)
km_n_high_desurv <- fig_median_survival_desurv_tcgacptac$data.survtable$n.risk[
  fig_median_survival_desurv_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_desurv_tcgacptac$data.survtable$strata == "factor=1"]
km_n_high_nmf <- fig_median_survival_std_desurvk_tcgacptac$data.survtable$n.risk[
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$time == 0 &
  fig_median_survival_std_desurvk_tcgacptac$data.survtable$strata == "factor=1"]
km_reclass_diff <- km_n_high_desurv - km_n_high_nmf
```

## Model Overview {.unnumbered}

We developed DeSurv, a survival-supervised deconvolution framework that learns two linked quantities from bulk transcriptomes paired with patient outcomes: gene programs and their survival associations (Fig.\ \ref{fig:schema}A, Methods). DeSurv integrates nonnegative matrix factorization with Cox proportional hazards modeling so that each gene program is shaped during discovery by both expression reconstruction and survival association. The survival objective acts on the gene-program matrix rather than directly on sample loadings, preserving the mixture-coefficient interpretation of loadings and allowing trained programs to score new samples by projection without retraining. A supervision parameter controls the balance between reconstruction and survival, and this parameter, the factorization rank, and regularization are selected by Bayesian optimization on cross-validated concordance (Fig.\ \ref{fig:schema}B, Methods).

Applied to the PDAC training cohorts from TCGA [@raphael2017integrated] and CPTAC [@ellis2013clinical] (Materials and Methods), cross-validated concordance varied across candidate ranks but plateaued from $k = 3$ through $k = 12$ (SI Appendix, Fig.\ \ref{fig:fig-cindex-by-k}), indicating that three supervised factors captured nearly as much prognostic information as twelve. Within this plateau, joint Bayesian optimization over rank and supervision strength with a one-standard-error rule (Methods) selected the parsimonious model, $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$ at $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$. The peak cross-validated C-index was $`r round(read_result("tar_k_selection_tcgacptac")$best_mean, 3)`$ across `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)` patients with `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)` events. By contrast, standard NMF rank-selection diagnostics, including reconstruction residuals, cophenetic correlation and silhouette width [@frigyesi2008nmf; @brunet2004metagenes], did not converge on a consistent rank in PDAC (SI Appendix, Fig.\ \ref{fig:fig-nmf-diagnostics}), motivating outcome-aware model selection. All rank and hyperparameter selection was performed within the TCGA and CPTAC training set, before any projection into validation cohorts. This $k = 3$ solution was dominated by a co-varying tumor--stroma axis, which we characterize next.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "model_schematic_final.pdf"))
```


## Survival supervision and standard NMF produce different factor structures in PDAC {.unnumbered}

To determine how survival supervision shaped the recovered factor structure, we fit standard NMF at the same rank selected for DeSurv ($k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$) and compared each method's factor-specific gene rankings with established PDAC gene programs (Fig.\ \ref{fig:pdac}A--B). Matching $k$ across methods isolates the effect of survival supervision on factor content from differences in model complexity; a sensitivity analysis using NMF's independently selected ranks (elbow $k = 5$, BO $k = 7$) is provided in SI Appendix, Table \ref{tab:tab-cindex-comparison}. We first asked which malignant, stromal, immune and other microenvironmental programs were represented in each factorization, then tested whether these structural differences corresponded to differences in survival contribution.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. In the DeSurv factorization, D1 correlated with Classical tumor programs and restCAF-like stromal signatures, D2 with proCAF-like programs, and D3 with Basal-like tumor programs (Fig. \ref{fig:pdac}A). Thus, DeSurv organized the dominant survival-associated signal around a tumor--stroma axis: higher D1 scores mark a Classical/restCAF-like, lower-risk state, whereas movement toward Basal-like tumor and proCAF-like stroma marks higher risk. This organization is consistent with evidence that survival stratification in PDAC improves when tumor-intrinsic and CAF subtypes are considered jointly [@peng2024determination; @elyada2019cross]. No DeSurv factor showed significant positive correlation with exocrine-associated programs, including DECODER Exocrine, Collisson Exocrine and Bailey aberrantly differentiated endocrine--exocrine (ADEX) programs (Fig. \ref{fig:pdac}A).

Standard NMF produced a different factor organization (Fig. \ref{fig:pdac}B). One factor, N2, was strongly correlated with exocrine-associated programs and negatively correlated with immune and stromal signatures. A second factor, N1, correlated with malignant-epithelial identity and carried both Classical and Basal-like markers. A third factor, N3, correlated with immune, fibroblast and extracellular matrix (ECM)-related programs, including Puleo Desmoplastic, SCISSORS proCAF and restCAF [@leary2023s], Maurer Immune and ECM [@Maurer2019], and DECODER Activated Stroma. Thus, at $k = 3$, standard NMF devoted one factor to exocrine-associated variation and aggregated tumor-intrinsic and microenvironmental programs into two broad factors, whereas DeSurv separated Basal-like tumor, proCAF-like stroma and a Classical/restCAF-like survival-aligned program.

NMF seeks a low-rank \emph{reconstruction} of the transcriptome; many factorizations can reconstruct comparably well, and the unsupervised objective has no incentive to prefer prognostically informative directions. We therefore quantified each factor's contribution using two metrics (Fig. \ref{fig:pdac}C; SI Appendix, Sections 10--11): a normalized reconstruction Shapley value, which partitions reconstruction contribution across factors and sums to $100\%$, and survival contribution, defined as the change in Cox partial log-likelihood ($\Delta\ell$) when that factor is dropped from a $k$-factor Cox model. NMF distributed reconstruction across factors in proportions unrelated to prognosis: all three NMF factors had negligible $\Delta\ell$ despite spanning a wide range of reconstruction contribution. By contrast, DeSurv concentrated nearly all survival contribution into D1. Specifically, D1 contributed $`r desurv_maxsurv_pct`$% of reconstruction but $\Delta\ell = `r desurv_maxsurv_dl`$, whereas the largest-reconstruction NMF factor, `r nmf_maxvar_label`, contributed $`r nmf_maxvar_pct`$% of reconstruction but only $\Delta\ell = `r nmf_maxvar_dl`$ to survival. The exocrine-associated NMF factor, N2, contributed $`r nmf_n2_pct`$% of reconstruction but only $\Delta\ell = `r nmf_n2_dl`$ to survival. The largest-reconstruction DeSurv factor, `r desurv_maxvar_label`, contributed $`r desurv_maxvar_pct`$% of reconstruction but added little survival information after accounting for D1. Thus, in PDAC, factors that best reconstruct expression are not necessarily the factors that best explain survival, consistent with tumor-purity analyses across cancer types [@aran2015systematic].

To further characterize the relationship between the two factorizations, we examined pairwise W-matrix gene-loading correlations across all 1,970 genes (Fig. \ref{fig:pdac}D; Spearman correlations over full loading profiles, since top-gene subsets can overstate preservation). At $k = 3$, NMF captured the malignant-epithelial compartment largely in a single factor, N1, which carried both Classical and Basal-like markers and therefore is labeled ``tumor'' rather than Classical. N1 mapped strongly onto DeSurv's Basal-like tumor factor, D3 (Spearman $\rho = 0.84$), while NMF's microenvironmental factor, N3, mapped onto DeSurv's proCAF-like stromal factor, D2 ($\rho = 0.87$). In contrast, DeSurv's most prognostic factor, D1, was only weakly correlated with all three NMF factors (all $\rho \le 0.36$). Because D1 couples Classical tumor and restCAF-like stromal signatures into a single survival-aligned program, it is not simply a reweighting of an NMF factor (SI Appendix, Table \ref{tab:projection-r2}). NMF's exocrine factor, N2, had no strong DeSurv counterpart (maximum $\rho = 0.61$), consistent with reduced emphasis on exocrine-associated variation under survival supervision. Together, these analyses show that standard NMF primarily separates broad tumor, stromal/immune and exocrine-associated structure, whereas DeSurv resolves a survival-aligned tumor--stroma program that standard NMF does not isolate.
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (A--B) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$. Only $|r| > 0.2$ shown. DeSurv concentrates survival signal into D1, whose Classical-tumor + restCAF-like-stroma pole marks lower risk, while suppressing exocrine variation; standard NMF devotes one factor to exocrine signals. (C) Per-factor contribution to reconstruction of the expression matrix (normalized reconstruction Shapley share, which partitions and sums to 100\\%) versus survival contribution ($\\Delta$ Cox partial log-likelihood upon factor removal). DeSurv concentrates survival signal into D1 while NMF's largest-reconstruction factors --- including the exocrine-associated factor N2 --- contribute negligibly to survival. (D) Pairwise correspondence between NMF and DeSurv factors, measured by Spearman rank correlation of W-matrix gene loadings over all genes. \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.height= "7in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig3_tcgacptac.pdf"))
```


## Survival-aligned programs generalize across independent PDAC cohorts with lower complexity {.unnumbered}

A prognostic factorization is useful only if it generalizes beyond the training cohort. Having established that survival supervision and standard NMF produce different factor structures in the training data, we next tested whether each method's learned programs transferred to independent cohorts without retraining. The five external validation cohorts (Dijk, Moffitt, PACA-AU array, PACA-AU RNA-seq and Puleo) consisted of non-metastatic, treatment-naive specimens (SI Appendix, Table \ref{tab:cohorts}). We computed factor scores in each validation dataset by projecting the gene programs learned in the training cohort onto each new expression matrix, $Z = \widetilde{W}^\top X_{\text{new}}$, yielding a continuous score for each factor in each sample.

External validation C-indices for NMF at independently selected ranks are summarized in SI Appendix, Table \ref{tab:tab-cindex-comparison}. Standard NMF at elbow-selected $k = 5$ and Bayesian-optimization-selected $k = 7$ substantially improved over NMF at $k = 3$, with $k = 7$ matching or exceeding DeSurv's per-cohort C-index in most cohorts. DeSurv's aim, however, is not maximal prediction at arbitrary rank but recovery of a compact, interpretable tumor--stroma program; the NMF result required more than twice as many factors and depended on applying DeSurv's tuning framework with $\alpha = 0$ in the joint optimization over rank and hyperparameters, whereas NMF's own rank-selection diagnostics (cophenetic correlation, silhouette width and reconstruction residuals) did not converge on this rank (SI Appendix, Fig.\ \ref{fig:fig-nmf-diagnostics}). DeSurv at $k = 3$ retained a significant association with survival after simultaneous adjustment for PurIST and DeCAF classifications (per-SD HR = `r hr_adj`, 95\% CI `r ci_adj_lo`--`r ci_adj_hi`, $P `r p_adj_text`$; Table \ref{tab:hr-adjusted}), indicating independent prognostic value as a lower-dimensional, biologically coherent program. Standard NMF at $k = 7$ attained an adjusted hazard ratio at least as large (Table \ref{tab:hr-adjusted}); we therefore do not claim predictive superiority for DeSurv, whose advantage is recovering this association as a compact, interpretable, source-attributable program rather than as a higher-dimensional, less-interpretable factorization.

```{r suprec-bench, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_stats")
.svb <- function(m) sprintf("%.2f", desurv_vs_supervised_stats$pooled[m, "pooled_C"])
```

Prediction alone did not distinguish DeSurv from simpler survival-supervised models. Freezing each method's leading prognostic direction on the training data and projecting it without refitting onto the five validation cohorts, DeSurv's pooled transfer concordance (C-index $= `r .svb("DeSurv (full LP)")`$) was on par with survival-supervised baselines fit directly on genes: penalized (lasso) Cox ($`r .svb("Sparse Cox")`$), Cox partial least squares ($`r .svb("Cox-PLS")`$) and supervised principal components ($`r .svb("Supervised PCA")`$). Unsupervised directions transferred poorly (standard NMF at $\alpha = 0$, $`r .svb("Unsup. NMF (a=0)")`$; the leading principal component, $`r .svb("Unsup. PCA (PC1)")`$), consistent with evidence that learned representations do not reliably out-predict penalized regression on genes for out-of-sample prognosis [@smith2020standard]. Because these survival-supervised scores are nearly interchangeable for prediction yet collapse the signal into a single, compartment-ambiguous axis, DeSurv's contribution is not higher accuracy but the decomposition of that same prognostic signal into interpretable, source-attributable programs.

When $\beta$ coefficients were re-fit within each validation cohort, DeSurv's three-factor decomposition was preferred by BIC over NMF's seven-factor decomposition in three of five cohorts, with the remaining two essentially tied (SI Appendix, Table \ref{tab:tab-bic-percohort}). Across the `r n_val_cohorts` independent PDAC cohorts ($n = `r n_val_total`$), the DeSurv linear predictor (oriented so that higher values indicate higher risk) showed a consistent survival association in a stratified Cox model (pooled HR per SD $= `r hr_pool`$, 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$, $P = `r p_pool`$; per-cohort forest plot in Fig. \ref{fig:val}A; Kaplan--Meier curves in SI Appendix, Fig. \ref{fig:fig-cutpoint-km}C--F). At the matched rank of $k = 3$, the analogous NMF linear predictor showed a weaker pooled association (HR per SD $= `r hr_pool_nmf`$, 95% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$, $P = `r p_pool_nmf`$), motivating a matched-rank comparison of dichotomized risk groups.

The continuous, stratified-Cox associations above are our primary analysis; to illustrate them clinically, we dichotomized validation samples into high- and low-risk groups using method-specific cutpoints independently optimized by cross-validation in the training data. The DeSurv linear predictor separated survival trajectories in the pooled validation cohort (HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:val}B). Applying the same procedure to NMF at $k = 3$ yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:val}C). The DeSurv cutpoint classified a larger validation subset as high risk than NMF at matched rank (`r km_n_high_desurv` versus `r km_n_high_nmf` high-risk patients; `r km_reclass_diff` reclassified) while preserving strong survival separation. The DeSurv high-risk group was enriched for both established poor-prognosis subtypes, Basal-like by PurIST and proCAF by DeCAF (Fisher's exact test; SI Appendix, Fig. \ref{fig:fig-subtype-overlap}), capturing consensus biology while adding independent prognostic information. By contrast, NMF risk groups at $k = 3$ showed Basal-like enrichment but no significant proCAF enrichment (SI Appendix, Fig. \ref{fig:fig-subtype-overlap}), consistent with NMF not isolating the tumor--stroma axis represented by D1.

Together, these results indicate that survival supervision yields a compact, biologically interpretable prognostic structure that generalizes across independent PDAC cohorts. Higher-rank NMF models can achieve similar per-cohort concordance, but only at greater complexity and without recovering the same axis; DeSurv instead identifies a lower-dimensional tumor--stroma program with prognostic value beyond PurIST and DeCAF.

```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC cohorts (pooled $n=", n_val_total, "$ patients across ", n_val_cohorts, " cohorts: ", paste(sprintf("%s $n=%d$", gsub("AU seq", "AU RNA-seq", gsub("PACA AU", "PACA-AU", gsub("_", " ", names(table(pool_df$dataset))))), as.integer(table(pool_df$dataset))), collapse = "; "), "). (A) Forest plot of univariate hazard ratios (95\\% CIs) from separate Cox models for each factor, showing each factor's marginal survival association in each validation cohort. DeSurv factor D1 shows a consistent protective association across all five cohorts (pooled HR = 0.76), while no standard NMF factor achieves a comparable pooled effect. Error bars are 95\\% confidence intervals; black diamonds denote inverse-variance-weighted pooled estimates. (B--C) Kaplan--Meier curves for pooled external validation samples stratified by the combined multivariate linear predictor ($\\hat{\\beta}_1 Z_1 + \\hat{\\beta}_2 Z_2 + \\hat{\\beta}_3 Z_3$, where coefficients are learned from training data), dichotomized at a cross-validated optimal cutpoint: DeSurv (B) and standard NMF (C). No validation-cohort survival outcomes were used in model training, scoring, or hyperparameter selection. \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig4_tcgacptac.pdf"))
```

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(knitr); library(kableExtra)
# extract_pool_df_with_covariates() and data_val_filtered_tcgacptac
# are defined in the setup chunks above.
load_result("val_latent_desurv_alpha0_tcgacptac")

pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(hr = round(ci_row[, "exp(coef)"], 2),
         ci_lo = round(ci_row[, "lower .95"], 2),
         ci_hi = round(ci_row[, "upper .95"], 2),
         p = coef_row[, "Pr(>|z|)"])
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

pool_desurv_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv_t1 <- pooled_cox_unadj_adj(pool_desurv_t1)
pool_nmf7_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7_t1 <- pooled_cox_unadj_adj(pool_nmf7_t1)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)", "Std NMF ($k = 7$, BO $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv_t1$unadj), fmt_hr(stats_nmf7_t1$unadj)),
  Adjusted = c(fmt_hr(stats_desurv_t1$adj), fmt_hr(stats_nmf7_t1$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per SD (95\\% CI), $P$",
                       "Adjusted HR per SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = paste0("Pooled external validation hazard ratios (per SD of the linear predictor) for DeSurv at $k = 3$ and standard NMF at BO-selected $k = 7$ ($\\alpha = 0$), across the five validation cohorts ($n = ", n_val_total, "$ patients, ", n_val_events, " events). Adjusted models include the PurIST tumor-intrinsic and DeCAF stromal molecular classifiers as covariates, with stratification by dataset. Both methods retain significant prognostic value after adjustment, with DeSurv's larger HR drop reflecting biological concordance with established subtypes and NMF $k = 7$'s smaller drop reflecting orthogonal residual content from its additional, less interpretable factors. \\label{tab:hr-adjusted}")) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

```{r d1-partition-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("d1_variance_partition_stats")
.d1p <- d1_variance_partition_stats
.d1pf <- function(p) if (p < 1e-4) "< 0.0001" else sprintf("= %.4f", p)
load_result("supervised_recovery_stats")
.d1p_sr_mean  <- supervised_recovery_stats$sup_mean_score_cor
.d1p_sr_unsup <- abs(supervised_recovery_stats$table$score_cor_D1[!supervised_recovery_stats$table$supervised])
```

## The tumor--stroma axis adds prognostic information beyond established PDAC classifiers {.unnumbered}

To establish that the dominant program is more than a continuous restatement of discrete subtype calls, we partitioned D1 against both classifiers across the pooled validation cohorts ($n = `r .d1p$meta$n`$, `r .d1p$meta$n_events` events). Binary PurIST and DeCAF calls together accounted for `r sprintf("%.0f", 100*.d1p$variance$explained_total)`\% of the within-cohort variance of the continuous D1 score, indicating that D1 grades tumor--stroma state more finely than the dichotomous subtypes with which it aligns. More importantly, D1 retained significant prognostic value after conditioning on both classifiers (adjusted HR per SD = `r sprintf("%.2f", .d1p$prognosis$d1_adj_hr)`, 95\% CI `r sprintf("%.2f", .d1p$prognosis$d1_adj_lo)`--`r sprintf("%.2f", .d1p$prognosis$d1_adj_hi)`, $P `r .d1pf(.d1p$prognosis$d1_adj_p)`$; partial likelihood-ratio $\chi^2 = `r sprintf("%.1f", .d1p$prognosis$lr_chisq)`$ on 1 df, $P `r .d1pf(.d1p$prognosis$lr_p)`$), improving stratified-Cox concordance over a PurIST-plus-DeCAF model from `r sprintf("%.3f", .d1p$prognosis$c_base)` to `r sprintf("%.3f", .d1p$prognosis$c_with_d1)`. The tumor--stroma axis therefore carries prognostic information that neither the tumor-intrinsic nor stromal classifier captures alone. This axis was also not specific to DeSurv's optimizer: three independent survival-supervised methods (supervised principal components, Cox partial least squares and sparse Cox) recovered the same D1 score in the training cohort (mean $|r| = `r sprintf("%.2f", .d1p_sr_mean)`$), whereas the leading unsupervised direction did not ($|r| = `r sprintf("%.2f", .d1p_sr_unsup)`$), identifying D1 as a property of survival supervision rather than of a single algorithm (SI Appendix, supervised-method recovery). After adjustment for standard clinical prognostic variables (stage, grade, age and sex), the DeSurv score retained a significant, independent association with survival in the fully-annotated external PACA-AU cohorts, and it remained prognostic after adjustment for tumor purity, arguing against a purity- or composition-driven artifact (SI Appendix, Table \ref{tab:clinical-adjustment}); the smaller Dijk and Moffitt cohorts were individually underpowered for multivariable analysis.

## The recovered programs show context-dependent prognostic associations in treated PDAC {.unnumbered}

```{r treated-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("treated_cohort_stats")
.tcs <- treated_cohort_stats
.osm <- function(coh, prog, mod){ r <- .tcs$os[.tcs$os$cohort==coh & .tcs$os$program==prog & .tcs$os$model==mod, ]
  if (!nrow(r)) return("---")
  sprintf("%.2f (95%% CI %.2f--%.2f, $P %s$)", r$hr, r$lo, r$hi,
          ifelse(r$p < 0.001, "< 0.001", paste0("= ", sprintf("%.3f", r$p)))) }
.tn  <- function(coh) .tcs$prevalence$n[.tcs$prevalence$cohort==coh]
# Benjamini-Hochberg correction over the 8 primary treated-cohort survival tests
.d2adj <- subset(.tcs$os, program=="D2" & model=="adjusted")
.d1mar <- subset(.tcs$os, program=="D1" & model=="marginal")
.pf2   <- .tcs$pfs[.tcs$pfs$program %in% c("D2","D3"), ]
.fam <- data.frame(
  lab = c(paste0("D2-OS-", .d2adj$cohort), paste0("D1-OS-", .d1mar$cohort), paste0(.pf2$program, "-PFS")),
  p   = c(.d2adj$p, .d1mar$p, .pf2$p))
.fam$bh <- p.adjust(.fam$p, "BH")
.bh <- function(lab){ v <- .fam$bh[.fam$lab==lab]; if(!length(v)) return("NA")
  if (v < 0.001) "< 0.001" else paste0("= ", sprintf("%.3f", v)) }
.pfsm <- function(prog){ r <- .tcs$pfs[.tcs$pfs$program==prog, ]
  sprintf("%.2f (95%% CI %.2f--%.2f, $P %s$)", r$hr, r$lo, r$hi,
          ifelse(r$p < 0.001, "< 0.001", paste0("= ", sprintf("%.3f", r$p)))) }
```

As an exploratory, hypothesis-generating analysis, we examined how the fixed, treatment-naive programs behave when read in treated cohorts; because treatment, stage and cohort are confounded in these data, the associations below are prognostic, not predictive. Because DeSurv scores new samples by projecting the trained gene-program matrix without re-estimating parameters, the same programs can be read directly in these cohorts and compared across disease settings. We projected the trained programs onto three bulk treated cohorts: the metastatic gemcitabine-backbone trials Rash ($n = `r .tn("Rash")`$) and Accept ($n = `r .tn("Accept")`$), and the borderline-resectable/locally advanced Linehan cohort (FOLFIRINOX $\pm$ CCR2 inhibitor, $n = `r .tn("Linehan")`$). Because these cohorts differ by stage and regimen, we analyzed each separately rather than as a pooled estimate (Fig. \ref{fig:treated}; SI Appendix, treated-cohort analysis). Because the elastic-net penalty shrank the trained survival coefficient for the proCAF program D2 to exactly zero in the treatment-naive model (an estimated selection outcome, not an imposed constraint), D2 contributes nothing to the trained risk score, and programs were evaluated per factor; any association of D2 in the treated cohorts therefore reflects a marginal re-evaluation of the projected program in a new clinical context rather than a property of the trained predictor.

The proCAF stromal program D2 was associated with longer overall survival in all three treated cohorts: Linehan, adjusted HR per SD = `r .osm("Linehan","D2","adjusted")`; Accept, `r .osm("Accept","D2","adjusted")`; and Rash, `r .osm("Rash","D2","adjusted")`. Thus, D2 showed a consistently protective direction across independent settings and chemotherapy backbones, reaching significance in the borderline-resectable/locally advanced cohort and remaining directionally concordant in the metastatic cohorts. This association was not attenuated by adjustment for the tumor-intrinsic and CAF classifiers; in Linehan, where D2 reached significance, the adjusted effect strengthened rather than weakened. D2 was also associated with longer progression-free survival in the metastatic setting and increased from pre- to post-treatment in paired biopsies, consistent with therapy-associated remodeling of the proCAF program (SI Appendix). Near-identical results under full-gene projection (Spearman $r \geq 0.86$ between top-gene and full-gene scores; SI Appendix) further supported the robustness of this association.

By contrast, D1 showed a setting-dependent association that tracked the basal--classical axis rather than a property unique to D1. D1 separated PurIST calls in every treated cohort (AUC `r sprintf("%.2f", min(.tcs$d1_auc, na.rm=TRUE))`--`r sprintf("%.2f", max(.tcs$d1_auc, na.rm=TRUE))`), remained prognostic in the less-advanced Linehan cohort with a marginal HR per SD of `r .osm("Linehan","D1","marginal")`, and attenuated in the metastatic cohorts. Established basal--classical classifiers showed the same pattern: PurIST, Moffitt and Bailey classifications were all non-significant in the metastatic cohorts (SI Appendix). These findings indicate stage-dependent attenuation of the basal--classical prognostic axis rather than a failure specific to D1. Basal-like prevalence in the metastatic cohorts was within the expected range (16--22\%), and the same axis was strongly prognostic in treatment-naive cohorts with similar prevalence ($P<10^{-8}$); the Linehan estimate was based on only six Basal-like cases and was correspondingly imprecise, and large real-world cohorts confirm that the basal subtype remains adverse in metastatic disease [@singh2024basal]. We therefore report D1 marginally, because adjustment for the collinear PurIST classifier would condition the same axis on itself.

D3, the Basal-like program, was not associated with overall survival in any treated cohort, consistent with the basal--classical contrast residing primarily in D1. However, D3 was associated with shorter progression-free survival in the metastatic setting, with an HR per SD of `r .pfsm("D3")`, consistent with the aggressive biology of Basal-like tumors. Across the eight primary treated-cohort survival tests, the significant associations---D2 with overall survival in the borderline-resectable/locally advanced cohort ($P_{\text{BH}} `r .bh("D2-OS-Linehan")`$) and D3 with progression-free survival ($P_{\text{BH}} `r .bh("D3-PFS")`$)---remained significant after Benjamini--Hochberg correction, whereas the marginal associations did not, consistent with their exploratory interpretation. The D1 score also independently validated against GATA6 RNA in situ hybridization (Spearman $\rho = `r sprintf("%.2f", .tcs$gata6$rho)`$, $P < 0.001$; SI Appendix). Because treatment was confounded with cohort and stage, these analyses address prognosis rather than treatment prediction.

```{r fig-treated, fig.cap = "Survival-supervised programs show context-dependent prognostic associations in treated PDAC. Trained DeSurv programs were projected without retraining onto three bulk treated cohorts (Linehan, borderline-resectable/locally-advanced, $n=66$; Rash, metastatic, $n=85$; Accept, metastatic, $n=70$) and analyzed per cohort. (A) Per-cohort overall-survival hazard ratios per SD for each program (D1 and D3 marginal; D2 PurIST + DeCAF-adjusted); D2 (proCAF) is protective in all three cohorts, D1 (basal--classical) is prognostic only in the less-advanced cohort, and D3 is null on overall survival. (B) The basal--classical prognostic axis is setting-dependent and concordant across measures: Classical-vs-Basal overall-survival hazard ratios for the PurIST, Moffitt, and Bailey classifiers in less-advanced versus metastatic disease, all attenuating together in metastatic disease. (C) The proCAF program D2 increases from pre- to post-treatment in paired biopsies (paired Wilcoxon test). (D) Orthogonal validation: the D1 score versus GATA6 RNA in situ hybridization (Spearman correlation). \\label{fig:treated}", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
knitr::include_graphics(file.path("..", "figures", "fig_treated_context.pdf"))
```


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

We asked whether survival supervision improves recovery of prognostic gene programs when prognostic variation is not the dominant source of transcriptomic variance. To address this, we designed simulations with known ground-truth latent structure, in which prognostic gene programs explained low variance relative to outcome-neutral background signals and survival times were generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Materials and Methods, Simulation Studies). These simulations probe this specific regime, in which survival-relevant programs are weaker than variance-dominant background structure, rather than representing all transcriptomic settings. To test whether incorporating survival information during factorization improves recovery, rather than estimating it only afterward, we compared DeSurv to standard NMF ($\alpha=0$), where factors are learned by reconstruction error alone and survival associations are estimated by a separately fit Cox model. Both methods were tuned via BO over cross-validated C-index, so that differences in performance reflect the role of supervision during factorization, not during model selection.

In the prognostic scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index (Fig.\ \ref{fig:sim}A) and substantially higher precision for recovering the true prognostic gene programs (Fig.\ \ref{fig:sim}B). Here precision measures how well the learned factors recover the true prognostic programs, computed per program against its best-matching factor and averaged (SI Appendix, Sections 6 and 9). The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This indicates that outcome supervision during factorization can be required to recover prognostic programs that do not dominate expression variance. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$). The gain in gene-program precision exceeded the gain in C-index, indicating that unsupervised factorization can achieve similar predictive performance while recovering different gene programs. Survival supervision therefore improves not only prediction but also recovery of the underlying prognostic biology. A similar decoupling appears empirically in PDAC at $k = 5$, where supervised (DeSurv) and unsupervised (NMF) factorizations yield substantially different factor-specific top genes (SI Appendix, Fig.\ \ref{fig:fig-desurv-k5-heatmap} [DeSurv $k = 5$] vs Fig.\ \ref{fig:fig-nmf-k5-heatmap} [NMF $k = 5$]), with similar C-indices in external cohorts.

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under a null scenario ($\beta = 0$) and a mixed scenario with partial overlap between variance-dominant and prognostic structure (SI Appendix, Fig.\ \ref{fig:fig-sim-null-mixed}). The benefit of survival supervision scaled with the separation between prognostic and variance-dominant structure: largest in the prognostic scenario ($\Delta = `r delta_sim_cindex`$), attenuated when variance and prognosis partially overlapped ($\Delta = `r delta_mixed_cindex`$, paired Wilcoxon $P `r mixed_wilcox_p`$), and absent when no survival signal existed, where both methods yielded C-index near 0.5. DeSurv also recovered the true rank ($k = 3$) more frequently than standard NMF, which systematically under-selected near $k = 2$ (Fig.\ \ref{fig:sim}C).

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Survival supervision improves recovery of true prognostic gene programs in controlled simulations. (A) Test-set concordance index across 100 simulation replicates ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$) comparing DeSurv to standard NMF ($\\alpha = 0$); paired Wilcoxon $P ", sim_wilcox_p, "$. (B) Precision (fraction of a factor's top-weighted genes, defined in main text, overlapping a true prognostic program) across replicates. In (A) and (B), boxes show median and interquartile range (IQR); whiskers extend to $1.5\\times$IQR; points beyond are individual replicates. The y-axis in (A) is truncated at 0.5 because C-index values below 0.5 indicate worse-than-random concordance. (C) Distribution of selected $k$ values across the same 100 replicates for DeSurv versus standard NMF. (D) Gaussian-process--predicted mean cross-validated C-index from Bayesian optimization over factorization rank ($k$) and supervision strength ($\\alpha$) in PDAC training data (TCGA + CPTAC), showing the broad plateau spanning $k = 3$--$12$ from which the one-standard-error rule selects $k = 3$ (Methods). \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

knitr::include_graphics(file.path("..", "figures", "fig2_tcgacptac.pdf"))

```

