<!-- REVISED RESULTS — prediction-validation structure (originally for PNAS; retained for Nature submission) -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- Supplementary Information figure order. VERIFIED against si_appendix.Rmd
     chunk order 2026-08-15; the previous version of this block was wrong in
     nearly every row and is what the line-by-line pass F11 caught. Re-verify
     with:  grep -n '^```{r fig-' paper/si_appendix.Rmd
     S1  fig-si-bo-tuning      BO tuning surface over k and alpha
     S2  fig-suprec            D1 recovered by independent supervised methods
     S3  fig-sim-null-mixed    null and mixed simulation scenarios
     S4  fig-cindex-by-k       training/CV C-index vs k sensitivity
     S5  fig-nmf-diagnostics   standard NMF rank selection heuristics
     S6  fig-nmf-k7-heatmap    NMF gene overlap at k=7 (broader correspondence)
     S7  fig-rpm               reference-program coverage/edges/multiplicity
                                by rank (companion to S6; added for R5)
     NOT in the SI: fig-treated is now main-text Fig. 4. fig-converge and
     fig-cutpoint-km no longer exist anywhere (cutpoint analysis removed in
     94c1149); do not look for them.
-->

```{r setup}
source("paper/load_precomputed.R")
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(magick)
library(gt)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)
library(gtable)

load_result("sim_figs_by_scenario")

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.sim_wilcox_p_raw <- wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                  paired = TRUE)$p.value
sim_wilcox_p <- if (is.na(.sim_wilcox_p_raw) || .sim_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.sim_wilcox_p_raw, 2))

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
.mixed_wilcox_p_raw <- wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                    paired = TRUE)$p.value
mixed_wilcox_p <- if (is.na(.mixed_wilcox_p_raw) || .mixed_wilcox_p_raw < 0.001) "< 0.001" else sprintf("= %.3g", signif(.mixed_wilcox_p_raw, 2))

# ── Variance vs survival statistics for PDAC training data ───────────────────
load_result("fig_variation_explained_tcgacptac")
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)
desurv_maxvar_label <- paste0("D", desurv_maxvar$factor)
nmf_maxvar_label    <- paste0("N", nmf_maxvar$factor)
# Exocrine NMF factor (N2) — referenced explicitly because of its biological
# role as the high-variance exocrine-associated factor
nmf_n2              <- nmf_vs[nmf_vs$factor == 2, ]
nmf_n2_pct          <- round(nmf_n2$variance_explained * 100, 1)
nmf_n2_dl           <- round(nmf_n2$delta_loglik, 2)
# Per-factor survival contribution for the variance-versus-prognosis sentence:
# D1 carries essentially all of it despite the smallest reconstruction share.
.dl <- setNames(desurv_vs$delta_loglik, paste0("D", desurv_vs$factor))
.vp <- setNames(round(100 * desurv_vs$variance_explained, 1), paste0("D", desurv_vs$factor))

# ── Clinical / purity adjustment, per cohort ─────────────────────────────────
.capur <- local({
  ca <- read_result("desurv_clinical_adjustment")
  t(vapply(ca, function(r) c(unadj = r$unadjusted$hr,
                             pur   = if (is.null(r$purity_adjusted)) NA_real_ else r$purity_adjusted$hr),
           numeric(2)))
})

# ── Whole-matrix reconstruction R^2, centered denominator ────────────────────
# The cached helper reports an UNCENTERED R^2 on a rank matrix, where the grand
# mean alone carries ~75% of the denominator. Quote the conventional centered
# value in prose (see R/reconstruction_helpers.R and SI Table \ref{tab:overall-r2}).
source("R/reconstruction_helpers.R")
.Xr2      <- as.matrix(read_result("tar_data_filtered_tcgacptac")$ex)
.fit_d_r2 <- read_result("tar_fit_desurv_tcgacptac")
.fit_n_r2 <- read_result("fit_std_desurvk_tcgacptac")
.r2c_desurv <- overall_recon_r2_centered(.fit_d_r2$W, .fit_d_r2$H, .Xr2)
.r2c_nmf    <- overall_recon_r2_centered(.fit_n_r2$W, .fit_n_r2$H, .Xr2)
.r2c_gap    <- 100 * (.r2c_nmf - .r2c_desurv)

# Size of the DeSurv projection support: union across factors of the
# BO-selected top genes. Used in Methods when contrasting comparator sparsity.
source("R/get_top_genes.R")
.n_support <- length(unique(unlist(get_top_genes(.fit_d_r2$W,
                    as.integer(read_result("tar_params_best_tcgacptac")$ntop))$top_genes)))

# ── Cross-method correspondence and projection R^2 ───────────────────────────
# Previously typed as literals in prose (0.84 / 0.87 / 0.36 / 0.09). Values are
# unchanged; they now come from the fits so they cannot drift out of date.
.Wd <- .fit_d_r2$W; .Wn <- .fit_n_r2$W
.corrC   <- cor(.Wd, .Wn, method = "spearman")
.corr_D2 <- max(abs(.corrC[2, ])); .corr_D3 <- max(abs(.corrC[3, ]))
.corr_D1 <- max(abs(.corrC[1, ]))
.r2_projD1 <- projection_r2(.Wd, .Wn)[1]

# Stromal-resolution accessors were removed with the count-based sentence:
# the main text now describes the pattern shown in Fig. 2a,b rather than
# quoting marker-gene counts, which read as weak in isolation. The numbers
# remain in code/23 and the SI (Stromal resolution).

# ── Tuning-budget control (code/03 sec. 5, code/04 sec. 1d) ─────────────────
# The matched-rank comparator above is NMF::nmf, which gets neither DeSurv's
# consensus initialization nor its tuning budget. To show the reorganization is
# not an artifact of that asymmetry, an unsupervised model (alpha = 0) was fit
# at the same rank using DeSurv's own optimizer, consensus initialization and
# an identical Bayesian-optimization budget. These are its correspondences.
.Wf <- read_result("tar_fit_desurv_a0k3_tcgacptac")$W
.corrF   <- cor(.Wd, .Wf, method = "spearman")
.corrF_D1 <- max(abs(.corrF[1, ]))
.corrF_D2 <- max(abs(.corrF[2, ]))
.corrF_D3 <- max(abs(.corrF[3, ]))

# Supervised risk scores versus the D2 stromal program (tuned comparators)
.ax <- read_result("desurv_vs_supervised_tuned")$axis_decomposition

# ── Orthogonal GATA6 validation (COMPASS) ────────────────────────────────────
# Values come from the cached treated-cohort statistics rather than being typed
# in. rho_excl is the self-correlation sensitivity with GATA6 removed from its
# own projection support (see code/11).

# Treated-cohort quantitative results reported in the Fig. 4 caption: the
# O'Kane/COMPASS arm-stratified common association (panel a), the paired
# post-treatment D2 shift (panel b), and the between-cohort heterogeneity of D1
# and the paired post-treatment D2 shift (panel b).
.tr    <- read_result("treated_cohort_stats")
.g6    <- .tr$gata6
.trok  <- .tr$okane$stratified
.trint <- .tr$okane$interaction_p
.trpr  <- .tr$paired

# Only two treated resources are used: the O'Kane/COMPASS arm-stratified D1
# transportability (panel a) and the Linehan paired pre/post D2 shift (panel b).
# No program-by-survival association is reported in bulk treated tissue.

# ── Helper: extract pooled risk_score data from validation latent targets ─────
extract_pool_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    data.frame(id = rownames(surv), time = surv$time, event = surv$event,
               dataset = entry$dataset, risk_score = entry$risk_score,
               stringsAsFactors = FALSE)
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df <- df[is.finite(df$time) & is.finite(df$risk_score), ]
  # Combined analysis: one record per unique patient (drop PACA-AU array
  # duplicates, keep RNA-seq), per the shared DeCAF-consistent rule.
  dedup_combined(df, dataset_col = "dataset", id_col = "id")
}

# ── Helper: fit stratified Cox on SD-standardized risk_score ─────────────────
pooled_cox_stats <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_fit <- coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df)
  cox_sum <- summary(cox_fit)
  list(
    hr    = round(cox_sum$coefficients[, "exp(coef)"], 2),
    ci_lo = round(cox_sum$conf.int[, "lower .95"], 2),
    ci_hi = round(cox_sum$conf.int[, "upper .95"], 2),
    p     = signif(cox_sum$coefficients[, "Pr(>|z|)"], 2)
  )
}

# ── Pooled PDAC validation statistics (DeSurv) ───────────────────────────────
load_result("val_latent_desurv_tcgacptac")
pool_df      <- extract_pool_df(val_latent_desurv_tcgacptac)
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))

# ── Combined per-factor associations (the diamonds in Fig. 3a) ───────────────
# The Results assert D1 is the only individual program associated with survival
# in the combined analysis; these are the numbers behind that claim. Computed
# here rather than typed, on the same de-duplicated, cohort-stratified footing
# as the pooled linear predictor above.
.pf <- local({
  rows <- lapply(val_latent_desurv_tcgacptac, function(e) {
    Z <- as.matrix(e$latent); sv <- e$survival
    data.frame(sample_id = rownames(sv), dataset = e$dataset,
               D1 = Z[, 1], D2 = Z[, 2], D3 = Z[, 3],
               time = sv$time, event = sv$event, stringsAsFactors = FALSE)
  })
  d <- do.call(rbind, rows); d <- d[is.finite(d$time) & d$time > 0, ]
  d <- dedup_combined(d, dataset_col = "dataset", id_col = "sample_id")
  out <- lapply(c("D1", "D2", "D3"), function(p) {
    d$z <- ave(d[[p]], d$dataset, FUN = function(x) as.numeric(scale(x)))
    m <- summary(coxph(Surv(time, event) ~ z + strata(dataset), data = d))
    data.frame(prog = p, hr = m$conf.int["z", 1], lo = m$conf.int["z", 3],
               hi = m$conf.int["z", 4], p = m$coefficients["z", 5])
  })
  do.call(rbind, out)
})
# Same convention for the standard NMF factors shown on the right of Fig. 3a
.pfn <- local({
  if (!exists("val_latent_std_desurvk_tcgacptac")) load_result("val_latent_std_desurvk_tcgacptac")
  rows <- lapply(val_latent_std_desurvk_tcgacptac, function(e) {
    Z <- as.matrix(e$latent); sv <- e$survival
    data.frame(sample_id = rownames(sv), dataset = e$dataset,
               N1 = Z[, 1], N2 = Z[, 2], N3 = Z[, 3],
               time = sv$time, event = sv$event, stringsAsFactors = FALSE)
  })
  d <- do.call(rbind, rows); d <- d[is.finite(d$time) & d$time > 0, ]
  d <- dedup_combined(d, dataset_col = "dataset", id_col = "sample_id")
  out <- lapply(c("N1", "N2", "N3"), function(p) {
    d$z <- ave(d[[p]], d$dataset, FUN = function(x) as.numeric(scale(x)))
    m <- summary(coxph(Surv(time, event) ~ z + strata(dataset), data = d))
    data.frame(prog = p, hr = m$conf.int["z", 1], lo = m$conf.int["z", 3],
               hi = m$conf.int["z", 4], p = m$coefficients["z", 5])
  })
  out <- do.call(rbind, out)
  stopifnot(all(out$p > 0.05))   # the sentence in the text asserts no association
  out
})
.pff <- function(p) { r <- .pf[.pf$prog == p, ]
  sprintf("$%.2f$, 95\\%% CI $%.2f$--$%.2f$", r$hr, r$lo, r$hi) }
.pfp <- function(p) { r <- .pf[.pf$prog == p, ]
  if (r$p < 0.001) "< 0.001" else sprintf("= %.2f", r$p) }

# ── Leave-one-cohort-out stability of the pooled hazard ────────────────
# Puleo contributes roughly half the combined validation set, so the pooled
# estimate should be shown not to rest on it. Computed here rather than typed.
# The concentration itself is stated in the Fig. 3 legend; the LOCO range below
# is what demonstrates the estimate does not depend on it.
.n_puleo   <- sum(grepl("^Puleo", pool_df$dataset))
.pct_puleo <- 100 * .n_puleo / n_val_total
.loco <- local({
  d <- pool_df
  fitone <- function(x) {
    x$rz <- x$risk_score / sd(x$risk_score)
    summary(coxph(Surv(time, event) ~ rz + strata(dataset), data = x))$conf.int["rz", "exp(coef)"]
  }
  vapply(sort(unique(d$dataset)), function(co) fitone(d[d$dataset != co, ]), numeric(1))
})
.loco_lo <- min(.loco); .loco_hi <- max(.loco)

# Build-time safeguard against pipeline drift: report the combined-cohort counts
# and flag any deviation from the expected de-duplicated totals. dedup_combined()
# already hard-asserts one record per unique patient.
message(sprintf("Unique validation patients: %d\nOS events: %d", n_val_total, n_val_events))
if (n_val_total != DESURV_EXPECTED_COMBINED_PATIENTS || n_val_events != DESURV_EXPECTED_COMBINED_EVENTS)
  warning(sprintf("Combined validation counts (%d patients, %d events) differ from expected (%d, %d): check PACA-AU de-duplication / cohort inputs.",
                  n_val_total, n_val_events, DESURV_EXPECTED_COMBINED_PATIENTS, DESURV_EXPECTED_COMBINED_EVENTS))
desurv_stats <- pooled_cox_stats(pool_df)

# ── Guard: the abstract's headline HR cannot be computed, so verify it ────────
# The abstract lives in paper.Rmd's YAML header, where inline R does not
# evaluate, so the pooled HR is necessarily typed there by hand. It has drifted
# once already (1.46 -> 1.42 in the Aug 2026 revision). Same shape as the
# combined-counts tripwire above: compare the typed value against the computed
# one and fail the build rather than silently shipping a stale abstract.
local({
  src <- tryCatch(readLines("paper/paper.Rmd", warn = FALSE),
                  error = function(e) character())
  pat <- "HR per SD, ([0-9.]+); 95% CI, ([0-9.]+)--([0-9.]+)"
  hit <- grep(pat, src, value = TRUE)
  if (!length(hit)) {
    warning("Abstract-HR guard did not match. If the abstract was rephrased, ",
            "update the pattern in 04_results_REVISED.Rmd so this check keeps firing.")
    return(invisible(NULL))
  }
  got  <- as.numeric(regmatches(hit[1], regexec(pat, hit[1]))[[1]][-1])
  want <- unname(c(desurv_stats$hr, desurv_stats$ci_lo, desurv_stats$ci_hi))
  if (!isTRUE(all.equal(got, want)))
    stop(sprintf(paste0("Abstract HR is stale. paper/paper.Rmd says %s (%s-%s); ",
                        "the computed pooled estimate is %s (%s-%s). Update the abstract."),
                 got[1], got[2], got[3], want[1], want[2], want[3]))
})

hr_pool      <- sprintf("%.2f", desurv_stats$hr)
ci_pool_lo   <- sprintf("%.2f", desurv_stats$ci_lo)
ci_pool_hi   <- sprintf("%.2f", desurv_stats$ci_hi)
p_pool       <- desurv_stats$p

# ── Between-cohort heterogeneity, C-index CIs, PH diagnostic (code/25) ───────
# The cohort-by-score interaction standardizes the linear predictor WITHIN
# cohort, the same convention as the per-factor forest estimates in `.pf` above
# and deliberately NOT the pooled-SD convention of pooled_cox_stats(), so the
# test cannot be reading a between-cohort difference in score scale. Cochran's
# Q and I^2 are computed from the same within-cohort-standardized per-cohort
# estimates and are descriptive companions to that test, not a substitute.
# `.het$ph` reproduces the SI proportional-hazards diagnostic verbatim; it is
# cached in code/25 only because inline R in the main text cannot reach a
# variable defined in si_appendix.Rmd, and code/25 stops if it ever disagrees.
.het   <- read_result("validation_heterogeneity_ci")
.hetp  <- .het$cindex[.het$cindex$cohort == "POOLED", ]
.brsp  <- read_result("desurv_brier_stats")$pooled
# Clinical-covariate adjustment is fitted per cohort in code/18; the SI table
# tab:clinical-adjustment has no pooled row and there is no cached pooled
# clinical-adjusted estimate, so the main text quotes the largest validation
# cohort (Puleo, 288 of the 570 unique patients) rather than inventing one.
.cladj <- read_result("desurv_clinical_adjustment")$Puleo_array$clinical_adjusted
# sprintf("%.1e") renders as "1.3e - 04" inside math mode; write the exponent out.
.fmt_sci <- function(p) { e <- floor(log10(p))
  sprintf("%.1f \\times 10^{%d}", p / 10^e, e) }

# ── Pooled adjusted (PurIST + DeCAF) statistics for DeSurv ───────────────────
load_result("data_val_filtered_tcgacptac")

extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(sample_id = rownames(surv), time = surv$time, event = surv$event,
                     dataset = entry$dataset, risk_score = entry$risk_score,
                     stringsAsFactors = FALSE)
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo; si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df <- df[is.finite(df$time) & is.finite(df$risk_score), ]
  # Combined analysis: one record per unique patient, per the shared rule.
  dedup_combined(df, dataset_col = "dataset", id_col = "sample_id")
}

pool_df_adj_complete <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
pool_df_adj_complete <- pool_df_adj_complete[
  complete.cases(pool_df_adj_complete[, c("PurIST", "DeCAF")]), ]
rs_sd_adj <- sd(pool_df_adj_complete$risk_score, na.rm = TRUE)
if (is.na(rs_sd_adj) || rs_sd_adj < 1e-12) rs_sd_adj <- 1
pool_df_adj_complete$risk_z <- pool_df_adj_complete$risk_score / rs_sd_adj
cox_adj_sum <- summary(coxph(
  Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
  data = pool_df_adj_complete))
hr_adj      <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "exp(coef)"])
ci_adj_lo   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "lower .95"])
ci_adj_hi   <- sprintf("%.2f", cox_adj_sum$conf.int["risk_z", "upper .95"])
p_adj_val   <- cox_adj_sum$coefficients["risk_z", "Pr(>|z|)"]
p_adj_text  <- if (p_adj_val < 0.001) "< 0.001" else paste0("= ", signif(p_adj_val, 2))

# ── Pooled PDAC validation statistics (NMF at matched rank) ──────────────────
load_result("val_latent_std_desurvk_tcgacptac")
pool_df_nmf  <- extract_pool_df(val_latent_std_desurvk_tcgacptac)
nmf_stats    <- pooled_cox_stats(pool_df_nmf)
hr_pool_nmf    <- sprintf("%.2f", nmf_stats$hr)
ci_pool_lo_nmf <- sprintf("%.2f", nmf_stats$ci_lo)
ci_pool_hi_nmf <- sprintf("%.2f", nmf_stats$ci_hi)
p_pool_nmf     <- nmf_stats$p

# ── KM dichotomized risk-group statistics (extracted from plot annotations) ───

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- read_result("tar_k_selection_tcgacptac")
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- read_result("desurv_bo_results_tcgacptac")$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)

```

## DeSurv directly incorporates survival information into gene program discovery {.unnumbered}

DeSurv learns gene programs by balancing reconstruction of the tumor transcriptome with association to patient survival, so that outcome information shapes which programs are recovered rather than only which programs are selected after an unsupervised factorization. A supervision parameter $\alpha$ controls the strength of survival supervision, with $\alpha = 0$ leaving a purely reconstruction-driven factorization, and the rank $k$ sets how many programs are extracted (Fig.\ \ref{fig:schema}a; Methods). After training, the gene weights and the Cox coefficients are fixed, so the same learned programs and the trained risk score apply unchanged to new tumors without refitting the factorization.

We trained DeSurv in pooled TCGA and CPTAC PDAC cohorts [@raphael2017integrated; @cao2021proteogenomic] ($n = `r ncol(read_result("tar_data_filtered_tcgacptac")$ex)`$, `r sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event)` events). Hyperparameters were selected by Bayesian optimization [@snoek2012practical; @shahriari2016taking], a sequential search algorithm that proposes each new setting from the results of previous ones to optimize a pre-specified objective, here cross-validated concordance (the C-index, the probability that the model ranks a pair of patients in the correct survival order; Methods). Concordance favored intermediate survival supervision over the unsupervised $\alpha = 0$ boundary but changed little across ranks (Supplementary Figs.\ \ref{fig:fig-si-bo-tuning} and \ref{fig:fig-cindex-by-k}). We therefore applied the one-standard-error rule, which takes the most parsimonious model within one standard error of the best, selecting $k = `r read_result("tar_k_selection_tcgacptac")$k_selected`$; the remaining hyperparameters, including $\alpha = `r round(read_result("tar_params_best_tcgacptac")$alpha, 3)`$, were retained from the overall best configuration, which had rank $k = `r read_result("tar_k_selection_tcgacptac")$k_best`$ (Fig.\ \ref{fig:schema}b; Methods; Supplementary Information, Bayesian Optimization). The reported combination was therefore never itself evaluated`r stopifnot(!any(abs(k3_hist$alpha_grid - read_result("tar_params_best_tcgacptac")$alpha) < 1e-9 & abs(k3_hist$lambda_grid - read_result("tar_params_best_tcgacptac")$lambda) < 1e-9 & abs(k3_hist$nu_grid - read_result("tar_params_best_tcgacptac")$nu) < 1e-9 & k3_hist$ntop == read_result("tar_params_best_tcgacptac")$ntop))`. In contrast, standard unsupervised rank criteria (reconstruction residuals, cophenetic correlation and silhouette width [@brunet2004metagenes; @frigyesi2008nmf]) gave conflicting rank choices in the same data (Supplementary Fig.\ \ref{fig:fig-nmf-diagnostics}). All model and hyperparameter selection was completed within the training cohorts.

The final fit was initialized from a consensus of `r read_result("program_stability_stats")$meta$n_restarts` random restarts (Methods). Individual restarts reached similar training concordance (median C-index `r sprintf("%.2f", read_result("program_stability_stats")$cindex[["median"]])`, range `r sprintf("%.2f", read_result("program_stability_stats")$cindex[["min"]])`--`r sprintf("%.2f", read_result("program_stability_stats")$cindex[["max"]])`) yet dispersed gene sets (median pairwise Jaccard of top-ranked genes `r sprintf("%.2f", median(read_result("program_stability_stats")$pairwise$top_jaccard[, "median"]))` against `r sprintf("%.2f", read_result("program_stability_stats")$meta$jaccard_chance)` by chance), so the consensus step, not any single restart, defines the reported programs. Consensus fits from each half of `r read_result("consensus_endpoint_stats")$meta$n_splits` split-half partitions of that restart pool agreed closely with the reported basis (median gene-loading correlations `r sprintf("%.2f", min(read_result("consensus_endpoint_stats")$vs_reported$loading[, "median"]))`--`r sprintf("%.2f", max(read_result("consensus_endpoint_stats")$vs_reported$loading[, "median"]))`, median top-ranked-gene Jaccard `r sprintf("%.2f", min(read_result("consensus_endpoint_stats")$vs_reported$top_jaccard[, "median"]))`--`r sprintf("%.2f", max(read_result("consensus_endpoint_stats")$vs_reported$top_jaccard[, "median"]))`; Supplementary Information, Stability of the recovered gene programs). These partitions assess sensitivity to restart-pool composition within the same training data; they do not independently repeat the full discovery procedure.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (a) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, the unsupervised limit) and survival prediction ($\\alpha$ close to 1). (b) Hyperparameters (rank $k$, supervision strength $\\alpha$, regularization $\\lambda$, $\\xi$ and $n_{\\text{top}}$) are searched by Bayesian optimization over cross-validated concordance; the rank is then chosen by a one-standard-error rule and the remaining hyperparameters are retained from the best configuration (Methods). Survival supervision acts on the gene-program matrix $W$ during discovery, rather than being applied only after an unsupervised factorization has been completed; following training-only model selection, the resulting gene weights are fixed for direct evaluation in external cohorts. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "model_schematic_final.pdf"))
```

## Survival supervision changes which gene programs are recovered {.unnumbered}

To compare the recovered gene programs at the same rank, we fitted DeSurv and standard Lee multiplicative NMF with $k = 3$. DeSurv resolved three distinct programs (D1--D3): D1, enriched for classical tumor and restCAF-like stromal signatures; D2, enriched for proCAF-like stroma; and D3, enriched for basal-like tumor programs (Fig.\ \ref{fig:pdac}a). Together, these programs define a tumor--stroma architecture organized around the survival-aligned D1 direction. In the COMPASS trial cohort, D1 scores correlated with GATA6 RNA in situ hybridization, independently supporting the classical tumor component of D1 (Spearman $\rho = `r sprintf("%.2f", .g6$rho)`$, $P `r if (.g6$p < 0.001) "< 0.001" else sprintf("= %.3f", .g6$p)`$, $n = `r .g6$n`$; Fig.\ \ref{fig:pdac}e).

Standard NMF, the unsupervised comparator, recovered a different set of programs (Fig.\ \ref{fig:pdac}b). Its malignant factor (N1) was enriched for both classical and basal-like tumor programs, its microenvironmental factor (N3) for multiple stromal and immune programs [@leary2023s; @Maurer2019], and a third factor (N2) for exocrine-associated variation. Both methods recovered the activated-stroma and proCAF-like reference programs but differed at the tumor-restraining end: among the displayed top-ranked genes, the restCAF and iCAF reference programs corresponded to a DeSurv program but to no standard NMF factor (Fig.\ \ref{fig:pdac}a,b). These are bulk gene-program correspondences, not identifications of CAF cell states (Supplementary Information, Stromal resolution).

This difference reflected selective reorganization rather than wholesale loss of expression structure. D1 accounted for only $`r sprintf("%.1f", .vp["D1"])`$% of the full model's reconstruction gain (Shapley shares [@shapley1953value]), the smallest share among the three programs. However, in post-fit unpenalized Cox models of the three training-cohort scores projected over all analyzed genes, omitting D1 and refitting the Cox model using the remaining scores lowered the maximized Cox partial log-likelihood, the fit criterion of the Cox model, by $`r sprintf("%.1f", .dl["D1"])`$, compared with $`r sprintf("%.1f", .dl["D3"])`$ for D3 and $`r sprintf("%.2f", .dl["D2"])`$ for D2 (Fig.\ \ref{fig:pdac}c). These conditional, nonadditive increments describe associations in the training data used for discovery and do not independently validate prognostic relevance. Thus, the programs that best reconstruct the transcriptome need not be those most relevant to survival [@aran2015systematic].

The principal NMF tumor and microenvironmental factors corresponded closely to D3 ($\rho = `r sprintf("%.2f", .corr_D3)`$) and D2 ($\rho = `r sprintf("%.2f", .corr_D2)`$), whereas D1 correlated only weakly with every matched-rank NMF factor ($\rho \leq `r sprintf("%.2f", .corr_D1)`$; gene-loading Spearman correlation over all analyzed genes; Fig.\ \ref{fig:pdac}d). DeSurv reconstructed the analysis matrix less completely than matched-rank NMF ($R^2$ `r sprintf("%.2f", .r2c_desurv)` versus `r sprintf("%.2f", .r2c_nmf)`, a difference of `r sprintf("%.1f", .r2c_gap)` percentage points; Supplementary Table \ref{tab:overall-r2}). Survival supervision therefore preserved most of the variance-dominant tumor and stromal structure while resolving a distinct survival-aligned program.

Tuning effort did not appear to explain this difference. We repeated the comparison with a matched-budget unsupervised control, which used DeSurv's optimizer, gene filter, cross-validation folds, Bayesian-optimization budget and consensus initialization, with rank fixed at three, $\alpha$ fixed at zero and the remaining hyperparameters re-optimized (Methods). The control again recovered D2 ($\rho = `r sprintf("%.2f", .corrF_D2)`$) and D3 ($\rho = `r sprintf("%.2f", .corrF_D3)`$) but similarly failed to recover D1 ($\rho \leq `r sprintf("%.2f", .corrF_D1)`$).
```{r fig-pdac, fig.width=7, fig.height=7, fig.cap = paste0("Survival supervision produces different factor structures in PDAC (TCGA + CPTAC training cohort, $n=", ncol(read_result("tar_data_filtered_tcgacptac")$ex), "$ patients, ", sum(read_result("tar_data_filtered_tcgacptac")$sampInfo$event), " events). (a,b) Spearman correlations between factor gene rankings (top 50 genes per factor) and established PDAC gene programs for DeSurv (a) and standard NMF (b) at $k = 3$. Both panels display the same rows, the union of reference programs correlating at Spearman $\\rho > 0.2$ with at least one factor of either method; grey cells denote programs with none of their genes in that method's displayed gene space, where the correlation is not computable. The rows labeled SCISSORS: restCAF and SCISSORS: proCAF display the SCISSORS iCAF and myCAF gene sets under the naming used by the DeCAF classification (Supplementary Information). These rows are correspondences between bulk gene programs and reference signatures, not direct identification of CAF cell states. (c) Per-factor reconstruction contribution (normalized Shapley share, summing to 100\\%) versus survival contribution, the decrease in the maximized unpenalized Cox partial log-likelihood when one factor is dropped from the three training-cohort scores projected over all analyzed genes and the Cox model is refitted using the remaining scores. These increments are conditional and nonadditive; D1 was estimated using these survival outcomes, so its increment locates the survival signal in the fitted model rather than testing whether that signal is real. (d) Pairwise correspondence between NMF and DeSurv factors (Spearman correlation of W-matrix gene loadings over all genes). (e) DeSurv D1 score versus GATA6 RNA in situ hybridization level in the COMPASS cohort (Spearman correlation; $n = 33$). \\label{fig:pdac}"), fig.env='figure*', fig.pos='t', out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig3_tcgacptac.pdf"))
```

## Frozen programs retain prognostic relevance across independent PDAC cohorts, with exploratory transport to treated disease {.unnumbered}

We next projected the frozen programs into five external validation datasets from four independent, non-metastatic, treatment-naive PDAC cohorts (Dijk, Moffitt, PACA-AU [array and RNA-seq], and Puleo; Supplementary Table \ref{tab:cohorts}). Each tumor was scored by projecting its expression onto the trained programs over the union of their top-ranked genes (`r .n_support` genes; $Z = \widetilde{W}^\top X_{\text{new}}$; Methods). Combined analyses represented each patient once, yielding $`r n_val_total`$ unique patients and `r n_val_events` deaths (Methods).

D1 had protective point estimates in all five validation datasets and was the only individual DeSurv program with a statistically significant combined association in the dataset-stratified analysis (HR per within-dataset SD `r .pff("D1")`, $P$ `r .pfp("D1")`). D3 had hazard ratios above 1 in four of the five datasets (`r .pff("D3")`, $P `r .pfp("D3")`$), while D2 was near-null overall (`r .pff("D2")`, $P `r .pfp("D2")`$) and varied more across datasets; neither reached statistical significance individually or in combination (Fig.\ \ref{fig:val}a). Nor did any of the three standard NMF factors at the same rank (HRs `r sprintf("%.2f", min(.pfn$hr))`--`r sprintf("%.2f", max(.pfn$hr))`; Fig.\ \ref{fig:val}a, right).

The trained DeSurv linear predictor combines D1 and D3, with coefficients oriented so that higher values indicate higher risk; the elastic-net penalty set the D2 coefficient to exactly zero`r stopifnot(read_result("tar_fit_desurv_tcgacptac")$beta[2] == 0)`. It was associated with overall survival in the dataset-stratified analysis (HR per pooled SD `r hr_pool`, 95\% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$, $P = `r p_pool`$; Table \ref{tab:hr-adjusted}), and leave-one-dataset-out estimates were similar (HRs `r sprintf("%.2f", .loco_lo)`--`r sprintf("%.2f", .loco_hi)`; Fig.\ \ref{fig:val}a). Scaled Schoenfeld residuals indicated attenuation over follow-up ($\chi^2 = `r sprintf("%.1f", .het$ph$chisq)`$, `r .het$ph$df` df, $P = `r .fmt_sci(.het$ph$p)`$; Supplementary Information, Proportional-hazards diagnostic), so it should be read as a time-averaged summary rather than a constant multiplicative effect.

Allowing cohort-specific slopes did not significantly improve fit over a common slope (likelihood-ratio $\chi^2 = `r sprintf("%.1f", .het$lrt$stat)`$, `r .het$lrt$df` df, $P = `r sprintf("%.2f", .het$lrt$p)`$; score standardized within each of the four cohorts). Discrimination across cohorts was moderate (event-weighted C-index $`r sprintf("%.2f", .hetp$cindex)`$, 95\% CI $`r sprintf("%.2f", .hetp$lo)`$--$`r sprintf("%.2f", .hetp$hi)`$). In a sensitivity analysis recalibrating the frozen score in sample within each validation dataset (apparent rather than out-of-sample error), the integrated Brier score, an event-weighted summary over dataset-specific follow-up horizons, was $`r sprintf("%.3f", .brsp["ibs"])`$ compared with $`r sprintf("%.3f", .brsp["ibs_null"])`$ for a Kaplan--Meier reference (Supplementary Information).

We also projected the frozen programs, without retraining, into two treated PDAC cohorts. In the O'Kane/COMPASS cohort of metastatic disease ($n = `r read_result("treated_cohort_stats")$okane$n`$, `r read_result("treated_cohort_stats")$okane$events` deaths), D1 remained associated with overall survival in a Cox model stratified by treatment arm (HR per SD $`r sprintf("%.2f", read_result("treated_cohort_stats")$okane$stratified["hr"])`$, 95\% CI $`r sprintf("%.2f", read_result("treated_cohort_stats")$okane$stratified["lo"])`$--$`r sprintf("%.2f", read_result("treated_cohort_stats")$okane$stratified["hi"])`$; Fig.\ \ref{fig:treated}a); these laser-capture microdissected profiles are epithelial-enriched, probing chiefly the tumor component of D1. In the Linehan cohort, the projected proCAF-like D2 score increased from pre- to post-treatment biopsies (mean change `r sprintf("%+.2f", read_result("treated_cohort_stats")$paired$mean_delta)` SD, paired Wilcoxon $P = `r sprintf("%.3f", read_result("treated_cohort_stats")$paired$wilcox_p)`$; Fig.\ \ref{fig:treated}b; Supplementary Information, Translational transportability), a shift that may reflect cell composition, cellular state or both.

```{r fig-val, fig.cap = paste0("Survival-aligned programs generalize across independent PDAC patient cohorts (five external validation datasets: Dijk, Moffitt GEO array, PACA-AU array, PACA-AU RNA-seq and Puleo; combined $n=", n_val_total, "$ unique patients, ", n_val_events, " deaths, with PACA-AU patients profiled on both platforms represented once via RNA-seq). (a) Forest plot of per-dataset univariate hazard ratios (95\\% CIs) for each DeSurv program (left) and standard NMF factor (right); the platform-specific PACA-AU array and RNA-seq estimates are shown separately, and the black diamonds denote the combined estimate over unique patients (per-factor dataset-stratified Cox on the de-duplicated patient-level data, with each program standardized within dataset and scored on the union top-gene support; PACA-AU array duplicates excluded). Puleo is the largest contributor, supplying $", .n_puleo, "$ of the $", n_val_total, "$ unique patients (", sprintf("%.0f", .pct_puleo), "\\%), which is why the combined estimate is also reported with each dataset omitted in turn; those leave-one-dataset-out estimates range from $", sprintf("%.2f", .loco_lo), "$ to $", sprintf("%.2f", .loco_hi), "$. (b) Correspondence ($|$Pearson $r|$) of tuned supervised risk scores (lasso Cox and supervised PCA) with the three DeSurv programs (D1--D3) in the training data; supervised scores concentrate on D1 and are only weakly correlated with the D2 stromal program. These score correspondences do not establish recovery of identical gene programs. No validation-cohort survival outcomes were used in training, scoring or hyperparameter selection. \\label{fig:val}"), fig.env='figure*', fig.pos='t', fig.width=7, fig.height=4.8, out.height= "4.5in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..", "figures", "fig4_tcgacptac.pdf"))
```

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(knitr); library(kableExtra)
# extract_pool_df_with_covariates() and data_val_filtered_tcgacptac
# are defined in the setup chunks above.
load_result("val_latent_desurv_alpha0_tcgacptac")

pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(hr = round(ci_row[, "exp(coef)"], 2),
         ci_lo = round(ci_row[, "lower .95"], 2),
         ci_hi = round(ci_row[, "upper .95"], 2),
         p = coef_row[, "Pr(>|z|)"],
         n = cx$n, nevent = cx$nevent)
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

pool_desurv_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv_t1 <- pooled_cox_unadj_adj(pool_desurv_t1)
pool_nmf7_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7_t1 <- pooled_cox_unadj_adj(pool_nmf7_t1)
# Tuning-budget control: unsupervised at DeSurv's rank, DeSurv's optimizer,
# consensus initialization and an identical BO budget (code/03 sec. 5).
load_result("val_latent_desurv_a0k3_tcgacptac")
pool_a0k3_t1 <- extract_pool_df_with_covariates(
  val_latent_desurv_a0k3_tcgacptac, data_val_filtered_tcgacptac)
stats_a0k3_t1 <- pooled_cox_unadj_adj(pool_a0k3_t1)

# The adjusted rows are all fit on the same classifier-complete subset; the
# caption reports that n once, so fail loudly if the methods ever diverge.
stopifnot(stats_desurv_t1$adj$n == stats_nmf7_t1$adj$n,
          stats_desurv_t1$adj$n == stats_a0k3_t1$adj$n)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)",
             "Matched-budget unsupervised control ($k = 3$, $\\alpha = 0$)",
             "Rank-optimized unsupervised control ($k = 7$, $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv_t1$unadj), fmt_hr(stats_a0k3_t1$unadj),
                 fmt_hr(stats_nmf7_t1$unadj)),
  Adjusted = c(fmt_hr(stats_desurv_t1$adj), fmt_hr(stats_a0k3_t1$adj),
               fmt_hr(stats_nmf7_t1$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per pooled SD (95\\% CI), $P$",
                       "Adjusted HR per pooled SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = paste0("Pooled external validation of DeSurv and unsupervised controls. Hazard ratios per pooled SD of the linear predictor across five validation datasets spanning four independent cohorts (", n_val_total, " unique patients, ", n_val_events, " deaths). Adjusted models include PurIST and DeCAF and are stratified by dataset; all ", stats_desurv_t1$adj$n, " patients had both classifier calls. The matched-budget control fixes rank at $k = 3$, whereas the rank-optimized control uses the rank selected by the same one-standard-error rule applied to its own Bayesian-optimization results ($k = 7$); both use the same fitting pipeline, consensus procedure and outcome-based tuning budget as DeSurv, with $\\alpha$ fixed at zero (Supplementary Table \\ref{tab:tab-cindex-comparison}). These are point estimates for the selected models, not tests of superiority and not a comparison of minimum ranks. \\label{tab:hr-adjusted}")) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"), font_size = 9)
```

```{r fig-treated, fig.cap = paste0("Transportability of the frozen DeSurv programs into independent treated PDAC cohorts. (a) Prognostic transportability of D1 in treated metastatic PDAC. Forest plot of hazard ratios per SD of the frozen D1 score within each O'Kane/COMPASS treatment arm (FOLFIRINOX, gemcitabine/nab-paclitaxel [GA], and GA/experimental), scored from laser-capture microdissected, epithelial-enriched profiles; the diamond is the common association estimated across arms from a stratified Cox model with treatment-arm-specific baseline hazards, HR $", sprintf("%.2f", .trok["hr"]), "$ (95\\% CI $", sprintf("%.2f", .trok["lo"]), "$--$", sprintf("%.2f", .trok["hi"]), "$, $P=", sprintf("%.3f", .trok["p"]), "$). D1 was protective in all three arms; the arm-by-score interaction was not significant (interaction $P=", sprintf("%.2f", .trint), "$), though the individual arms are underpowered for detecting heterogeneity. Error bars denote 95\\% confidence intervals. (b) The proCAF program D2 before and after treatment in paired Linehan biopsies ($n=", .trpr$n, "$, paired Wilcoxon $P=", sprintf("%.3f", .trpr$wilcox_p), "$); grey lines are individual patients, the red line the group mean ($", sprintf("%+.2f", .trpr$mean_delta), "$ SD). Because treatment, stage and cohort were not independently varied, these analyses support prognostic and biological transportability rather than treatment-predictive effects. \\label{fig:treated}"), fig.env='figure*', fig.pos='t', fig.width=7.2, fig.height=3.4, out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(magick::image_read("figures/fig_treated_context.pdf", density = 300))
```

```{r d1-partition-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("d1_variance_partition_stats")
.d1p <- d1_variance_partition_stats
.d1pf <- function(p) if (p < 1e-4) "< 0.0001" else sprintf("= %.4f", p)
load_result("supervised_recovery_stats")
.d1p_sr_mean  <- supervised_recovery_stats$sup_mean_score_cor
.d1p_sr_unsup <- abs(supervised_recovery_stats$table$score_cor_D1[!supervised_recovery_stats$table$supervised])
.srt2 <- supervised_recovery_stats$table
.d1p_srcor <- function(m) sprintf("%.2f", abs(.srt2$score_cor_D1[.srt2$method == m]))
```

## D1 retains prognostic information beyond established PDAC classifications {.unnumbered}

We first asked whether D1 simply recapitulated established PDAC subtype classifications. Across the pooled validation cohorts ($n = `r .d1p$meta$n`$, `r .d1p$meta$n_events` events), binary PurIST and DeCAF classifications together explained `r sprintf("%.0f", 100*.d1p$variance$explained_total)`\% of the within-dataset variance in the continuous D1 score. D1 nevertheless remained associated with survival after adjustment for both classifiers (HR per within-dataset SD `r sprintf("%.2f", .d1p$prognosis$d1_adj_hr)`, 95\% CI $`r sprintf("%.2f", .d1p$prognosis$d1_adj_lo)`$--$`r sprintf("%.2f", .d1p$prognosis$d1_adj_hi)`$, $P `r .d1pf(.d1p$prognosis$d1_adj_p)`$). Adding D1 to a dataset-stratified Cox model with additive PurIST and DeCAF terms improved fit (partial likelihood-ratio $\chi^2 = `r sprintf("%.1f", .d1p$prognosis$lr_chisq)`$, 1 df, $P `r .d1pf(.d1p$prognosis$lr_p)`$). D1 therefore overlaps substantially with established tumor-intrinsic and stromal classifications while retaining prognostic information beyond those binary calls.

Because D1 spans tumor and stromal programs, we also asked whether it primarily reflected tumor purity. In the two datasets with purity estimates for all patients, D1 correlated only weakly with purity (Spearman $\rho = `r sprintf("%.2f", read_result("desurv_clinical_adjustment")$PACA_AU_array$purity_cor$D1)`$ and $`r sprintf("%.2f", read_result("desurv_clinical_adjustment")$PACA_AU_seq$purity_cor$D1)`$), and adjustment for purity left the linear predictor's survival association essentially unchanged (HR per SD: PACA-AU array, `r sprintf("%.2f", .capur["PACA_AU_array","unadj"])` versus `r sprintf("%.2f", .capur["PACA_AU_array","pur"])`; RNA-seq, `r sprintf("%.2f", .capur["PACA_AU_seq","unadj"])` versus `r sprintf("%.2f", .capur["PACA_AU_seq","pur"])`; Supplementary Table \ref{tab:clinical-adjustment}). Within the PACA-AU datasets, these findings argue against measured tumor purity as the main explanation for the observed associations, although they do not establish the cellular identity of D1's restCAF-like component (Supplementary Information, Stromal resolution: which reference programs are recovered from bulk).

## Method comparisons distinguish survival prediction from gene program recovery {.unnumbered}

```{r rpm-stats, echo=FALSE, message=FALSE, warning=FALSE}
.rpm <- read_result("reference_program_multiplicity")
.rpm_get <- function(id, field) .rpm$anchors[[field]][.rpm$anchors$anchor_id == id]
.rpm_co3 <- .rpm_get("desurv_k3", "coverage")
.rpm_ed3 <- .rpm_get("desurv_k3", "edges")
.rpm_mu3 <- sprintf("%.2f", .rpm_get("desurv_k3", "multiplicity"))
.rpm_co7 <- .rpm_get("nmf_k7_unsupervised", "coverage")
.rpm_ed7 <- .rpm_get("nmf_k7_unsupervised", "edges")
.rpm_mu7 <- sprintf("%.3f", .rpm_get("nmf_k7_unsupervised", "multiplicity"))
```

We next asked whether comparable prognostic information could be recovered without survival-supervised factorization. At DeSurv's rank, the matched-budget unsupervised control did not show a statistically significant survival association (Table \ref{tab:hr-adjusted}).

The rank-optimized unsupervised control (the same pipeline at $\alpha = 0$ with rank selected by the same one-standard-error rule applied to its own Bayesian-optimization results, $k = 7$) carried comparable external prognostic information (Table \ref{tab:hr-adjusted}; Supplementary Table \ref{tab:tab-cindex-comparison}). Reference-program coverage increased from `r .rpm_co3` for DeSurv at $k = 3$ to `r .rpm_co7` for the $k = 7$ control, while multiplicity remained low (`r .rpm_mu3` versus `r .rpm_mu7`; Supplementary Figs.\ \ref{fig:fig-nmf-k7-heatmap} and \ref{fig:fig-rpm}), indicating broader reference-program coverage with little repeated mapping in the seven-factor control. Seven is where the one-standard-error rule landed for the unsupervised search, not a general minimum rank for comparable prognostic information. Survival supervision was not required to obtain a prognostic signal, but it altered which programs were identified and how that signal was decomposed.

```{r suprec-bench, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_stats")
load_result("desurv_vs_supervised_tuned")
.svb  <- function(m) sprintf("%.2f", desurv_vs_supervised_stats$pooled[m, "pooled_C"])
.svbt <- function(m) sprintf("%.2f", desurv_vs_supervised_tuned$prediction[m, "POOLED"])
```

We next asked whether D1 was specific to the DeSurv optimizer. We applied three other survival-supervised approaches to the training datasets (one supervised component each for supervised principal components and Cox partial least squares, one penalized fit for sparse Cox; Methods). All three produced patient scores strongly correlated with D1 (Pearson $|r| = `r .d1p_srcor("Supervised PCA")`$, $`r .d1p_srcor("Cox-PLS")`$ and $`r .d1p_srcor("Sparse Cox")`$, respectively), whereas the leading unsupervised principal component was only weakly correlated with D1 ($|r| = `r sprintf("%.2f", .d1p_sr_unsup)`$) and instead tracked the higher-variance D2/D3 structure (Supplementary Information, The D1 axis is not specific to DeSurv). This indicates a shared survival-associated direction in the patient scores, not that the three methods recover the same gene programs.

We then asked what DeSurv adds beyond recovery of this shared prognostic direction. Independently tuned supervised principal components and lasso-penalized Cox regression had pooled external C-indices of $`r .svbt("Supervised PCA")`$ and $`r .svbt("Sparse Cox")`$, versus $`r .svbt("DeSurv full LP")`$ for DeSurv. For this comparison, DeSurv used per-program gene supports and unpenalized Cox coefficients re-estimated on the training scores, rather than the union-support predictor used for primary external validation. The selected genes of the supervised predictors, however, were enriched for basal-like tumor programs in every cross-validation seed and for none of the stromal, immune or exocrine programs examined, and their risk scores correlated only weakly with the separately resolved proCAF-like D2 program (lasso Cox, $|r| = `r sprintf("%.2f", .ax["Sparse Cox", "D2"])`$; supervised PCA, $|r| = `r sprintf("%.2f", .ax["Supervised PCA", "D2"])`$; Fig.\ \ref{fig:val}b). DeSurv places the shared prognostic direction within separately scored tumor and stromal programs and provides program-specific gene lists for biological investigation and future classifier development.

## Survival supervision improves program recovery when prognostic signal contributes little expression variation {.unnumbered}

```{r sim-precision, echo=FALSE, message=FALSE, warning=FALSE}
load_result("sim_figs_by_scenario")
.sim_ids <- sapply(sim_figs_by_scenario, `[[`, "scenario_id"); .sim_an <- sapply(sim_figs_by_scenario, `[[`, "analysis_id")
.sim_prim <- sim_figs_by_scenario[[which(.sim_ids == "R0_easy" & .sim_an == "bo_tune_ntop")[1]]]$k_hist$data
.sim_prec <- tapply(.sim_prim$precision, .sim_prim$method, median, na.rm = TRUE)
stopifnot(all(tapply(.sim_prim$precision, .sim_prim$method, function(x) sum(!is.na(x))) == 100))
# Global marker recall (no factor matching) and the null-scenario supervision strength.
# Both reproduce the Supplementary Information accessors (chunk si-null-alpha) exactly,
# so the two sections cannot drift.
.sim_rec <- tapply(.sim_prim$marker_recall_all, .sim_prim$method, median, na.rm = TRUE)
.null_i  <- which(.sim_ids == "R00_null" & .sim_an == "bo_tune_ntop")[1]
.null_alpha <- with(sim_figs_by_scenario[[.null_i]]$k_hist$data, alpha[method == "DeSurv"])
stopifnot(length(.null_alpha) == 100)
```

We used simulations with known latent structure to determine when survival supervision improves gene program recovery. Expression profiles were generated for $p = 3{,}000$ genes and $n = 200$ samples from $k = 3$ true programs, each with its own marker genes and all drawn from the same loading and activity distributions (Methods, Simulation studies). In the primary scenario, survival depended on the marker genes of one program only, and those genes contributed less transcriptomic variation than a shared background signal excluded from the survival-generating gene set. DeSurv and its unsupervised limit ($\alpha = 0$) were tuned by the same cross-validation procedure, giving a controlled comparison with and without survival supervision.

Under this setting, DeSurv more accurately recovered the genes defining the true prognostic program, whereas unsupervised NMF was more strongly influenced by the dominant background structure (Fig.\ \ref{fig:sim}a,b). Across 100 simulated datasets, median matched-factor precision, the proportion of the best-matching learned program's selected genes that belonged to the true prognostic gene set, was `r sprintf("%.2f", .sim_prec[["DeSurv"]])` for DeSurv versus `r sprintf("%.2f", .sim_prec[["NMF"]])` for unsupervised NMF. Global marker recall, the proportion of the true survival-driving marker genes appearing in the union of all learned programs' top-ranked gene sets, was `r sprintf("%.2f", .sim_rec[["DeSurv"]])` versus `r sprintf("%.2f", .sim_rec[["NMF"]])` (Supplementary Information). Median test-set C-index was $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P `r sim_wilcox_p`$), and DeSurv recovered the true rank ($k = 3$) more frequently (Fig.\ \ref{fig:sim}c).

Performance depended on how prognostic signal aligned with expression structure. In the mixed setting, the benefit of supervision differed between recovery of program-specific markers and recovery of shared background genes included in the survival-generating set (Supplementary Fig.\ \ref{fig:fig-sim-null-mixed}). Under the null, test-set concordance was near chance for both methods. There, the selected supervision strength dispersed across its admissible range rather than concentrating at zero (median $\alpha = `r round(median(.null_alpha), 2)`$, interquartile range $`r round(quantile(.null_alpha, .25), 2)`$--$`r round(quantile(.null_alpha, .75), 2)`$ across `r length(.null_alpha)` simulated datasets), consistent with weak discrimination among supervision strengths when concordance is near chance.

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = paste0("Survival supervision improves recovery of prognostic gene programs when survival signal is not variance-dominant. (a) Test-set concordance across 100 simulated datasets comparing DeSurv with unsupervised NMF ($\\alpha = 0$; $p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$); medians ", desurv_sim_median, " versus ", nmf_sim_median, " (paired Wilcoxon $P ", sim_wilcox_p, "$). (b) Recovery of genes defining the true prognostic program, quantified as the proportion of selected genes in the best-matching learned program that belonged to the ground-truth prognostic gene set; medians ", sprintf("%.2f", .sim_prec[["DeSurv"]]), " versus ", sprintf("%.2f", .sim_prec[["NMF"]]), ". The best-matching program is selected using the known simulated gene set; this evaluates recovery with access to ground truth. Global marker recall, which uses no matching, is reported in the Supplementary Information. In (a) and (b), boxes show median and interquartile range; whiskers extend to $1.5\\times$IQR; points are individual simulated datasets. (c) Distribution of selected ranks across simulated datasets for DeSurv versus unsupervised NMF. (d) The same matched-factor precision metric as in (b), computed for two signal regimes: a prognostic program whose genes carry low transcriptomic variance relative to the shared background excluded from the survival-generating gene set (separated from the dominant variance) versus one that partially overlaps the dominant variance. DeSurv recovers the prognostic program more accurately in both regimes, whereas the unsupervised solution is more strongly influenced by the dominant expression structure; the null scenario (no prognostic program) is shown in the Supplementary Information. The concordance gap in (a) is not an expected discrimination gain in real data, where the two approaches perform comparably (Table \\ref{tab:hr-adjusted}). \\label{fig:sim}"), fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

knitr::include_graphics(file.path("..", "figures", "fig2_tcgacptac.pdf"))

```
