---
title: "Survival-supervised deconvolution identifies a reproducible tumor--stroma prognostic axis in pancreatic cancer"

abstract: |
  Pancreatic ductal adenocarcinoma (PDAC) transcriptomes contain heterogeneous tumor and microenvironmental programs with differing relevance to survival, yet standard deconvolution methods learn these programs without using survival outcomes. We developed DeSurv, a survival-supervised matrix factorization framework in which survival helps shape gene programs as they are learned. In pooled TCGA and CPTAC cohorts, DeSurv recovered a three-program tumor--stroma architecture comprising a dominant survival-aligned classical/restCAF-like program together with separately resolved proCAF-like stromal and basal-like tumor programs. The frozen DeSurv predictor generalized across five external validation datasets spanning four independent patient cohorts (HR per SD, 1.46; 95% CI, 1.27--1.67) and remained prognostic after PurIST and DeCAF adjustment. Higher-rank NMF and conventional supervised models achieved comparable discrimination, but did not recover the same biologically coherent tumor--stroma organization. Thus, survival-supervised deconvolution aligns biological program discovery with clinically relevant variation without requiring superior predictive accuracy.

keywords:
  - nonnegative matrix factorization
  - survival analysis
  - survival-supervised learning
  - tumor deconvolution
  - pancreatic cancer

params:
  tar_store: "store_PKG_VERSION=20260107bugfix_GIT_BRANCH=main"
  bo_label: ~
  run_label: ~
  val_label: ~

bibliography: references_30102025.bib
csl: nature.csl
link-citations: true

output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
    number_sections: false
    fig_caption: true
    toc: false

geometry: margin=1in
fontsize: 11pt

header-includes:
  - \providecommand{\pandocbounded}[1]{#1}
  - \usepackage{authblk}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{xr-hyper}
  - \externaldocument{si_appendix}
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{setspace}
  - \usepackage{orcidlink}
  - \DeclareMathOperator*{\argmin}{argmin}
  - \DeclareMathOperator*{\argmax}{argmax}
  - \DeclareMathOperator\diag{diag}
  - \renewcommand{\algorithmicrequire}{\textbf{Input:}}
  - \renewcommand{\algorithmicensure}{\textbf{Output:}}
  # authblk affiliation formatting
  - \renewcommand\Authsep{, }
  - \renewcommand\Authand{, }
  - \renewcommand\Authands{, }
  - \setlength{\affilsep}{0.5em}
  # author + affiliation block (rendered by \maketitle)
  - \author[a]{Amber M.\ Young\,\orcidlink{0000-0001-7042-6961}}
  - \author[b]{Alisa Yurovsky\,\orcidlink{0000-0002-7500-7907}}
  - \author[c,d]{Xianlu Laura Peng\,\orcidlink{0000-0002-8684-5762}}
  - \author[c,d,e]{Jen Jen Yeh\,\orcidlink{0000-0003-2542-6211}}
  - \author[a]{Didong Li\,\orcidlink{0000-0001-9146-705X}}
  - \author[a,d,*]{Naim U.\ Rashid\,\orcidlink{0000-0002-7274-8259}}
  - \affil[a]{Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[b]{Department of Biomedical Informatics, Stony Brook University, Stony Brook, NY 11794, USA}
  - \affil[c]{Department of Pharmacology, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[d]{Lineberger Comprehensive Cancer Center, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[e]{Department of Surgery, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - '\affil[*]{Corresponding author: nur2@email.unc.edu}'
  - '\renewcommand{\author}[2][]{}'
---
```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  error     = FALSE,
  fig.align = "center",
  fig.path  = "figures/",
  dev       = "cairo_pdf",
  dpi       = 300
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::theme_set(ggplot2::theme_gray(base_size = 10))
}

set_fig_font <- function(plot_obj, size = 10) {
  if (inherits(plot_obj, "ggplot")) {
    plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
  } else {
    plot_obj
  }
}

```

```{r load-precomputed, include=FALSE}
source("paper/load_precomputed.R")
```

<!-- NOTE: Nature Cancer editors write the Editorial Summary post-acceptance.
     Authors should NOT draft one for submission. -->

```{r child="paper/02_introduction_REVISED.Rmd"}

```

# Results {.unnumbered}

```{r child="paper/04_results_REVISED.Rmd"}

```

# Discussion {.unnumbered}

```{r child="paper/05_discussion_REVISED.Rmd"}

```

# Methods {.unnumbered}

```{r child="paper/03_methods_REVISED.Rmd"}

```

# Data Availability {.unnumbered}

The training and validation datasets used in this study are publicly available or accessible through controlled-access repositories, and are de-identified; no Institutional Review Board (IRB) approval was required for their analysis. Training data were obtained from the Genomic Data Commons: TCGA-PAAD [@data_tcga_paad] and CPTAC-3 [@data_cptac3]. Validation cohorts were obtained from ArrayExpress (Dijk, E-MTAB-6830 [@data_dijk_emtab6830]; Puleo, E-MTAB-6134 [@data_puleo_emtab6134]), the Gene Expression Omnibus (GEO; Moffitt, GSE71729 [@data_moffitt_gse71729]), and the International Cancer Genome Consortium (ICGC): the PACA-AU array cohort is publicly available from the Gene Expression Omnibus (GSE36924), and the PACA-AU RNA-seq cohort from the European Genome--phenome Archive (EGA) study EGAS00001000154 under controlled access [@data_paca_au_egas]. Expression data and molecular classifications (PurIST, DeCAF) for the validation cohorts were originally compiled for the DeCAF study [@peng2024determination]; cohort details and sample sizes after quality control are summarized in Supplementary Information, Section 6. The treated bulk-tissue cohorts used in the exploratory translational analysis (Rash-Accept and the Linehan FOLFIRINOX~$\pm$~CCR2-inhibitor series), and the COMPASS cohort used for the GATA6 RNA in situ hybridization comparison, are restricted-access clinical-trial datasets that are not publicly deposited; raw data are available from the respective study investigators under a data-use agreement, subject to the governing trial consents and institutional approvals. To preserve reproducibility without redistributing restricted data, the derived per-cohort summary statistics reported in the main text and Supplementary Information are provided in the reproducibility repository (file \texttt{results/treated\_cohort\_stats.rds}), so that every treated-cohort value can be inspected and verified. Code used to generate the COMPASS D1--GATA6 comparison (main-text Fig.\ \ref{fig:pdac}E) is also provided in the reproducibility repository; reproduction of the patient-level analysis requires authorized access to the COMPASS data.

# Code Availability {.unnumbered}

All analyses were performed in R (version 4.6.0; R Core Team) using DeSurv (v1.0.1; this work), \texttt{NMF} [@gaujoux2010flexible], \texttt{glmnet} [@simon2011regularization], \texttt{survival}, \texttt{ggplot2}, and \texttt{cowplot}; Bayesian optimization for hyperparameter selection was implemented within DeSurv. The DeSurv R package is available at \url{https://github.com/rashidlab/DeSurv} (archived at Zenodo, DOI \texttt{10.5281/zenodo.20150929}). Reproducibility code, processed data, and the exact session-info file are available at \url{https://github.com/rashidlab/DeSurv-paper} (archived at Zenodo, DOI \texttt{10.5281/zenodo.20150946}); this includes \texttt{code/11\_treated\_cohort\_analysis.R}, which generates the treated-cohort summary statistics above.

# Acknowledgements {.unnumbered}

This work was supported by the National Cancer Institute grants U01 CA274298, P50 CA257911, T32 CA106209, and R01 CA288145, and by the Department of Defense Peer Reviewed Cancer Research Program (HT9425241103100121). We thank the patients and research teams who contributed samples and clinical data to the public cohorts used here (TCGA, CPTAC, Dijk, Moffitt, Puleo, PACA-AU).

\textbf{Inclusion and Diversity.} The patient cohorts analyzed in this study were curated by the original investigators or consortia (TCGA, CPTAC, ICGC, and the original Moffitt, Dijk, and Puleo studies) and our access was limited to the de-identified expression and survival data they provide; we did not stratify analyses by sex, race, or ethnicity because consistent demographic annotations were not available across all cohorts. Sex and other demographic variables that could modify the prognostic relevance of the gene programs identified here remain important directions for follow-up work as more comprehensively annotated cohorts become available.

# Author Contributions {.unnumbered}

Conceptualization: A.M.Y., N.U.R. Methodology: A.M.Y., A.Y., D.L., N.U.R. Software: A.M.Y. Formal analysis: A.M.Y. Investigation: A.M.Y. Data curation: A.M.Y., X.L.P. Visualization: A.M.Y. Writing -- original draft: A.M.Y., N.U.R. Writing -- review and editing: A.M.Y., A.Y., X.L.P., J.J.Y., D.L., N.U.R. Resources (PDAC domain expertise and biological interpretation): X.L.P., J.J.Y. Supervision: N.U.R. Funding acquisition: A.M.Y., N.U.R. All authors reviewed and approved the final manuscript.

# Competing Interests {.unnumbered}

N.U.R. and J.J.Y. are named as inventors on US patents US20220127676A1 and US20250066856A1, which relate to pancreatic cancer subtype classification. The PurIST and DeCAF classifiers were used only as published, independently derived comparators: their subtype calls were obtained with the released classifiers and included as fixed covariate annotations, and the adjustment analyses using these calls are objective and reproducible from the deposited code. They are not components of DeSurv. The other authors declare no competing interests.

# References

```{r versioning, echo=FALSE, eval=FALSE}
version_info <- tryCatch(
  targets::tar_read("version_info"),
  error = function(e) NULL
)

if (is.null(version_info)) {
  desurv_version <- tryCatch(
    as.character(utils::packageVersion("DeSurv")),
    error = function(e) "unknown"
  )
  desurv_repo <- normalizePath("../DeSurv", mustWork = FALSE)
  desurv_branch <- tryCatch(
    if (dir.exists(desurv_repo)) {
      trimws(system(
        sprintf("git -C %s rev-parse --abbrev-ref HEAD", shQuote(desurv_repo)),
        intern = TRUE,
        ignore.stderr = TRUE
      ))
    } else {
      "unknown"
    },
    error = function(e) "unknown"
  )
  desurv_commit <- tryCatch(
    if (dir.exists(desurv_repo)) {
      trimws(system(
        sprintf("git -C %s rev-parse HEAD", shQuote(desurv_repo)),
        intern = TRUE,
        ignore.stderr = TRUE
      ))
    } else {
      "unknown"
    },
    error = function(e) "unknown"
  )
  paper_branch <- tryCatch(
    trimws(system("git rev-parse --abbrev-ref HEAD", intern = TRUE, ignore.stderr = TRUE)),
    error = function(e) "unknown"
  )
  paper_commit <- tryCatch(
    trimws(system("git rev-parse HEAD", intern = TRUE, ignore.stderr = TRUE)),
    error = function(e) "unknown"
  )
} else {
  desurv_version <- version_info$desurv_pkg_version
  desurv_branch <- version_info$desurv_git_branch
  desurv_commit <- version_info$desurv_git_commit
  paper_branch <- version_info$paper_git_branch
  paper_commit <- version_info$paper_git_commit
}

if (!length(desurv_version) || !nzchar(desurv_version)) desurv_version <- "unknown"
if (!length(desurv_branch) || !nzchar(desurv_branch)) desurv_branch <- "unknown"
if (!length(desurv_commit) || !nzchar(desurv_commit)) desurv_commit <- "unknown"
if (!length(paper_branch) || !nzchar(paper_branch)) paper_branch <- "unknown"
if (!length(paper_commit) || !nzchar(paper_commit)) paper_commit <- "unknown"

cat(
  "DeSurv package version: ", desurv_version, "\n",
  "DeSurv git branch: ", desurv_branch, "\n",
  "DeSurv git commit: ", desurv_commit, "\n",
  "Paper git branch: ", paper_branch, "\n",
  "Paper git commit: ", paper_commit, "\n",
  sep = ""
)
```
