---
title: "Survival-guided matrix factorization identifies generalizable prognostic programs in pancreatic cancer"

abstract: |
  The transcriptional programs that explain the most variance in bulk tumor expression data are often not those that predict patient survival. Standard nonnegative matrix factorization (NMF) discovers latent structure by minimizing reconstruction error, which may be driven by high-variance but prognostically neutral signals such as tissue composition. Here we present DeSurv, a survival-supervised deconvolution framework that jointly integrates NMF with Cox proportional hazards modeling, shaping gene programs to capture outcome-relevant structure during discovery rather than only in post hoc evaluation. In simulations with known latent structure, DeSurv places more true prognostic genes among the top-weighted genes of each factor than standard NMF (precision, defined as the fraction of top-weighted factor genes that are true prognostic markers, 0.50 vs. 0.07), with the largest advantage when variance and prognosis diverge. In pancreatic ductal adenocarcinoma (PDAC), DeSurv recovers a compact, three-factor decomposition in which a single tumor--stroma coupling drives prognosis while exocrine-compositional signals are suppressed. These survival-aligned factors generalize to five independent PDAC cohorts without retraining, with stronger out-of-sample survival stratification than their unsupervised counterparts (pooled validation hazard ratio (HR) per standard deviation (SD) = 1.50, 95% confidence interval (CI) 1.31--1.72, P < 0.001), demonstrating that outcome supervision during factorization yields more generalizable transcriptional signatures than the standard discover-then-evaluate paradigm. By prioritizing transcriptional programs that survive cross-cohort generalization, DeSurv supports the development of generalizable prognostic biomarkers for the bulk-transcriptomic cohorts that dominate clinical cancer research.

keywords:
  - nonnegative matrix factorization
  - survival analysis
  - semi-supervised learning
  - tumor deconvolution
  - pancreatic cancer

params:
  tar_store: "store_PKG_VERSION=20260107bugfix_GIT_BRANCH=main"
  bo_label: ~
  run_label: ~
  val_label: ~

bibliography: references_30102025.bib
csl: nature.csl
link-citations: true

output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
    number_sections: false
    fig_caption: true
    toc: false

geometry: margin=1in
fontsize: 11pt

header-includes:
  - \providecommand{\pandocbounded}[1]{#1}
  - \usepackage{authblk}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{setspace}
  - \usepackage{orcidlink}
  - \DeclareMathOperator*{\argmin}{argmin}
  - \DeclareMathOperator*{\argmax}{argmax}
  - \DeclareMathOperator\diag{diag}
  - \renewcommand{\algorithmicrequire}{\textbf{Input:}}
  - \renewcommand{\algorithmicensure}{\textbf{Output:}}
  # authblk affiliation formatting
  - \renewcommand\Authsep{, }
  - \renewcommand\Authand{, }
  - \renewcommand\Authands{, }
  - \setlength{\affilsep}{0.5em}
  # author + affiliation block (rendered by \maketitle)
  - \author[a]{Amber M.\ Young\,\orcidlink{0000-0001-7042-6961}}
  - \author[b]{Alisa Yurovsky\,\orcidlink{0000-0002-7500-7907}}
  - \author[c,d]{Xianlu Laura Peng\,\orcidlink{0000-0002-8684-5762}}
  - \author[c,d,e]{Jen Jen Yeh\,\orcidlink{0000-0003-2542-6211}}
  - \author[a]{Didong Li\,\orcidlink{0000-0001-9146-705X}}
  - \author[a,d,*]{Naim U.\ Rashid\,\orcidlink{0000-0002-7274-8259}}
  - \affil[a]{Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[b]{Department of Biomedical Informatics, Stony Brook University, Stony Brook, NY 11794, USA}
  - \affil[c]{Department of Pharmacology, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[d]{Lineberger Comprehensive Cancer Center, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[e]{Department of Surgery, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - '\affil[*]{Corresponding author: nur2@email.unc.edu}'
---
```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  error     = FALSE,
  fig.align = "center",
  fig.path  = "figures/",
  dev       = "cairo_pdf",
  dpi       = 300
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::theme_set(ggplot2::theme_gray(base_size = 10))
}

set_fig_font <- function(plot_obj, size = 10) {
  if (inherits(plot_obj, "ggplot")) {
    plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
  } else {
    plot_obj
  }
}

```

```{r load-precomputed, include=FALSE}
source("paper/load_precomputed.R")
```

<!-- ============================================================
     EDITORIAL SUMMARY (Nature Cancer convention, ~3 sentences,
     plain-language for the broader cancer research community).
     ============================================================ -->

\noindent\textbf{Editorial Summary.} Unsupervised matrix factorisation of bulk tumour transcriptomes typically recovers the gene programs that explain the most expression variance — often tissue-composition signals — rather than the programs that predict patient outcomes. Young et al. introduce DeSurv, a deconvolution framework that couples nonnegative matrix factorisation to a Cox survival model so that prognostically informative programs are learned during discovery rather than retrofitted afterwards. Applied to pancreatic ductal adenocarcinoma, DeSurv recovers a compact three-program decomposition centred on a tumour--stroma axis that generalises across five independent cohorts and retains prognostic value after adjustment for established molecular classifiers.

\bigskip

```{r child="paper/02_introduction_REVISED.Rmd"}

```

# Results {.unnumbered}

```{r child="paper/04_results_REVISED.Rmd"}

```

# Discussion {.unnumbered}

```{r child="paper/05_discussion_REVISED.Rmd"}

```

# Methods {.unnumbered}

```{r child="paper/03_methods_REVISED.Rmd"}

```

# Data and Code Availability {.unnumbered}

The DeSurv R package is available at \url{https://github.com/rashidlab/DeSurv} (archived at Zenodo, DOI \texttt{10.5281/zenodo.20150929}). Reproducibility code, processed data, and session-info for this manuscript are available at \url{https://github.com/rashidlab/DeSurv-paper} (archived at Zenodo, DOI \texttt{10.5281/zenodo.20150946}). All training and validation cohorts are publicly available: TCGA-PAAD and CPTAC-3 from the Genomic Data Commons; Dijk (E-MTAB-6830) and Puleo (E-MTAB-6134) from ArrayExpress; Moffitt (GSE71729) from the Gene Expression Omnibus; and PACA-AU (EGA study EGAS00001000154) from the ICGC Data Portal.

# Acknowledgements {.unnumbered}

This work was supported by the National Cancer Institute grants U01 CA274298, P50 CA257911, T32 CA106209, and R01 CA288145, and by the Department of Defense Peer Reviewed Cancer Research Program (HT9425241103100121). We thank the patients and research teams who contributed samples and clinical data to the public cohorts used here (TCGA, CPTAC, Dijk, Moffitt, Puleo, PACA-AU).

\textbf{Inclusion and Diversity.} The patient cohorts analyzed in this study were curated by the original investigators or consortia (TCGA, CPTAC, ICGC, and the original Moffitt, Dijk, and Puleo studies) and our access was limited to the de-identified expression and survival data they provide; we did not stratify analyses by sex, race, or ethnicity because consistent demographic annotations were not available across all cohorts. Sex and other demographic variables that could modify the prognostic relevance of the gene programs identified here remain important directions for follow-up work as more comprehensively annotated cohorts become available.

# Author Contributions {.unnumbered}

Conceptualization: A.M.Y., N.U.R. Methodology: A.M.Y., A.Y., D.L., N.U.R. Software: A.M.Y. Formal analysis: A.M.Y. Investigation: A.M.Y. Data curation: A.M.Y., X.L.P. Visualization: A.M.Y. Writing -- original draft: A.M.Y., N.U.R. Writing -- review and editing: A.M.Y., A.Y., X.L.P., J.J.Y., D.L., N.U.R. Resources (PDAC domain expertise and biological interpretation): X.L.P., J.J.Y. Supervision: N.U.R. Funding acquisition: A.M.Y., N.U.R. All authors reviewed and approved the final manuscript.

# Competing Interests {.unnumbered}

N.U.R. and J.J.Y. are named as inventors on US patents US20220127676A1 and US20250066856A1, which relate to pancreatic cancer subtype classification (establishment of Basal and Classical subtypes and their prediction) derived from prior work and were not used directly in this study. The other authors declare no competing interests.

# References

```{r versioning, echo=FALSE, eval=FALSE}
version_info <- tryCatch(
  targets::tar_read("version_info"),
  error = function(e) NULL
)

if (is.null(version_info)) {
  desurv_version <- tryCatch(
    as.character(utils::packageVersion("DeSurv")),
    error = function(e) "unknown"
  )
  desurv_repo <- normalizePath("../DeSurv", mustWork = FALSE)
  desurv_branch <- tryCatch(
    if (dir.exists(desurv_repo)) {
      trimws(system(
        sprintf("git -C %s rev-parse --abbrev-ref HEAD", shQuote(desurv_repo)),
        intern = TRUE,
        ignore.stderr = TRUE
      ))
    } else {
      "unknown"
    },
    error = function(e) "unknown"
  )
  desurv_commit <- tryCatch(
    if (dir.exists(desurv_repo)) {
      trimws(system(
        sprintf("git -C %s rev-parse HEAD", shQuote(desurv_repo)),
        intern = TRUE,
        ignore.stderr = TRUE
      ))
    } else {
      "unknown"
    },
    error = function(e) "unknown"
  )
  paper_branch <- tryCatch(
    trimws(system("git rev-parse --abbrev-ref HEAD", intern = TRUE, ignore.stderr = TRUE)),
    error = function(e) "unknown"
  )
  paper_commit <- tryCatch(
    trimws(system("git rev-parse HEAD", intern = TRUE, ignore.stderr = TRUE)),
    error = function(e) "unknown"
  )
} else {
  desurv_version <- version_info$desurv_pkg_version
  desurv_branch <- version_info$desurv_git_branch
  desurv_commit <- version_info$desurv_git_commit
  paper_branch <- version_info$paper_git_branch
  paper_commit <- version_info$paper_git_commit
}

if (!length(desurv_version) || !nzchar(desurv_version)) desurv_version <- "unknown"
if (!length(desurv_branch) || !nzchar(desurv_branch)) desurv_branch <- "unknown"
if (!length(desurv_commit) || !nzchar(desurv_commit)) desurv_commit <- "unknown"
if (!length(paper_branch) || !nzchar(paper_branch)) paper_branch <- "unknown"
if (!length(paper_commit) || !nzchar(paper_commit)) paper_commit <- "unknown"

cat(
  "DeSurv package version: ", desurv_version, "\n",
  "DeSurv git branch: ", desurv_branch, "\n",
  "DeSurv git commit: ", desurv_commit, "\n",
  "Paper git branch: ", paper_branch, "\n",
  "Paper git commit: ", paper_commit, "\n",
  sep = ""
)
```
