---
title: "Tumor--stroma coupling defines a generalizable prognostic program in pancreatic cancer"

abstract: |
  Pancreatic ductal adenocarcinoma (PDAC) outcomes reflect coordinated tumor-intrinsic and stromal states, but bulk-transcriptomic deconvolution methods split this biology across latent factors that fail to generalize across cohorts. We developed DeSurv, a survival-supervised matrix factorization framework that jointly learns gene programs and their survival association. In pooled TCGA and CPTAC PDAC cohorts, DeSurv recovered a compact three-program structure dominated by a tumor-stroma axis linking Classical tumor identity and tumor-restraining CAF (restCAF) stroma with a Basal-like tumor and tumor-promoting CAF (proCAF) high-risk pole. This score generalized across five independent PDAC validation cohorts (Dijk, Moffitt, PACA-AU array and RNA-seq, Puleo), with each standard-deviation shift toward the Basal-like/proCAF pole conferring a pooled hazard ratio of 1.50 (95% CI 1.31 to 1.72, P < 0.001), and retained independent prognostic value after adjustment for PurIST and DeCAF. These findings nominate tumor-stroma coupling as a reproducible prognostic state in PDAC.

keywords:
  - nonnegative matrix factorization
  - survival analysis
  - semi-supervised learning
  - tumor deconvolution
  - pancreatic cancer

params:
  tar_store: "store_PKG_VERSION=20260107bugfix_GIT_BRANCH=main"
  bo_label: ~
  run_label: ~
  val_label: ~

bibliography: references_30102025.bib
csl: nature.csl
link-citations: true

output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
    number_sections: false
    fig_caption: true
    toc: false

geometry: margin=1in
fontsize: 11pt

header-includes:
  - \providecommand{\pandocbounded}[1]{#1}
  - \usepackage{authblk}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{setspace}
  - \usepackage{orcidlink}
  - \DeclareMathOperator*{\argmin}{argmin}
  - \DeclareMathOperator*{\argmax}{argmax}
  - \DeclareMathOperator\diag{diag}
  - \renewcommand{\algorithmicrequire}{\textbf{Input:}}
  - \renewcommand{\algorithmicensure}{\textbf{Output:}}
  # authblk affiliation formatting
  - \renewcommand\Authsep{, }
  - \renewcommand\Authand{, }
  - \renewcommand\Authands{, }
  - \setlength{\affilsep}{0.5em}
  # author + affiliation block (rendered by \maketitle)
  - \author[a]{Amber M.\ Young\,\orcidlink{0000-0001-7042-6961}}
  - \author[b]{Alisa Yurovsky\,\orcidlink{0000-0002-7500-7907}}
  - \author[c,d]{Xianlu Laura Peng\,\orcidlink{0000-0002-8684-5762}}
  - \author[c,d,e]{Jen Jen Yeh\,\orcidlink{0000-0003-2542-6211}}
  - \author[a]{Didong Li\,\orcidlink{0000-0001-9146-705X}}
  - \author[a,d,*]{Naim U.\ Rashid\,\orcidlink{0000-0002-7274-8259}}
  - \affil[a]{Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[b]{Department of Biomedical Informatics, Stony Brook University, Stony Brook, NY 11794, USA}
  - \affil[c]{Department of Pharmacology, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[d]{Lineberger Comprehensive Cancer Center, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - \affil[e]{Department of Surgery, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
  - '\affil[*]{Corresponding author: nur2@email.unc.edu}'
---
```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  error     = FALSE,
  fig.align = "center",
  fig.path  = "figures/",
  dev       = "cairo_pdf",
  dpi       = 300
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::theme_set(ggplot2::theme_gray(base_size = 10))
}

set_fig_font <- function(plot_obj, size = 10) {
  if (inherits(plot_obj, "ggplot")) {
    plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
  } else {
    plot_obj
  }
}

```

```{r load-precomputed, include=FALSE}
source("paper/load_precomputed.R")
```

<!-- NOTE: Nature Cancer editors write the Editorial Summary post-acceptance.
     Authors should NOT draft one for submission. -->

```{r child="paper/02_introduction_REVISED.Rmd"}

```

# Results {.unnumbered}

```{r child="paper/04_results_REVISED.Rmd"}

```

# Discussion {.unnumbered}

```{r child="paper/05_discussion_REVISED.Rmd"}

```

# Methods {.unnumbered}

```{r child="paper/03_methods_REVISED.Rmd"}

```

# Data and Code Availability {.unnumbered}

The DeSurv R package is available at \url{https://github.com/rashidlab/DeSurv} (archived at Zenodo, DOI \texttt{10.5281/zenodo.20150929}). Reproducibility code, processed data, and session-info for this manuscript are available at \url{https://github.com/rashidlab/DeSurv-paper} (archived at Zenodo, DOI \texttt{10.5281/zenodo.20150946}). All training and validation cohorts are publicly available: TCGA-PAAD and CPTAC-3 from the Genomic Data Commons; Dijk (E-MTAB-6830) and Puleo (E-MTAB-6134) from ArrayExpress; Moffitt (GSE71729) from the Gene Expression Omnibus; and PACA-AU (EGA study EGAS00001000154) from the ICGC Data Portal.

# Acknowledgements {.unnumbered}

This work was supported by the National Cancer Institute grants U01 CA274298, P50 CA257911, T32 CA106209, and R01 CA288145, and by the Department of Defense Peer Reviewed Cancer Research Program (HT9425241103100121). We thank the patients and research teams who contributed samples and clinical data to the public cohorts used here (TCGA, CPTAC, Dijk, Moffitt, Puleo, PACA-AU).

\textbf{Inclusion and Diversity.} The patient cohorts analyzed in this study were curated by the original investigators or consortia (TCGA, CPTAC, ICGC, and the original Moffitt, Dijk, and Puleo studies) and our access was limited to the de-identified expression and survival data they provide; we did not stratify analyses by sex, race, or ethnicity because consistent demographic annotations were not available across all cohorts. Sex and other demographic variables that could modify the prognostic relevance of the gene programs identified here remain important directions for follow-up work as more comprehensively annotated cohorts become available.

# Author Contributions {.unnumbered}

Conceptualization: A.M.Y., N.U.R. Methodology: A.M.Y., A.Y., D.L., N.U.R. Software: A.M.Y. Formal analysis: A.M.Y. Investigation: A.M.Y. Data curation: A.M.Y., X.L.P. Visualization: A.M.Y. Writing -- original draft: A.M.Y., N.U.R. Writing -- review and editing: A.M.Y., A.Y., X.L.P., J.J.Y., D.L., N.U.R. Resources (PDAC domain expertise and biological interpretation): X.L.P., J.J.Y. Supervision: N.U.R. Funding acquisition: A.M.Y., N.U.R. All authors reviewed and approved the final manuscript.

# Competing Interests {.unnumbered}

N.U.R. and J.J.Y. are named as inventors on US patents US20220127676A1 and US20250066856A1, which relate to pancreatic cancer subtype classification (establishment of Basal and Classical subtypes and their prediction) derived from prior work and were not used directly in this study. The other authors declare no competing interests.

# References

```{r versioning, echo=FALSE, eval=FALSE}
version_info <- tryCatch(
  targets::tar_read("version_info"),
  error = function(e) NULL
)

if (is.null(version_info)) {
  desurv_version <- tryCatch(
    as.character(utils::packageVersion("DeSurv")),
    error = function(e) "unknown"
  )
  desurv_repo <- normalizePath("../DeSurv", mustWork = FALSE)
  desurv_branch <- tryCatch(
    if (dir.exists(desurv_repo)) {
      trimws(system(
        sprintf("git -C %s rev-parse --abbrev-ref HEAD", shQuote(desurv_repo)),
        intern = TRUE,
        ignore.stderr = TRUE
      ))
    } else {
      "unknown"
    },
    error = function(e) "unknown"
  )
  desurv_commit <- tryCatch(
    if (dir.exists(desurv_repo)) {
      trimws(system(
        sprintf("git -C %s rev-parse HEAD", shQuote(desurv_repo)),
        intern = TRUE,
        ignore.stderr = TRUE
      ))
    } else {
      "unknown"
    },
    error = function(e) "unknown"
  )
  paper_branch <- tryCatch(
    trimws(system("git rev-parse --abbrev-ref HEAD", intern = TRUE, ignore.stderr = TRUE)),
    error = function(e) "unknown"
  )
  paper_commit <- tryCatch(
    trimws(system("git rev-parse HEAD", intern = TRUE, ignore.stderr = TRUE)),
    error = function(e) "unknown"
  )
} else {
  desurv_version <- version_info$desurv_pkg_version
  desurv_branch <- version_info$desurv_git_branch
  desurv_commit <- version_info$desurv_git_commit
  paper_branch <- version_info$paper_git_branch
  paper_commit <- version_info$paper_git_commit
}

if (!length(desurv_version) || !nzchar(desurv_version)) desurv_version <- "unknown"
if (!length(desurv_branch) || !nzchar(desurv_branch)) desurv_branch <- "unknown"
if (!length(desurv_commit) || !nzchar(desurv_commit)) desurv_commit <- "unknown"
if (!length(paper_branch) || !nzchar(paper_branch)) paper_branch <- "unknown"
if (!length(paper_commit) || !nzchar(paper_commit)) paper_commit <- "unknown"

cat(
  "DeSurv package version: ", desurv_version, "\n",
  "DeSurv git branch: ", desurv_branch, "\n",
  "DeSurv git commit: ", desurv_commit, "\n",
  "Paper git branch: ", paper_branch, "\n",
  "Paper git commit: ", paper_commit, "\n",
  sep = ""
)
```
