---
title: "Supplementary Information"
subtitle: "DeSurv identifies a replicable tumor--stroma prognostic architecture in pancreatic cancer through survival-supervised matrix factorization"
author: Amber M. Young, Alisa Yurovsky, Xianlu Laura Peng, Jen Jen Yeh, Didong Li, Naim U. Rashid
header-includes:
- \providecommand{\pandocbounded}[1]{#1}
- \usepackage{booktabs}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{xr-hyper}
- \externaldocument{paper}
- \usepackage{amsmath}
- \usepackage{amsfonts}
- \usepackage{amssymb}
- \usepackage{bbm}
- \usepackage{algorithm}
- \usepackage{algpseudocode}
- \usepackage{setspace}
- \usepackage{enumitem}
- \usepackage[nolists,nomarkers]{endfloat}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator\diag{diag}
- \renewcommand{\algorithmicrequire}{\textbf{Input:}}
- \renewcommand{\algorithmicensure}{\textbf{Output:}}
- \renewcommand{\thefigure}{S\arabic{figure}}
- \renewcommand{\thetable}{S\arabic{table}}
- \renewcommand{\thealgorithm}{S\arabic{algorithm}}
- \renewcommand{\theequation}{S\arabic{equation}}
bibliography: references_30102025.bib
csl: nature.csl
link-citations: true
reference-section-title: "SI References"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    keep_tex: true
  html_document: default
---
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\newpage

\part{Supplementary Methods}

\section{Model details}
Let $X \in \mathbb{R}_{\ge 0}^{p \times n}$ denote the nonnegative gene
expression matrix with $p$ genes and $n$ subjects. DeSurv approximates
$X \approx WH$ with $W \in \mathbb{R}_{\ge 0}^{p \times k}$ and
$H \in \mathbb{R}_{\ge 0}^{k \times n}$, and links the shared program matrix
$W$ to survival via an elastic-net–penalized Cox model. Supervision is placed on the program matrix $W$, the basis that transfers to new cohorts, whereas the sample loadings $H$ serve only to reconstruct the training expression matrix; new samples are therefore scored by projection onto the learned programs, $Z = W^\top X$, rather than by re-estimating loadings.

Let $y \in \mathbb{R}^n_{> 0}$ be the observed survival times,
$\delta \in \{0,1\}^n$ the event indicators, and
$Z = W^\top X \in \mathbb{R}^{k \times n}$ the sample-wise factor scores. We write $Z_i$ for the $i$th column of $Z$, and
$n_{\text{event}} = \sum_{i=1}^n \delta_i$. For convenience we recall the
DeSurv loss from Equation 1 in the main text and rewrite it as
\begin{align}
  \mathcal{L}(W,H,\beta) &=\nonumber \\
  &(1-\alpha)\left(\frac{1}{2\lVert X \rVert_F^2}\lVert X - WH \rVert_F^2 + \frac{\lambda_H}{2nk} \lVert H \rVert_F^2\right) \nonumber \\
  -&\alpha \left(\frac{2}{n_{event}}\ell(W,\beta) - \lambda \{ \xi \lVert \beta \rVert_1 + \tfrac{1-\xi}{2} \lVert \beta \rVert_2^2 \} \right) \nonumber
\end{align}

where $\lVert \cdot \rVert_F$ is the Frobenius norm and
$\beta \in \mathbb{R}^k$ are the Cox coefficients. The Cox log-partial
likelihood $\ell(W,\beta)$ is
\begin{equation}
\ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left[
  Z_i^\top \beta
  -
  \log \left(
    \sum_{j : y_j \ge y_i}
    \exp(Z_j^\top \beta)
  \right)
\right].
\end{equation}
The inner sum runs over the risk set $\{j : y_j \ge y_i\}$ of subjects still at risk at the observed time $y_i$; tied event times are handled by the Breslow approximation in the \texttt{glmnet} Coxnet routine used for the $\beta$ update (below).
Hyperparameters satisfy $\alpha \in [0,1)$,
$\lambda_H \ge 0$, $\lambda \ge 0$, and $\xi \in [0,1]$.
The constants $1/(2\lVert X \rVert_F^2)$, $2/n_{\text{event}}$, and $1/(2nk)$ place the reconstruction, loadings-penalty, and survival terms on comparable numerical scales for stable optimization. The elastic-net mixing parameter $\xi$ is the argument \texttt{nu} in the DeSurv package.


\section{Optimization Algorithm}

\subsection{Block coordinate descent scheme}
Optimization proceeds by block coordinate descent (BCD; Algorithm \ref{alg:desurv}), which at each outer iteration updates $H$, then $W$, then $\beta$ while holding the other blocks fixed.

\begin{algorithm}[htbp]
\caption{DeSurv block coordinate descent}
\label{alg:desurv}
    \begin{algorithmic}[1]
        \Require $X \in \mathbb{R}_{\geq 0}^{p \times n}$, survival times $y \in \mathbb{R}^n_{> 0}$, event indicators $\delta \in \{0,1\}^{n}$, tolerance $tol$, maximum iterations $maxit$, supervision parameter $\alpha$, rank $k$, penalties $\lambda_H$, $\lambda$, and $\xi$
        \Ensure Fitted DeSurv parameters $(W,H,\beta)$

        \State Initialize $W^{(0)}, H^{(0)}$ with positive entries (e.g., $W^{(0)},H^{(0)} \sim \mathrm{Uniform}(0,\max X)$)
        \State Initialize $\beta^{(0)} = \mathbf{0}$ (the value used in all reported fits)
        \State $loss = \mathcal{L}(W^{(0)},H^{(0)},\beta^{(0)})$
        \State $eps = \infty$, $t = 0$
        \While{$eps \geq tol$ \textbf{and} $t < maxit$}
            \State \textbf{(H-update)}
            \State \hspace{0.5cm} Update $H^{(t+1)}$ from $H^{(t)}$ using the multiplicative rule in Eq. \ref{eqn:Hupdate},
            \State \hspace{0.5cm} holding $W^{(t)}$ and $\beta^{(t)}$ fixed.
            \State \textbf{(W-update)}
            \State \hspace{0.5cm} Update $W^{(t+1)}$ from $W^{(t)}$ using the hybrid multiplicative rule in Eq. \ref{eqn:W_update},
            \State \hspace{0.5cm} holding $H^{(t+1)}$ and $\beta^{(t)}$ fixed, and perform backtracking to ensure
            \State \hspace{0.5cm} $\mathcal{L}$ does not increase.
            \State \textbf{($\beta$-update)}
            \State \hspace{0.5cm} Update $\beta^{(t+1)}$ by solving the elastic-net Cox subproblem using cyclic coordinate descent to numerical tolerance
            \State \hspace{0.5cm} (Eq. \ref{eqn:beta_update}), holding $W^{(t+1)}$ and $H^{(t+1)}$ fixed.
            \State $lossNew = \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})$
            \State $eps = |lossNew - loss| / |loss|$
            \State $loss = lossNew$
            \State $t = t + 1$
        \EndWhile
        \State \Return $\hat{W} = W^{(t)}, \hat{H} = H^{(t)}, \hat{\beta}=\beta^{(t)}$
    \end{algorithmic}
\end{algorithm}

\subsection{Update for $H$}
Conditional on $W$, the loss $\mathcal{L}(W,H,\beta)$ reduces to a
convex quadratic function of $H$. We adopt the standard NMF multiplicative
update with $\ell_2$ penalty:
\begin{equation}
\label{eqn:Hupdate}
H \leftarrow
\max\!\left(
  H \odot \frac{W^\top X}{
    W^\top W H + \dfrac{\lVert X \rVert_F^2}{nk}\,\lambda_H H + \varepsilon_H
  },
  \varepsilon_H
\right),
\end{equation}
where $\odot$ denotes elementwise multiplication, all divisions are
elementwise, and $\varepsilon_H > 0$ is a small floor to prevent entries from
becoming exactly zero. This update is equivalent to a majorization–minimization
step and guarantees a nonincreasing reconstruction term conditional on $W$
[@seung2001algorithms; @pascualmontano2006nonsmooth]. The $\varepsilon_H$-floor
ensures that iterates remain in the interior of the nonnegative orthant, which
keeps the multiplicative $H$-update numerically well-defined.


\subsection{Update for $W$}
\label{sec:wupdate}
For $W$, we construct a hybrid multiplicative update that balances the
contributions of the NMF and Cox gradients. Let
\[
\nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H)
=
\frac{1}{\lVert X \rVert_F^2}(W H - X)H^\top,
\]
and let
\[
\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta)
=
\frac{2}{n_{event}}\nabla_W \ell(W,\beta),
\]
where $\nabla_W \ell(W,\beta)$ denotes the Cox gradient with respect to $W$. A closed-form expression for $\nabla_W \ell$ is
\begin{equation}
\nabla_W \ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left(
  x_i
  -\frac{
      \sum_{j : y_j \ge y_i}
        x_j \exp(x_j^\top W \beta)
    }{
      \sum_{j : y_j \ge y_i}
        \exp(x_j^\top W \beta)
    }
\right)\beta^\top,
\end{equation}
where $x_i$ denotes the $i$th column of $X$.

We define a gradient-scaling heuristic
\begin{equation}
\label{eq:supp-zeta}
\zeta^{(t)}
=
\frac{
  \left\lVert (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W^{(t)},H^{(t+1)}) \right\rVert_F + \varepsilon_\zeta
}{
  \left\lVert \alpha \nabla_W \mathcal{L}_{Cox}(W^{(t)},\beta^{(t)}) \right\rVert_F + \varepsilon_\zeta
},
\end{equation}
with a numerical stabilizer $\varepsilon_\zeta = 10^{-12}$ in each norm, and clip
$\zeta^{(t)}$ to avoid extreme values:
$\zeta^{(t)} \leftarrow \min(\zeta^{(t)}, \zeta_{\max})$
with $\zeta_{\max} = 10^6$ in all experiments. The scaling removes differences in
the magnitudes of the weighted reconstruction and survival gradients before the
additional factor $\alpha$ below controls the relative Cox contribution. Because
the Cox gradient entering $\zeta$ is already weighted by $\alpha$, the additional
explicit factor of $\alpha$ in the $W$ update below makes the scaled Cox
contribution approximately $\alpha$ times the weighted reconstruction-gradient
norm when the cap is inactive, up to the numerical stabilizer; the two factors of
$\alpha$ in that update are therefore not a duplication.

```{r zeta-trace, echo=FALSE, message=FALSE, warning=FALSE}
.zt <- read_result("zeta_trace")
```

In the reported consensus fit, $\zeta$ decreased from
$`r sprintf("%.2f", .zt$trace$zeta[1])`$ after the first iteration to
$`r sprintf("%.3f", .zt$final$zeta)`$ at convergence, and the scaled Cox
contribution was then `r sprintf("%.1f", 100 * .zt$final$cox_to_nmf_ratio)`\%
of the reconstruction term in the numerator. The cap was reached only at
initialization, where $\beta = 0$ makes the Cox gradient itself zero, so the
cap contributed no update; across the `r .zt$endpoints$n` single-run endpoints
$\zeta$ ranged from $`r sprintf("%.3f", .zt$endpoints$min)`$ to
$`r sprintf("%.3f", .zt$endpoints$max)`$ and was never capped
(the $\zeta$-trace script, Table \ref{tab:scripts}, reconstructs the production
optimization path from the cached seed fits and checks it against the cached
final fit to a declared tolerance).

The multiplicative factor for $W$ is then
\begin{equation}
R^{(t)}
=
\frac{
  \frac{(1-\alpha)}{\lVert X \rVert_F^2} X H^{(t+1)\top}
  +\alpha\,\zeta^{(t)}\,\frac{2\alpha}{n_{\mathrm{event}}}\,\nabla_W \ell(W^{(t)},\beta^{(t)})
}{
  \frac{(1-\alpha)}{\lVert X \rVert_F^2} W^{(t)} H^{(t+1)}H^{(t+1)\top}
  +\varepsilon_W
},
\end{equation}
$R^{(t)}$ is clamped elementwise to $[0.2, 5]$ before use, which bounds the
change of any entry of $W$ in a single step, and the proposed update is
\begin{equation}
\label{eqn:W_update}
W^{(t+1)}
=
\max\!\left(
  W^{(t)} \odot R^{(t)},
  \varepsilon_W
\right),
\end{equation}
with a small floor $\varepsilon_W > 0$. When $\alpha=0$, this reduces
to the standard multiplicative update for NMF [@seung2001algorithms];
for $\alpha>0$, the Cox gradient perturbs the update in a direction that
decreases the supervised loss.

Because the combined multiplicative step does not, by itself, guarantee a
nonincrease in the full loss $\mathcal{L}$, we embed it in a backtracking
line search.

\begin{algorithm}[H]
\caption{$W$ update with backtracking}
\label{alg:backtrack}
    \begin{algorithmic}[1]
        \Require $W$ at the previous iteration $t$, $W^{(t)}$; the multiplicative update term at the current iteration $R^{(t)}$, the max number of backtracking updates $max\_bt$, the initial step $s_0 \in (0,1]$, the backtracking factor $\gamma \in (0,1)$
        \Ensure $W^{(t+1)}$
        \State $s = s_0$
        \State b = 1
        \State $flag\_accept = FALSE$
        \While{$b \leq max\_bt$}
            \State $W^{(t+1)}_{cand} = \max\!\left( W^{(t)} \odot [(1-s) + s R^{(t)}],\ \varepsilon_W \right)$
            \State \textbf{(Column normalization)}
            \State \hspace{0.5cm} Compute $D = \mathrm{diag}\!\big(\lVert W^{(t+1)}_{(:,1)cand}\rVert_2,\dots,\lVert W^{(t+1)}_{(:,k)cand}\rVert_2\big)$
            \State \hspace{0.5cm} and set $W^{(t+1)}_{cand} \leftarrow W^{(t+1)}_{cand} D^{-1}$, $H^{(t+1)}_{cand} \leftarrow D H^{(t+1)}$, and $\beta^{(t)}_{cand} \leftarrow D \beta^{(t)}$
            \If{$\mathcal{L}(W^{(t+1)}_{cand},H^{(t+1)}_{cand},\beta^{(t)}_{cand}) \leq \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)})$}
              \State $W^{(t+1)} = W^{(t+1)}_{cand}$
              \State $H^{(t+1)} = H^{(t+1)}_{cand}$
              \State $\beta^{(t)} = \beta^{(t)}_{cand}$
              \State $flag\_accept = TRUE$
              \State break
            \EndIf
            \State $s = s \gamma$
            \State $b = b + 1$
        \EndWhile
        \If{$flag\_accept = FALSE$}
            \State $W^{(t+1)}=W^{(t)}$
        \EndIf
    \end{algorithmic}
\end{algorithm}

In all reported fits $s_0 = 0.5$, $\gamma = 0.5$ and $max\_bt = 10$
(the values passed by the package's R wrapper to the optimizer).
The column normalization inside the line search is described in Normalization
of $W$ below; it is applied to each candidate before the objective is compared,
so the accepted update is the normalized one.

\subsection{Update for $\beta$}

Conditional on $(W,H)$, the loss in $\beta$ reduces to a convex
elastic-net–penalized Cox problem; the common factor $\alpha > 0$ multiplies both
the log-partial-likelihood and the penalty in the joint loss (main-text Equation 1) and is
dropped here, as it does not affect the minimizer:
\[
\min_{\beta \in \mathbb{R}^k}
\left\{
  -\frac{2}{n_{\mathrm{event}}}\,\ell(W,\beta)
  +
  \lambda\left(
    \xi \lVert \beta \rVert_1
    +\frac{1-\xi}{2} \lVert \beta \rVert_2^2
  \right)
\right\}.
\]
We solve this subproblem by cyclic coordinate descent following
[@simon2011regularization]. Writing $\ell(\beta) = \ell(W,\beta)$ and
$\tilde{\eta}_i = Z_i^\top \beta$, the update for coordinate $r$ has the
closed form
\begin{equation}
\label{eqn:beta_update}
\hat{\beta}_r
=
\frac{
  S\!\left(
    \dfrac{1}{n_{\mathrm{event}}}
    \sum_{i=1}^n w(\tilde{\eta})_i
    z_{i,r}
    \left[
      v(\tilde{\eta})_i - \sum_{j \ne r} z_{i,j} \beta_j
    \right],
    \lambda \xi
  \right)
}{
  \frac{1}{n_{\mathrm{event}}} \sum_{i=1}^n w(\tilde{\eta})_i z_{i,r}^2 + \lambda(1-\xi)
},
\end{equation}
where $S(a,\tau) = \mathrm{sign}(a)\max(|a|-\tau,0)$ is the soft-thresholding
operator, and $w(\tilde{\eta})$, $v(\tilde{\eta})$ are standard local
quadratic approximations of the Cox log-partial likelihood [@simon2011regularization];
the factor $1/n_{\mathrm{event}}$ arises because the $2/n_{\mathrm{event}}$ weight
on the log-partial likelihood in the objective cancels the one-half of the
quadratic approximation. We iterate coordinate updates until the subproblem converges to numerical
tolerance, yielding a minimizer in $\beta$ for the current $W$.


\subsection{Normalization of W}

After each accepted $W$ update, we normalize the columns of $W$ as
$W \leftarrow W D^{-1}$, and adjust $H$ and $\beta$ as
\[
H \leftarrow D H, \qquad \beta \leftarrow D \beta,
\]
where $D$ is the diagonal matrix of column $\ell_2$-norms of $W$.
This preserves the reconstruction $WH$ and the linear predictor $W\beta$; because
the accompanying rescaling of $H$ and $\beta$ can alter the penalization terms,
the normalized proposal is evaluated under the full penalized objective within
the backtracking procedure for the $W$ update rather than assumed loss-invariant. Normalization
prevents degeneracy in the scale-nonidentifiable factorization and keeps
columns of $W$ comparable in magnitude and interpretable as gene programs.


\section{Supervision on \(W\) versus \(H\)}

Classical nonnegative matrix factorization (NMF) is non-identifiable: for any
invertible matrix \(R\) with \(W R \ge 0\) and \(R^{-1} H \ge 0\), the factorization
\((W, H)\) can be transformed to \((W R, R^{-1} H)\) without changing the
reconstruction error, yielding multiple valid solutions
[@donoho2004nmf; @laurberg2008uniqueness; @gillis2014nmf]. Although normalizing
columns of \(W\) or rows of \(H\) resolves the trivial scaling ambiguity,
nonnegative mixing transformations persist, so factorizations started from
different initializations need not agree
[@brunet2004metagenes; @kim2007sparse; @gillis2012accelerated]. This applies to
DeSurv as well: individual restarts of the reported model are dispersed, and the
final fit is obtained by a consensus procedure over many restarts
(Stability of the recovered gene programs).

In DeSurv, survival supervision enters through the projection $Z = W^\top X$ in the partial log-likelihood, so that the gradient of the DeSurv loss with respect to the programs is
\[
\nabla \mathcal{L} = (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H) - \alpha\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta),
\]
so the update for $W$ in Equation \ref{eqn:W_update} combines the reconstruction and supervision terms, and the gene-program definitions are shaped directly by their association with survival.

By contrast, if supervision were applied to the sample loadings \(H\), the model
could reduce the Cox loss by redistributing patient-specific coefficients while
leaving the gene-level programs \(W\) nearly unchanged, a behavior documented for
supervised topic models, where supervision applied only to the coefficient
matrix modifies \(H\) but not the basis \(W\) [@blei2007slda].

Supervising \(W\) also provides a practical advantage for **portability**:
because \(W\) defines gene-level programs, new samples can be scored directly by the projection \(Z_{\text{new}} = W^\top X_{\text{new}}\). Supervising \(H\)
would not yield such a mapping: obtaining sample loadings for external datasets
would require an additional sample-specific estimation step, a nonnegative least-squares (NNLS) problem for each new sample.

Supervising through \(W\) is therefore intended to shape the gene programs
toward prognostic directions and to yield signatures that can be scored
unchanged in new cohorts; whether they do so is an empirical question addressed
in Part II.

Supplementary Table \ref{tab:architecture} places DeSurv beside the unsupervised
factorization it extends, the two published survival-supervised factorizations,
and the two conventional supervised predictors used as comparators, by where
survival enters each method, what object is learned, and what must happen when
a new sample arrives. It records architecture only; it does not rank the methods.

```{r tab-architecture, echo=FALSE, results='asis'}
# Every cell is sourced: DeSurv, standard NMF, supervised PCA and penalized Cox
# from this manuscript's Methods; CoxNMF from its published objective
# (arXiv 2008.03776v2, Objective Function section, verified 2026-08-15);
# SurvNMF from its full text (HAL hal-04975434, JdS 2025, verified 2026-09-05).
# Where a source does not describe a behavior the cell says so; a consequence
# of an objective is stated as an inference in the table note, never as a cell.
.arch <- data.frame(
  Method = c("Standard NMF", "DeSurv (this work)", "CoxNMF", "SurvNMF",
             "Supervised principal components", "Penalized Cox regression"),
  `Learned representation` = c(
    "Gene programs $W$ and sample loadings $H$",
    "Gene programs $W$, sample loadings $H$ and Cox coefficients $\\beta$ on $Z = W^\\top X$",
    "Basis $W$ (genes) and coefficients $H$ (samples), with Cox coefficients $\\beta$ on $H$",
    "Sample-side latent matrix and feature matrix (the former is denoted $W$ in SurvNMF's transposed notation; full text)",
    "Principal components of genes passing a univariate Cox score threshold",
    "One coefficient vector over genes (a single linear predictor)"),
  `Where survival enters` = c(
    "Not in the factorization; Cox fitted afterwards on projected scores (Methods)",
    "Cox partial likelihood on $Z = W^\\top X$; its gradient acts on $W$ (Methods)",
    "Cox partial likelihood whose covariates are the columns of $H$ (published objective)",
    "Cox model fitted jointly on the sample-side latent matrix (full text)",
    "Gene screening by Cox score before the components are formed (Methods)",
    "Directly: the objective is the penalized Cox likelihood (Methods)"),
  `Gene-program basis directly supervised?` = c(
    "No", "Yes", "No: the Cox term does not depend on $W$ (published objective)",
    "No: the Cox term is on the sample side (full text)",
    "No: components are unsupervised on the screened genes", "No gene programs are learned"),
  `New-sample scoring` = c(
    "In this study, projection onto the fixed $W$, as for DeSurv (Methods)",
    "Projection $Z = \\widetilde{W}^\\top X_{\\text{new}}$ with $\\widetilde{W}$ frozen (Methods)",
    "Not described in the source",
    "Held-out loadings re-solved by constrained least squares (HALS) with the feature matrix fixed (full text)",
    "Fixed linear map over the screened genes (Methods)",
    "Fixed linear predictor (Methods)"),
  `External latent-variable re-estimation?` = c(
    "No (projection)", "No", "Not described in the source",
    "Yes, for held-out samples; reported for simulated data only (full text)",
    "No", "No"),
  check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(.arch, format = "latex", escape = FALSE, booktabs = TRUE,
  align = "llllll",
  caption = paste0(
    "Where survival supervision enters, and what a new sample requires, for DeSurv and five comparators. ",
    "Cells for DeSurv, standard NMF, supervised principal components and penalized Cox regression are taken from the Methods of this manuscript; ",
    "CoxNMF cells from its published objective function and SurvNMF cells from its full text; both works are cited in Methods, Hyperparameter selection and cross-validation. ",
    "``Not described in the source'' means the cited work does not state the behavior; no other cell is inferred. ",
    "The table records architecture and does not rank the methods. ",
    "Note: CoxNMF does not describe a procedure for scoring new samples. The implication that sample-side loadings would need to be obtained for new samples ",
    "follows from its published objective and is an inference, not a reported implementation. ",
    "\\label{tab:architecture}")) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale_down"), font_size = 9) |>
  kableExtra::column_spec(1, width = "2.2cm") |>
  kableExtra::column_spec(2:6, width = "4.1cm")
```


\section{Cross-validation procedure}
Algorithm \ref{alg:cv} describes the cross validation procedure with $F$ folds and $R$ initializations.
We define the C-index using comparable pairs
$\mathcal{P} = \{(i,j) : y_i < y_j,\ \delta_i = 1\}$ and linear predictors
$\hat{\eta}_i$:
\begin{equation}
\label{eqn:cindex}
\hat{c} =
\frac{1}{|\mathcal{P}|}
\sum_{(i,j)\in\mathcal{P}}
\left[
  \mathbbm{1}(\hat{\eta}_i > \hat{\eta}_j)
  + \tfrac{1}{2}\mathbbm{1}(\hat{\eta}_i = \hat{\eta}_j)
\right].
\end{equation}

\begin{algorithm}[H]
\caption{Cross-validation for DeSurv pipeline}
\label{alg:cv}
\begin{algorithmic}[1]
\Require
\Statex Number of folds $F$
\Statex Number of initializations $R$
\Statex Observed data ($X$, $y$, $\delta$)
\Statex Hyperparameters $k$, $\alpha$, $\lambda$, $\xi$, $\lambda_H$, and number of top genes per factor $n_{top}$
\Ensure The cross-validated C-index

\State Divide subjects into $F$ folds.
\For{$f = 1$ to $F$}
    \State Split data into training and validation $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(X_{(f)},y_{(f)}, \delta_{(f)})$ sets
    \For{$r = 1$ to $R$}
        \State set.seed$(r)$
        \State Apply Algorithm \ref{alg:desurv} with inputs $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(k, \alpha, \lambda, \xi, \lambda_H)$
        \State Obtain $\hat{W}$ and $\hat{\beta}$ as output from Algorithm \ref{alg:desurv}
        \State Restrict $\hat{W}$ to the top $n_{top}$ genes per factor (by factor-specificity score) and compute the estimated linear predictor on this support:
        $
        \hat{\eta} = X_{(f)}^T\hat{W}\hat{\beta}
        $
        \State Compute C-index, denoted $\hat{c}_{(f)r}$, according to Equation \ref{eqn:cindex}
    \EndFor
    \State \textbf{end loop over $r$}
    \State Compute average C-index across initializations:
    $
    \hat{c}_{(f)} = \frac{1}{R}\sum_{r=1}^R\hat{c}_{(f)r}
    $
\EndFor
\State \textbf{end loop over $f$}
\State Compute final cross-validated C-index:
$
\hat{c} = \frac{1}{F}\sum_{f=1}^F \hat{c}_{(f)}
$
\State \Return Final estimate $\hat{c}$

\end{algorithmic}
\end{algorithm}


\section{Bayesian Optimization}
\label{sec:bo}

We treat hyperparameter tuning as a black-box optimization problem over a
bounded domain. Let $\psi$ collect the tuned parameters (at minimum
$\psi = (k,\alpha,\lambda,\xi)$, and optionally discrete choices
such as the signature size $n_{\text{top}}$). For a given $\psi$ we run the
cross-validation procedure in Algorithm \ref{alg:cv}, averaging the C-index
across folds and random initializations to form the objective
\[
\hat{c}(\psi) = \frac{1}{F} \sum_{f=1}^F \hat{c}_{(f)}(\psi).
\]
Because $\hat{c}(\psi)$ is expensive to evaluate and non-differentiable with
respect to $\psi$, we use Bayesian optimization (BO) to adaptively select
candidate configurations.

We place a Gaussian-process (GP) prior on the response surface
$f(\psi) \sim \mathcal{GP}(0, k(\psi,\psi'))$ with a Matérn kernel and
automatic relevance determination length-scales. Continuous parameters are
rescaled to $[0,1]$ and penalties are optimized on a log scale; integer-valued
coordinates (e.g., $k$ or $n_{\text{top}}$) are proposed in the continuous
space and rounded inside the evaluator. Following an initial batch of evaluations,
the GP is refit each time a new observation is added and used to guide acquisition.

At iteration $t$, let $\mu_t(\psi)$ and $\sigma_t(\psi)$ denote the GP
posterior mean and standard deviation, and let $f_\star$ be the best observed
score. We select new candidates by maximizing expected improvement (EI),
\[
\operatorname{EI}(\psi) =
\left(\mu_t(\psi) - f_\star - \kappa\right)\Phi(Z(\psi))
  + \sigma_t(\psi)\phi(Z(\psi)),
\quad
Z(\psi) = \frac{\mu_t(\psi) - f_\star - \kappa}{\sigma_t(\psi)},
\]
where $\Phi$ and $\phi$ are the standard normal cdf/pdf and $\kappa \ge 0$
controls the exploration margin. EI is evaluated over a candidate pool sampled
from the bounded search region, and the maximizer is evaluated next. The loop
continues until the iteration budget is exhausted or the improvement plateaus.

For the PDAC application, the objective was the mean 5-fold cross-validated
Harrell C-index (Algorithm \ref{alg:cv}), with folds stratified by cohort under
a fixed seed (123); each configuration was evaluated from 30 random
initializations and its C-index averaged across them. The bounded search domain
was $k \in [2, 12]$, $\alpha \in [0, 1]$ (the proposal bounds; the model domain
is $\alpha \in [0, 1)$, package validation clips proposals at $\alpha \geq 1$, and the
largest evaluated value was $`r sprintf("%.3f", max(read_result("desurv_bo_results_tcgacptac")$history$alpha_grid, na.rm = TRUE))`$), $\lambda \in [10^{-3}, 10^{3}]$
(optimized on a $\log_{10}$ scale), $\xi \in [0, 1]$, and
$n_{\text{top}} \in [50, 300]$, with the loadings ridge penalty fixed at
$\lambda_H = 0$. Bayesian optimization used an initial design of 50 evaluations
followed by 100 GP-guided iterations (150 total evaluations), maximizing
expected improvement over a candidate pool of $4{,}000$ points sampled from the
bounded domain at each step with an exploration margin $\kappa = 0.01$. The
unsupervised comparator fixed $\alpha = 0$ and tuned the remaining coordinates
identically.

For model selection, we take the highest observed $\hat{c}(\psi)$ as the
optimal configuration. When fold-level standard errors are available, we apply
a one-standard-error rule for $k$: for each rank we retain the configuration
with the highest observed mean cross-validated C-index, set the threshold one
fold-level standard error below the overall best configuration's mean, and
select the smallest rank whose retained configuration meets it.

```{r k-selection-detail, echo=FALSE, message=FALSE, warning=FALSE}
.ks  <- read_result("tar_k_selection_tcgacptac")
.pbk <- read_result("tar_params_best_tcgacptac")
.boh <- read_result("desurv_bo_results_tcgacptac")$history; .boh <- .boh[.boh$status == "ok", ]
.bo3 <- .boh[round(.boh$k_grid) == .ks$k_selected, ]; .bo3best <- .bo3[which.max(.bo3$mean_cindex), ]
.cvg <- read.csv("results/cv_grid/cv_grid_summary.csv")
.cvg <- .cvg[.cvg$k == .ks$k_selected & !is.na(.cvg$ntop) & .cvg$ntop == as.integer(round(.pbk$ntop)), ]
.cvg_near <- .cvg[which.min(abs(.cvg$alpha - .pbk$alpha)), ]
```

Two details of this rule matter for reading the reported numbers. First, the
comparison is between the best configuration observed at each rank: the highest
mean cross-validated C-index anywhere in the search was
$`r sprintf("%.3f", .ks$best_mean)`$ at $k = `r .ks$k_best`$ (fold-level standard
error $`r sprintf("%.3f", .ks$best_se)`$), and the best configuration observed at
$k = `r .ks$k_selected`$ reached $`r sprintf("%.3f", .bo3best$mean_cindex)`$, above
the threshold of $`r sprintf("%.3f", .ks$lcb_threshold)`$. Second, the final model
is a hybrid of the two: the rule fixed $k = `r .ks$k_selected`$, while the remaining
hyperparameters ($\alpha = `r sprintf("%.3f", .pbk$alpha)`$,
$\lambda = `r sprintf("%.3f", .pbk$lambda)`$, $\xi = `r sprintf("%.3f", .pbk$nu)`$,
$n_{\text{top}} = `r as.integer(round(.pbk$ntop))`$) were retained from the overall
best configuration at $k = `r .ks$k_best`$. The final configuration was therefore
not itself a point evaluated in the search, and neither number above is its
cross-validated performance. The nearest evaluated point is in the
fixed-hyperparameter sensitivity grid (Cross-validated C-index across
factorization rank), which holds the same $\lambda$, $\xi$ and $n_{\text{top}}$:
at $k = `r .ks$k_selected`$, $\alpha = `r .cvg_near$alpha`$, the five-fold
cross-validated C-index was $`r sprintf("%.3f", .cvg_near$mean_cindex)`$ (standard
error $`r sprintf("%.3f", .cvg_near$se_cindex)`$). The final DeSurv model
is refit at the selected parameters using the consensus-based initialization
(Consensus initialization for the final model), and the BO-chosen
$n_{\text{top}}$ is used to
truncate each program for downstream analysis.

The parameter $n_{\text{top}}$ determines the factor-specific gene support used for projection and, in the final consensus procedure, the gene sets from each restart that enter the consensus initialization. After factorization, each sample's score for factor $j$ is computed as $\widetilde{W}_j^\top x$, where $\widetilde{W}_j$ retains only the $n_{\text{top}}$ genes with the largest factor-specificity scores $s_{ij}$ (Supplementary Information, Section \ref{sec:gene-program-correlation}); genes outside this set still shape the factorization but do not enter the linear predictor. BO selects $n_{\text{top}}$ jointly with $k$, $\alpha$, $\lambda$, and $\xi$ by optimizing cross-validated C-index. When $\alpha = 0$ (no survival supervision), survival does not enter the factorization objective; it is used for the post hoc Cox fit and for tuning $n_{\text{top}}$, which sets the projection support and, through the consensus initialization, can still influence the final unsupervised fit indirectly.

Supplementary Fig. \ref{fig:fig-si-bo-tuning} illustrates the tuned response surface for the two coordinates of primary interest in the PDAC training cohort: the cross-validated C-index over factorization rank $k$ and supervision strength $\alpha$. Performance is high across a broad region at intermediate supervision and moderate rank, with a second high region at $k \ge 9$ and $\alpha \approx 0.2$--$0.3$, and degrades toward the unsupervised corner ($\alpha = 0$), consistent with a benefit of survival supervision; the second region is consistent with the flat rank dependence of the supervised curve in Section \ref{sec:cvk}. This is an empirical model-selection sensitivity surface, not a simulation result.

```{r fig-si-bo-tuning, fig.cap="Bayesian-optimization tuning surface for DeSurv in the PDAC training cohort (TCGA + CPTAC). Gaussian-process--predicted mean cross-validated C-index over factorization rank ($k$) and supervision strength ($\\alpha$; $\\alpha = 0$ corresponds to the unsupervised limit). The response surface shows a broad region of high cross-validated performance at intermediate supervision and moderate rank, a second high region at $k \\ge 9$ and $\\alpha \\approx 0.2$--$0.3$, and discrimination degrading toward the unsupervised ($\\alpha = 0$) boundary.", echo=FALSE, fig.env='figure*', fig.pos='t', out.height= '3.6in', out.width='4.5in', fig.align='center'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_bo_tuning_surface_tcgacptac.pdf", density = 300))

```

\subsection{Consensus initialization for the final model}
\label{sec:consensus-init}

The final DeSurv model at the selected hyperparameters was initialized from a consensus of multiple independent runs to reduce sensitivity to random starts. We first fit $R = 100$ DeSurv models to the training data from independent random nonnegative initializations at the selected $(k,\alpha,\lambda,\xi,n_{\text{top}})$ (seeds 1--100). For each run, the top $n_{\text{top}}$ genes of every learned program were extracted by factor-specificity score, and a gene-by-gene consensus matrix was accumulated across runs, recording how often each pair of genes appeared together among the top genes of the same program.

Genes appearing in the top sets of fewer than $\lceil 0.3R \rceil$ of the $R$ runs were treated as inactive and excluded from clustering. The consensus matrix over the remaining genes was partitioned into $k$ groups by agglomerative hierarchical clustering (average linkage), and each cluster defined one column of the initial gene-program matrix $W_0$; inactive genes and otherwise-empty entries were set to a small positive floor ($\sqrt{\epsilon_{\text{mach}}}$) to preserve strict positivity. The initial loadings $H_0$ were obtained by nonnegative least squares given $W_0$ and the training expression matrix, and the initial Cox coefficients $\beta_0$ were set to zero. This consensus seed used only training-cohort data and was applied solely to initialize the single final refit after hyperparameter selection; it used no validation-cohort data and therefore introduced no information leakage.

\section{PDAC Datasets}

We analyzed PDAC cohorts from TCGA-PAAD and CPTAC-PDAC as training data
[@raphael2017integrated;@cao2021proteogenomic]. External validation used
independent cohorts from Dijk, Moffitt, PACA-AU (array and RNA-seq), and Puleo
[@dijk2020unsupervised;@moffitt2015virtual;@Bailey2016;@puleo2018stratification].
Training models were fit on the combined TCGA and CPTAC expression matrices
after harmonizing gene identifiers; validation datasets were processed with the
same gene set and transformations before projection, scoring and downstream validation analyses.

Dataset-specific inclusion and preprocessing steps were as follows. For
TCGA-PAAD, we excluded samples lacking tumor grade and mapped features to gene
symbols. For CPTAC, we retained samples annotated as PDAC by histology. For
Dijk, all 90 samples passed quality-control filters and were retained. For
Moffitt, only primary tumor specimens were kept. For PACA-AU array and RNA-seq,
we retained primary PDAC tumors and
collapsed duplicated gene symbols by median. Puleo
required no additional dataset-specific filtering beyond survival QC.

Across cohorts, samples without valid survival outcomes (nonpositive time,
missing time, or missing event indicator) were removed. Expression matrices
were analyzed on a log scale: RNA-seq cohorts as log$_2$(TPM + 1); microarray cohorts in platform-normalized log-scale intensities. For each training cohort (TCGA-PAAD and CPTAC) separately, genes were ranked
in descending order of mean expression and, independently, of variance; the
3,000 genes with the lowest sum of these two ranks were retained, selecting
genes that are both highly expressed and highly variable.
The intersection of these two gene sets yielded 1,970 genes used for model
training. Validation datasets were restricted to this same 1,970-gene set.

A within-sample rank transformation was then applied to every dataset to
mitigate platform and scale effects. Specifically, for each sample the
expression values across the 1,970 retained genes were replaced by their
ascending ranks (ties resolved by average ranks), so that every sample carries
the same nonnegative rank distribution over the shared gene set; ranks are
inherently nonnegative and therefore preserve NMF compatibility. Validation
cohorts were rank-transformed by exactly the same per-sample procedure before
projection, with no use of validation outcomes.

Table \ref{tab:cohorts} summarizes the cohorts, platforms, sample sizes after
quality control, and patient-level metadata used in this study. Overall survival
time and event indicators were available for all cohorts. PurIST
[@rashid2020purity] and DeCAF [@peng2024determination] molecular
classifications, applied to bulk expression data as described in those
references, were used as covariates in the covariate-adjusted external
validation analysis. Gene expression data and molecular
classifications for the validation cohorts were originally compiled for the
DeCAF study [@peng2024determination]; training data were obtained directly
from GDC.

```{r cohorts-events, echo=FALSE, message=FALSE, warning=FALSE}
.coh <- local({
  tr <- read_result("tar_data_filtered_tcgacptac")$sampInfo
  vl <- read_result("val_latent_desurv_tcgacptac")
  ev <- list(TCGA_PAAD = tr$event[tr$dataset == "TCGA_PAAD"],
             CPTAC     = tr$event[tr$dataset == "CPTAC"])
  for (x in vl) ev[[x$dataset]] <- x$survival$event
  ev$train_tot <- tr$event
  ev$val_tot   <- unlist(lapply(vl, function(x) x$survival$event))
  N    <- sapply(ev, length)
  E    <- sapply(ev, function(z) sum(z == 1, na.rm = TRUE))
  rate <- round(100 * E / N)
  cell <- sprintf("%d (%d\\%%)", E, rate); names(cell) <- names(ev)
  vr <- rate[c("Dijk","Moffitt_GEO_array","PACA_AU_array","PACA_AU_seq","Puleo_array")]
  # Combined unique-patient validation set (DeCAF-consistent PACA-AU de-duplication)
  pooled <- do.call(rbind, lapply(vl, function(x) data.frame(
    id = rownames(x$survival), dataset = x$dataset, event = x$survival$event,
    stringsAsFactors = FALSE)))
  pooled <- dedup_combined(pooled, "dataset", "id")
  Ncomb <- nrow(pooled); Ecomb <- sum(pooled$event == 1, na.rm = TRUE)
  list(N = N, cell = cell, comb_N = Ncomb,
       comb_cell = sprintf("%d (%d\\%%)", Ecomb, round(100 * Ecomb / Ncomb)),
       train_lo = min(rate[c("TCGA_PAAD","CPTAC")]), train_hi = max(rate[c("TCGA_PAAD","CPTAC")]),
       val_lo = min(vr), val_hi = max(vr), val_pooled = unname(rate["val_tot"]))
})
```

\begin{table}[ht]
\centering
\caption{Summary of PDAC cohorts used in this study. $N$ denotes the number of
samples after all quality-control filters (histology, survival validity,
dataset-specific exclusions). Patient-level metadata: OS = overall survival
(time and event indicator); PurIST = tumor subtype classification
(Rashid et al. 2020); DeCAF = CAF subtype classification
(Peng et al. 2026). Expression data and molecular classifications
for the validation cohorts were originally compiled for the DeCAF study
(Peng et al. 2026). Events denotes the number of overall-survival events (deaths) with the corresponding event rate; outcome frequency was lower and more homogeneous in the training cohorts (`r .coh$train_lo`--`r .coh$train_hi`\%) than across the validation cohorts (`r .coh$val_lo`--`r .coh$val_hi`\%, pooled `r .coh$val_pooled`\%), a case-mix difference relevant to the interpretation of external-validation performance.}
\label{tab:cohorts}
\footnotesize
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrllll}
\toprule
Cohort & Platform & $N$ & Events & Role & Accession & Metadata & Reference \\
\midrule
TCGA-PAAD        & RNA-seq    & `r .coh$N[["TCGA_PAAD"]]` & `r .coh$cell[["TCGA_PAAD"]]`  & Training   & GDC (TCGA-PAAD) & OS, PurIST, DeCAF & Raphael et al. (2017) \\
CPTAC            & RNA-seq    & `r .coh$N[["CPTAC"]]` & `r .coh$cell[["CPTAC"]]`  & Training   & GDC (CPTAC-3)    & OS, PurIST, DeCAF & Cao et al. (2021) \\
\addlinespace
Dijk             & RNA-seq    &  `r .coh$N[["Dijk"]]` & `r .coh$cell[["Dijk"]]`  & Validation & E-MTAB-6830      & OS, PurIST, DeCAF & Dijk et al. (2020) \\
Moffitt          & Microarray & `r .coh$N[["Moffitt_GEO_array"]]` & `r .coh$cell[["Moffitt_GEO_array"]]`  & Validation & GSE71729         & OS, PurIST, DeCAF & Moffitt et al. (2015) \\
PACA-AU (array)  & Microarray &  `r .coh$N[["PACA_AU_array"]]` & `r .coh$cell[["PACA_AU_array"]]`  & Validation & GSE36924         & OS, PurIST, DeCAF & Bailey et al. (2016) \\
PACA-AU (RNA-seq) & RNA-seq   &  `r .coh$N[["PACA_AU_seq"]]` & `r .coh$cell[["PACA_AU_seq"]]`  & Validation & EGAS00001000154  & OS, PurIST, DeCAF & Bailey et al. (2016) \\
Puleo            & Microarray & `r .coh$N[["Puleo_array"]]` & `r .coh$cell[["Puleo_array"]]` & Validation & E-MTAB-6134      & OS, PurIST, DeCAF & Puleo et al. (2018) \\
\addlinespace
\multicolumn{2}{l}{Training total} & `r .coh$N[["train_tot"]]` & `r .coh$cell[["train_tot"]]` & & & & \\
\multicolumn{2}{l}{Validation platform profiles total} & `r .coh$N[["val_tot"]]` & `r .coh$cell[["val_tot"]]` & & & & \\
\multicolumn{2}{l}{Combined unique-patient validation set} & `r .coh$comb_N` & `r .coh$comb_cell` & & & & \\
\bottomrule
\end{tabular}%
}

{\footnotesize PACA-AU contains 46 patients profiled on both the array and RNA-seq platforms; combined (pooled) analyses represent each patient once, retaining the RNA-seq profile for duplicated patients, consistent with the DeCAF preprocessing rule. Platform-specific per-dataset analyses retain both records.}
\end{table}

\section{Survival Analysis}
\label{sec:survival}

Survival analyses used overall survival time with event indicators as provided
by each cohort. For fitted DeSurv models, sample-level program scores are the
raw projections $Z = \widetilde{W}^\top X$ onto the frozen top-$n_{\text{top}}$
support, and the linear predictor is $\hat{\eta} = Z \hat{\beta}$ with the
trained coefficients; no re-estimation and no standardization enters the score
itself. The elastic-net penalty set the trained coefficient for D2 to exactly
zero, so $\hat{\eta}$ is a weighted sum of the D1 and D3 scores. Prognostic
performance was summarized using the concordance index as defined above
(Cross-validation procedure). When multiple datasets were pooled, Cox models were
stratified by dataset to allow cohort-specific baseline hazards.

Every reported linear predictor is this same linear functional of expression up
to a positive constant; the manuscript uses three scalings of it, and the map
from scaling to reported quantity is as follows. (i) During cross-validated
tuning the collapsed weight vector $\widetilde{W}\hat{\beta}$ is scaled to unit
$\ell_2$ norm (main text, Methods); this changes $\hat{\eta}$ by a positive
constant only and leaves every concordance index unchanged. (ii) The headline
pooled hazard ratio, the covariate-adjusted hazard ratios (main-text Table
\ref{tab:hr-adjusted}) and the leave-one-dataset-out estimates divide
$\hat{\eta}$ by its standard deviation over the pooled, de-duplicated validation
set, so they are per pooled standard deviation. (iii) The per-factor forest
estimates (main-text Fig. \ref{fig:val}a), the cohort-specific hazard ratios of
Table \ref{tab:het-ci} and the between-cohort heterogeneity test standardize each
score to zero mean and unit standard deviation within its own cohort, so they are
per within-cohort standard deviation. Hazard ratios under (ii) and (iii) are
therefore not interchangeable, whereas C-indices are identical under all three.

\subsection{Independence of discovery and validation}

DeSurv
is trained with survival supervision only; it never uses PurIST, DeCAF, or any molecular
subtype label during factorization or hyperparameter selection. The PurIST and DeCAF
subtype calls were fixed before the DeSurv analysis (loaded as precomputed per-cohort
annotations) and enter only as covariates in the adjusted Cox models, not as inputs to
DeSurv. All rank and hyperparameter selection (including the one-standard-error rule) was
performed by cross-validation entirely within the TCGA and CPTAC training data;
no external validation cohort was used for training or tuning.
Validation-cohort expression enters only at scoring, a fixed-basis projection
($Z = \widetilde{W}^\top X_{\text{new}}$) that uses no validation outcome, and
validation-cohort survival outcomes were used only for the final, out-of-sample
performance evaluation.

\section{Gene Program Correlation Analysis}
\label{sec:gene-program-correlation}

To assess the biological identity of learned factors, we evaluated the
correspondence between factor-specific gene rankings and established PDAC gene
programs. Each column $j$ of the gene program matrix $W$ was first scaled by its
column maximum, $W^*_{ij} = W_{ij} / \max_i W_{ij}$, placing all factors on a
common $[0,1]$ scale, and the specificity score for gene $i$ in factor $j$ was
defined as
$$s_{ij} = W^*_{ij} - \max_{j' \neq j} W^*_{ij'}.$$
Genes were ranked in descending order of $s_{ij}$. This ranking defines both the
top-$n_{\text{top}}$ genes used to score samples (Supplementary Data: factor-specific
gene lists) and, for the top 50 genes per factor, the loadings used to quantify
factor--program correspondence.

For each factor $j$ and each reference gene program $G_k$, we computed the
Spearman rank correlation between the continuous factor loadings $W_{\cdot j}$
(restricted to the union of the per-factor top-50 gene sets intersected with the
reference programs) and a binary indicator vector $\mathbf{1}[\text{gene} \in G_k]$
(main-text Fig.\ 2a,b; Methods). $P$ values were derived from the asymptotic
Spearman test and adjusted for multiple comparisons using the Benjamini--Hochberg
procedure within each factor column independently; as in the main text, these
adjusted $P$ values are displayed as descriptive marks, because the gene
universe is itself selected by loading. Reference programs were
included in the visualization only if at least one factor yielded a Spearman
correlation exceeding $0.2$.

The reference universe comprises 45 published PDAC gene programs (Table
\ref{tab:ref-universe}): the PurIST basal-like and classical tumor signatures
[@rashid2020purity]; the DECODER compartment signatures [@peng2019novo]; the
Moffitt tumor, stroma and immune programs [@moffitt2015virtual]; the Collisson
subtypes [@collisson2011subtypes]; the Bailey subtypes, including ADEX
(aberrantly differentiated endocrine exocrine) [@Bailey2016]; the Chan-Seng-Yue
subtypes [@chansengyue2020transcription]; the Maurer ECM and immune programs
[@Maurer2019]; the five Puleo whole-tumor components together with Puleo's two
tumor-compartment signatures [@puleo2018stratification]; the Elyada iCAF and
myCAF signatures [@elyada2019cross]; the SCISSORS fibroblast and tumor
signatures [@leary2023s]; and the DeCAF proCAF and restCAF programs
[@peng2024determination]. Nine source entries were excluded before analysis: six that duplicate a
retained list gene for gene (a second copy of the Moffitt activated and normal
stroma lists, the SCISSORS CAF-versus-perivascular variants of its apCAF, iCAF
and myCAF lists, and a second copy of its perivascular list), the two nine-gene
PurISS iCAF and myCAF lists, and Bailey's non-unique gene set. The SCISSORS iCAF
and myCAF signatures are displayed under the DeCAF nomenclature, restCAF and
proCAF (Stromal resolution: which reference programs are recovered from bulk).
Puleo's whole-tumor ``pure basal-like'' component and its tumor-compartment
``basal-like'' signature are distinct programs and are labeled as such in every
figure. Many programs place only a minority of their genes inside the 1,970-gene
analysis set; the correlation for a program is computed over the genes it shares
with that set, and a program with no such genes cannot be represented.

```{r tab-ref-universe, echo=FALSE, results='asis'}
.ru <- read_result("reference_program_multiplicity")$universe
.src <- c(PurIST = "Rashid et al. 2020", DECODER = "Peng et al. 2019",
          Moffitt = "Moffitt et al. 2015", Collisson = "Collisson et al. 2011",
          Bailey = "Bailey et al. 2016", `Chan-Seng-Yue` = "Chan-Seng-Yue et al. 2020",
          Maurer = "Maurer et al. 2019", Puleo = "Puleo et al. 2018",
          Elyada = "Elyada et al. 2019", SCISSORS = "Leary et al. 2023",
          deCAF = "Peng et al. 2026")
stopifnot(all(.ru$source_group %in% names(.src)), nrow(.ru) == 45)
.rut <- data.frame(`Displayed alias` = .ru$display_label,
                   Source = unname(.src[.ru$source_group]),
                   `Genes` = .ru$n_genes,
                   `In analysis set` = .ru$n_in_analysis,
                   check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(.rut, format = "latex", booktabs = TRUE, linesep = "",
  align = "llrr",
  caption = "The 45-program reference universe used for every factor-to-program correspondence analysis (main-text Fig. 2a,b; Supplementary Figs. \\ref{fig:fig-nmf-k7-heatmap} and \\ref{fig:fig-rpm}), generated from the same renamed and de-duplicated object the analyses read. Displayed alias is the row label used in the figures; Genes is the size of the published list; In analysis set is the number of those genes among the 1,970 analyzed genes, over which the Spearman correlation is computed. The SCISSORS restCAF and proCAF rows are the SCISSORS iCAF and myCAF signatures under DeCAF nomenclature. The two Puleo basal-like rows are different programs: the whole-tumor pure basal-like component and the tumor-compartment basal-like signature. \\label{tab:ref-universe}") |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 7)
```

\section{Reconstruction and Survival Contribution per Factor}

NMF does not provide an orthogonal variance decomposition; it seeks a low-rank
reconstruction of the expression matrix $X \approx WH$. We therefore measure
each factor's contribution to reconstruction rather than to ``variance
explained.'' For a subset of factors $S \subseteq \{1, \ldots, k\}$, define the
reconstruction value
$$v(S) = \|X\|_F^2 - \Big\| X - \sum_{j \in S} W_{\cdot j} H_{j \cdot} \Big\|_F^2,$$
the reduction in residual error relative to the null (zero-factor) model, where
$W_{\cdot j} H_{j \cdot}$ is the rank-one outer-product contribution of factor $j$ (the outer product of column $j$ of $W$ with row $j$ of $H$). The
reconstruction Shapley value [@shapley1953value] of factor $j$,
$$\phi_j = \sum_{S \subseteq \{1,\ldots,k\} \setminus \{j\}}
\frac{|S|!\,(k - |S| - 1)!}{k!}\,\big(v(S \cup \{j\}) - v(S)\big),$$
is its average marginal reconstruction gain over all orderings of the factors.

By the Shapley efficiency axiom, $\sum_j \phi_j = v(\{1,\ldots,k\})$, the total
reconstruction gain of the full model, so the normalized shares
$\phi_j / \sum_l \phi_l$ form an additive decomposition of reconstruction that
sums to one; individual shares may be signed in principle but were all positive
in the reported PDAC models. This is a genuine additive decomposition, in contrast to per-factor
variance fractions, which for non-orthogonal factorizations such as NMF do not
partition (cross-terms between rank-one contributions are non-zero). The Shapley
shares are computed exactly by enumerating the $2^k$ subsets (trivial at the
small ranks considered here).

The survival contribution of factor $j$ was quantified as the Type III partial
likelihood increment: the reduction in Cox partial log-likelihood when factor $j$
is omitted from the full $k$-factor model,
$$\Delta\ell_j = \ell_{\text{full}}(\hat{\beta}_{\text{full}}) - \ell_{-j}(\hat{\beta}_{-j}),$$
where $\ell_{\text{full}}(\hat{\beta}_{\text{full}})$ is the maximized unpenalized
Cox partial log-likelihood of the full $k$-factor model and
$\ell_{-j}(\hat{\beta}_{-j})$ is that of the reduced model refit on the remaining
$k - 1$ factors after omitting factor $j$, each evaluated at its own maximum
partial likelihood estimate. Larger $\Delta\ell_j$ indicates a greater conditional contribution of factor $j$ to Cox model fit given the remaining factors. Unlike the reconstruction Shapley shares, these Type III increments are
conditional (each is evaluated against the full complement of the remaining
$k-1$ factors) and are nonadditive: they do not partition the total model partial
likelihood and are not comparable across factors as variance-style shares. We
therefore report them as per-factor conditional increments rather than as an
additive survival decomposition.

```{r recon-surv-shares, echo=FALSE}
.vs <- read_result("fig_variation_explained_tcgacptac")$data
.vs_share <- function(m) paste(sprintf("%.1f\\%%", 100 * .vs$variance_explained[.vs$method == m][order(.vs$factor[.vs$method == m])]), collapse = "/")
.vs_dl1 <- .vs$delta_loglik[.vs$method == "DeSurv" & .vs$factor == 1]
.vs_nmf_max_dl <- max(.vs$delta_loglik[.vs$method == "NMF"])
stopifnot(.vs_dl1 > 50, .vs_nmf_max_dl < 2)
```

These two quantities are plotted against each other in
main text, Fig.\ 2c. In PDAC at $k = 3$, the reconstruction shares are
$`r .vs_share("NMF")`$ for NMF (N1/N2/N3) and $`r .vs_share("DeSurv")`$ for DeSurv
(D1/D2/D3), while the survival contribution $\Delta\ell$ concentrates in DeSurv
D1 ($\Delta\ell = `r sprintf("%.1f", .vs_dl1)`$) with all NMF factors small
($\Delta\ell \le `r sprintf("%.1f", .vs_nmf_max_dl)`$).

\section{Factor Correspondence Analysis}
\label{sec:factor-correspondence}

To assess which biological programs are preserved, altered, or suppressed
under survival supervision, we quantified the pairwise correspondence between
NMF and DeSurv factor loadings. For each pair of factors
$(j_{\text{NMF}},\, j_{\text{DeSurv}})$, we computed the Spearman rank
correlation between the corresponding columns of the NMF gene program matrix
$W_{\text{NMF}}$ and the DeSurv gene program matrix $W_{\text{DeSurv}}$ over all
genes. We use the full loading profile rather than a top-gene subset, which
would impose a threshold and overstate apparent preservation.
Spearman correlation was used to capture monotone correspondence in gene loading
ranks while remaining robust to scale differences between the two methods.
Results are displayed as a heatmap with NMF factors as rows and DeSurv factors
as columns (main text, Fig.\ 2d).

\section{Analysis scripts and cached objects}
\label{sec:scripts}

Numerical results from the analyses are read at render time from cached
analysis objects rather than transcribed manually. Table \ref{tab:scripts} indexes the scripts referred
to in this document by the analysis they perform, so that the prose above and
below can name analyses rather than files.

```{r tab-scripts, echo=FALSE, results='asis'}
.sc <- data.frame(
  Script = c("code/03", "code/04", "code/05", "code/06", "code/07", "code/11", "code/14",
             "code/15", "code/16", "code/17", "code/18", "code/19", "code/20", "code/21",
             "code/23", "code/24", "code/25", "code/26"),
  Analysis = c(
    "Bayesian optimization of hyperparameters; the matched-budget refinement with $\\alpha$ pinned at 0",
    "Final 100-restart consensus fits (DeSurv, Lee NMF at matched and elbow rank, unsupervised controls)",
    "External validation: frozen projection, linear predictor and per-dataset C-index",
    "Fixed-hyperparameter $k \\times \\alpha$ cross-validation grid and its external-validation curve",
    "Simulation studies (three scenarios, 100 replicates each)",
    "Treated-cohort analyses (O'Kane/COMPASS arm-stratified D1; Linehan paired D2)",
    "Variance partition of the D1 score; additive and interaction adjustment for PurIST and DeCAF",
    "Single-component supervised comparators and recovery of the D1 axis",
    "Subsample stability of top-gene sets across methods",
    "Proper-scoring sensitivity (integrated Brier score, IPA)",
    "Clinical-covariate and tumor-purity adjustment by cohort",
    "Independently tuned supervised comparators, axis decomposition and compartment attribution",
    "Dispersion of individual random restarts",
    "Reproducibility of the consensus endpoint from disjoint restart halves",
    "Stromal reference-program resolution (top-gene and full-gene views)",
    "Reference-program coverage, edges and multiplicity by rank; the reference-universe table",
    "Between-cohort heterogeneity, proportional-hazards diagnostic and bootstrap C-index intervals",
    "Reconstruction of the gradient-scaling factor $\\zeta$ along the production optimization path"),
  `Read in` = c(
    paste0("Sections ", "\\ref{sec:bo}, \\ref{sec:matched}"), paste0("Sections ", "\\ref{sec:consensus-init}, \\ref{sec:progstab}"), paste0("Sections ", "\\ref{sec:survival}, \\ref{sec:het}"), paste0("Sections ", "\\ref{sec:cvk}, \\ref{sec:coverage}"),
    paste0("Section ", "\\ref{sec:sim}"), paste0("Section ", "\\ref{si:treated}"), paste0("Section ", "\\ref{sec:varpart}"), paste0("Section ", "\\ref{sec:d1axis}"),
    paste0("Section ", "\\ref{sec:progstab}"), paste0("Section ", "\\ref{sec:brier}"), paste0("Section ", "\\ref{sec:purity}"), paste0("Section ", "\\ref{sec:tuned}"),
    paste0("Section ", "\\ref{sec:progstab}"), paste0("Section ", "\\ref{sec:progstab}"), paste0("Section ", "\\ref{sec:stroma}"), paste0("Sections ", "\\ref{sec:gene-program-correlation}, \\ref{sec:coverage}"),
    paste0("Sections ", "\\ref{sec:ph}, \\ref{sec:het}"), paste0("Sections ", "\\ref{sec:wupdate}, \\ref{sec:progstab}")),
  check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(.sc, format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
  caption = "Index of analysis scripts. Scripts are numbered as in the code repository; helper functions shared by several scripts (de-duplication of PACA-AU patients, validation scoring, elbow selection) are called from the scripts listed. Section numbers refer to this document. \\label{tab:scripts}") |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 8) |>
  kableExtra::column_spec(2, width = "9.5cm")
```

\clearpage

\newpage

\part{Supplementary Results}

<!-- Supplementary numbering convention: all SI figures labeled "Fig. S1", "Fig. S2", etc.
     (not "Figure 1"). Tables: "Table S1", etc. Internal scenario code names
     (R0_easy, R00_null, R_mixed) are replaced with descriptive labels (prognostic,
     null, mixed) consistent with main-text terminology. -->
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo    = FALSE,
                      fig.path = "figures/",
                      dev     = "png",
                      dpi     = 300)
source("paper/load_precomputed.R")
library(dplyr)
library(ggplot2)
library(cowplot)
if (!exists("set_fig_font", inherits = TRUE)) {
  set_fig_font <- function(plot_obj, size = 10) {
    if (inherits(plot_obj, "ggplot")) {
      plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
    } else {
      plot_obj
    }
  }
}
```


\section{Cross-method reconstructability and reconstruction fidelity}
\label{sec:recon}

The factor-correspondence heatmap of Section \ref{sec:factor-correspondence} (main-text Fig.\ 2d) summarizes pairwise similarity; to make a direction-aware
statement of how much of each factor is \emph{linearly reconstructable} from the
other method's basis, we regressed each factor's gene-loading vector onto the full
set of the other method's loading vectors and report the ordinary least-squares
$R^2$ (Table \ref{tab:projection-r2}). A high $R^2$ means the factor is well
reconstructed as a linear combination of the other method's factors; a low $R^2$
means it is not linearly reconstructable from that basis. This is a statement
about linear reconstructability of gene-loading vectors, not about which method
discovered a program first.

```{r tab-projection-r2, echo=FALSE, results='asis'}
source("R/reconstruction_helpers.R")
.fit_d  <- read_result("tar_fit_desurv_tcgacptac")
.fit_n  <- read_result("fit_std_desurvk_tcgacptac")
.r2_N <- projection_r2(.fit_n$W, .fit_d$W)   # NMF factor -> DeSurv basis
.r2_D <- projection_r2(.fit_d$W, .fit_n$W)   # DeSurv factor -> NMF basis
.pr <- data.frame(
  Factor = c("NMF N1 (tumor)", "NMF N2 (exocrine)", "NMF N3 (microenviron.)",
             "DeSurv D1 (classical/restCAF-like)", "DeSurv D2 (proCAF-like)", "DeSurv D3 (basal-like)"),
  R2 = sprintf("%.2f", c(.r2_N, .r2_D)),
  stringsAsFactors = FALSE)
colnames(.pr) <- c("Factor", "$R^2$ from the other method's basis")
knitr::kable(.pr, format = "latex", escape = FALSE, booktabs = TRUE,
  caption = sprintf("Cross-method projection $R^2$: how reconstructable each factor's gene-loading vector is from the other method's full basis (OLS $R^2$). DeSurv retains the bulk of the unsupervised structure (NMF N1 and N3 carry over at $R^2 \\ge %.2f$) and reorganizes it: the survival-aligned D1 program is essentially unreconstructable from the NMF basis ($R^2 = %.2f$), while the exocrine NMF factor N2 is poorly reconstructed from the DeSurv basis ($R^2 = %.2f$). \\label{tab:projection-r2}", min(.r2_N[1], .r2_N[3]), .r2_D[1], .r2_N[2])) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

DeSurv D1 is reconstructed at $R^2 = `r sprintf("%.2f", .r2_D[1])`$ from the NMF basis: its
gene-loading vector is not well reconstructed as a linear combination of the
matched-rank NMF factors. The biologically dominant tumor (N1) and stromal (N3)
programs, by contrast, are largely reconstructable from the DeSurv basis
($R^2 = `r sprintf("%.2f", .r2_N[1])`$ and $`r sprintf("%.2f", .r2_N[3])`$), so most of the unsupervised structure is
retained; the principal reorganization is that the DeSurv basis resolves tumor
signal into basal- and classical-aligned programs while reconstructing the
exocrine factor (N2) less well ($R^2 = `r sprintf("%.2f", .r2_N[2])`$).

```{r overall-r2-compute, echo=FALSE, message=FALSE, warning=FALSE}
# Computed before the prose below so both the sentence and the table read the
# same objects. Two denominators: see R/reconstruction_helpers.R.
.X <- as.matrix(read_result("tar_data_filtered_tcgacptac")$ex)
.r2u <- c(overall_recon_r2(.fit_n$W, .fit_n$H, .X),
          overall_recon_r2(.fit_d$W, .fit_d$H, .X))
.r2c <- c(overall_recon_r2_centered(.fit_n$W, .fit_n$H, .X),
          overall_recon_r2_centered(.fit_d$W, .fit_d$H, .X))
.gap_u <- 100 * (.r2u[1] - .r2u[2]); .gap_c <- 100 * (.r2c[1] - .r2c[2])
```

At the whole-matrix level, this reorganization carries a real but bounded reconstruction cost (Table \ref{tab:overall-r2}). On the conventional centered $R^2$, DeSurv reconstructs `r sprintf("%.2f", .r2c[2])` of the analysis matrix against `r sprintf("%.2f", .r2c[1])` for unsupervised NMF, a difference of `r sprintf("%.1f", .gap_c)` percentage points and a relative loss of about `r sprintf("%.0f", 100 * .gap_c / (100 * .r2c[1]))`\%; this centered figure is the one reported in the main text. On the uncentered scale the gap is only `r sprintf("%.1f", .gap_u)` percentage points, but that scale flatters any nonnegative low-rank fit to a rank matrix.

```{r tab-overall-r2, echo=FALSE, results='asis'}
.or <- data.frame(
  Model = c("NMF ($k = 3$)", "DeSurv ($k = 3$)"),
  R2c = sprintf("%.3f", .r2c),
  R2u = sprintf("%.3f", .r2u),
  Resid = sprintf("%.3f", 1 - .r2c),
  stringsAsFactors = FALSE)
colnames(.or) <- c("Model", "Centered $R^2$", "Uncentered $R^2$", "Residual fraction (centered)")
knitr::kable(.or, format = "latex", escape = FALSE, booktabs = TRUE,
  caption = sprintf("Whole-matrix reconstruction $R^2$ across the 1,970-gene analysis matrix for unsupervised NMF versus DeSurv at $k = 3$. Two denominators are shown. The centered $R^2$ is the conventional definition, taken about the grand mean, and is the one to read as fraction of variation reconstructed. The uncentered $R^2$ divides by $\\sum X^2$; because the analysis matrix holds within-sample ranks with grand mean $p/2$, the mean alone accounts for about 75\\%% of that denominator, so uncentered values are high for any nonnegative low-rank fit and are reported for completeness. On the centered definition DeSurv gives up %.1f percentage points of reconstruction relative to NMF (%.1f on the uncentered scale). \\label{tab:overall-r2}", .gap_c, .gap_u)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

\section{The D1 axis is not specific to DeSurv}
\label{sec:d1axis}

```{r suprec-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("supervised_recovery_stats")
.sr <- supervised_recovery_stats
.srt <- .sr$table
.srcor <- function(m) abs(.srt$score_cor_D1[.srt$method == m])
```

If the D1 tumor--stroma axis were an artifact of DeSurv's particular
optimizer, independent survival-supervised methods should not recover it. We
therefore fit three other supervised dimension-reduction methods on the same
training cohort ($n = `r .sr$meta$n`$, `r .sr$meta$n_events` events), namely supervised
principal components (Cox-screened genes followed by PCA), one-component Cox
partial least squares against martingale residuals, and a sparse Cox model
(lasso-penalized; `r .sr$meta$n_sparse_cox_genes` genes selected in this single fit), and compared
their leading survival-aligned direction to the DeSurv D1 score. All three
recovered D1 closely (per-patient $|r| = `r sprintf("%.2f", .srcor("Supervised PCA"))`$
for supervised PCA, $`r sprintf("%.2f", .srcor("Cox-PLS"))`$ for Cox-PLS, and
$`r sprintf("%.2f", .srcor("Sparse Cox"))`$ for sparse Cox; mean
$`r sprintf("%.2f", .sr$sup_mean_score_cor)`$; Fig. \ref{fig:suprec}), and each aligned
with D1 rather than with the stromal D2 or basal-like D3 programs (absolute
correlations with D2 and D3 at most
$`r sprintf("%.2f", max(abs(unlist(.srt[.srt$supervised, c("score_cor_D2", "score_cor_D3")]))))`$).

By contrast, the leading \emph{unsupervised} direction
(the first principal component) was only weakly correlated with D1
($|r| = `r sprintf("%.2f", .srcor("Unsupervised PCA"))`$) and instead tracked the
high-variance D2/D3 stroma--basal axis. This D1 axis is thus shared across survival-supervised methods, not an idiosyncrasy of the DeSurv
algorithm. Two additional findings are consistent with this conclusion: D1 is unreconstructable from the
unsupervised NMF basis (Table \ref{tab:projection-r2}), and Lee NMF at matched rank does not recover the restCAF and iCAF marker genes that appear among DeSurv's defining genes (Stromal resolution: which reference programs are recovered from bulk).

```{r fig-suprec, echo=FALSE, fig.width = 8, fig.height = 3.6, fig.cap = "Independent survival-supervised methods recover the D1 axis; an unsupervised method does not. (a) Absolute correlation (across training-cohort patients) of each method's leading direction with the DeSurv D1 score, in the training cohort: supervised principal components, Cox partial least squares, and sparse Cox all recover D1 (green), whereas the leading unsupervised principal component does not (grey). (b) Sample-level agreement with D1 for Cox-PLS (supervised, green) versus the first unsupervised principal component (grey). \\label{fig:suprec}", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
knitr::include_graphics(file.path("..", "figures", "fig_supervised_recovery.pdf"), error = FALSE)
```

```{r suptuned-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_tuned")
.stv    <- desurv_vs_supervised_tuned
.stvp   <- function(m) sprintf("%.2f", .stv$prediction[m, "POOLED"])
.stva   <- function(m, f) sprintf("%.2f", abs(.stv$axis_decomposition[m, f]))
.stvhit <- function(m) unname(.stv$supervised_attribution_across_seeds[[m]]["BasalTumor"])
.stvg   <- function(m, q) unname(.stv$genes_per_seed[[m]][q])
.stvc   <- function(m) sprintf("%.2f", .stv$stromal_carriage[m])
.stvns  <- .stv$meta$nseed
```

\section{Robustness to tuned supervised comparators and compartment attribution}
\label{sec:tuned}

The preceding section establishes D1 as a survival-associated expression direction shared across supervised methods; here we ask what DeSurv adds beyond recovering that direction, using canonical, independently tuned comparators in place of the simple single-component versions above: supervised principal components (\texttt{superpc}; Bair--Tibshirani, with cross-validated screening threshold and component number) and penalized (lasso) Cox (\texttt{glmnet}, a single $\lambda_{\min}$ model). Cox partial least squares was not included in this comparison because a validated multi-component implementation was not available. All methods used the same 1,970-gene universe selected from the training data by expression and variance, and each performed its own feature selection; DeSurv was scored with its Bayesian-optimization-selected top-270-per-factor programs; all scoring used the frozen, rank-transformed transfer with no validation refitting.

External discrimination was similar: the pooled frozen-transfer C-index was $`r .stvp("DeSurv full LP")`$ for the DeSurv linear predictor, $`r .stvp("Supervised PCA")`$ for supervised PCA, and $`r .stvp("Sparse Cox")`$ for penalized Cox. Consistent with the score-level recovery above, both tuned methods aligned with the D1 direction (training $|r| = `r .stva("Sparse Cox","D1")`$ for penalized Cox and $`r .stva("Supervised PCA","D1")`$ for supervised PCA) and with the DeSurv linear predictor ($|r| = `r .stva("Sparse Cox","fullLP")`$ for penalized Cox and $`r .stva("Supervised PCA","fullLP")`$ for supervised PCA), but were only weakly correlated with the activated-stroma/proCAF program D2 ($|r| = `r .stva("Sparse Cox","D2")`$ for penalized Cox and $`r .stva("Supervised PCA","D2")`$ for supervised PCA): a single supervised risk axis captures the dominant tumor-axis signal but not the separate stromal program.

The compartment attribution made this explicit. A sparse supervised risk score selects genes whose expression rises with risk, whereas the high-risk end of D1 is defined by loss of classical-tumor and restCAF-like expression rather than by gain of a D1 gene set, so the genes such a score selects need not be D1's defining genes. Empirically they were basal-like tumor markers: across `r .stvns` cross-validation seeds the genes selected by both methods were enriched (hypergeometric $P < 0.01$) for the basal-like tumor program (in `r .stvhit("Sparse Cox")`/`r .stvns` seeds for penalized Cox and `r .stvhit("Supervised PCA")`/`r .stvns` for supervised PCA) and in no seed for the activated-stroma/proCAF, restraining-stroma, immune, or exocrine compartments. The supervised risk scores did nonetheless retain some stromal signal: each correlated modestly with an external activated-stroma/proCAF signature ($|r| = `r .stvc("Sparse Cox")`$ for penalized Cox and $`r .stvc("Supervised PCA")`$ for supervised PCA), but well below DeSurv's D2 program ($|r| = `r .stvc("DeSurv D2 (proCAF factor)")`$), which tracks that signature directly.

The distinction is therefore representational: DeSurv preserves stromal context as an explicit program even when that program does not contribute independently to the fitted linear predictor.

```{r brier-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_brier_stats")
.br <- desurv_brier_stats; .brp <- .br$pooled; .bripa <- range(.br$per_cohort$ipa)
```

\section{Proper-scoring sensitivity of the external validation}
\label{sec:brier}

The main-text external validation reports discrimination with Harrell's C-index
under a Cox proportional-hazards model. Harrell's C ranks patients by a single
time-invariant score, so when hazards are non-proportional, as the diagnostic
below shows for this predictor, its value depends on the censoring distribution
and need not reflect discrimination at any particular horizon
[@uno2011cstatistics; @antolini2005timedependent], and it is not a proper
scoring rule for predicted risks [@blanche2019cindex]. We therefore additionally
evaluated the DeSurv linear predictor, its direction frozen from training, with
a proper score, the integrated Brier score (IBS), in each validation cohort,
recalibrating only its scalar slope and baseline hazard within each cohort and
integrating over that cohort's follow-up to the 80th percentile of its event
times.

Across the `r c("one","two","three","four","five","six")[nrow(.br$per_cohort)]` validation platform datasets (PACA-AU array restricted to its array-only patients) the DeSurv predictor achieved an
event-weighted integrated Brier score of `r sprintf("%.3f", .brp["ibs"])` versus
`r sprintf("%.3f", .brp["ibs_null"])` for a Kaplan--Meier reference, an index of
prediction accuracy (IPA $= 1 - \mathrm{IBS}/\mathrm{IBS}_{\text{null}}$) of
`r sprintf("%.1f", 100*.brp["ipa"])`\% (per-cohort range
`r sprintf("%.1f", 100*.bripa[1])`--`r sprintf("%.1f", 100*.bripa[2])`\%). Because each
cohort's IBS is integrated over its own event-time horizon, this pooled value is a
descriptive event-weighted summary rather than a single common-horizon proper score,
and the per-cohort IPA range is the more directly interpretable quantity.

Two caveats limit the interpretation of this analysis. First, the recalibration
step estimates both the scalar slope and the baseline hazard from each cohort's
own outcomes, so unlike the C-index and hazard-ratio analyses, which use a fully
frozen predictor, this sensitivity evaluates the transported score direction
under local recalibration rather than a frozen absolute-risk model. Second, the
scores are computed on the same data used to fit that recalibration, with no
cross-validation or resampling, so the reported IBS and IPA are apparent
in-sample values and overstate out-of-sample accuracy. Both caveats weaken the evidentiary value of the apparent improvement, which is already small. This
modest proper-score improvement over the null is consistent with the moderate
discrimination (C $= `r sprintf("%.2f", .brp["cindex"])`$) and with DeSurv's
contribution being a separately resolved tumor- and stroma-associated decomposition
rather than higher predictive accuracy.

\section{Proportional-hazards diagnostic}
\label{sec:ph}

```{r si-ph-diagnostic, echo=FALSE, message=FALSE, warning=FALSE}
# Single owner: the proportional-hazards diagnostic is computed once, in
# code/25_validation_heterogeneity_ci.R, on the de-duplicated patient-level
# table (shared PACA-AU rule, 55cbdaf) with survival::cox.zph on the
# cohort-stratified model. The main text reads the same cached object, so the
# two documents cannot drift apart. code/25 stops if its result ever departs
# from the value originally computed here (chi-square 14.7 on 1 df).
.ph_cache <- read_result("validation_heterogeneity_ci")$ph
.ph_chisq <- .ph_cache$chisq
.ph_df    <- .ph_cache$df
.ph_p     <- .ph_cache$p
.ph_slope <- .ph_cache$slope_time
```

Because the pooled hazard ratio assumes proportional hazards, we tested the scaled Schoenfeld residuals of the validation model, a Cox regression of overall survival on the standardized DeSurv linear predictor with a separate baseline hazard for each dataset (stratification), using the Grambsch--Therneau test. The proportional-hazards assumption was not supported for the standardized DeSurv linear predictor ($\chi^2 = `r sprintf("%.1f", .ph_chisq)`$, $\mathrm{df} = `r .ph_df`$, $P = `r local({e <- floor(log10(.ph_p)); sprintf("%.1f \\times 10^{%d}", .ph_p / 10^e, e)})`$), with the association attenuating over follow-up: the estimated time-varying coefficient declines with follow-up (slope $`r sprintf("%.2f", .ph_slope)`$ on Kaplan--Meier-transformed time) from a positive value. The reported pooled hazard ratio is therefore a time-averaged summary of that association, not a constant multiplicative effect. The horizon-integrated proper-scoring (IBS/IPA) sensitivity above complements it by integrating prediction accuracy over each cohort's event-time horizon rather than through a single hazard-ratio summary.

\section{Between-cohort heterogeneity and C-index confidence intervals}
\label{sec:het}

```{r het-ci-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("validation_heterogeneity_ci")
.hci  <- validation_heterogeneity_ci
.hcic <- .hci$cindex
.hcip <- .hcic[.hcic$cohort == "POOLED", ]
.hdisp <- c(Dijk = "Dijk", Moffitt = "Moffitt", `PACA-AU` = "PACA-AU", Puleo = "Puleo")
```

The five validation platform datasets span four independent patient cohorts, because the PACA-AU array and RNA-seq series profile a partially overlapping set of the same ICGC patients. All analyses in this section are patient-level and de-duplicated by the rule used throughout the manuscript: a patient profiled on both PACA-AU platforms is represented once, by the RNA-seq record, leaving `r .hci$paca$n_after_dedup` PACA-AU patients (`r .hci$paca$n_array_only` array-only and `r .hci$paca$n_rnaseq` RNA-seq) and `r .hci$totals$n` unique patients with `r .hci$totals$events` deaths in total.

The DeSurv linear predictor was standardized \emph{within} cohort (centered and scaled to unit standard deviation separately in each cohort), the same convention used for the per-factor forest estimates in main-text Fig. \ref{fig:val}a and not the pooled-standard-deviation convention used for the headline pooled hazard ratio (Survival Analysis). Within-cohort standardization removes the scale degree of freedom (every cohort's score then has mean zero and unit standard deviation by construction), so a cohort-by-score interaction tests only whether the slope per within-cohort standard deviation differs between cohorts, and not whether the score is more dispersed in one cohort than another. Both models are Cox regressions of overall survival with a separate baseline hazard for each cohort (stratification): the common-slope model has the standardized predictor as its single covariate, and the cohort-specific-slope model adds the predictor-by-cohort interaction. Comparing them by likelihood-ratio test gave $\chi^2 = `r sprintf("%.2f", .hci$lrt$stat)`$ on `r .hci$lrt$df` degrees of freedom ($P = `r sprintf("%.2f", .hci$lrt$p)`$); standardizing within platform rather than within cohort, which removes the small residual location difference between the two PACA-AU platforms, gives $\chi^2 = `r sprintf("%.2f", .hci$sensitivity$lrt_within_platform_standardization$stat)`$ on `r .hci$sensitivity$lrt_within_platform_standardization$df` degrees of freedom ($P = `r sprintf("%.2f", .hci$sensitivity$lrt_within_platform_standardization$p)`$).

Cochran's $Q$ over the four cohort-specific log hazard ratios and their standard errors was $`r sprintf("%.2f", .hci$heterogeneity$Q)`$ on `r .hci$heterogeneity$df` degrees of freedom ($P = `r sprintf("%.2f", .hci$heterogeneity$p)`$), giving $I^2 = `r sprintf("%.0f", 100*.hci$heterogeneity$I2)`$\% under $I^2 = \max\{0, (Q - \mathrm{df})/Q\}$. $Q$ and $I^2$ are descriptive summaries of dispersion among the four estimates and are reported as complements to the interaction test, not as a substitute for it. The likelihood-ratio test is the inferential statement, and with four cohorts both are low-powered against modest heterogeneity.

Confidence intervals for the C-index were obtained by a stratified nonparametric bootstrap on the same de-duplicated table, resampling patients with replacement \emph{within} cohort so that no replicate mixes patients across cohorts or reinstates a de-duplicated PACA-AU record ($B = `r .hci$meta$B`$ replicates, seed `r .hci$meta$seed`, percentile intervals). Concordance was evaluated as Harrell's C with higher scores treated as higher risk, the convention of the external-validation C-indices, and the heterogeneity script (Table \ref{tab:scripts}) verifies that this reproduces the per-dataset C-indices of Supplementary Table \ref{tab:tab-cindex-comparison} exactly. The combined value is the event-weighted mean of the cohort C-indices, the same convention as the proper-scoring and tuned-comparator analyses, and therefore never forms a concordance pair across cohorts. The combined C-index was $`r sprintf("%.3f", .hcip$cindex)`$ (95\% CI $`r sprintf("%.3f", .hcip$lo)`$--$`r sprintf("%.3f", .hcip$hi)`$); the per-cohort hazard ratios and C-indices are given in Table \ref{tab:het-ci}.

\section{Variance partition of the D1 score}
\label{sec:varpart}

The main text reports the fraction of the D1 score's variance explained by the
binary PurIST and DeCAF classifications. D1 scores are first
standardized within cohort, and an ordinary least-squares model is then fitted
with cohort as a fixed effect together with PurIST, DeCAF and their interaction.
The quantity reported is the ordinary $R^2$ of that model, not the adjusted
$R^2$; on $n = 570$ with eight estimated parameters the adjusted value is about
one percentage point lower. Because the scores are standardized within cohort
before fitting, the cohort main effect absorbs essentially no variance, so the
within-cohort normalization of the reported statistic is a formality rather than
a correction. The interaction term contributes negligibly, so the additive and
interaction models give the same figure to the reported precision. Prognostic
increments over the classifiers are assessed separately, by a partial
likelihood-ratio test and the change in concordance from adding D1 to a
cohort-stratified Cox model containing both classifiers.

```{r d1-interaction, echo=FALSE, message=FALSE, warning=FALSE}
load_result("d1_variance_partition_stats")
.d1i <- d1_variance_partition_stats$interaction
.pfmt <- function(p) if (p < 0.001) "$<0.001$" else paste0("$", sprintf("%.3f", p), "$")
```

D1 could also be a smoothed coding of the four joint PurIST-by-DeCAF subtype
cells, so that its classical and restCAF-like arms would simply re-express one
cell of a two-by-two. The main text reports the additive adjustment (D1 added to
a cohort-stratified Cox model containing PurIST and DeCAF as main effects). As a
sensitivity analysis we repeated that test with the PurIST-by-DeCAF interaction
also in the base model; the statistics below are from this
interaction model and differ slightly from the main-text additive ones for that
reason. The interaction term itself is not prognostic
(HR $`r sprintf("%.2f", .d1i$int_hr)`$, $P = `r sprintf("%.2f", .d1i$int_p)`$),
so there was no evidence that the joint classification improved prognostic fit over the additive one. Adding D1 does improve it: it gives a partial likelihood-ratio statistic of
$`r sprintf("%.2f", .d1i$lr_chisq)`$ on 1 degree of freedom
($P = `r sprintf("%.4f", .d1i$lr_p)`$), an adjusted hazard ratio per standard
deviation of $`r sprintf("%.2f", .d1i$d1_hr)`$
($`r sprintf("%.2f", .d1i$d1_lo)`$--$`r sprintf("%.2f", .d1i$d1_hi)`$), and a
concordance increase from $`r sprintf("%.3f", .d1i$c_base)`$ to
$`r sprintf("%.3f", .d1i$c_with_d1)`$. Re-coding the two classifiers as a single
saturated four-level categorical variable is the same model reparameterized and
returns the identical statistic
($\chi^2 = `r sprintf("%.2f", .d1i$cell_lr_chisq)`$). D1's prognostic information is therefore not captured by the saturated four-level joint classification.

\section{Adjustment for clinical covariates and tumor purity}
\label{sec:purity}

The association of the frozen DeSurv linear predictor with survival was estimated in cohort-specific Cox models before and after adjustment for the clinical covariates that cohort annotates and, where available, for tumor purity (Table \ref{tab:clinical-adjustment}). Purity annotation is present in only a subset of cohorts and is derived by different estimators across them (TCGA-PAAD purity was estimated by ABSOLUTE; the PACA-AU and Moffitt purity fields were used as distributed with the cohort annotations, which do not identify the estimator), so it is used here solely as an adjustment covariate; wherever purity could be measured on a full cohort the adjusted hazard ratio was essentially unchanged from the unadjusted one, providing no evidence that the association was explained by measured tumor purity. Consistently, the D1 score itself correlated only weakly with purity in the two full-cohort datasets (Spearman $\rho = `r sprintf("%.2f", read_result("desurv_clinical_adjustment")$PACA_AU_array$purity_cor$D1)`$ in PACA-AU array and $`r sprintf("%.2f", read_result("desurv_clinical_adjustment")$PACA_AU_seq$purity_cor$D1)`$ in PACA-AU RNA-seq), the values reported in the main text.



```{r clinical-adj, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
load_result("desurv_clinical_adjustment")
.ca  <- desurv_clinical_adjustment
.pp  <- function(k) sprintf("%.2f versus %.2f", .ca[[k]]$unadjusted$hr, .ca[[k]]$purity_adjusted$hr)
.caf <- function(m) if (is.null(m)) "---" else sprintf("%.2f (%.2f--%.2f), %s ($n$=%d, %d ev)", m$hr, m$lo, m$hi,
  if (m$p < 0.001) "$<0.001$" else paste0("$", sprintf("%.3f", m$p), "$"), m$n, m$events)
.ord  <- c("PACA_AU_array","PACA_AU_seq","Puleo_array","Dijk","Moffitt_GEO_array","TCGA_PAAD","CPTAC")
.disp <- c(PACA_AU_array="PACA-AU array", PACA_AU_seq="PACA-AU RNA-seq", Puleo_array="Puleo",
           Dijk="Dijk", Moffitt_GEO_array="Moffitt", TCGA_PAAD="TCGA (train)", CPTAC="CPTAC (train)")
.carows <- lapply(.ord, function(k){ r <- .ca[[k]]
  covs <- if (!is.null(r$clinical_adjusted)) paste(r$clinical_adjusted$covars, collapse=", ") else "---"
  data.frame(Cohort=.disp[[k]], Covariates=covs,
             Unadjusted=.caf(r$unadjusted), `Clinical-adjusted`=.caf(r$clinical_adjusted),
             `Purity-adjusted`=.caf(r$purity_adjusted), check.names=FALSE, stringsAsFactors=FALSE) })
.catab <- do.call(rbind, .carows)
knitr::kable(.catab, format="latex", escape=FALSE, booktabs=TRUE,
  caption=paste0("Hazard ratios (per SD of the DeSurv linear predictor) for overall survival, before and after adjustment for standard clinical covariates and, where annotated, tumor purity, fitted separately in each cohort by complete-case analysis (per-cell $n$ and events shown). The Covariates column lists the clinical variables entered (stage, grade, age, sex, nodal status, or resection margin, as available). After clinical adjustment the DeSurv score remains independently prognostic in Puleo, PACA-AU array and both training cohorts, and is attenuated to non-significance in PACA-AU RNA-seq, Dijk and Moffitt, the three smallest complete-case sets. Where tumor-purity annotation was available, purity adjustment left the estimate essentially unchanged wherever it could be measured on the full cohort (PACA-AU array ", .pp("PACA_AU_array"), ", PACA-AU RNA-seq ", .pp("PACA_AU_seq"), ", TCGA ", .pp("TCGA_PAAD"), "; unadjusted versus purity-adjusted), providing no evidence that the association was explained by measured tumor purity. The smaller Dijk and Moffitt cohorts are individually underpowered, and their estimates are correspondingly unstable: Moffitt is non-significant both unadjusted and after adjustment, and its purity-adjusted model rests on only $n = ", .ca$Moffitt_GEO_array$purity_adjusted$n, "$ of its ", .ca$Moffitt_GEO_array$unadjusted$n, " patients. Training-cohort (TCGA, CPTAC) hazard ratios are in-sample. \\label{tab:clinical-adjustment}")) |>
  kableExtra::kable_styling(latex_options=c("HOLD_position","scale_down"), font_size=8)
```

\section{Empirical optimization stability across random initializations}
\label{sec:optstab}

```{r converge-diag, echo=FALSE}
.sf    <- read_result("desurv_seed_fits_tcgacptac")
.fits  <- .sf$fits
.R     <- length(.fits)
.h     <- .fits[[1]]$hyper
.tol   <- .h$tol; .maxit <- .h$maxit
.niter <- vapply(.fits, function(f) length(f$lossit), numeric(1))
.conv  <- vapply(.fits, function(f) isTRUE(f$convergence), logical(1))
.nan   <- vapply(.fits, function(f) isTRUE(f$nan_flag), logical(1))
.mono  <- vapply(.fits, function(f){ d <- diff(f$lossit);
            all(d <= 1e-8 * abs(head(f$lossit, -1)) + 1e-12) }, logical(1))
.n_conv  <- sum(.conv)
.n_maxit <- sum(.niter >= .maxit)
.tol_tex <- sprintf("10^{%d}", as.integer(round(log10(.tol))))
```

The idealized alternating scheme is not accompanied by a formal convergence guarantee for the fitted models. With $\lambda_H = 0$, as in the reported analyses, strict convexity of the $H$-subproblem and uniqueness of its block minimizer are not guaranteed a priori, so convergence results that require unique block minimizers do not apply directly; in addition, the implemented algorithm combines column normalization, gradient balancing between the reconstruction and Cox terms, and a Coxnet approximation of the $\beta$ update. We therefore assessed optimization behavior empirically across $R = `r .R`$ independent random initializations of the selected PDAC model, with a stopping tolerance of $`r .tol_tex`$ and a maximum of `r format(.maxit, big.mark = ",")` iterations.

`r .n_conv` of the `r .R` runs terminated on the stopping criterion rather than the iteration cap`r stopifnot(.n_conv == .R, .n_maxit == 0, !any(.nan), all(.mono))`, none failed numerically, and in every run the recorded penalized objective decreased monotonically to a numerically stable value; the median number of iterations was `r median(.niter)` (range `r min(.niter)`--`r max(.niter)`). The size of the decrease from a random start varies widely between initializations, but that quantity primarily reflects the starting objective value rather than agreement among the final solutions. Whether independent runs of the reported procedure recover the same gene programs is addressed in Stability of the recovered gene programs below.




\section{Stability of the recovered gene programs}
\label{sec:progstab}

```{r program-stability, echo=FALSE, message=FALSE, warning=FALSE}
.ps  <- read_result("program_stability_stats")
.pe  <- read_result("consensus_endpoint_stats")
.pj  <- .ps$pairwise$top_jaccard[, "median"]      # restart vs restart
.pl  <- .ps$pairwise$loading[, "median"]
.ci  <- .ps$cindex
.ej  <- .pe$vs_reported$top_jaccard[, "median"]   # consensus fit vs reported basis
.el  <- .pe$vs_reported$loading[, "median"]
.hl  <- .pe$half_vs_half$loading[, "median"]
.hs  <- .pe$half_vs_half$score[, "median"]
.rng <- function(v) sprintf("%.2f--%.2f", min(v), max(v))
```

The previous section documents that the optimizer converges. It does not
establish that the programs it converges to are the same ones each time, and
that is the property the manuscript relies on when it applies a frozen
decomposition across cohorts. We therefore measured stability at the
level of the recovered programs rather than the objective, using the same
$R = `r .ps$meta$n_restarts`$ cached initializations. No model was refitted for the restart analysis.

Individual restarts are highly dispersed. Comparing pairs of independent
random restarts, and matching their columns one to one by absolute Spearman
correlation of gene loadings, the median overlap between the top
$`r .ps$meta$ntop`$ genes of matched programs was
$`r sprintf("%.2f", median(.pj))`$ by Jaccard index, against a chance baseline of
$`r sprintf("%.3f", .ps$meta$jaccard_chance)`$ for two independent draws of that
size from `r format(.ps$meta$n_genes, big.mark=",")` genes. Matched gene-loading
correlations were moderate (`r .rng(.pl)` across the three programs). Over the
same runs, training concordance was nearly constant (median
$`r sprintf("%.3f", .ci["median"])`$, standard deviation
$`r sprintf("%.3f", .ci["sd"])`$). Near-identical predictive performance from
bases that share little gene membership is consistent with a non-identifiable objective, and it is precisely why the reported model is not a single run.

The reported procedure is reproducible with respect to initialization. The reported basis is obtained by
aggregating restarts into a consensus initialization and then fitting from it.
To test whether that procedure returns the same programs, we partitioned the
cached restarts into two disjoint halves, built a consensus initialization from
each half independently, fitted from each at the production settings, and
compared the results (`r .pe$meta$n_splits` random splits). Two independently
seeded consensus fits agreed closely with each other (gene-loading correlation
`r .rng(.hl)`, patient-score correlation `r .rng(.hs)`), and each agreed still
more closely with the reported basis (gene-loading correlation `r .rng(.el)`,
top-$`r .ps$meta$ntop`$ Jaccard `r .rng(.ej)`). Splitting the restarts halves the
number aggregated relative to the reported procedure, so these figures are
conservative.

The fitted basis also moved materially from the consensus initialization
$W_0$: with programs corresponding by position (confirmed against the full
correlation matrix), gene-loading correlations between $W_0$ and the final $W$
were `r sprintf("%.2f", min(.zt$w0$r))`--`r sprintf("%.2f", max(.zt$w0$r))` and
top-$`r .zt$w0$ntop`$ Jaccard overlaps were
`r sprintf("%.2f", min(.zt$w0$jaccard))`--`r sprintf("%.2f", max(.zt$w0$jaccard))`.
The consensus therefore supplies a reproducible
initialization rather than fixing the final basis.

The optimizer alone is not identifiable. The consensus step is what makes the
reported decomposition reproducible rather than one draw from a wide
distribution, and reproducibility claims in this manuscript refer to that
end-to-end procedure.

Stability under data perturbation is a separate and weaker property. The
analyses above vary the initialization while holding the data fixed. Refitting
on 25 repeated 80\% subsamples of the training cohort, with each DeSurv refit a
single run from one random initialization at the production hyperparameters
rather than the 100-restart consensus procedure, and comparing the top-100 gene
sets of the D1-matched factor across subsamples instead gives a mean pairwise
Jaccard of
$`r sprintf("%.2f", read_result("desurv_vs_supervised_stats")$stability[["DeSurv (D1)"]])`$
for DeSurv, against
$`r sprintf("%.2f", read_result("desurv_vs_supervised_stats")$stability[["Cox-PLS"]])`$
for Cox partial least squares,
$`r sprintf("%.2f", read_result("desurv_vs_supervised_stats")$stability[["Supervised PCA"]])`$
for supervised principal components and
$`r sprintf("%.2f", read_result("desurv_vs_supervised_stats")$stability[["Sparse Cox"]])`$
for sparse Cox. DeSurv is mid-range. The 100-gene list is shorter than the
270-gene projection support so that the comparison is on a common footing with
the single-vector comparators, whose gene rankings are not factor-specific. This
comparison favors dense methods by construction, because a score that spreads
weight across many genes has a more stable top-100 list than one built from a
selective basis, as the sparse Cox value illustrates; it also disfavors DeSurv
relative to the consensus procedure used for the reported model, because each
refit here is a single run. Gene-level resampling stability addresses a complementary property distinct
from reproducibility of the frozen end-to-end procedure.

\section{Simulation results under null and mixed scenarios}
\label{sec:sim}

To characterize the operating characteristics of DeSurv across different signal regimes, we evaluated performance under three simulation scenarios. In the prognostic scenario (main text Fig. \ref{fig:sim}), the factor-specific survival signal contributed relatively little transcriptomic variation compared with the shared outcome-neutral background signal, and survival depended on 150 factor-specific marker genes ($\beta_1 = 2$, $\beta_2 = \beta_3 = 0$). In the null scenario, survival times were generated independently of gene expression ($\beta = 0$), so no prognostic structure existed. In the mixed scenario, survival depended on 300 genes per survival-driving factor, of which 50% were factor-specific markers and 50% were background genes with loadings shared across all factors. All scenarios used $p = 3{,}000$ genes, $n = 200$ samples, $k = 3$ true factors, and identical gamma distribution parameters for the gene loading matrix $W$ (see Methods). Performance was evaluated using the concordance index (C-index) on the held-out 30\% test split of each replicate, precision of recovered prognostic gene sets, and the distribution of selected ranks across simulation replicates.

```{r si-null-alpha, echo=FALSE}
load_result("sim_figs_by_scenario")
sids <- sapply(sim_figs_by_scenario, `[[`, "scenario_id")
aids <- sapply(sim_figs_by_scenario, `[[`, "analysis_id")
null_i <- which(sids == "R00_null" & aids == "bo_tune_ntop")
null_alpha <- with(sim_figs_by_scenario[[null_i]]$k_hist$data, alpha[method == "DeSurv"])
null_alpha_n <- length(null_alpha); null_alpha_median <- median(null_alpha)
null_alpha_q1 <- quantile(null_alpha, .25); null_alpha_q3 <- quantile(null_alpha, .75)
null_alpha_min <- min(null_alpha); null_alpha_max <- max(null_alpha)
prim_i <- which(sids == "R0_easy" & aids == "bo_tune_ntop")[1]
.prim_d <- sim_figs_by_scenario[[prim_i]]$k_hist$data
.prim_prec <- tapply(.prim_d$precision, .prim_d$method, median, na.rm = TRUE)
.prim_rec  <- tapply(.prim_d$marker_recall_all, .prim_d$method, median, na.rm = TRUE)
```

In the prognostic scenario, median matched-factor precision was `r sprintf("%.2f", .prim_prec[["DeSurv"]])` for DeSurv versus `r sprintf("%.2f", .prim_prec[["NMF"]])` for unsupervised NMF, and global marker recall, the fraction of the survival-driving marker genes recovered anywhere in the union of all learned factors' top-$n_{\text{top}}$ gene sets, was `r sprintf("%.2f", .prim_rec[["DeSurv"]])` versus `r sprintf("%.2f", .prim_rec[["NMF"]])`. Recall is insensitive to factor matching because it searches the union of all learned factors rather than a single matched factor; like precision, it depends on each method's selected $n_{\text{top}}$. It complements the matched-factor precision reported in main-text Fig. \ref{fig:sim}.

Under the null scenario (Fig. \ref{fig:sim-null-mixed}a--b), DeSurv showed no advantage over unsupervised NMF ($\alpha = 0$): both methods yielded C-index values near 0.5, and selected ranks concentrated at low values (NMF favoring $k = 2$), consistent with the absence of survival signal. Because the survival objective is uninformative under the null, the cross-validated concordance is essentially flat in $\alpha$ and the Bayesian-optimized supervision strength is unidentified, dispersing across its admissible range rather than concentrating near zero (median $\alpha = `r round(null_alpha_median, 2)`$, IQR $`r round(null_alpha_q1, 2)`$--$`r round(null_alpha_q3, 2)`$; range $`r round(null_alpha_min, 2)`$--$`r round(null_alpha_max, 2)`$ across `r null_alpha_n` replicates). The null-scenario specificity thus reflects the flat concordance surface itself, not the optimizer driving $\alpha$ to zero.

Under the mixed scenario (Fig. \ref{fig:sim-null-mixed}c--f), DeSurv's advantage was concentrated in marker-gene recovery. Precision against the factor-specific marker genes alone (150 genes; Fig. \ref{fig:sim-null-mixed}d left) substantially favored DeSurv, whereas precision against the full 300-gene survival set including background genes (Fig. \ref{fig:sim-null-mixed}d right) was more modestly improved, indicating similar background-gene recovery between methods. The learned factor aligned with the true marker program also received a larger Cox coefficient ($|\beta|$) under DeSurv (Fig. \ref{fig:sim-null-mixed}e), and C-index was modestly higher (Fig. \ref{fig:sim-null-mixed}c). Across the three scenarios the pattern was graded: DeSurv's advantage was greatest when variance and prognosis diverged (prognostic), attenuated when they partially overlapped (mixed, as background-gene dilution reduced the discrimination gain), and absent when no survival signal existed (null). These simulations indicate that DeSurv is most advantageous when prognostic structure is not aligned with the dominant sources of transcriptomic variation.

```{r fig-sim-null-mixed, fig.width = 6.5, fig.height = 7, fig.cap = "DeSurv performance under null and mixed simulation scenarios. (a,b) Null scenario ($\\beta = 0$): both methods yield C-index $\\approx 0.5$ (a) and comparable rank distributions (b), so both methods show no out-of-sample prognostic discrimination under the null; precision is omitted as no true survival genes exist. (c--f) Mixed scenario: survival depends on factor-specific marker genes (50\\%) and shared background genes (50\\%). (c) C-index is modestly higher for DeSurv. (d) Precision by gene-set definition: marker-only precision (150 genes; left facet) substantially favors DeSurv; full 300-gene survival-set precision (right facet) shows a more modest improvement. (e) The matched factor receives a larger fitted Cox coefficient $|\\beta|$ under DeSurv. (f) Selected-rank distributions. DeSurv (blue); NMF (red, $\\alpha = 0$); both Bayesian-optimized. \\label{fig:sim-null-mixed}", fig.env='figure*', fig.pos='t', out.height = "6.5in", out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_sim_null_mixed_tcgacptac.pdf", density = 300))
```


\section{Cross-validated C-index across factorization rank}
\label{sec:cvk}

To evaluate how the number of latent factors $k$ affects prognostic performance, as a sensitivity analysis we performed an exhaustive grid search over $k \in \{2, 3, \ldots, 12\}$ and supervision strength $\alpha \in \{0, 0.05, \ldots, 0.95\}$ using 5-fold cross-validation on the TCGA+CPTAC training cohort. For each $k$, the best-performing $\alpha$ was selected by maximizing the mean CV C-index. The unsupervised limit ($\alpha = 0$) was evaluated at the same $k$ values for comparison. Results are shown for two gene-selection settings: $n_{\text{top}} = 270$ top genes per factor and $n_{\text{top}} = \text{ALL}$ (all genes retained), both at $\lambda = 0.349$ and $\xi = 0.056$. External validation C-index is shown descriptively; validation outcomes were not used to select $k$, $\alpha$, $\lambda$, $\xi$, or $n_{\text{top}}$.

Fig. \ref{fig:fig-cindex-by-k} shows the resulting C-index curves for both training (CV) and external validation (pooled across held-out cohorts), with shaded ribbons indicating $\pm 1$ standard error. Rank was selected on the training cross-validated concordance. For DeSurv that curve is close to flat across the whole range examined, spanning about two standard errors from $k = 2$ to $k = 12$, with its nominal maximum at $k = 9$ rather than at the selected rank. This fixed-$\lambda$, $\xi$, $n_{\text{top}}$ grid is a different exercise from the full Bayesian-optimization search over all tuned hyperparameters, whose highest observed mean C-index occurred at $k = `r .ks$k_best`$; applying the same one-standard-error rule to this fixed-hyperparameter sensitivity grid also selects $k = 3$. The selection depends on the gene-selection setting: under panel (b), with all genes retained, the same rule selects $k = 2$. On the training cross-validated curve DeSurv had a higher point estimate than its unsupervised limit at every $k$ examined. The external-validation curve, shown only descriptively, does not show that ordering uniformly: the unsupervised limit is higher at $k = 6$ and at $k = 10$--$12$, and the supervised curve dips at $k = 6$, where the single grid fit at its selected $\alpha$ is the same fit whose factors correspond to almost no reference program (Reference-program coverage, edges and multiplicity). These rank-wise points are single fits, not consensus fits. At the selected rank the validation curve is consistent with the training-based choice, with DeSurv at $k = 3$ attaining concordance comparable to the rank-optimized unsupervised control ($k = 7$; Table \ref{tab:tab-cindex-comparison}).

```{r fig-cindex-by-k, fig.width = 10, fig.height = 10, fig.cap="Training (CV) and external validation C-index as a function of factorization rank $k$ for DeSurv (blue) and its unsupervised limit (red, $\\alpha = 0$). Shaded ribbons show $\\pm 1$ SE. For each $k$, the best $\\alpha$ was selected by maximizing the mean CV C-index. Validation C-index is pooled across all external cohorts. (a) Results using $n_{\\text{top}} = 270$ top genes per factor ($\\lambda = 0.349$, $\\xi = 0.056$). (b) Results retaining all genes ($n_{\\text{top}} = \\text{ALL}$; $\\lambda = 0.349$, $\\xi = 0.056$).", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
# This figure is produced by the cv_grid pipeline (separate store);
# read directly from the file output.
# Use magick to render the PDF as a raster (avoids LaTeX path issues)
# NOTE: DL flagged duplicate legend across panels A and B. The legend is baked
# into the pre-rendered PDF; consolidating it requires regenerating the source
# figure in inst/cv_grid_validation_analysis_amber_optimal.R on HPC.
img <- magick::image_read("figures/cv_cindex_by_k_primary.pdf", density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

\section{Standard NMF rank selection heuristics}

Standard unsupervised rank selection criteria (reconstruction residuals, cophenetic correlation coefficient, and mean silhouette width) yielded inconsistent guidance for selecting the factorization rank $k$ in the PDAC training cohort (TCGA + CPTAC). Reconstruction residuals decreased smoothly without a visually distinct elbow (Fig. \ref{fig:nmf-diagnostics}a); the Kneedle elbow rule (maximum deviation of the normalized curve from its chord) applied to this curve nonetheless returns $k = 5$, which is the rank used for the elbow-selected Lee NMF comparator in Table \ref{tab:tab-cindex-comparison}. Cophenetic correlation fluctuated without a distinct transition point (Fig. \ref{fig:nmf-diagnostics}b). Mean silhouette width favored very low ranks that conflicted with other criteria (Fig. \ref{fig:nmf-diagnostics}c). These inconsistencies illustrate the ambiguity of rank selection under unsupervised criteria in this dataset; the main text selects rank on cross-validated prognostic performance.

```{r fig-nmf-diagnostics, fig.width = 6, fig.height = 3.5, fig.cap = "Standard unsupervised NMF rank selection heuristics for the PDAC training cohort (TCGA + CPTAC). (a) Reconstruction residuals as a function of factorization rank $k$. (b) Cophenetic correlation coefficient as a function of $k$. (c) Mean silhouette width of the basis, coefficient and consensus matrices as a function of $k$. These criteria yield inconsistent guidance, illustrating the ambiguity of rank selection under unsupervised criteria; the main text selects rank on cross-validated prognostic performance. \\label{fig:nmf-diagnostics}", fig.env='figure*', fig.pos='t', out.height= "3.5in", out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_diagnostics_tcgacptac.pdf", density = 300))
```




\section{Factor structure of the rank-optimized unsupervised control ($k = 7$)}

Throughout the main text, both DeSurv and Lee NMF were fit at the same rank ($k = 3$), enabling a controlled comparison of the effect of survival supervision on factor structure. When the unsupervised limit of the same pipeline selects its own rank, Bayesian optimization at $\alpha = 0$ selects $k = 7$ (the rank-optimized unsupervised control), whereas the one-standard-error rule applied to the supervised DeSurv BO results selects $k = 3$. We therefore examine the factor structure of this independently selected $k = 7$ solution to characterize how the additional factors relate to established programs when supervision is absent (Fig. \ref{fig:fig-nmf-k7-heatmap}).

At BO-selected $k = 7$ under $\alpha = 0$ (Fig. \ref{fig:fig-nmf-k7-heatmap}), the rank-optimized unsupervised control corresponds to a broader set of PDAC gene signatures than the matched-rank $k = 3$ solution, with some programs represented by more than one factor; Supplementary Fig. \ref{fig:fig-rpm} quantifies the accompanying changes in coverage and multiplicity.

```{r fig-nmf-k7-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene-program correspondence for the rank-optimized unsupervised control ($k = 7$, BO-selected rank at $\\alpha = 0$), computed with the same Spearman-correlation analysis and reference programs as main-text Fig.\\ 2a,b (top 50 genes per factor; reference programs with $r > 0.2$ for at least one factor are shown). Compared with the matched-rank $k = 3$ solution, the $k = 7$ solution shows broader correspondence to established PDAC programs, with some programs represented by more than one factor. Supplementary Fig.\\ \\ref{fig:fig-rpm} quantifies the accompanying changes in coverage and multiplicity. \\label{fig:fig-nmf-k7-heatmap}", fig.env='figure', fig.pos='t'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_k7_heatmap_tcgacptac.pdf", density = 300))
```

\subsection{Reference-program coverage, edges and multiplicity}
\label{sec:coverage}

```{r rpm-si-stats, echo=FALSE, message=FALSE, warning=FALSE}
.rpmsi <- read_result("reference_program_multiplicity")
.rpmsi_get <- function(id, field) .rpmsi$anchors[[field]][.rpmsi$anchors$anchor_id == id]
.rpmsi_universe <- .rpmsi$metadata$reference_universe_size
.rpmsi_co3 <- .rpmsi_get("desurv_k3", "coverage")
.rpmsi_ed3 <- .rpmsi_get("desurv_k3", "edges")
.rpmsi_mu3 <- sprintf("%.2f", .rpmsi_get("desurv_k3", "multiplicity"))
.rpmsi_co_n3 <- .rpmsi_get("nmf_k3_matched_rank", "coverage")
.rpmsi_ed_n3 <- .rpmsi_get("nmf_k3_matched_rank", "edges")
.rpmsi_mu_n3 <- sprintf("%.2f", .rpmsi_get("nmf_k3_matched_rank", "multiplicity"))
.rpmsi_co7 <- .rpmsi_get("nmf_k7_unsupervised", "coverage")
.rpmsi_ed7 <- .rpmsi_get("nmf_k7_unsupervised", "edges")
.rpmsi_mu7 <- sprintf("%.3f", .rpmsi_get("nmf_k7_unsupervised", "multiplicity"))
```

Fig.\ \ref{fig:fig-nmf-k7-heatmap} (and main-text Fig.\ 2a,b) display factor-to-reference-program correspondence as a heatmap. We additionally summarize this same correspondence across factorization rank $k$ using three quantities computed from the same correlation matrix (top 50 genes per factor by factor-specificity score, Spearman correlation of factor loadings against binary reference-program membership, $r > 0.2$ representation threshold; Supplementary Information, Section \ref{sec:gene-program-correlation}), over the fixed `r .rpmsi_universe`-program reference universe. Coverage is the number of reference programs with at least one factor at $r > 0.2$. Edges is the total number of (factor, reference-program) pairs at $r > 0.2$. Reference-program multiplicity is edges divided by coverage, defined only when coverage $> 0$, and reports edges per represented program. A low multiplicity is not by itself evidence that fewer programs are represented, so multiplicity is always reported alongside coverage rather than in place of it.

At each rank $k = 2, \ldots, 12$, these quantities are computed for two single fits at fixed $\lambda$, $\xi$ and $n_{\text{top}} = 270$ from the $k \times \alpha$ cross-validation grid: the unsupervised fit ($\alpha = 0$) and the fit at the $\alpha$ maximizing mean cross-validated C-index, the same $\alpha$-selection convention used for Supplementary Fig.\ \ref{fig:fig-cindex-by-k}. These rank-wise points are single fits at fixed hyperparameters, not the 100-restart consensus fits used elsewhere in the manuscript. Three consensus-fit anchors are overlaid: DeSurv ($k = 3$), standard NMF at matched rank ($k = 3$; Lee NMF), and the rank-optimized unsupervised control ($k = 7$, $\alpha = 0$).

At these anchors, DeSurv reaches coverage `r .rpmsi_co3`, edges `r .rpmsi_ed3`, multiplicity `r .rpmsi_mu3`; matched-rank standard NMF reaches coverage `r .rpmsi_co_n3`, edges `r .rpmsi_ed_n3`, multiplicity `r .rpmsi_mu_n3`; and the rank-optimized unsupervised control at $k = 7$ reaches coverage `r .rpmsi_co7`, edges `r .rpmsi_ed7`, multiplicity `r .rpmsi_mu7`.

```{r rpm-collapse, echo=FALSE}
.rpc <- .rpmsi$per_config[.rpmsi$per_config$arm == "supervised", ]
.rpc_low <- .rpc[.rpc$coverage <= 2, ]
stopifnot(identical(sort(.rpc_low$k), c(6, 8)))
.rpc6 <- .rpc_low[.rpc_low$k == 6, ]; .rpc8 <- .rpc_low[.rpc_low$k == 8, ]
.rpu <- .rpmsi$per_config[.rpmsi$per_config$arm == "unsupervised", ]
.rp_neighbors <- min(c(.rpu$max_rho[.rpu$k %in% c(6, 8)], .rpc$max_rho[.rpc$k %in% c(5, 7, 9)]))
stopifnot(.rp_neighbors >= 0.5, max(.rpc_low$max_rho) < 0.3)   # the sentence says "barely above 0.2" vs "at least" the neighbor minimum
```

The supervised curve collapses at two ranks: at $k = `r .rpc6$k`$ ($\alpha = `r .rpc6$alpha`$) and $k = `r .rpc8$k`$ ($\alpha = `r .rpc8$alpha`$) the single grid fit reaches coverage `r .rpc6$coverage` and `r .rpc8$coverage` (edges `r .rpc6$edges` and `r .rpc8$edges`), because the largest correlation between any of its factors and any reference program is only $`r sprintf("%.2f", .rpc6$max_rho)`$ and $`r sprintf("%.2f", .rpc8$max_rho)`$, barely above the $0.2$ representation threshold, whereas the unsupervised fits at the same ranks and the supervised fits at the neighboring ranks $k = 5, 7, 9$ all reach at least $`r sprintf("%.2f", .rp_neighbors)`$. These are single fits from one random start at fixed hyperparameters, and the $k = 6$ fit is also the one whose external concordance dips in Supplementary Fig. \ref{fig:fig-cindex-by-k}; their collapse is the single-run dispersion documented in Stability of the recovered gene programs, not a property of the consensus fits. Across the remaining ranks, coverage increases with rank considerably more than multiplicity does, indicating that the additional factors at higher rank extend correspondence to more reference programs rather than densely duplicating correspondence to the same ones. Through $k = 7$ the dominant change is increased coverage with modest repeated representation. This is not a claim about predictive performance, which is addressed separately (Table \ref{tab:hr-adjusted}; Supplementary Table \ref{tab:tab-cindex-comparison}).

```{r fig-rpm, fig.width = 6, fig.height = 8, fig.cap = "Reference-program coverage, edges and multiplicity by factorization rank $k$, for the unsupervised ($\\alpha = 0$) and supervised (best $\\alpha$ by cross-validated C-index) curves, computed from the same factor-to-reference-program correspondence analysis as Fig.\\ \\ref{fig:fig-nmf-k7-heatmap} and main-text Fig.\\ 2a,b (top 50 genes per factor, $r > 0.2$ threshold, same reference-program universe). Rank-wise points are single fits at fixed $\\lambda$, $\\xi$ and $n_{\\text{top}}$, not the 100-restart consensus fits used elsewhere; markers denote the three consensus-fit anchors (DeSurv $k = 3$, matched-rank standard NMF $k = 3$, rank-optimized unsupervised control $k = 7$). \\label{fig:fig-rpm}", fig.env='figure', fig.pos='t', out.width='0.72\\textwidth'}
knitr::include_graphics(file.path("..", "figures", "fig_reference_program_multiplicity.pdf"), error = FALSE)
```

```{r tab-cindex-comparison, echo=FALSE, results='asis'}
library(dplyr)
library(knitr)
library(kableExtra)
load_result("val_cindex_desurv_tcgacptac")
load_result("val_cindex_std_desurvk_tcgacptac")
load_result("val_cindex_std_elbowk_tcgacptac")
load_result("val_cindex_desurv_alpha0_tcgacptac")
load_result("val_cindex_desurv_a0k3_tcgacptac")

fmt_cindex <- function(tbl, col_name) {
  tbl %>%
    dplyr::select(dataset, cindex) %>%
    dplyr::rename(!!col_name := cindex)
}

cindex_wide <- fmt_cindex(val_cindex_desurv_tcgacptac, "DeSurv ($k = 3$)") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_desurvk_tcgacptac, "Lee NMF ($k = 3$)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_desurv_a0k3_tcgacptac, "Matched-budget control ($k = 3$, $\\alpha = 0$)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_elbowk_tcgacptac, "Lee NMF ($k = 5$, elbow)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_desurv_alpha0_tcgacptac, "Rank-optimized control ($k = 7$, $\\alpha = 0$)"), by = "dataset") %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))) %>%
  dplyr::rename(Dataset = dataset) %>%
  dplyr::mutate(Dataset = c(Dijk = "Dijk", Moffitt_GEO_array = "Moffitt", PACA_AU_array = "PACA-AU array",
                            PACA_AU_seq = "PACA-AU RNA-seq", Puleo_array = "Puleo")[Dataset])

# Explicit \label is the canonical referenced name (main paper + SI \ref it);
# push knitr's auto chunk-label to a throwaway name so it neither collides with
# the explicit one nor is silently relied upon in its place.
knitr::opts_current$set(label = "tab-cindex-comparison-auto")
kable(cindex_wide, format = "latex", booktabs = TRUE, linesep = "", escape = FALSE,
      caption = "External validation C-index by method and validation dataset. DeSurv at the one-standard-error-selected rank ($k = 3$) is compared with Lee NMF at matched rank ($k = 3$) and at elbow-selected rank ($k = 5$), with the matched-budget unsupervised control ($k = 3$, $\\alpha = 0$) and with the rank-optimized unsupervised control ($k = 7$, $\\alpha = 0$). The unsupervised columns are not all produced by one implementation: the two Lee NMF columns are \\texttt{NMF::nmf}, whereas the two controls are the DeSurv optimizer run at $\\alpha = 0$ with the same consensus-initialization procedure as the supervised model. The matched-budget column is the tuning-budget control: it holds rank, optimizer, consensus-initialization procedure and Bayesian-optimization budget fixed while re-optimizing the remaining hyperparameters with $\\alpha = 0$. It therefore varies only supervision and does not improve on \\texttt{NMF::nmf} at the same rank. Higher values indicate better discrimination. All columns are scored on the same union-of-top-genes support ($n_{\\text{top}} = 270$). \\label{tab:tab-cindex-comparison}") |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

```{r tab-het-ci, echo=FALSE, results='asis'}
.hpc <- .hci$per_cohort
.hrow <- data.frame(
  Cohort = c(unname(.hdisp[.hpc$cohort]), "Combined"),
  `$n$` = c(.hpc$n, .hci$totals$n),
  Deaths = c(.hpc$events, .hci$totals$events),
  `HR per within-cohort SD (95\\% CI)` = c(
    sprintf("%.2f (%.2f--%.2f)", .hpc$hr, .hpc$lo, .hpc$hi), "---"),
  `C-index (95\\% CI)` = sprintf("%.3f (%.3f--%.3f)", .hcic$cindex, .hcic$lo, .hcic$hi),
  check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(.hrow, format = "latex", escape = FALSE, booktabs = TRUE, align = "lrrcc",
  caption = paste0("Cohort-specific external-validation estimates for the frozen DeSurv linear predictor, on the de-duplicated patient-level table (", .hci$totals$n, " unique patients, ", .hci$totals$events, " deaths; PACA-AU represented once, RNA-seq retained). Hazard ratios are per standard deviation of the linear predictor computed within that cohort, the convention used for the interaction test, so they are not comparable to the pooled-standard-deviation hazard ratio in main-text Table \\ref{tab:hr-adjusted}. C-index confidence intervals are percentile intervals from ", .hci$meta$B, " stratified bootstrap replicates (seed ", .hci$meta$seed, ") resampling within cohort; the combined C-index is the event-weighted mean of the four cohort values. \\label{tab:het-ci}")) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

```{r a0k3-control, echo=FALSE, message=FALSE, warning=FALSE}
.pf <- read_result("tar_params_best_a0k3_tcgacptac")
.pd <- read_result("tar_params_best_tcgacptac")
.Wd_c <- read_result("tar_fit_desurv_tcgacptac")$W
.Wf_c <- read_result("tar_fit_desurv_a0k3_tcgacptac")$W
.Wn_c <- read_result("fit_std_desurvk_tcgacptac")$W
.cf <- abs(cor(.Wd_c, .Wf_c, method = "spearman"))
.cn <- abs(cor(.Wd_c, .Wn_c, method = "spearman"))
```

\subsection{Stromal resolution: which reference programs are recovered from bulk}
\label{sec:stroma}

```{r stroma-res, echo=FALSE, message=FALSE, warning=FALSE}
load_result("stroma_resolution_stats")
.sr2 <- stroma_resolution_stats
.srN <- .sr2$top_n; .srF <- .sr2$full_rho
```

Bulk deconvolution has conventionally separated activated from normal stroma,
whereas the finer cancer-associated fibroblast states, and the proCAF/restCAF
classification derived from them, were established in single-cell data. We
therefore asked which stromal reference programs each $k = 3$ factorization
actually resolves. Two views are reported because they do not
agree, and the defensible claim depends on which question is asked.

In the top-gene view shown in main-text Fig.\ 2a,b, which asks whether signature
genes are enriched among the 50 genes that define each factor, the
activated-stroma program is recovered comparably by all methods
(`r .srN["DECODER_ActivatedStroma","DeSurv"]`, `r .srN["DECODER_ActivatedStroma","NMF_k3"]`
and `r .srN["DECODER_ActivatedStroma","fair_a0k3"]` of
`r .sr2$sig_size["DECODER_ActivatedStroma"]` genes for DeSurv, standard NMF and the
matched-budget control). The restraining-stroma programs behave differently:
`r .srN["SCISSORS_restCAF","DeSurv"]` of `r .sr2$sig_size["SCISSORS_restCAF"]`
restCAF genes and `r .srN["Elyada_iCAF","DeSurv"]` of
`r .sr2$sig_size["Elyada_iCAF"]` iCAF genes appear among DeSurv's defining genes,
against `r .srN["SCISSORS_restCAF","NMF_k3"]` and `r .srN["Elyada_iCAF","NMF_k3"]`
for standard NMF. The proCAF programs, by contrast, are recovered comparably by
all methods (`r .srN["SCISSORS_proCAF","DeSurv"]` versus
`r .srN["SCISSORS_proCAF","NMF_k3"]` of `r .sr2$sig_size["SCISSORS_proCAF"]` genes).

In the full-gene view, which correlates loadings across all 1,970 analyzed genes
with signature membership, no method separates restCAF from proCAF: every
correspondence lies between
$`r sprintf("%.2f", min(.srF[c("SCISSORS_restCAF","Elyada_iCAF","SCISSORS_proCAF","Elyada_myCAF"),]))`$
and
$`r sprintf("%.2f", max(.srF[c("SCISSORS_restCAF","Elyada_iCAF","SCISSORS_proCAF","Elyada_myCAF"),]))`$,
whereas the activated-stroma axis is recovered by all methods at
$`r sprintf("%.2f", max(.srF["DECODER_ActivatedStroma",]))`$. This view is far
more dilute for a 15--25 gene signature among 1,970 genes and correspondingly
low powered, but it bounds what can be claimed. We therefore report the
restraining-stroma difference as marker-gene presence among the defining genes
of each factorization, and do not claim to have resolved the single-cell CAF
states from bulk expression.

Note also that the SCISSORS iCAF and myCAF reference signatures are displayed as
restCAF and proCAF in main-text Fig.\ 2, following the naming used by the DeCAF
classification; the underlying gene sets are the single-cell-derived ones.

\subsection{Matched-budget unsupervised control at the same rank}
\label{sec:matched}

The matched-rank comparison in the main text uses Lee multiplicative NMF, which
receives neither DeSurv's consensus initialization nor its hyperparameter
tuning. That asymmetry leaves open whether the difference in factor structure reflects
tuning effort rather than supervision. We therefore fit an unsupervised control that holds everything but supervision fixed: Bayesian optimization was run with $\alpha$ pinned to $0$ and the rank pinned to DeSurv's selected $k = `r .pd$k`$, with the same optimizer, search budget, gene filter and cross-validation folds, re-optimizing only the remaining hyperparameters; the resulting model was then fit by the same 100-restart consensus procedure used for DeSurv.

The control
selected $\lambda = `r sprintf("%.4f", .pf$lambda)`$,
$\xi = `r sprintf("%.4f", .pf$nu)`$ and $n_{\text{top}} = `r .pf$ntop`$; note
that its gene support was selected near the upper bound of the search range ($300$), whereas the supervised model settled at
$n_{\text{top}} = `r .pd$ntop`$ well inside it. Because $\alpha = 0$ removes the
survival term, the selected $\lambda$ is not comparable like-for-like with the
supervised model's.

The control does not recover D1. Its best-matching factor reaches Spearman
$\rho = `r sprintf("%.2f", max(.cf[1, ]))`$ against D1, slightly lower than the
$\rho = `r sprintf("%.2f", max(.cn[1, ]))`$ attained by \texttt{NMF::nmf} at the
same rank, while still recovering D2
($\rho = `r sprintf("%.2f", max(.cf[2, ]))`$) and D3
($\rho = `r sprintf("%.2f", max(.cf[3, ]))`$). Its external discrimination is
likewise no better than \texttt{NMF::nmf} at $k = 3$
(Table \ref{tab:tab-cindex-comparison}). A sensitivity analysis holding the
supervised model's own $\lambda$, $\xi$ and $n_{\text{top}}$ fixed and varying
only $\alpha$ gave the same conclusion. Matching the tuning budget therefore does not close the representational gap, supporting survival supervision rather than tuning effort alone as the source of the reorganization reported in the main text.

Lee NMF at elbow-selected $k = 5$ and the rank-optimized unsupervised control at $k = 7$ both substantially improve over Lee NMF at $k = 3$ across all cohorts (Table \ref{tab:tab-cindex-comparison}); additional unsupervised factors capture prognostically relevant variation missed at the matched rank. At $k = 7$, the rank-optimized control exceeds DeSurv's C-index in all five validation datasets, so unsupervised factorization given more factors discriminates at least as well as the supervised three-program model. The tradeoff is between that higher discrimination and a less parsimonious factorization: the $k = 7$ control represents a broader set of established programs across seven factors (Fig. \ref{fig:fig-nmf-k7-heatmap}; Supplementary Fig. \ref{fig:fig-rpm}) rather than concentrating the prognostic signal in three programs. Both the $k = 7$ control and the DeSurv linear predictor retain significant prognostic value after adjustment for established classifiers (main-text Table \ref{tab:hr-adjusted}); DeSurv does so with fewer factors.


\section{Supplementary Data: factor-specific gene lists}

As Supplementary Data, Tables \ref{tab:gene-list-270-D1}--\ref{tab:gene-list-270-D3} list the top 270 genes for each of the three DeSurv factors, ranked by the factor specificity score $s_{ij}$ defined in Supplementary Information, Section \ref{sec:gene-program-correlation}. The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\text{top}} = 270$); these are the genes used to compute sample-level factor scores for survival prediction.

```{r support-disjoint, echo=FALSE}
.nt_sup <- as.integer(round(read_result("tar_params_best_tcgacptac")$ntop))
.tg_sup <- DeSurv::desurv_get_top_genes(read_result("tar_fit_desurv_tcgacptac")$W, .nt_sup)
.pos_sup <- colSums(.tg_sup$diffs > 0)
.uni_sup <- length(unique(unlist(.tg_sup$top_genes)))
```

Gene symbols in these tables are retained as distributed with the source expression matrices compiled for the training and validation cohorts (Section 6) and were not updated to current HGNC nomenclature; several therefore appear under previous symbols (for example SEPT11, WISP1, CYR61 and H1F0). Because the analysis set was selected for high mean expression as well as high variance, abundantly expressed ribosomal-protein and housekeeping genes are eligible for the factor supports, and some appear among the D2 and D3 defining genes (for example RPS26 and RPS28 in D2; RPS16, RPS21, RPL28 and GAPDH in D3); they enter the supports through the factor-specificity ranking and are not excluded post hoc. In the fitted model, each factor had at least $`r .nt_sup`$ genes with positive factor specificity (`r paste(format(.pos_sup, big.mark = ",", trim = TRUE), collapse = ", ")` for D1, D2 and D3). Because this specificity can be positive for at most one factor, the three top-$`r .nt_sup`$ sets were disjoint, yielding an `r .uni_sup`-gene projection support. Factor D1 is enriched for classical tumor and restCAF-like stromal genes, Factor D2 for proCAF-associated stromal genes, and Factor D3 for basal-like tumor genes.

```{r tab-gene-list-270, echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)
library(tidyr)
library(dplyr)

genes <- read.csv("paper/gene_lists_top270_k3.csv", stringsAsFactors = FALSE)

# Keep only DeSurv factors
desurv_genes <- genes %>%
  dplyr::filter(factor %in% c("D1", "D2", "D3"))

# For each factor, split 270 genes into 6 columns of 45
build_factor_table <- function(df, fct) {
  g <- df %>% dplyr::filter(factor == fct) %>% dplyr::arrange(rank) %>% dplyr::pull(gene)
  data.frame(
    a = g[1:45], b = g[46:90], c = g[91:135],
    d = g[136:180], e = g[181:225], f = g[226:270],
    stringsAsFactors = FALSE
  )
}

factor_labels <- c("D1", "D2", "D3")
factor_captions <- c(
  "Factor D1 (classical tumor + restCAF-like stroma)",
  "Factor D2 (proCAF-like stroma)",
  "Factor D3 (basal-like tumor)"
)
factor_descriptions <- c(
  "classical tumor and restCAF-like stroma",
  "proCAF-like stroma",
  "basal-like tumor"
)

for (i in seq_along(factor_labels)) {
  tbl <- build_factor_table(desurv_genes, factor_labels[i])
  # Make knitr's auto-generated table label unique per iteration so the three
  # loop tables don't all collide on tab:tab-gene-list-270 (multiply-defined).
  # The referenced per-factor labels (tab:gene-list-270-D<i>) are set in the caption.
  knitr::opts_current$set(label = paste0("tab-gene-list-270-", factor_labels[i]))
  cat("")
  kable(tbl, format = "latex", booktabs = TRUE, linesep = "",
        col.names = c("1--45", "46--90", "91--135", "136--180", "181--225", "226--270"),
        caption = paste0("Top 270 genes for DeSurv factor ", factor_labels[i], ", ",
          factor_descriptions[i], ", ranked by factor specificity score $s_{ij}$.",
          if (i == 1) paste0(" The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\\text{top}}$).") else "",
          " \\label{tab:gene-list-270-", factor_labels[i], "}"),
        escape = FALSE) %>%
    kable_styling(font_size = 8, latex_options = c("HOLD_position")) %>%
    add_header_above(setNames(6, factor_captions[i]), bold = TRUE) %>%
    cat()
  cat("")
}
```

\section{Translational transportability into independent treated PDAC cohorts}
\label{si:treated}

```{r treated-setup, echo=FALSE, message=FALSE, warning=FALSE}
# Treated-cohort accessors. Only two treated resources are used: the
# O'Kane/COMPASS metastatic cohort (arm-stratified D1 transportability) and the
# Linehan paired pre/post biopsies (treatment-associated change in D2).
load_result("treated_cohort_stats")
.tcs  <- treated_cohort_stats
.pfmt <- function(p) if (is.na(p)) "---" else if (p < 0.001) "$<0.001$" else sprintf("%.3f", p)
.oka  <- function(a) { r <- .tcs$okane$per_arm[.tcs$okane$per_arm$arm==a, ]
  sprintf("$%.2f$, 95\\%% CI $%.2f$--$%.2f$", r$hr, r$lo, r$hi) }
```

The programs were learned in treatment-naive, non-metastatic PDAC, the setting in which the gene-program matrix $W$ and Cox coefficients $\beta$ were trained. As a translational test we projected the fixed programs, without retraining, into independently treated PDAC and asked whether they retained clinically relevant associations. We used two resources. The first is the O'Kane/COMPASS cohort of treated metastatic PDAC, laser-capture microdissected and therefore epithelial-enriched, which supports a transportability test of the tumor-associated $D_1$ direction under a model stratified by treatment arm (Fig.\ \ref{fig:treated}a). The second is the borderline-resectable or locally advanced Linehan cohort, in which paired pre- and post-treatment biopsies allow a within-patient test of whether the stromal $D_2$ program changes with therapy (Fig.\ \ref{fig:treated}b).

Each analysis is confined to a single cohort; no estimate is pooled across cohorts. Within O'Kane/COMPASS, arm-specific estimates are reported alongside a common $D_1$ slope from a Cox model stratified by treatment arm, which allows arm-specific baseline hazards. The elastic-net penalty shrank the trained survival coefficient for $D_2$ to exactly zero in the treatment-naive model (an estimated selection outcome, not an imposed constraint), so the paired analysis is reported as a change in program level and not as evidence that $D_2$ predicts survival.



\subsection{Prognostic transportability of D1 in treated metastatic disease}

To evaluate transportability in treated metastatic disease while retaining treatment-arm structure, we projected the frozen DeSurv programs into the O'Kane/COMPASS cohort and first estimated $D_1$ associations separately within each arm. Higher $D_1$ scores had protective point estimates in all three treatment arms: FOLFIRINOX (HR per SD `r .oka("FFX")`), gemcitabine/nab-paclitaxel (HR `r .oka("GA")`) and the GA/experimental arm (HR `r .oka("GA/experimental")`; Fig.\ \ref{fig:treated}a). An arm-stratified Cox model allowing treatment-specific baseline hazards yielded a common association of HR `r sprintf("$%.2f$, 95\\%% CI $%.2f$--$%.2f$", .tcs$okane$stratified["hr"], .tcs$okane$stratified["lo"], .tcs$okane$stratified["hi"])` ($P = `r .pfmt(.tcs$okane$stratified["p"])`$; `r .tcs$okane$n` patients, `r .tcs$okane$events` deaths), with no evidence of heterogeneity across treatment arms (interaction $P = `r sprintf("%.2f", .tcs$okane$interaction_p)`$). Because these specimens were epithelial-enriched by laser-capture microdissection, this analysis specifically tests transportability of the tumor-associated $D_1$ direction rather than the stromal programs.

\subsection{Treatment-associated change in D2}

In the `r .tcs$paired$n` Linehan patients with paired pre- and post-treatment biopsies, $D_2$ increased significantly from pre- to post-treatment (paired Wilcoxon $P = `r sprintf("%.3f", .tcs$paired$wilcox_p)`$; mean $\Delta = `r sprintf("%+.2f", .tcs$paired$mean_delta)`$ SD; Fig.\ \ref{fig:treated}b), indicating a shift toward proCAF-associated bulk expression. This bulk analysis cannot distinguish a change in fibroblast state from altered tissue composition.

\subsection{Caveats}

These analyses are exploratory and prognostic, not predictive: treatment, stage and cohort are confounded, the treated cohorts are modestly sized, and no analysis plan was preregistered. Each analysis is confined to a single cohort, and the O'Kane/COMPASS common slope is estimated within that cohort from a model with arm-specific baseline hazards; it is not a pooled estimate across cohorts. We do not test a survival association for $D_2$ in the paired Linehan cohort, because its trained coefficient is zero. No $D_2$ analysis is possible in O'Kane/COMPASS, whose specimens are laser-capture epithelium rather than bulk tissue. Robust treated-cohort validation, and any assessment of whether the programs predict differential treatment benefit, will require larger harmonized series with adequately powered treatment-by-program interaction analyses.
