---
title: "Supplementary Information"
subtitle: "Survival-supervised deconvolution identifies a reproducible tumor--stroma prognostic axis in pancreatic cancer"
author: Amber M. Young, Alisa Yurovsky, Xianlu Laura Peng, Jen Jen Yeh, Didong Li, Naim U. Rashid
header-includes:
- \providecommand{\pandocbounded}[1]{#1}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{xr-hyper}
- \externaldocument{paper}
- \usepackage{amsmath}
- \usepackage{amsfonts}
- \usepackage{amssymb}
- \usepackage{bbm}
- \usepackage{algorithm}
- \usepackage{algpseudocode}
- \usepackage{setspace}
- \usepackage{enumitem}
- \usepackage[nolists,nomarkers]{endfloat}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator\diag{diag}
- \renewcommand{\algorithmicrequire}{\textbf{Input:}}
- \renewcommand{\algorithmicensure}{\textbf{Output:}}
- \renewcommand{\thefigure}{S\arabic{figure}}
- \renewcommand{\thetable}{S\arabic{table}}
- \renewcommand{\thealgorithm}{S\arabic{algorithm}}
bibliography: references_30102025.bib
csl: nature.csl
link-citations: true
reference-section-title: "SI References"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    keep_tex: true
  html_document: default
---
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\newpage

\part{Supplementary Methods}

\section{Model details}
Let $X \in \mathbb{R}_{\ge 0}^{p \times n}$ denote the nonnegative gene
expression matrix with $p$ genes and $n$ subjects. DeSurv approximates
$X \approx WH$ with $W \in \mathbb{R}_{\ge 0}^{p \times k}$ and
$H \in \mathbb{R}_{\ge 0}^{k \times n}$, and links the shared program matrix
$W$ to survival via an elastic-net–penalized Cox model. Supervision is placed on the program matrix $W$, the basis that transfers to new cohorts, whereas the sample loadings $H$ serve only to reconstruct the training expression matrix; new samples are therefore scored by projection onto the learned programs, $Z = W^\top X$, rather than by re-estimating loadings.

Let $y \in \mathbb{R}^n_{> 0}$ be the observed survival times,
$\delta \in \{0,1\}^n$ the event indicators, and
$Z = W^\top X \in \mathbb{R}^{k \times n}$ the sample-wise factor scores. We write $Z_i$ for the $i$th column of $Z$, and
$n_{\text{event}} = \sum_{i=1}^n \delta_i$. For convenience we recall the
DeSurv loss from Equation 1 in the main text and rewrite it as
\begin{align}
  \mathcal{L}(W,H,\beta) &=\nonumber \\
  &(1-\alpha)\left(\frac{1}{2\lVert X \rVert_F^2}\lVert X - WH \rVert_F^2 + \frac{\lambda_H}{2nk} \lVert H \rVert_F^2\right) \nonumber \\
  -&\alpha \left(\frac{2}{n_{event}}\ell(W,\beta) - \lambda \{ \xi \lVert \beta \rVert_1 + \tfrac{1-\xi}{2} \lVert \beta \rVert_2^2 \} \right) \nonumber
\end{align}

where $\lVert \cdot \rVert_F$ is the Frobenius norm and
$\beta \in \mathbb{R}^k$ are the Cox coefficients. The Cox log-partial
likelihood $\ell(W,\beta)$ is
\begin{equation}
\ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left[
  Z_i^\top \beta
  -
  \log \left(
    \sum_{j : y_j \ge y_i}
    \exp(Z_j^\top \beta)
  \right)
\right].
\end{equation}
The inner sum runs over the risk set $\{j : y_j \ge y_i\}$ of subjects still at risk at the observed time $y_i$; tied event times are handled by the Breslow approximation in the \texttt{glmnet} Coxnet routine used for the $\beta$ update (below).
Hyperparameters satisfy $\alpha \in [0,1)$,
$\lambda_H \ge 0$, $\lambda \ge 0$, and $\xi \in [0,1]$.
The constants $1/(2\lVert X \rVert_F^2)$, $2/n_{\text{event}}$, and $1/(2nk)$ place the reconstruction, loadings-penalty, and survival terms on comparable numerical scales for stable optimization.


\section{Optimization Algorithm}

\subsection{Block coordinate descent scheme}
Optimization proceeds by block coordinate descent (BCD; Algorithm \ref{alg:desurv}), which at each outer iteration updates $H$, then $W$, then $\beta$ while holding the other blocks fixed.

\begin{algorithm}[htbp]
\caption{DeSurv block coordinate descent}
\label{alg:desurv}
    \begin{algorithmic}[1]
        \Require $X \in \mathbb{R}_{\geq 0}^{p \times n}$, survival times $y \in \mathbb{R}^n_{> 0}$, event indicators $\delta \in \{0,1\}^{n}$, tolerance $tol$, maximum iterations $maxit$, supervision parameter $\alpha$, rank $k$, penalties $\lambda_H$, $\lambda$, and $\xi$
        \Ensure Fitted DeSurv parameters $(W,H,\beta)$

        \State Initialize $W^{(0)}, H^{(0)}$ with positive entries (e.g., $W^{(0)},H^{(0)} \sim \mathrm{Uniform}(0,\max X)$)
        \State Initialize $\beta^{(0)}$ (e.g., $\beta^{(0)} = \mathbf{1}$)
        \State $loss = \mathcal{L}(W^{(0)},H^{(0)},\beta^{(0)})$
        \State $eps = \infty$, $t = 0$
        \While{$eps \geq tol$ \textbf{and} $t < maxit$}
            \State \textbf{(H-update)}
            \State \hspace{0.5cm} Update $H^{(t+1)}$ from $H^{(t)}$ using the multiplicative rule in Eq. \ref{eqn:Hupdate},
            \State \hspace{0.5cm} holding $W^{(t)}$ and $\beta^{(t)}$ fixed.
            \State \textbf{(W-update)}
            \State \hspace{0.5cm} Update $W^{(t+1)}$ from $W^{(t)}$ using the hybrid multiplicative rule in Eq. \ref{eqn:W_update},
            \State \hspace{0.5cm} holding $H^{(t+1)}$ and $\beta^{(t)}$ fixed, and perform backtracking to ensure
            \State \hspace{0.5cm} $\mathcal{L}$ does not increase.
            \State \textbf{($\beta$-update)}
            \State \hspace{0.5cm} Update $\beta^{(t+1)}$ by solving the elastic-net Cox subproblem using cyclic coordinate descent to numerical tolerance
            \State \hspace{0.5cm} (Eq. \ref{eqn:beta_update}), holding $W^{(t+1)}$ and $H^{(t+1)}$ fixed.
            \State $lossNew = \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})$
            \State $eps = |lossNew - loss| / |loss|$
            \State $loss = lossNew$
            \State $t = t + 1$
        \EndWhile
        \State \Return $\hat{W} = W^{(t)}, \hat{H} = H^{(t)}, \hat{\beta}=\beta^{(t)}$
    \end{algorithmic}
\end{algorithm}

\subsection{Update for $H$}
Conditional on $W$, the loss $\mathcal{L}(W,H,\beta)$ reduces to a
convex quadratic function of $H$. We adopt the standard NMF multiplicative
update with $\ell_2$ penalty:
\begin{equation}
\label{eqn:Hupdate}
H \leftarrow
\max\!\left(
  H \odot \frac{W^\top X}{
    W^\top W H + \dfrac{\lVert X \rVert_F^2}{nk}\,\lambda_H H + \varepsilon_H
  },
  \varepsilon_H
\right),
\end{equation}
where $\odot$ denotes elementwise multiplication, all divisions are
elementwise, and $\varepsilon_H > 0$ is a small floor to prevent entries from
becoming exactly zero. This update is equivalent to a majorization–minimization
step and guarantees a nonincreasing reconstruction term conditional on $W$
[@seung2001algorithms; @pascualmontano2006nonsmooth]. The $\varepsilon_H$-floor
ensures that iterates remain in the interior of the nonnegative orthant, which
keeps the multiplicative $H$-update numerically well-defined.


\subsection{Update for $W$}
For $W$, we construct a hybrid multiplicative update that balances the
contributions of the NMF and Cox gradients. Let
\[
\nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H)
=
\frac{1}{\lVert X \rVert_F^2}(W H - X)H^\top,
\]
and let
\[
\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta)
=
\frac{2}{n_{event}}\nabla_W \ell(W,\beta).
\]
where $\nabla_W \ell(W,\beta)$ denotes the Cox gradient with respect to $W$. A closed-form expression for $\nabla_W \ell$ is
\begin{equation}
\nabla_W \ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left(
  x_i
  -\frac{
      \sum_{j : y_j \ge y_i}
        x_j \exp(x_j^\top W \beta)
    }{
      \sum_{j : y_j \ge y_i}
        \exp(x_j^\top W \beta)
    }
\right)\beta^\top,
\end{equation}
where $x_i$ denotes the $i$th column of $X$.

We define a gradient-scaling heuristic
\begin{equation}
\label{eq:supp-zeta}
\zeta^{(t)}
=
\frac{
  \left\lVert (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W^{(t)},H^{(t+1)}) \right\rVert_F^2
}{
  \left\lVert \alpha \nabla_W \mathcal{L}_{Cox}(W^{(t)},\beta^{(t)}) \right\rVert_F^2
},
\end{equation}
and clip $\zeta^{(t)}$ to avoid extreme values:
$\zeta^{(t)} \leftarrow \min(\zeta^{(t)}, \zeta_{\max})$
with $\zeta_{\max} = 10^6$ in all experiments.

The multiplicative factor for $W$ is then
\begin{equation}
R^{(t)}
=
\frac{
  \frac{(1-\alpha)}{\lVert X \rVert_F^2} X H^{(t+1)\top}
  +\frac{2\alpha}{n_{\mathrm{event}}}\,\zeta^{(t)}\,\nabla_W \ell(W^{(t)},\beta^{(t)})
}{
  \frac{(1-\alpha)}{\lVert X \rVert_F^2} W^{(t)} H^{(t+1)}H^{(t+1)\top}
  +\varepsilon_W
},
\end{equation}
and the proposed update is
\begin{equation}
\label{eqn:W_update}
W^{(t+1)}
=
\max\!\left(
  W^{(t)} \odot R^{(t)},
  \varepsilon_W
\right),
\end{equation}
with a small floor $\varepsilon_W > 0$. When $\alpha=0$, this reduces
to the standard multiplicative update for NMF [@seung2001algorithms];
for $\alpha>0$, the Cox gradient perturbs the update in a direction that
decreases the supervised loss.

Because the combined multiplicative step does not, by itself, guarantee a
nonincrease in the full loss $\mathcal{L}$, we embed it in a backtracking
line search.

\begin{algorithm}[H]
\caption{$W$ update with backtracking}
\label{alg:backtrack}
    \begin{algorithmic}[1]
        \Require Value of $W$ and the previous iteration $t$: $W^{(t)}$, the multiplicative update term at the current iteration $R^{(t)}$, the max number of backtracking updates $max\_bt$, the backtracking parameter $\rho \in (0,1)$
        \Ensure $W^{(t+1)}$
        \State $\theta =1$
        \State b = 1
        \State $flag\_accept = FALSE$
        \While{$b \leq max\_bt$}
            \State $W^{(t+1)}_{cand} = \max\!\left( W^{(t)} \odot [(1-\theta) + \theta R^{(t)}],\ \varepsilon_W \right)$
            \State \textbf{(Column normalization)}
            \State \hspace{0.5cm} Compute $D = \mathrm{diag}\!\big(\lVert W^{(t+1)}_{(:,1)cand}\rVert_2,\dots,\lVert W^{(t+1)}_{(:,k)cand}\rVert_2\big)$
            \State \hspace{0.5cm} and set $W^{(t+1)}_{cand} \leftarrow W^{(t+1)}_{cand} D^{-1}$, $H^{(t+1)}_{cand} \leftarrow D H^{(t+1)}$, and $\beta^{(t)}_{cand} \leftarrow D \beta^{(t)}$
            \If{$\mathcal{L}(W^{(t+1)}_{cand},H^{(t+1)},\beta^{(t)}) \leq \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)})$}
              \State $W^{(t+1)} = W^{(t+1)}_{cand}$
              \State $H^{(t+1)} = H^{(t+1)}_{cand}$
              \State $\beta^{(t)} = \beta^{(t)}_{cand}$
              \State $flag\_accept = TRUE$
              \State break
            \EndIf
            \State $\theta = \theta*\rho$
            \State $b = b + 1$
        \EndWhile
        \If{$flag\_accept = FALSE$}
            \State $W^{(t+1)}=W^{(t)}$
        \EndIf
    \end{algorithmic}
\end{algorithm}

Column normalization preserves the reconstruction $WH$ and linear predictor
$W\beta$. Because the accompanying rescaling can alter penalization terms, the
normalized proposal is evaluated using the full penalized objective within the
backtracking procedure, which accepts the update only if it does not increase
$\mathcal{L}$.

\subsection{Update for $\beta$}

Conditional on $(W,H)$, the loss in $\beta$ reduces to a convex
elastic-net–penalized Cox problem; the common factor $\alpha > 0$ multiplies both
the log-partial-likelihood and the penalty in the joint loss (Equation 1) and is
dropped here, as it does not affect the minimizer:
\[
\min_{\beta \in \mathbb{R}^k}
\left\{
  -\frac{2}{n_{\mathrm{event}}}\,\ell(W,\beta)
  +
  \lambda\left(
    \xi \lVert \beta \rVert_1
    +\frac{1-\xi}{2} \lVert \beta \rVert_2^2
  \right)
\right\}.
\]
We solve this subproblem by cyclic coordinate descent following
[@simon2011regularization]. Writing $\ell(\beta) = \ell(W,\beta)$ and
$\tilde{\eta}_i = Z_i^\top \beta$, the update for coordinate $r$ has the
closed form
\begin{equation}
\label{eqn:beta_update}
\hat{\beta}_r
=
\frac{
  S\!\left(
    \dfrac{1}{n}
    \sum_{i=1}^n w(\tilde{\eta})_i
    z_{i,r}
    \left[
      v(\tilde{\eta})_i - \sum_{j \ne r} z_{i,j} \beta_j
    \right],
    \lambda \xi
  \right)
}{
  \frac{1}{n} \sum_{i=1}^n w(\tilde{\eta})_i z_{i,r}^2 + \lambda(1-\xi)
},
\end{equation}
where $S(a,\tau) = \mathrm{sign}(a)\max(|a|-\tau,0)$ is the soft-thresholding
operator, and $w(\tilde{\eta})$, $v(\tilde{\eta})$ are standard local
quadratic approximations of the Cox log-partial likelihood [@simon2011regularization].
We iterate coordinate updates until the subproblem converges to numerical
tolerance, yielding a minimizer in $\beta$ for the current $W$.


\subsection{Normalization of W}

After each accepted $W$ update, we normalize the columns of $W$ as
$W \leftarrow W D^{-1}$, and adjust $H$ and $\beta$ as
\[
H \leftarrow D H, \qquad \beta \leftarrow D \beta,
\]
where $D$ is the diagonal matrix of column $\ell_2$-norms of $W$.
This preserves the reconstruction $WH$ and the linear predictor $W\beta$; because
the accompanying rescaling of $H$ and $\beta$ can alter the penalization terms,
the normalized proposal is evaluated under the full penalized objective within
the backtracking procedure for the $W$ update rather than assumed loss-invariant. Normalization
prevents degeneracy in the scale-nonidentifiable factorization and keeps
columns of $W$ comparable in magnitude and interpretable as gene programs.


\section{Supervision on \(W\) versus \(H\)}

Classical nonnegative matrix factorization (NMF) is non-identifiable: for any
invertible matrix \(R\) with \(W R \ge 0\) and \(R^{-1} H \ge 0\), the factorization
\((W, H)\) can be transformed to \((W R, R^{-1} H)\) without changing the
reconstruction error, yielding multiple valid solutions
[@donoho2004nmf; @laurberg2008uniqueness; @gillis2014nmf]. Although normalizing
columns of \(W\) or rows of \(H\) resolves the trivial scaling ambiguity,
nonnegative mixing transformations persist, and empirical studies consistently
show that the sample-loading matrix \(H\) is more sensitive to initialization
and local minima than the program matrix \(W\)
[@brunet2004metagenes; @kim2007sparse; @gillis2012accelerated].

In DeSurv, survival supervision enters through the projection $Z = W^\top X$ in the partial log-likelihood, so that the gradient of the DeSurv loss with respect to the programs is
\[
\nabla \mathcal{L} = (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H) - \alpha\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta),
\]
and therefore the update rule for $W$ as in Equation \ref{eqn:W_update} relies on both the reconstruction term and the supervision term. As such, the gene-program definitions are directly shaped by their association with survival. This mechanism amplifies
genes aligned with the hazard gradient and suppresses those that dilute
prognostic structure, favoring programs enriched for survival-relevant variation.

By contrast, if supervision were applied to the sample loadings \(H\), the model
could reduce the Cox loss by redistributing patient-specific coefficients while
leaving the gene-level programs \(W\) nearly unchanged, a behavior documented in
multiple supervised variants of NMF and supervised topic models, where
supervision applied only to the coefficient matrix modifies \(H\) but not the
basis \(W\) [@cai2011graph; @blei2007slda].

Supervising \(W\) also provides a practical advantage for **portability**:
because \(W\) defines gene-level programs, new samples can be embedded by the
closed-form projection \(Z_{\text{new}} = W^\top X_{\text{new}}\). In contrast,
supervising \(H\) would not yield such a mapping; instead, obtaining sample
loadings for external datasets would require solving a nonnegative least-squares
(NNLS) problem for each new sample, introducing optimization noise and reducing
reproducibility.

Together, these considerations motivate supervising through \(W\), which shapes
the gene programs toward prognostic directions, stabilizes solutions across
initializations, and yields biologically interpretable signatures that
generalize across cohorts.


\section{Cross-validation procedure}
Algorithm \ref{alg:cv} describes the cross validation procedure with $F$ folds and $R$ initializations.
We define the c-index using comparable pairs
$\mathcal{P} = \{(i,j) : y_i < y_j,\ \delta_i = 1\}$ and linear predictors
$\hat{\eta}_i$:
\begin{equation}
\label{eqn:cindex}
\hat{c} =
\frac{1}{|\mathcal{P}|}
\sum_{(i,j)\in\mathcal{P}}
\left[
  \mathbbm{1}(\hat{\eta}_i > \hat{\eta}_j)
  + \tfrac{1}{2}\mathbbm{1}(\hat{\eta}_i = \hat{\eta}_j)
\right].
\end{equation}

\begin{algorithm}[H]
\caption{Cross-validation for DeSurv pipeline}
\label{alg:cv}
\begin{algorithmic}[1]
\Require
\Statex Number of folds $F$
\Statex Number of initializations $R$
\Statex Observed data ($X$, $y$, $\delta$)
\Statex Hyperparameters $k$, $\alpha$, $\lambda$, $\xi$, $\lambda_H$, and number of top genes per factor $n_{top}$
\Ensure The cross-validated c-index

\State Divide subjects into $F$ folds.
\For{$f = 1$ to $F$}
    \State Split data into training and validation $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(X_{(f)},y_{(f)}, \delta_{(f)})$ sets
    \For{$r = 1$ to $R$}
        \State set.seed$(r)$
        \State Apply Algorithm \ref{alg:desurv} with inputs $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(k, \alpha, \lambda, \xi, \lambda_H)$
        \State Obtain $\hat{W}$ and $\hat{\beta}$ as output from Algorithm \ref{alg:desurv}
        \State Restrict $\hat{W}$ to the top $n_{top}$ genes per factor (by factor-specificity score) and compute the estimated linear predictor on this support:
        $
        \hat{\eta} = X_{(f)}^T\hat{W}\hat{\beta}
        $
        \State Compute c-index, denoted $\hat{c}_{(f)r}$, according to Equation \ref{eqn:cindex}
    \EndFor
    \State \textbf{end loop over $r$}
    \State Compute average c-index across initializations:
    $
    \hat{c}_{(f)} = \frac{1}{R}\sum_{r=1}^R\hat{c}_{(f)r}
    $
\EndFor
\State \textbf{end loop over $f$}
\State Compute final cross-validated c-index:
$
\hat{c} = \frac{1}{F}\sum_{f=1}^F \hat{c}_{(f)}
$
\State \Return Final estimate $\hat{c}$

\end{algorithmic}
\end{algorithm}


\section{Bayesian Optimization}

We treat hyperparameter tuning as a black-box optimization problem over a
bounded domain. Let $\theta$ collect the tuned parameters (at minimum
$\theta = (k,\alpha,\lambda,\xi)$, and optionally discrete choices
such as the signature size $n_{\text{top}}$). For a given $\theta$ we run the
cross-validation procedure in Algorithm \ref{alg:cv}, averaging the c-index
across folds and random initializations to form the objective
\[
\hat{c}(\theta) = \frac{1}{F} \sum_{f=1}^F \hat{c}_{(f)}(\theta).
\]
Because $\hat{c}(\theta)$ is expensive to evaluate and non-differentiable with
respect to $\theta$, we use Bayesian optimization (BO) to adaptively select
candidate configurations.

We place a Gaussian-process (GP) prior on the response surface
$f(\theta) \sim \mathcal{GP}(0, k(\theta,\theta'))$ with a Mat\'ern kernel and
automatic relevance determination length-scales. Continuous parameters are
rescaled to $[0,1]$ and penalties are optimized on a log scale; integer-valued
coordinates (e.g., $k$ or $n_{\text{top}}$) are proposed in the continuous
space and rounded inside the evaluator. After an initial batch of evaluations,
the GP is refit after each new observation and used to guide acquisition.

At iteration $t$, let $\mu_t(\theta)$ and $\sigma_t(\theta)$ denote the GP
posterior mean and standard deviation, and let $f_\star$ be the best observed
score. We select new candidates by maximizing expected improvement (EI),
\[
\operatorname{EI}(\theta) =
\left(\mu_t(\theta) - f_\star - \kappa\right)\Phi(Z(\theta))
  + \sigma_t(\theta)\phi(Z(\theta)),
\quad
Z(\theta) = \frac{\mu_t(\theta) - f_\star - \kappa}{\sigma_t(\theta)},
\]
where $\Phi$ and $\phi$ are the standard normal cdf/pdf and $\kappa \ge 0$
controls the exploration margin. EI is evaluated over a candidate pool sampled
from the bounded search region, and the maximizer is evaluated next. The loop
continues until the iteration budget is exhausted or the improvement plateaus.

For the PDAC application, the objective was the mean 5-fold cross-validated
Harrell C-index (Algorithm \ref{alg:cv}), with folds stratified by cohort under
a fixed seed (123); each configuration was evaluated from 30 random
initializations and its C-index averaged across them. The bounded search domain
was $k \in [2, 12]$, $\alpha \in [0, 1]$, $\lambda \in [10^{-3}, 10^{3}]$
(optimized on a $\log_{10}$ scale), $\xi \in [0, 1]$, and
$n_{\text{top}} \in [50, 300]$, with the loadings ridge penalty fixed at
$\lambda_H = 0$. Bayesian optimization used an initial design of 50 evaluations
followed by 100 GP-guided iterations (150 total evaluations), maximizing
expected improvement over a candidate pool of $4{,}000$ points sampled from the
bounded domain at each step with an exploration margin $\kappa = 0.01$. The
unsupervised comparator fixed $\alpha = 0$ and tuned the remaining coordinates
identically.

For model selection, we take the highest observed $\hat{c}(\theta)$ as the
optimal configuration. When fold-level standard errors are available, we apply
a one-standard-error rule for $k$, selecting the smallest rank whose mean
performance lies within one standard error of the best. The final DeSurv model
is refit at the selected parameters using the consensus-based initialization
described above, and the BO-chosen $n_{\text{top}}$ (when tuned) is used to
truncate each program for downstream analysis.

The parameter $n_{\text{top}}$ controls which genes enter the survival linear predictor. After factorization, each sample's score for factor $j$ is computed as $\widetilde{W}_j^\top x$, where $\widetilde{W}_j$ retains only the $n_{\text{top}}$ genes with the largest factor-specificity scores $s_{ij}$ (Supplementary Information, Section \ref{sec:gene-program-correlation}); genes outside this set still shape the factorization but do not enter the linear predictor. BO selects $n_{\text{top}}$ jointly with $k$, $\alpha$, $\lambda$, and $\xi$ by optimizing cross-validated C-index. When $\alpha = 0$ (no survival supervision), gene selection and survival outcomes are decoupled.

Supplementary Fig. \ref{fig:fig-si-bo-tuning} illustrates the tuned response surface for the two most influential coordinates in the PDAC training cohort: the cross-validated C-index over factorization rank $k$ and supervision strength $\alpha$. Performance is maximized at intermediate supervision and moderate rank and degrades toward the unsupervised corner ($\alpha = 0$), consistent with a benefit of survival supervision; this is an empirical model-selection sensitivity surface, not a simulation result.

```{r fig-si-bo-tuning, fig.cap="Bayesian-optimization tuning surface for DeSurv in the PDAC training cohort (TCGA + CPTAC). Gaussian-process--predicted mean cross-validated C-index over factorization rank ($k$) and supervision strength ($\\alpha$; $\\alpha = 0$ corresponds to standard NMF). The response surface shows a broad region of high cross-validated performance at intermediate supervision and moderate rank, with discrimination degrading toward the unsupervised ($\\alpha = 0$) boundary. This is empirical model-selection tuning on the training data, not a simulation result.", echo=FALSE, fig.env='figure*', fig.pos='t', out.height= '3.6in', out.width='4.5in', fig.align='center'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_bo_tuning_surface_tcgacptac.pdf", density = 300))

```

\subsection{Consensus initialization for the final model}

The final DeSurv model at the selected hyperparameters was initialized from a consensus of multiple independent runs to reduce sensitivity to random starts. We first fit $R = 100$ DeSurv models to the training data from independent random nonnegative initializations at the selected $(k,\alpha,\lambda,\xi,n_{\text{top}})$ (seed 1--100). For each run, the top $n_{\text{top}}$ genes of every learned program were extracted by factor-specificity score, and a gene-by-gene consensus matrix was accumulated across runs, recording how often each pair of genes appeared together among the top genes of the same program. Genes appearing in the top sets of fewer than $\lceil 0.3R \rceil$ of the $R$ runs were treated as inactive and excluded from clustering. The consensus matrix over the remaining genes was partitioned into $k$ groups by agglomerative hierarchical clustering (average linkage), and each cluster defined one column of the initial gene-program matrix $W_0$; inactive genes and otherwise-empty entries were set to a small positive floor ($\sqrt{\epsilon_{\text{mach}}}$) to preserve strict positivity. The initial loadings $H_0$ were obtained by nonnegative least squares given $W_0$ and the training expression matrix, and the initial Cox coefficients $\beta_0$ were set to zero. This consensus seed used only training-cohort expression and was applied solely to initialize the single final refit after hyperparameter selection; it used no validation-cohort data and therefore introduced no information leakage.

\section{PDAC Datasets}

We analyzed PDAC cohorts from TCGA-PAAD and CPTAC-PDAC as training data
[@raphael2017integrated;@cao2021proteogenomic]. External validation used
independent cohorts from Dijk, Moffitt, PACA-AU (array and RNA-seq), and Puleo
[@dijk2020unsupervised;@moffitt2015virtual;@Bailey2016;@puleo2018stratification].
Training models were fit on the combined TCGA and CPTAC expression matrices
after harmonizing gene identifiers; validation datasets were processed with the
same gene set and transformations before projection, scoring and downstream validation analyses.

Dataset-specific inclusion and preprocessing steps were as follows. For
TCGA-PAAD, we excluded samples lacking tumor grade and mapped features to gene
symbols. For CPTAC, we retained samples annotated as PDAC by histology. For
Dijk, all 90 samples passed quality-control filters and were retained. For
Moffitt, only primary tumor specimens were kept. For PACA-AU array and RNA-seq,
we retained primary PDAC tumors and
collapsed duplicated gene symbols by median; the RNA-seq cohort received a
``\_seq'' suffix on sample IDs to avoid collisions with the array cohort. Puleo
required no additional dataset-specific filtering beyond survival QC.

Across cohorts, samples without valid survival outcomes (nonpositive time,
missing time, or missing event indicator) were removed. Expression matrices
were analyzed on a log scale: RNA-seq cohorts as log$_2$(TPM + 1); microarray cohorts in platform-normalized log-scale intensities. For each training cohort (TCGA-PAAD and CPTAC) separately, genes were ranked
by mean expression and by variance independently; the 3,000 genes with the
lowest sum of these two ranks were retained, selecting genes that are both
highly expressed and highly variable.
The intersection of these two gene sets yielded 1,970 genes used for model
training. Validation datasets were restricted to this same 1,970-gene set.
A within-sample rank transformation was then applied to every dataset to
mitigate platform and scale effects. Specifically, for each sample the
expression values across the 1{,}970 retained genes were replaced by their
ascending ranks (ties resolved by average ranks), so that every sample carries
the same nonnegative rank distribution over the shared gene set; ranks are
inherently nonnegative and therefore preserve NMF compatibility. Validation
cohorts were rank-transformed by exactly the same per-sample procedure before
projection, with no use of validation outcomes.

Table \ref{tab:cohorts} summarizes the cohorts, platforms, sample sizes after
quality control, and patient-level metadata used in this study. Overall survival
time and event indicators were available for all cohorts. PurIST
[@rashid2020purity] and DeCAF [@peng2024determination] molecular
classifications, applied to bulk expression data as described in those
references, were used as covariates in the covariate-adjusted external
validation analysis. Gene expression data and molecular
classifications for the validation cohorts were originally compiled for the
DeCAF study [@peng2024determination]; training data were obtained directly
from GDC.

```{r cohorts-events, echo=FALSE, message=FALSE, warning=FALSE}
.coh <- local({
  tr <- read_result("tar_data_filtered_tcgacptac")$sampInfo
  vl <- read_result("val_latent_desurv_tcgacptac")
  ev <- list(TCGA_PAAD = tr$event[tr$dataset == "TCGA_PAAD"],
             CPTAC     = tr$event[tr$dataset == "CPTAC"])
  for (x in vl) ev[[x$dataset]] <- x$survival$event
  ev$train_tot <- tr$event
  ev$val_tot   <- unlist(lapply(vl, function(x) x$survival$event))
  N    <- sapply(ev, length)
  E    <- sapply(ev, function(z) sum(z == 1, na.rm = TRUE))
  rate <- round(100 * E / N)
  cell <- sprintf("%d (%d\\%%)", E, rate); names(cell) <- names(ev)
  vr <- rate[c("Dijk","Moffitt_GEO_array","PACA_AU_array","PACA_AU_seq","Puleo_array")]
  # Combined unique-patient validation set (DeCAF-consistent PACA-AU de-duplication)
  pooled <- do.call(rbind, lapply(vl, function(x) data.frame(
    id = rownames(x$survival), dataset = x$dataset, event = x$survival$event,
    stringsAsFactors = FALSE)))
  pooled <- dedup_combined(pooled, "dataset", "id")
  Ncomb <- nrow(pooled); Ecomb <- sum(pooled$event == 1, na.rm = TRUE)
  list(N = N, cell = cell, comb_N = Ncomb,
       comb_cell = sprintf("%d (%d\\%%)", Ecomb, round(100 * Ecomb / Ncomb)),
       train_lo = min(rate[c("TCGA_PAAD","CPTAC")]), train_hi = max(rate[c("TCGA_PAAD","CPTAC")]),
       val_lo = min(vr), val_hi = max(vr), val_pooled = unname(rate["val_tot"]))
})
```

\begin{table}[ht]
\centering
\caption{Summary of PDAC cohorts used in this study. $N$ denotes the number of
samples after all quality-control filters (histology, survival validity,
dataset-specific exclusions). Patient-level metadata: OS = overall survival
(time and event indicator); PurIST = tumor subtype classification
(Rashid et al. 2020); DeCAF = CAF subtype classification
(Peng et al. 2026). Expression data and molecular classifications
for the validation cohorts were originally compiled for the DeCAF study
(Peng et al. 2026). Events denotes the number of overall-survival events (deaths) with the corresponding event rate; outcome frequency was lower and more homogeneous in the training cohorts (`r .coh$train_lo`--`r .coh$train_hi`\%) than across the validation cohorts (`r .coh$val_lo`--`r .coh$val_hi`\%, pooled `r .coh$val_pooled`\%), a case-mix difference relevant to the interpretation of external-validation performance.}
\label{tab:cohorts}
\footnotesize
\begin{tabular}{llrrllll}
\toprule
Cohort & Platform & $N$ & Events & Role & Accession & Metadata & Reference \\
\midrule
TCGA-PAAD        & RNA-seq    & `r .coh$N[["TCGA_PAAD"]]` & `r .coh$cell[["TCGA_PAAD"]]`  & Training   & GDC (TCGA-PAAD) & OS, PurIST, DeCAF & Raphael et al. (2017) \\
CPTAC            & RNA-seq    & `r .coh$N[["CPTAC"]]` & `r .coh$cell[["CPTAC"]]`  & Training   & GDC (CPTAC-3)    & OS, PurIST, DeCAF & Cao et al. (2021) \\
\addlinespace
Dijk             & RNA-seq    &  `r .coh$N[["Dijk"]]` & `r .coh$cell[["Dijk"]]`  & Validation & E-MTAB-6830      & OS, PurIST, DeCAF & Dijk et al. (2020) \\
Moffitt          & Microarray & `r .coh$N[["Moffitt_GEO_array"]]` & `r .coh$cell[["Moffitt_GEO_array"]]`  & Validation & GSE71729         & OS, PurIST, DeCAF & Moffitt et al. (2015) \\
PACA-AU (array)  & Microarray &  `r .coh$N[["PACA_AU_array"]]` & `r .coh$cell[["PACA_AU_array"]]`  & Validation & GSE36924         & OS, PurIST, DeCAF & Bailey et al. (2016) \\
PACA-AU (RNA-seq) & RNA-seq   &  `r .coh$N[["PACA_AU_seq"]]` & `r .coh$cell[["PACA_AU_seq"]]`  & Validation & EGAS00001000154  & OS, PurIST, DeCAF & Bailey et al. (2016) \\
Puleo            & Microarray & `r .coh$N[["Puleo_array"]]` & `r .coh$cell[["Puleo_array"]]` & Validation & E-MTAB-6134      & OS, PurIST, DeCAF & Puleo et al. (2018) \\
\addlinespace
\multicolumn{2}{l}{Training total} & `r .coh$N[["train_tot"]]` & `r .coh$cell[["train_tot"]]` & & & & \\
\multicolumn{2}{l}{Validation platform profiles total} & `r .coh$N[["val_tot"]]` & `r .coh$cell[["val_tot"]]` & & & & \\
\multicolumn{2}{l}{Combined unique-patient validation set} & `r .coh$comb_N` & `r .coh$comb_cell` & & & & \\
\bottomrule
\end{tabular}

{\footnotesize PACA-AU contains 46 patients profiled on both the array and RNA-seq platforms; combined (pooled) analyses represent each patient once, retaining the RNA-seq profile for duplicated patients, consistent with the DeCAF preprocessing rule. Platform-specific per-dataset analyses retain both records.}
\end{table}

\section{Survival Analysis}

Survival analyses used overall survival time with event indicators as provided
by each cohort. For fitted DeSurv models, sample-level program scores were
computed as $Z = \widetilde{W}^\top X$ and standardized using the training mean and
standard deviation before constructing the linear predictor
$\hat{\eta} = Z \hat{\beta}$. For the per-factor associations shown in the
validation forest plot, each individual factor score was standardized to zero
mean and unit standard deviation within its own cohort, so that per-factor
hazard ratios are expressed per one standard-deviation increment; the
multivariable linear predictor was likewise reported per standard deviation. Prognostic performance was summarized using the
concordance index as defined above. For visualization and group-level testing,
we generated Kaplan--Meier curves and log-rank tests for risk groups defined by
a single cutpoint selected by cross-validation in the TCGA and CPTAC training
cohort, applied unchanged to the pooled validation samples after standardizing
each linear predictor by the training mean and standard deviation. When multiple datasets were pooled, Cox models were stratified
by dataset to allow cohort-specific baseline hazards. P-values for Kaplan--Meier group comparisons correspond to two-sided log-rank tests; continuous-score associations were evaluated using Cox models stratified by dataset.

\section{Gene Program Correlation Analysis}
\label{sec:gene-program-correlation}

To assess the biological identity of learned factors, we evaluated the
correspondence between factor-specific gene rankings and established PDAC gene
programs. Each column $j$ of the gene program matrix $W$ was first scaled by its
column maximum, $W^*_{ij} = W_{ij} / \max_i W_{ij}$, placing all factors on a
common $[0,1]$ scale, and the specificity score for gene $i$ in factor $j$ was
defined as
$$s_{ij} = W^*_{ij} - \max_{j' \neq j} W^*_{ij'}.$$
Genes were ranked in descending order of $s_{ij}$. This ranking defines both the
top-$n_{\text{top}}$ genes used to score samples (Supplementary Data: factor-specific
gene lists) and, for the top 50 genes per factor, the loadings used to quantify
factor--program correspondence.

For each factor $j$ and each reference gene program $G_k$, we computed the
Spearman rank correlation between the continuous factor loadings $W_{\cdot j}$
(restricted to the union of the per-factor top-50 gene sets intersected with the
reference programs) and a binary indicator vector $\mathbf{1}[\text{gene} \in G_k]$
(main-text Fig.\ 2a,b; Methods). $P$ values were derived from the asymptotic
Spearman test and adjusted for multiple comparisons using the Benjamini--Hochberg
procedure within each factor column independently. Reference programs were
included in the visualization only if at least one factor yielded a Spearman
correlation exceeding $0.2$.

Reference gene programs displayed spanned classical and basal-like tumor
signatures and activated/immune stromal programs from Moffitt et al.\
[@moffitt2015virtual]; classical and exocrine subtypes from Collisson et al.\
[@collisson2011subtypes]; classical-tumor, basal-like-tumor, exocrine, and
activated-stroma compartment signatures from DECODER [@peng2019novo]; the pure
classical, immune classical, desmoplastic, and basal-like tumor
microenvironment programs from Puleo et al.\ [@puleo2018stratification]; the ECM
and immune programs from Maurer et al.\ [@Maurer2019]; the iCAF
cancer-associated fibroblast signature from Elyada et al.\ [@elyada2019cross];
the proCAF, restCAF, and panCAF fibroblast subtypes from SCISSORS
[@leary2023s]; the proCAF and restCAF programs from the DeCAF study
[@peng2024determination]; the classical A subtype from Chan-Seng-Yue et al.\
[@chansengyue2020transcription]; and the ADEX (aberrantly differentiated
endocrine exocrine) program from Bailey et al.\ [@Bailey2016]. Bailey's ADEX
program is retained because it is unique, whereas redundant Bailey programs,
periductal subtypes, and MSI subtypes were dropped as exact duplicates of, or
low-correlation relative to, other retained programs.

\section{Reconstruction and Survival Contribution per Factor}

NMF does not provide an orthogonal variance decomposition; it seeks a low-rank
reconstruction of the expression matrix $X \approx WH$. We therefore measure
each factor's contribution to reconstruction rather than to ``variance
explained.'' For a subset of factors $S \subseteq \{1, \ldots, k\}$, define the
reconstruction value
$$v(S) = \|X\|_F^2 - \Big\| X - \sum_{j \in S} W_{\cdot j} H_{j \cdot} \Big\|_F^2,$$
the reduction in residual error relative to the null (zero-factor) model, where
$W_{\cdot j} H_{j \cdot}$ is the rank-one outer-product contribution of factor $j$ (the outer product of column $j$ of $W$ with row $j$ of $H$). The
reconstruction Shapley value of factor $j$,
$$\phi_j = \sum_{S \subseteq \{1,\ldots,k\} \setminus \{j\}}
\frac{|S|!\,(k - |S| - 1)!}{k!}\,\big(v(S \cup \{j\}) - v(S)\big),$$
is its average marginal reconstruction gain over all orderings of the factors.
By the Shapley efficiency axiom, $\sum_j \phi_j = v(\{1,\ldots,k\})$, the total
reconstruction gain of the full model, so the normalized shares
$\phi_j / \sum_l \phi_l$ form an additive decomposition of reconstruction that
sums to one; individual shares may be signed in principle but were all positive
in the reported PDAC models. This is a genuine additive decomposition, in contrast to per-factor
variance fractions, which for non-orthogonal factorizations such as NMF do not
partition (cross-terms between rank-one contributions are non-zero). The Shapley
shares are computed exactly by enumerating the $2^k$ subsets (trivial at the
small ranks considered here).

The survival contribution of factor $j$ was quantified as the Type III partial
likelihood increment: the reduction in Cox partial log-likelihood when factor $j$
is omitted from the full $k$-factor model,
$$\Delta\ell_j = \ell_{\text{full}}(\hat{\beta}_{\text{full}}) - \ell_{-j}(\hat{\beta}_{-j}),$$
where $\ell_{\text{full}}(\hat{\beta}_{\text{full}})$ is the maximized unpenalized
Cox partial log-likelihood of the full $k$-factor model and
$\ell_{-j}(\hat{\beta}_{-j})$ is that of the reduced model refit on the remaining
$k - 1$ factors after omitting factor $j$, each evaluated at its own maximum
partial likelihood estimate. Larger $\Delta\ell_j$ indicates a greater
marginal contribution of factor $j$ to survival prediction given all other
factors. Unlike the reconstruction Shapley shares, these Type III increments are
conditional (each is evaluated against the full complement of the remaining
$k-1$ factors) and are nonadditive: they do not partition the total model partial
likelihood and are not comparable across factors as variance-style shares. We
therefore report them as per-factor conditional increments rather than as an
additive survival decomposition. These two quantities are plotted against each other in
main text, Fig.\ 2c. In PDAC at $k = 3$, the reconstruction shares are
$41.8\%/18.6\%/39.7\%$ for NMF (N1/N2/N3) and $18.8\%/25.2\%/56.0\%$ for DeSurv
(D1/D2/D3), while the survival contribution $\Delta\ell$ concentrates in DeSurv
D1 ($\Delta\ell = 66.4$) with all NMF factors negligible --- the factors that
contribute most to reconstruction are not the most prognostic.

\section{Factor Correspondence Analysis}

To assess which biological programs are preserved, altered, or suppressed
under survival supervision, we quantified the pairwise correspondence between
NMF and DeSurv factor loadings. For each pair of factors
$(j_{\text{NMF}},\, j_{\text{DeSurv}})$, we computed the Spearman rank
correlation between the corresponding columns of the NMF gene program matrix
$W_{\text{NMF}}$ and the DeSurv gene program matrix $W_{\text{DeSurv}}$ over all
genes. We use the full loading profile rather than a top-gene subset, which
would impose an arbitrary threshold and overstate apparent preservation.
Spearman correlation was used to capture monotone correspondence in gene loading
ranks while remaining robust to scale differences between the two methods.
Results are displayed as a heatmap with NMF factors as rows and DeSurv factors
as columns (main text, Fig.\ 2d).

\subsection{Cross-method reconstructability and reconstruction fidelity}

The correlation heatmap summarizes pairwise similarity; to make a direction-aware
statement of how much of each factor is \emph{linearly reconstructable} from the
other method's basis, we regressed each factor's gene-loading vector onto the full
set of the other method's loading vectors and report the ordinary least-squares
$R^2$ (Table \ref{tab:projection-r2}). A high $R^2$ means the factor is well
reconstructed as a linear combination of the other method's factors; a low $R^2$
means it is not linearly reconstructable from that basis. This is a statement
about linear reconstructability of gene-loading vectors, not about which method
discovered a program first.

```{r tab-projection-r2, echo=FALSE, results='asis'}
source("R/reconstruction_helpers.R")
.fit_d  <- read_result("tar_fit_desurv_tcgacptac")
.fit_n  <- read_result("fit_std_desurvk_tcgacptac")
.r2_N <- projection_r2(.fit_n$W, .fit_d$W)   # NMF factor -> DeSurv basis
.r2_D <- projection_r2(.fit_d$W, .fit_n$W)   # DeSurv factor -> NMF basis
.pr <- data.frame(
  Factor = c("NMF N1 (tumor)", "NMF N2 (exocrine)", "NMF N3 (microenviron.)",
             "DeSurv D1 (classical/restCAF)", "DeSurv D2 (proCAF)", "DeSurv D3 (basal-like)"),
  R2 = sprintf("%.2f", c(.r2_N, .r2_D)),
  Interpretation = c("well reconstructed", "poorly reconstructed", "well reconstructed",
                     "not linearly reconstructable", "reconstructable ($\\approx$N3)", "reconstructable ($\\approx$N1)"),
  stringsAsFactors = FALSE)
colnames(.pr) <- c("Factor", "$R^2$ from other basis", "Linear reconstructability")
knitr::kable(.pr, format = "latex", escape = FALSE, booktabs = TRUE,
  caption = sprintf("Cross-method projection $R^2$: how reconstructable each factor's gene-loading vector is from the other method's full basis (OLS $R^2$). DeSurv retains the bulk of the unsupervised structure (NMF N1 and N3 carry over at $R^2 \\ge %.2f$) and reorganizes it: the survival-aligned D1 program is essentially unreconstructable from the NMF basis ($R^2 = %.2f$), i.e. not well reconstructed as a linear combination of the NMF factors, while the exocrine NMF factor N2 is poorly reconstructed from the DeSurv basis ($R^2 = %.2f$). \\label{tab:projection-r2}", min(.r2_N[1], .r2_N[3]), .r2_D[1], .r2_N[2])) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

The headline is DeSurv D1: $R^2 = `r sprintf("%.2f", .r2_D[1])`$ from the NMF basis, indicating that
its gene-loading vector is not well reconstructed as a linear combination of the
matched-rank NMF factors. The biologically dominant tumor (N1) and stromal (N3)
programs, by contrast, are largely reconstructable from the DeSurv basis
($R^2 = `r sprintf("%.2f", .r2_N[1])`$ and $`r sprintf("%.2f", .r2_N[3])`$), so most of the unsupervised structure is
retained; the principal reorganization is that the DeSurv basis resolves tumor
signal into basal- and classical-aligned programs while reconstructing the
exocrine factor (N2) less well ($R^2 = `r sprintf("%.2f", .r2_N[2])`$).

At the whole-matrix level, this reorganization costs little reconstruction fidelity (Table \ref{tab:overall-r2}).

```{r tab-overall-r2, echo=FALSE, results='asis'}
.X <- as.matrix(read_result("tar_data_filtered_tcgacptac")$ex)
.or <- data.frame(
  Model = c("NMF ($k = 3$)", "DeSurv ($k = 3$)"),
  R2 = sprintf("%.3f", c(overall_recon_r2(.fit_n$W, .fit_n$H, .X),
                         overall_recon_r2(.fit_d$W, .fit_d$H, .X))),
  Resid = sprintf("%.3f", c(1 - overall_recon_r2(.fit_n$W, .fit_n$H, .X),
                            1 - overall_recon_r2(.fit_d$W, .fit_d$H, .X))),
  stringsAsFactors = FALSE)
colnames(.or) <- c("Model", "Reconstruction $R^2$", "Residual fraction")
knitr::kable(.or, format = "latex", escape = FALSE, booktabs = TRUE,
  caption = "Whole-matrix reconstruction $R^2$ for unsupervised NMF versus DeSurv at $k = 3$. DeSurv sacrifices only $\\approx 1.5$ percentage points of reconstruction $R^2$ to gain its prognostic structure. \\label{tab:overall-r2}") |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

DeSurv gives up only about $1.5$ percentage points of reconstruction $R^2$
relative to unsupervised NMF (Table \ref{tab:overall-r2}) --- very little
reconstruction fidelity is sacrificed in exchange for the prognostic
reorganization documented above.

\subsection{The D1 axis is recovered by survival supervision, not by DeSurv specifically}

```{r suprec-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("supervised_recovery_stats")
.sr <- supervised_recovery_stats
.srt <- .sr$table
.srcor <- function(m) abs(.srt$score_cor_D1[.srt$method == m])
```

If the D1 tumor--stroma axis were an artifact of DeSurv's particular
optimizer, independent survival-supervised methods should not recover it. We
therefore fit three other supervised dimension-reduction methods on the same
training cohort ($n = `r .sr$meta$n`$, `r .sr$meta$n_events` events) --- supervised
principal components (Cox-screened genes followed by PCA), one-component Cox
partial least squares against martingale residuals, and a sparse Cox model
(lasso-penalized, `r .sr$meta$n_sparse_cox_genes` selected genes) --- and compared
their leading survival-aligned direction to the DeSurv D1 score. All three
recovered D1 closely (per-patient $|r| = `r sprintf("%.2f", .srcor("Supervised PCA"))`$
for supervised PCA, $`r sprintf("%.2f", .srcor("Cox-PLS"))`$ for Cox-PLS, and
$`r sprintf("%.2f", .srcor("Sparse Cox"))`$ for sparse Cox; mean
$`r sprintf("%.2f", .sr$sup_mean_score_cor)`$), and each aligned specifically with
D1 rather than the stromal D2 or basal-like D3 programs
(Fig. \ref{fig:suprec}). By contrast, the leading \emph{unsupervised} direction
(the first principal component, a variance-dominant unsupervised direction) was nearly orthogonal to D1
($|r| = `r sprintf("%.2f", .srcor("Unsupervised PCA"))`$) and instead tracked the
high-variance D2/D3 stroma--basal axis. This D1 axis is thus a property of
survival supervision shared across methods --- consistent with its being
unreconstructable from the unsupervised NMF basis (Table \ref{tab:projection-r2})
and with standard NMF's failure to enrich for the proCAF stroma at matched rank
--- and not an idiosyncrasy of the DeSurv algorithm.

```{r fig-suprec, echo=FALSE, fig.width = 8, fig.height = 3.6, fig.cap = "Independent survival-supervised methods recover the D1 axis; an unsupervised method does not. (A) Absolute correlation (across training-cohort patients) of each method's leading direction with the DeSurv D1 score, in the training cohort: supervised principal components, Cox partial least squares, and sparse Cox all recover D1 (green), whereas the leading unsupervised principal component does not (grey). (B) Sample-level agreement with D1 for Cox-PLS (supervised, green) versus the first unsupervised principal component (grey). \\label{fig:suprec}", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
knitr::include_graphics(file.path("..", "figures", "fig_supervised_recovery.pdf"), error = FALSE)
```

```{r suptuned-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_tuned")
.stv    <- desurv_vs_supervised_tuned
.stvp   <- function(m) sprintf("%.2f", .stv$prediction[m, "POOLED"])
.stva   <- function(m, f) sprintf("%.2f", abs(.stv$axis_decomposition[m, f]))
.stvhit <- function(m) unname(.stv$supervised_attribution_across_seeds[[m]]["BasalTumor"])
.stvg   <- function(m, q) unname(.stv$genes_per_seed[[m]][q])
.stvc   <- function(m) sprintf("%.2f", .stv$stromal_carriage[m])
.stvns  <- .stv$meta$nseed
```

\subsection{Robustness to tuned supervised comparators and compartment attribution}

The recovery above establishes the D1 axis as a genuine survival direction shared across methods; here we ask what DeSurv adds beyond recovering it, using canonical, independently tuned comparators in place of the simple single-component versions above: supervised principal components (\texttt{superpc}; Bair--Tibshirani, with cross-validated screening threshold and component number) and penalized (lasso) Cox (\texttt{glmnet}, a single $\lambda_{\min}$ model). Cox partial least squares was not re-run in this pass, as a validated multi-component implementation was unavailable in our environment; the two canonical methods are sufficient. Every method began from the same 1{,}970-gene training-only expression/variance universe and performed its own feature selection; DeSurv was scored with its Bayesian-optimization-selected top-270-per-factor programs; all scoring used the frozen, rank-transformed transfer with no validation refitting.

Prediction parity was preserved: the pooled frozen-transfer C-index was $`r .stvp("DeSurv full LP")`$ for the DeSurv linear predictor, $`r .stvp("Supervised PCA")`$ for supervised PCA, and $`r .stvp("Sparse Cox")`$ for penalized Cox. Consistent with the score-level recovery above, both tuned methods aligned with the D1 direction (training $|r| = `r .stva("Sparse Cox","D1")`$ for penalized Cox and $`r .stva("Supervised PCA","D1")`$ for supervised PCA) and with the DeSurv linear predictor ($|r| = `r .stva("Sparse Cox","fullLP")`$ for penalized Cox and $`r .stva("Supervised PCA","fullLP")`$ for supervised PCA), but were essentially orthogonal to the activated-stroma/proCAF program D2 ($|r| = `r .stva("Sparse Cox","D2")`$ for penalized Cox and $`r .stva("Supervised PCA","D2")`$ for supervised PCA): a single supervised risk axis captures the dominant tumour-axis signal but not the separate stromal program.

The compartment attribution made this explicit. Because D1 runs from a classical-tumour, restraining-stroma protective pole to a basal-like high-risk pole, a sparse supervised risk score aligned with D1 is naturally composed of basal-like tumour markers. Accordingly, across `r .stvns` cross-validation seeds the genes selected by both methods were reproducibly enriched (hypergeometric $P < 0.01$) for the basal-like tumour program (in `r .stvhit("Sparse Cox")`/`r .stvns` seeds for penalized Cox and `r .stvhit("Supervised PCA")`/`r .stvns` for supervised PCA) and in no seed for the activated-stroma/proCAF, restraining-stroma, immune, or exocrine compartments. The supervised risk scores did nonetheless retain some stromal signal: each correlated modestly with an external activated-stroma/proCAF signature ($|r| = `r .stvc("Sparse Cox")`$ for penalized Cox and $`r .stvc("Supervised PCA")`$ for supervised PCA), but well below DeSurv's D2 program ($|r| = `r .stvc("DeSurv D2 (proCAF factor)")`$), which tracks that signature directly. The supervised risk scores retained modest stromal signal but did not resolve it as a distinct program. DeSurv instead placed the dominant survival-associated direction within a three-program representation containing separately measurable classical/restCAF-like, proCAF-like and basal-like programs. Thus, the distinction is representational: DeSurv preserves stromal context as an explicit program even when that program does not contribute independently to the fitted linear predictor.

```{r brier-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_brier_stats")
.br <- desurv_brier_stats; .brp <- .br$pooled; .bripa <- range(.br$per_cohort$ipa)
```

\subsection*{Proper-scoring sensitivity of the external validation}

The main-text external validation reports discrimination with Harrell's C-index
under a Cox proportional-hazards model. Because Harrell's C can be optimistic when
hazards are non-proportional [@birolo2025beyondcox; @hao2021dnnsurv], we additionally
evaluated the DeSurv linear predictor---its direction frozen from training---with
a proper score, the integrated Brier score (IBS), in each validation cohort,
recalibrating only its scalar slope and baseline hazard
within each cohort and scoring over that cohort's follow-up. For a single
time-invariant risk score under proportional hazards the monotone score-to-survival
map preserves the concordance ordering, so Antolini's time-dependent concordance
coincides with Harrell's C; the additive quantity is therefore the proper score, not
a second concordance. Across the `r nrow(.br$per_cohort)` cohorts the DeSurv predictor achieved an
event-weighted integrated Brier score of `r sprintf("%.3f", .brp["ibs"])` versus
`r sprintf("%.3f", .brp["ibs_null"])` for a Kaplan--Meier reference, an index of
prediction accuracy (IPA $= 1 - \mathrm{IBS}/\mathrm{IBS}_{\text{null}}$) of
`r sprintf("%.1f", 100*.brp["ipa"])`\% (per-cohort range
`r sprintf("%.1f", 100*.bripa[1])`--`r sprintf("%.1f", 100*.bripa[2])`\%). Because each
cohort's IBS is integrated over its own event-time horizon, this pooled value is a
descriptive event-weighted summary rather than a single common-horizon proper score,
and the per-cohort IPA range is the more directly interpretable quantity. This
modest proper-score improvement over the null is consistent with the moderate
discrimination (C $= `r sprintf("%.2f", .brp["cindex"])`$) and with DeSurv's
contribution being a separately resolved tumour- and stroma-associated decomposition
rather than higher predictive accuracy.

\subsection*{Proportional-hazards diagnostic}

```{r si-ph-diagnostic, echo=FALSE, message=FALSE, warning=FALSE}
.phdf <- do.call(rbind, lapply(val_latent_desurv_tcgacptac, function(x)
  data.frame(risk_score = x$risk_score, time = x$survival$time,
             event = x$survival$event, dataset = x$dataset)))
.phdf$risk_z <- .phdf$risk_score / sd(.phdf$risk_score)
.phz <- survival::cox.zph(
  survival::coxph(survival::Surv(time, event) ~ risk_z + strata(dataset), data = .phdf))
.ph_chisq <- .phz$table["GLOBAL", "chisq"]
.ph_df    <- .phz$table["GLOBAL", "df"]
.ph_p     <- .phz$table["GLOBAL", "p"]
```

Because the pooled hazard ratio assumes proportional hazards, we tested the scaled Schoenfeld residuals of the stratified validation model ($\mathrm{Surv}(t,\delta) \sim \text{risk\_z} + \mathrm{strata}(\text{dataset})$) with \texttt{survival::cox.zph}. The proportional-hazards assumption was not supported for the standardized DeSurv linear predictor ($\chi^2 = `r sprintf("%.1f", .ph_chisq)`$, $\mathrm{df} = `r .ph_df`$, $P = `r sprintf("%.1e", .ph_p)`$), with the score's effect attenuating over follow-up rather than reversing. The reported pooled hazard ratio should therefore be read as a time-averaged summary of a consistently prognostic score rather than a constant multiplicative effect; this time-varying behavior is precisely why we additionally report the horizon-integrated proper-scoring (IBS/IPA) sensitivity above, which integrates prediction accuracy over each cohort's event-time horizon (to the 80th percentile of event times) rather than through a single hazard-ratio summary.

\subsection*{Independence of discovery and validation}

To make the separation between discovery, adjustment, and validation unambiguous: DeSurv
is trained with survival supervision only; it never uses PurIST, DeCAF, or any molecular
subtype label during factorization or hyperparameter selection. The PurIST and DeCAF
subtype calls were fixed before the DeSurv analysis (loaded as precomputed per-cohort
annotations) and enter only as covariates in the adjusted Cox models, not as inputs to
DeSurv. All rank and hyperparameter selection (including the one-standard-error rule) was
performed by cross-validation entirely within the TCGA and CPTAC training data;
no external validation cohort was used for training, cutpoint selection, or tuning.
Validation-cohort expression enters only at scoring, a fixed-basis projection
($Z = \widetilde{W}^\top X_{\text{new}}$) that uses no validation outcome, and
validation-cohort survival outcomes were used only for the final, out-of-sample
performance evaluation.

\subsection*{Adjustment for clinical covariates and tumor purity}

```{r clinical-adj, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
load_result("desurv_clinical_adjustment")
.ca  <- desurv_clinical_adjustment
.caf <- function(m) if (is.null(m)) "---" else sprintf("%.2f (%.2f--%.2f), %s ($n$=%d, %d ev)", m$hr, m$lo, m$hi,
  if (m$p < 0.001) "$<0.001$" else paste0("$", sprintf("%.3f", m$p), "$"), m$n, m$events)
.ord  <- c("PACA_AU_array","PACA_AU_seq","Puleo_array","Dijk","Moffitt_GEO_array","TCGA_PAAD","CPTAC")
.disp <- c(PACA_AU_array="PACA-AU array", PACA_AU_seq="PACA-AU RNA-seq", Puleo_array="Puleo",
           Dijk="Dijk", Moffitt_GEO_array="Moffitt", TCGA_PAAD="TCGA (train)", CPTAC="CPTAC (train)")
.carows <- lapply(.ord, function(k){ r <- .ca[[k]]
  covs <- if (!is.null(r$clinical_adjusted)) paste(r$clinical_adjusted$covars, collapse=", ") else "---"
  data.frame(Cohort=.disp[[k]], Covariates=covs,
             Unadjusted=.caf(r$unadjusted), `Clinical-adjusted`=.caf(r$clinical_adjusted),
             `Purity-adjusted`=.caf(r$purity_adjusted), check.names=FALSE, stringsAsFactors=FALSE) })
.catab <- do.call(rbind, .carows)
knitr::kable(.catab, format="latex", escape=FALSE, booktabs=TRUE,
  caption="Adjusted hazard ratios (per SD of the DeSurv linear predictor) for overall survival, before and after adjustment for standard clinical covariates and, where annotated, tumor purity, fitted separately in each cohort by complete-case analysis (per-cell $n$ and events shown). The Covariates column lists the clinical variables entered (stage, grade, age, sex, nodal status, or resection margin, as available). The DeSurv score remains independently prognostic after clinical adjustment in the fully-annotated external PACA-AU cohorts and both training cohorts; where tumor-purity annotation was available, it also remained prognostic after purity adjustment in the adequately-powered cohorts (both PACA-AU cohorts and TCGA), consistent with the association not being driven solely by purity or composition. The smaller Dijk and Moffitt cohorts were individually underpowered (non-significant even unadjusted; the Moffitt purity-adjusted model, $n = 29$, likewise). Training-cohort (TCGA, CPTAC) hazard ratios are in-sample. \\label{tab:clinical-adjustment}") |>
  kableExtra::kable_styling(latex_options=c("HOLD_position","scale_down"), font_size=8)
```

\clearpage

\newpage

\part{Supplementary Results}

<!-- Supplementary numbering convention: all SI figures labeled "Fig. S1", "Fig. S2", etc.
     (not "Figure 1"). Tables: "Table S1", etc. Internal scenario code names
     (R0_easy, R00_null, R_mixed) are replaced with descriptive labels (prognostic,
     null, mixed) consistent with main-text terminology. -->
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo    = FALSE,
                      fig.path = "figures/",
                      dev     = "png",
                      dpi     = 300)
source("paper/load_precomputed.R")
library(dplyr)
library(ggplot2)
library(cowplot)
if (!exists("set_fig_font", inherits = TRUE)) {
  set_fig_font <- function(plot_obj, size = 10) {
    if (inherits(plot_obj, "ggplot")) {
      plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
    } else {
      plot_obj
    }
  }
}
```


# Empirical optimization stability across random initializations

```{r converge-diag, echo=FALSE}
.sf    <- read_result("desurv_seed_fits_tcgacptac")
.fits  <- .sf$fits
.R     <- length(.fits)
.h     <- .fits[[1]]$hyper
.tol   <- .h$tol; .maxit <- .h$maxit
.niter <- vapply(.fits, function(f) length(f$lossit), numeric(1))
.conv  <- vapply(.fits, function(f) isTRUE(f$convergence), logical(1))
.nan   <- vapply(.fits, function(f) isTRUE(f$nan_flag), logical(1))
.mono  <- vapply(.fits, function(f){ d <- diff(f$lossit);
            all(d <= 1e-8 * abs(head(f$lossit, -1)) + 1e-12) }, logical(1))
.n_conv  <- sum(.conv)
.n_maxit <- sum(.niter >= .maxit)
.tol_tex <- sprintf("10^{%d}", as.integer(round(log10(.tol))))
```

The idealized alternating scheme is not accompanied by a formal convergence guarantee for the fitted models, both because the reported analyses set the loadings ridge penalty to zero ($\lambda_H = 0$) and because the implemented algorithm combines column normalization, gradient balancing between the reconstruction and Cox terms, and a Coxnet approximation of the $\beta$ update. We therefore evaluated optimization stability empirically across $R = `r .R`$ independent random initializations of the selected PDAC model. The stopping tolerance was $`r .tol_tex`$ with a maximum of `r format(.maxit, big.mark=",")` iterations. Of the $`r .R`$ runs, $`r .n_conv`$ terminated after satisfying the stopping criterion and $`r .n_maxit`$ reached the iteration limit; the median number of iterations was $`r median(.niter)`$ (range, $`r min(.niter)`$--$`r max(.niter)`$). `r if (sum(.nan) == 0) "No numerical failures were observed." else sprintf("Numerical failures were observed in %d runs.", sum(.nan))` `r if (all(.mono)) "In every run the recorded full penalized objective decreased monotonically to a numerically stable value." else sprintf("The recorded full penalized objective decreased monotonically in %d of the %d runs.", sum(.mono), .R)` Supplementary Fig.\ \ref{fig:fig-converge} shows five runs selected at random (seed 147). These analyses document empirical optimization stability for the reported application rather than providing a formal convergence guarantee.


```{r fig-converge, fig.width = 5, fig.height= 3, fig.cap="Empirical optimization-stability diagnostic for the DeSurv objective across five randomly selected multi-start initializations on the PDAC training cohort (TCGA + CPTAC). Each line shows the relative decrease in the objective from its initial value, $(\\ell_0 - \\ell_t)/\\ell_0$ (in units of $10^{-5}$; the reconstruction-dominated total objective changes by a small relative amount), as a function of iteration number. In these runs the recorded full objective decreases monotonically and stabilizes within the iteration budget. This figure is an empirical diagnostic of optimization behavior for the reported application rather than a formal convergence proof.", fig.env='figure*', fig.pos='t', out.height= "3in", out.width='5in', fig.align='center'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_converge_tcgacptac.pdf", density = 300))

```


# Simulation results under null and mixed scenarios

To characterize the operating characteristics of DeSurv across different signal regimes, we evaluated performance under three simulation scenarios. In the prognostic scenario (main text Fig. \ref{fig:sim}), prognostic programs explained low variance relative to outcome-neutral background signals, and survival depended on 150 factor-specific marker genes ($\beta_1 = 2$, $\beta_2 = \beta_3 = 0$). In the null scenario, survival times were generated independently of gene expression ($\beta = 0$), so no prognostic structure existed. In the mixed scenario, survival depended on 300 genes per lethal factor, of which 50% were factor-specific markers and 50% were background genes with loadings shared across all factors. All scenarios used $G = 3{,}000$ genes, $N = 200$ samples, $K = 3$ true factors, and identical gamma distribution parameters for the gene loading matrix $W$ (see Materials and Methods). Performance was evaluated using cross-validated concordance index (C-index), precision of recovered prognostic gene sets, and the distribution of selected ranks across simulation replicates.

```{r si-null-alpha, echo=FALSE}
load_result("sim_figs_by_scenario")
sids <- sapply(sim_figs_by_scenario, `[[`, "scenario_id")
aids <- sapply(sim_figs_by_scenario, `[[`, "analysis_id")
null_i <- which(sids == "R00_null" & aids == "bo_tune_ntop")
null_alpha <- with(sim_figs_by_scenario[[null_i]]$k_hist$data, alpha[method == "DeSurv"])
null_alpha_n <- length(null_alpha); null_alpha_median <- median(null_alpha)
null_alpha_q1 <- quantile(null_alpha, .25); null_alpha_q3 <- quantile(null_alpha, .75)
null_alpha_min <- min(null_alpha); null_alpha_max <- max(null_alpha)
```

Under the null scenario (Fig. \ref{fig:sim-null-mixed}A--B), DeSurv showed no advantage over unsupervised NMF ($\alpha = 0$): both methods yielded C-index values near 0.5, and selected ranks concentrated at low values (NMF favoring $k = 2$), consistent with the absence of survival signal. DeSurv therefore did not recover spurious prognostic structure when none existed. Because the survival objective is uninformative under the null, the cross-validated concordance is essentially flat in $\alpha$ and the Bayesian-optimized supervision strength is unidentified, dispersing across its admissible range rather than concentrating near zero (median $\alpha = `r round(null_alpha_median, 2)`$, IQR $`r round(null_alpha_q1, 2)`$--$`r round(null_alpha_q3, 2)`$; range $`r round(null_alpha_min, 2)`$--$`r round(null_alpha_max, 2)`$ across `r null_alpha_n` replicates). The null-scenario specificity thus reflects the flat concordance surface itself, not the optimizer driving $\alpha$ to zero.

Under the mixed scenario (Fig. \ref{fig:sim-null-mixed}C--F), DeSurv's advantage was concentrated in marker-gene recovery. Precision against the factor-specific marker genes alone (150 genes; Fig. \ref{fig:sim-null-mixed}D left) substantially favored DeSurv, whereas precision against the full 300-gene survival set including background genes (Fig. \ref{fig:sim-null-mixed}D right) was more modestly improved, indicating similar background-gene recovery between methods. The learned factor aligned with the true marker program also received a larger Cox coefficient ($|\beta|$) under DeSurv (Fig. \ref{fig:sim-null-mixed}E), and C-index was modestly higher (Fig. \ref{fig:sim-null-mixed}C). Across the three scenarios the pattern was graded: DeSurv's advantage was greatest when variance and prognosis diverged (prognostic), attenuated when they partially overlapped (mixed, as background-gene dilution reduced the discrimination gain), and absent when no survival signal existed (null). DeSurv is therefore most useful when the dominant transcriptional axes are expected to be prognostically neutral, as in cancers with high tumor-purity variation or dominant tissue-composition signals.

```{r fig-sim-null-mixed, fig.width = 6.5, fig.height = 7, fig.cap = "DeSurv performance under null and mixed simulation scenarios. (a--b) Null scenario ($\\beta = 0$): both methods yield C-index $\\approx 0.5$ (a) and comparable rank distributions (b), confirming DeSurv does not recover spurious prognostic structure; precision is omitted as no true survival genes exist. (c--f) Mixed scenario: survival depends on factor-specific marker genes (50\\%) and shared background genes (50\\%). (c) C-index is modestly higher for DeSurv. (d) Precision by gene-set definition: marker-only precision (150 genes; left facet) substantially favors DeSurv; full 300-gene survival-set precision (right facet) shows a more modest improvement. (e) The matched factor receives a substantially larger Cox coefficient $|\\beta|$ under DeSurv, confirming preferential amplification of marker-specific signal. (f) Selected-rank distributions. DeSurv (blue); NMF (red, $\\alpha = 0$); both Bayesian-optimized. \\label{fig:sim-null-mixed}", fig.env='figure*', fig.pos='t', out.height = "6.5in", out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_sim_null_mixed_tcgacptac.pdf", density = 300))
```


# Cross-validated C-index across factorization rank

To evaluate how the number of latent factors $k$ affects prognostic performance, as a sensitivity analysis we performed an exhaustive grid search over $k \in \{2, 3, \ldots, 12\}$ and supervision strength $\alpha \in \{0, 0.05, \ldots, 0.95\}$ using 5-fold cross-validation on the TCGA+CPTAC training cohort. For each $k$, the best-performing $\alpha$ was selected by maximizing the mean CV C-index. The unsupervised NMF baseline ($\alpha = 0$) was evaluated at the same $k$ values for comparison. Results are shown for two gene-selection settings: $n_{\text{top}} = 270$ top genes per factor and $n_{\text{top}} = \text{ALL}$ (all genes retained), both at $\lambda = 0.349$ and $\xi = 0.056$. External validation C-index is shown descriptively; validation outcomes were not used to select $k$, $\alpha$, $\lambda$, $\xi$, or $n_{\text{top}}$.

Fig. \ref{fig:fig-cindex-by-k} shows the resulting C-index curves for both training (CV) and external validation (pooled across held-out cohorts), with shaded ribbons indicating $\pm 1$ standard error. Rank was selected on the training cross-validated concordance: for DeSurv this increased from $k = 2$ and then plateaued from $k = 3$ onward, and the one-standard-error rule accordingly selected the parsimonious $k = 3$, while DeSurv exceeded unsupervised NMF at matched rank throughout. The external-validation curve is shown only descriptively (validation outcomes were used for no part of rank or hyperparameter selection) and is consistent with this training-based choice, with DeSurv at $k = 3$ attaining concordance comparable to standard NMF at higher ranks (e.g., $k = 7$; Table \ref{tab:tab-cindex-comparison}). We therefore read this as a training/cross-validation sensitivity analysis of rank, not as an external model-selection argument.

```{r fig-cindex-by-k, fig.width = 10, fig.height = 10, fig.cap="Training (CV) and external validation C-index as a function of factorization rank $k$ for DeSurv (blue) and standard NMF (red, $\\alpha = 0$). Shaded ribbons show $\\pm 1$ SE. For each $k$, the best $\\alpha$ was selected by maximizing the mean CV C-index. Validation C-index is pooled across all external cohorts. (a) Results using $n_{\\text{top}} = 270$ top genes per factor ($\\lambda = 0.349$, $\\xi = 0.056$). (b) Results retaining all genes ($n_{\\text{top}} = \\text{ALL}$; $\\lambda = 0.349$, $\\xi = 0.056$).", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
# This figure is produced by the cv_grid pipeline (separate store);
# read directly from the file output.
# Use magick to render the PDF as a raster (avoids LaTeX path issues)
# NOTE: DL flagged duplicate legend across panels A and B. The legend is baked
# into the pre-rendered PDF; consolidating it requires regenerating the source
# figure in inst/cv_grid_validation_analysis_amber_optimal.R on HPC.
img <- magick::image_read("figures/cv_cindex_by_k_primary.pdf", density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

# Standard NMF rank selection heuristics

Standard unsupervised rank selection criteria (reconstruction residuals, cophenetic correlation coefficient, and mean silhouette width) yielded inconsistent guidance for selecting the factorization rank $k$ in the PDAC training cohort (TCGA + CPTAC). Reconstruction residuals decreased smoothly without a clear elbow (Fig. \ref{fig:nmf-diagnostics}A). Cophenetic correlation fluctuated without a distinct transition point (Fig. \ref{fig:nmf-diagnostics}B). Mean silhouette width favored very low ranks that conflicted with other criteria (Fig. \ref{fig:nmf-diagnostics}C). These contradictions motivated the survival-supervised rank selection approach described in the main text.

```{r fig-nmf-diagnostics, fig.width = 6, fig.height = 3.5, fig.cap = "Standard unsupervised NMF rank selection heuristics for the PDAC training cohort (TCGA + CPTAC). (a) Reconstruction residuals as a function of factorization rank $k$. (b) Cophenetic correlation coefficient as a function of $k$. (c) Mean silhouette width across multiple distance metrics as a function of $k$. These criteria yield inconsistent guidance, motivating the survival-supervised model-selection approach described in the main text. \\label{fig:nmf-diagnostics}", fig.env='figure*', fig.pos='t', out.height= "3.5in", out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_diagnostics_tcgacptac.pdf", density = 300))
```

# Cutpoint selection and validation of dichotomized risk groups

To convert the continuous DeSurv linear predictor into a clinically interpretable binary risk stratification, we selected an optimal z-score cutpoint via cross-validation on the TCGA+CPTAC training cohort. For each candidate cutpoint in $\{-2.0, -1.8, \ldots, 2.0\}$, we computed the mean absolute log-rank z-statistic across held-out folds and selected the cutpoint maximizing this metric (Fig. \ref{fig:fig-cutpoint-km}a). We then applied this cutpoint to external validation cohorts by standardizing each patient's linear predictor using the training mean and standard deviation. Fig. \ref{fig:fig-cutpoint-km}b shows the resulting Kaplan--Meier curves for the pooled external validation samples. To respect between-cohort differences in baseline hazard, the reported inference is cohort-stratified: the hazard ratio is from a Cox model with the risk group as covariate and dataset as a stratifying factor ($\mathrm{Surv}(t,\delta) \sim \text{group} + \mathrm{strata}(\text{dataset})$), and the accompanying log-rank test is stratified by dataset. The Kaplan--Meier curves themselves are shown pooled for visualization. Panels C--F display per-cohort validation KM curves for Dijk, Moffitt, PACA-AU (de-duplicated across array and RNA-seq platforms, retaining RNA-seq for dual-platform patients), and Puleo, respectively. Across cohorts, the DeSurv-derived high-risk group showed directionally shorter survival, indicating that the training-derived cutpoint generalizes to independent datasets.

```{r fig-cutpoint-km, fig.width = 10, fig.height = 11, fig.cap = "Cutpoint selection and external validation of dichotomized risk groups. (a) Cross-validated mean absolute log-rank z-statistic as a function of z-score cutpoint; red dashed line indicates the selected optimum. (b) Pooled validation Kaplan--Meier curves (log-rank test). (c--d) Per-cohort validation KM curves: (c) Dijk, (d) Moffitt. (e--f) Per-cohort validation KM curves: (e) PACA-AU (de-duplicated across array and RNA-seq platforms), (f) Puleo.", fig.env='figure*', fig.pos='t', out.height = "8in", out.width='\\textwidth', echo=FALSE}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_cutpoint_km_tcgacptac.pdf", density = 300))
```


<!-- ## Analysis of scRNA-seq {.unnumbered} -->
<!-- Next, we verify our findings at the cellular level using the Elyada scRNA-seq data [@elyada2019cross]. We found that our DeSurv factor 1 signature was expressed primarily in iCAF and B cells, factor 2 was expressed in Acinar and Classical PDAC 2, and factor 3 was expressed primarily in basal-like PDAC cells (Figure \ref{fig:fig-sc}B-E). This is consistent with our ORA analysis which found that factor 1 was mostly immune and factor 3 was basal-like, and with the overlap we saw with published gene lists that identified factor 1 as iCAF/classical/immune and factor 3 as basal-like. Interestingly, we do not see a large overlap of factor 1 with the classical PDAC cell types despite some overlap with classical gene lists. Need more here... -->

<!-- ```{r fig-sc, fig.cap="A. Cell type clusters from the Elyada scRNA-seq data. B-D. VAM scores for the DeSurv factor signatures in each cell shown on the UMAP of cells in the Elyada-sc data. E. Heatmap of average VAM scores by cell type and factor signature in the Elyada-sc data.", fig.env='figure*', fig.pos='t', out.height= "8in", out.width='\\textwidth'} -->

<!-- tar_load(fig_sc_plot_tcgacptac) -->
<!-- fig_sc_plot_tcgacptac -->
<!-- ``` -->

# Standard NMF factor structure at independently selected rank

Throughout the main text, both DeSurv and standard NMF were fit at the same rank ($k = 3$), enabling a controlled comparison of the effect of survival supervision on factor structure. When standard NMF selects its own rank independently, Bayesian optimization of the unsupervised ($\alpha = 0$) baseline selects $k = 7$, whereas supervised DeSurv BO selects $k = 3$. We examine here the NMF factor structure at this independently selected $k = 7$ to illustrate the additional fragmentation that arises when supervision is absent (Fig. \ref{fig:fig-nmf-k7-heatmap}).

At BO-selected $k = 7$ under $\alpha = 0$ (Fig. \ref{fig:fig-nmf-k7-heatmap}), standard NMF shows increased fragmentation relative to the matched-rank $k = 3$ solution: multiple factors partially overlap the same PDAC gene signatures, and no additional factor maps cleanly onto a distinct established PDAC program; the added factors fragment existing signatures rather than resolving new biology.

```{r fig-nmf-k7-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene-program correspondence for standard NMF at $k = 7$ (BO-selected rank with $\\alpha = 0$), computed with the same Spearman-correlation analysis and reference programs as main-text Fig.\\ 2a,b (top 50 genes per factor; reference programs with $r > 0.2$ for at least one factor are shown). Compared with the matched-rank $k = 3$ solution, the 7-factor solution shows increased fragmentation: multiple factors correlate with the same established PDAC signatures and no additional factor maps cleanly onto a distinct program. \\label{fig:fig-nmf-k7-heatmap}", fig.env='figure', fig.pos='t'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_k7_heatmap_tcgacptac.pdf", density = 300))
```

```{r tab-cindex-comparison, echo=FALSE, results='asis'}
library(dplyr)
library(knitr)
library(kableExtra)
load_result("val_cindex_desurv_tcgacptac")
load_result("val_cindex_std_desurvk_tcgacptac")
load_result("val_cindex_std_elbowk_tcgacptac")
load_result("val_cindex_desurv_alpha0_tcgacptac")

fmt_cindex <- function(tbl, col_name) {
  tbl %>%
    dplyr::select(dataset, cindex) %>%
    dplyr::rename(!!col_name := cindex)
}

cindex_wide <- fmt_cindex(val_cindex_desurv_tcgacptac, "DeSurv (K=3)") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_desurvk_tcgacptac, "Std NMF (K=3)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_elbowk_tcgacptac, "Std NMF (K=5, elbow)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_desurv_alpha0_tcgacptac, "Std NMF (K=7, BO $\\alpha$=0)"), by = "dataset") %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))) %>%
  dplyr::rename(Cohort = dataset) %>%
  dplyr::mutate(Cohort = gsub("_", "\\_", Cohort, fixed = TRUE))

# Explicit \label is the canonical referenced name (main paper + SI \ref it);
# push knitr's auto chunk-label to a throwaway name so it neither collides with
# the explicit one nor is silently relied upon in its place.
knitr::opts_current$set(label = "tab-cindex-comparison-auto")
kable(cindex_wide, format = "latex", booktabs = TRUE, linesep = "", escape = FALSE,
      caption = "External validation C-index by method and cohort. DeSurv at BO-selected rank ($k = 3$) is compared with standard NMF at matched rank ($k = 3$), elbow-selected rank ($k = 5$), and BO-selected rank ($k = 7$, $\\alpha = 0$). Higher values indicate better discrimination. \\label{tab:tab-cindex-comparison}") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

Standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ both substantially improve over NMF at $k = 3$ across all cohorts (Table \ref{tab:tab-cindex-comparison}); additional unsupervised factors capture prognostically relevant variation missed at the matched rank. At $k = 7$, NMF matches or exceeds DeSurv's per-cohort C-index in most cohorts, indicating that unsupervised factorization can approach supervised concordance given enough factors. This comes at the cost of a less parsimonious and less interpretable factorization: the $k = 7$ NMF solution shows increased fragmentation (Fig. \ref{fig:fig-nmf-k7-heatmap}), distributing the same prognostic signal across a rank-sensitive set of overlapping factors rather than concentrating it. The DeSurv linear predictor, by contrast, retains significant prognostic value after adjustment for established classifiers with only three factors (main-text Table \ref{tab:hr-adjusted}).


\section*{Supplementary Data: factor-specific gene lists}
\addcontentsline{toc}{section}{Supplementary Data: factor-specific gene lists}

As Supplementary Data, Tables \ref{tab:gene-list-270-D1}--\ref{tab:gene-list-270-D3} list the top 270 genes for each of the three DeSurv factors, ranked by the factor specificity score $s_{ij}$ defined in Supplementary Information, Section \ref{sec:gene-program-correlation}. The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\text{top}} = 270$); these are the genes used to compute sample-level factor scores for survival prediction. Factor D1 is enriched for classical tumor and restCAF-like stromal genes, Factor D2 for proCAF-associated stromal genes, and Factor D3 for basal-like tumor genes.

```{r tab-gene-list-270, echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)
library(tidyr)
library(dplyr)

genes <- read.csv("paper/gene_lists_top270_k3.csv", stringsAsFactors = FALSE)

# Keep only DeSurv factors
desurv_genes <- genes %>%
  dplyr::filter(factor %in% c("D1", "D2", "D3"))

# For each factor, split 270 genes into 6 columns of 45
build_factor_table <- function(df, fct) {
  g <- df %>% dplyr::filter(factor == fct) %>% dplyr::arrange(rank) %>% dplyr::pull(gene)
  data.frame(
    a = g[1:45], b = g[46:90], c = g[91:135],
    d = g[136:180], e = g[181:225], f = g[226:270],
    stringsAsFactors = FALSE
  )
}

factor_labels <- c("D1", "D2", "D3")
factor_captions <- c(
  "Factor D1 (classical tumor + restCAF-like stroma)",
  "Factor D2 (proCAF stroma)",
  "Factor D3 (basal-like tumor)"
)

for (i in seq_along(factor_labels)) {
  tbl <- build_factor_table(desurv_genes, factor_labels[i])
  # Make knitr's auto-generated table label unique per iteration so the three
  # loop tables don't all collide on tab:tab-gene-list-270 (multiply-defined).
  # The referenced per-factor labels (tab:gene-list-270-D<i>) are set in the caption.
  knitr::opts_current$set(label = paste0("tab-gene-list-270-", factor_labels[i]))
  cat("")
  kable(tbl, format = "latex", booktabs = TRUE, linesep = "",
        col.names = c("1--45", "46--90", "91--135", "136--180", "181--225", "226--270"),
        caption = paste0("Top 270 genes for DeSurv factor ", factor_labels[i],
          " (", factor_captions[i], "), ranked by factor specificity score $s_{ij}$.",
          if (i == 1) paste0(" The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\\text{top}}$).") else "",
          " \\label{tab:gene-list-270-", factor_labels[i], "}"),
        escape = FALSE) %>%
    kable_styling(font_size = 8, latex_options = c("HOLD_position")) %>%
    add_header_above(setNames(6, factor_captions[i]), bold = TRUE) %>%
    cat()
  cat("")
}
```

\section{Translational transportability into independently treated PDAC cohorts}
\label{si:treated}

```{r treated-setup, echo=FALSE, message=FALSE, warning=FALSE}
load_result("treated_cohort_stats")
.tcs  <- treated_cohort_stats
.pfmt <- function(p) if (is.na(p)) "---" else if (p < 0.001) "$<0.001$" else sprintf("%.3f", p)
.os   <- function(coh, prog, mod) { r <- .tcs$os[.tcs$os$cohort==coh & .tcs$os$program==prog & .tcs$os$model==mod, ]
  if (!nrow(r)) return("---"); sprintf("$%.2f$ ($%.2f$--$%.2f$), %s", r$hr, r$lo, r$hi, .pfmt(r$p)) }
.ax   <- function(coh, clf) { r <- .tcs$axis[.tcs$axis$cohort==coh & .tcs$axis$classifier==clf, ]
  if (!nrow(r)) return("---"); sprintf("$%.2f$ ($%.2f$--$%.2f$), %s", r$hr, r$lo, r$hi, .pfmt(r$p)) }
.prev <- function(coh) .tcs$prevalence[.tcs$prevalence$cohort==coh, ]
.oka  <- function(a) { r <- .tcs$okane$per_arm[.tcs$okane$per_arm$arm==a, ]
  sprintf("$%.2f$ ($%.2f$--$%.2f$)", r$hr, r$lo, r$hi) }
```

The programs were learned in treatment-naive, non-metastatic PDAC, the setting in which the gene-program matrix $W$ and Cox coefficients $\beta$ were trained. As a translational test we projected the fixed programs, without retraining, into independently treated PDAC cohorts and asked whether they retained clinically relevant associations. We used two complementary resources. First, the O'Kane/COMPASS cohort of treated metastatic PDAC (laser-capture microdissected and therefore epithelial-enriched), which supports a transportability test of the tumor-associated $D_1$ direction (Fig.\ \ref{fig:treated}A). Second, three bulk-tissue treated cohorts spanning two disease settings --- the metastatic gemcitabine-backbone trials Rash ($n = `r .prev("Rash")$n`$, gemcitabine + erlotinib) and Accept ($n = `r .prev("Accept")$n`$, gemcitabine $\pm$ afatinib) and the borderline-resectable / locally advanced Linehan cohort ($n = `r .prev("Linehan")$n`$; FOLFIRINOX $\pm$ the CCR2 inhibitor PF-04136309) --- in which, because tissue is bulk, both the PurIST (tumor-intrinsic) and DeCAF (stromal) classifiers are computable and the full three-program structure, including the stromal $D_2$ program, can be evaluated (Fig.\ \ref{fig:treated}C; Table \ref{tab:hr-adjusted} gives the same dual adjustment used in the main validation). Because the cohorts differ in stage and regimen, they are analysed per cohort and never pooled. The elastic-net penalty shrank the trained survival coefficient for the proCAF program $D_2$ to exactly zero in the treatment-naive model (an estimated selection outcome, not an imposed constraint), so any association of $D_2$ in the treated cohorts reflects a marginal re-evaluation of the projected program in a new clinical context rather than a property of the trained predictor.

```{r fig-treated, fig.cap = "Transportability of the frozen DeSurv programs into independently treated PDAC cohorts. (A) Prognostic transportability of D1 in treated metastatic PDAC. Forest plot of hazard ratios per SD of the frozen D1 score within each O'Kane/COMPASS treatment arm (FOLFIRINOX, gemcitabine/nab-paclitaxel [GA], and GA/experimental); the diamond is the common association from an arm-stratified Cox model with treatment-specific baseline hazards (not a pooled-arm estimate). D1 was protective in all three arms with no evidence of heterogeneity by treatment arm (interaction $P=0.40$). (B) The proCAF program D2 before and after treatment in paired Linehan biopsies (paired Wilcoxon test). (C) Per-cohort overall-survival hazard ratios per SD for each program in the bulk treated cohorts (Linehan, Rash, Accept); D1 and D3 are marginal, D2 is adjusted for PurIST and DeCAF. Error bars denote 95\\% confidence intervals. \\label{fig:treated}", fig.env='figure*', fig.pos='t', fig.width=7.2, fig.height=6, out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(magick::image_read("figures/fig_treated_context.pdf", density = 300))
```

\subsection{Prognostic transportability of D1 in treated metastatic disease}

To evaluate transportability in treated metastatic disease without combining treatment arms, we projected the frozen DeSurv programs into the O'Kane/COMPASS cohort and estimated $D_1$ associations separately within each arm. $D_1$ was associated with lower mortality in the same direction in FOLFIRINOX (HR per SD `r .oka("FFX")`), gemcitabine/nab-paclitaxel (HR `r .oka("GA")`) and the GA/experimental arm (HR `r .oka("GA/experimental")`; Fig.\ \ref{fig:treated}A). An arm-stratified Cox model allowing treatment-specific baseline hazards yielded a common association of HR `r sprintf("$%.2f$ ($%.2f$--$%.2f$)", .tcs$okane$stratified["hr"], .tcs$okane$stratified["lo"], .tcs$okane$stratified["hi"])` ($P = `r .pfmt(.tcs$okane$stratified["p"])`$; `r .tcs$okane$n` patients, `r .tcs$okane$events` deaths), with no evidence of heterogeneity across treatment arms (interaction $P = `r sprintf("%.2f", .tcs$okane$interaction_p)`$). Because these specimens were epithelial-enriched by laser-capture microdissection, this analysis specifically tests transportability of the tumor-associated $D_1$ direction rather than the stromal programs.

\subsection{Program associations in bulk treated cohorts}

Factor scores were computed by projection using the same trained, top-gene-truncated $\widetilde{W}$ ($n_{\text{top}} = `r .tcs$ntop`$ per factor, union across factors) and within-sample rank convention as the main-paper validation cohorts ($Z_{\text{new}} = \widetilde{W}^\top X_{\text{new}}$), z-standardized within cohort; results were near-identical under a full-gene projection (Spearman $r \geq 0.86$), so the treated-cohort findings do not depend on the projection convention. Because the three treated cohorts differ in disease setting and regimen (Linehan borderline-resectable/locally-advanced on FOLFIRINOX~$\pm$~CCR2 inhibitor; Rash and Accept metastatic on gemcitabine backbones), and the basal--classical program $D_1$ is heterogeneous across them, we report each cohort separately rather than as a pooled estimate. $D_1$ is reported \emph{marginally}: because $D_1$ tracks the basal--classical contrast and is collinear with the PurIST classifier (it separates PurIST calls at AUC $`r sprintf("%.2f", min(.tcs$d1_auc, na.rm=TRUE))`$--$`r sprintf("%.2f", max(.tcs$d1_auc, na.rm=TRUE))`$), adjusting $D_1$ for PurIST conditions the same axis on itself and over-adjusts; the marginal estimate is the appropriate measure of its prognostic value. $D_2$ (a stromal program, not strongly collinear with the classifiers) is reported both marginal and PurIST + DeCAF-adjusted.

\begin{table}[H]
\centering
\caption{Per-cohort overall-survival hazard ratios per SD in the three bulk treated PDAC cohorts. $D_1$ marginal; $D_2$ marginal and PurIST + DeCAF-adjusted; $D_3$ marginal. Sample sizes: Linehan $n=`r .prev("Linehan")$n`$ (`r .prev("Linehan")$events` events), Rash $n=`r .prev("Rash")$n`$ (`r .prev("Rash")$events` events), Accept $n=`r .prev("Accept")$n`$ (`r .prev("Accept")$events` events); basal-like prevalence Linehan `r .prev("Linehan")$pct_basal`\% (`r .prev("Linehan")$n_basal` cases), Rash `r .prev("Rash")$pct_basal`\%, Accept `r .prev("Accept")$pct_basal`\%. \label{tab:treated-os}}
\begin{tabular}{lcccc}
\toprule
Cohort (setting) & $D_1$ (marginal) & $D_2$ (marginal) & $D_2$ (+PurIST+DeCAF) & $D_3$ (marginal) \\
\midrule
Linehan (BR/LA)     & `r .os("Linehan","D1","marginal")` & `r .os("Linehan","D2","marginal")` & `r .os("Linehan","D2","adjusted")` & `r .os("Linehan","D3","marginal")` \\
Rash (metastatic)   & `r .os("Rash","D1","marginal")`    & `r .os("Rash","D2","marginal")`    & `r .os("Rash","D2","adjusted")`    & `r .os("Rash","D3","marginal")` \\
Accept (metastatic) & `r .os("Accept","D1","marginal")`  & `r .os("Accept","D2","marginal")`  & `r .os("Accept","D2","adjusted")`  & `r .os("Accept","D3","marginal")` \\
\bottomrule
\end{tabular}
\end{table}

The proCAF stromal program $D_2$ was protective in all three cohorts (both marginally and after adjustment, with significance reached in Linehan), with a consistent direction across the independent settings and chemotherapy backbones; it was significant in Linehan and non-significant but directionally concordant in the metastatic cohorts (Accept and Rash). Because $D_2$ is stromal and not strongly collinear with the classifiers, its association was not an artifact of adjustment: in Linehan it was significant both before and after adjustment, and across all three cohorts the adjusted estimate strengthened rather than attenuated relative to the marginal one, indicating that the $D_2$ association was not explained by PurIST or DeCAF adjustment. The basal--classical program $D_1$ was prognostic in the less-advanced Linehan cohort (marginal HR `r sub(",.*","",.os("Linehan","D1","marginal"))`) but not in the metastatic cohorts, a setting-dependence of the basal--classical axis itself (the same attenuation is seen for the PurIST, Moffitt and Bailey classifiers in these cohorts) rather than a failure of $D_1$. $D_3$ (basal-like) was not associated with overall survival in any treated cohort, consistent with the basal--classical contrast residing in $D_1$.

\subsection{Treatment-associated change in D2}

In the `r .tcs$paired$n` Linehan patients with paired pre- and post-treatment biopsies, $D_2$ increased significantly from pre- to post-treatment (paired Wilcoxon `r .pfmt(.tcs$paired$wilcox_p)`; mean $\Delta = `r sprintf("%+.2f", .tcs$paired$mean_delta)`$ SD; Fig.\ \ref{fig:treated}B), consistent with treatment-associated remodeling of the proCAF program.

\subsection{Arm-separated exploratory analyses}

We also examined each treatment arm separately across overall survival, time to progression, binary response and continuous depth of response (Supplementary Table \ref{tab:arm-separated}). These within-arm comparisons are underpowered for between-arm (treatment-specific) contrasts: the arms are small, and the few nominal associations were scattered across programs and endpoints and did not survive correction for multiple testing. The prognostic association instead prevails in the better-powered pooled and arm-stratified analyses (the O'Kane arm-stratified D1 effect and the consistent protective direction of D2 across the bulk cohorts), whereas differential effects by treatment arm are not resolvable in these data.

```{r tab-arm-separated, echo=FALSE, results='asis'}
.as <- .tcs$arm_separated
.est <- ifelse(.as$measure == "rho",
               sprintf("$\\rho$ %.2f", .as$estimate),
               sprintf("%s %.2f (%.2f--%.2f)", .as$measure, .as$estimate, .as$lo, .as$hi))
.ne  <- ifelse(is.na(.as$events), sprintf("%d", .as$n), sprintf("%d (%d)", .as$n, .as$events))
.pf  <- function(p) ifelse(p < 0.001, "$<0.001$", sprintf("%.3f", p))
cat("\\begin{table}[H]\\footnotesize\\centering\n",
    "\\caption{Arm-separated exploratory associations of the projected DeSurv programs (per SD) with clinical endpoints in the treated RASH/ACCEPT cohorts, analysed within each treatment arm. Overall survival and time to progression are Cox hazard ratios (HR); binary response (PR/SD vs PD) is a logistic odds ratio (OR); depth of response is a Spearman correlation ($\\rho$). Benjamini--Hochberg (BH) adjustment is across all unique arm-separated tests shown. \\label{tab:arm-separated}}\n",
    "\\begin{tabular}{lllccc}\n\\toprule\n",
    "Arm & Program & Endpoint & $n$ (events) & Estimate (95\\% CI) & $P$ ($P_{\\mathrm{BH}}$) \\\\\n\\midrule\n", sep = "")
for (i in seq_len(nrow(.as))) cat(sprintf("%s & %s & %s & %s & %s & %s (%s) \\\\\n",
    .as$arm[i], .as$program[i], .as$endpoint[i], .ne[i], .est[i], .pf(.as$p[i]), .pf(.as$p_bh[i])))
cat("\\bottomrule\n\\end{tabular}\n\\end{table}\n")
```

{\footnotesize Disease-control (CR/PR/SD vs PD) analyses were identical to the binary response analyses because no complete responses occurred in these treatment arms, and are therefore not reported separately or counted twice in the multiplicity adjustment. These analyses were exploratory: nominal $P$ values are reported for transparency, but no association remained significant after correction for multiple testing, and estimates from the smallest treatment arms are imprecise and should not be interpreted as evidence of treatment-specific effects.}

\subsection{Caveats}

These analyses are exploratory and prognostic, not predictive: treatment, stage and cohort are confounded, the treated cohorts are modestly sized, and no analysis plan was preregistered. The programs are analysed per cohort and never pooled, and $D_1$ is reported marginally in the bulk cohorts because it is collinear with the PurIST classifier. Robust treated-cohort validation, and any assessment of whether the programs predict differential treatment benefit, will require larger harmonized series with adequately powered treatment-by-program interaction analyses.
