---
title: "SI Appendix"
subtitle: "Survival-guided matrix factorization identifies reproducible prognostic programs in pancreatic cancer"
author: Amber M. Young, Alisa Yurovsky, Xianlu Laura Peng, Jen Jen Yeh, Didong Li, Naim U. Rashid
header-includes:
- \providecommand{\pandocbounded}[1]{#1}
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{amsmath}
- \usepackage{amsfonts}
- \usepackage{amssymb}
- \usepackage{bbm}
- \usepackage{algorithm}
- \usepackage{algpseudocode}
- \usepackage{setspace}
- \usepackage{enumitem}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator\diag{diag}
- \renewcommand{\algorithmicrequire}{\textbf{Input:}}
- \renewcommand{\algorithmicensure}{\textbf{Output:}}
- \renewcommand{\thefigure}{S\arabic{figure}}
- \renewcommand{\thetable}{S\arabic{table}}
- \renewcommand{\thealgorithm}{S\arabic{algorithm}}
bibliography: references_30102025.bib
reference-section-title: "SI References"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    keep_tex: true
  html_document: default
---
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\newpage

\part{Supplementary Methods}

\section{Model details}
Let $X \in \mathbb{R}_{\ge 0}^{p \times n}$ denote the nonnegative gene
expression matrix with $p$ genes and $n$ subjects. DeSurv approximates
$X \approx WH$ with $W \in \mathbb{R}_{\ge 0}^{p \times k}$ and
$H \in \mathbb{R}_{\ge 0}^{k \times n}$, and links the shared program matrix
$W$ to survival via an elastic-net–penalized Cox model.

Let $y \in \mathbb{R}^n_{> 0}$ be the observed survival times,
$\delta \in \{0,1\}^n$ the event indicators, and
$Z = W^\top X \in \mathbb{R}^{k \times n}$ the sample-wise factor scores. We write $Z_i$ for the $i$th column of $Z$, and
$n_{\text{event}} = \sum_{i=1}^n \delta_i$. For convenience we recall the
DeSurv loss from Equation 1 in the main text and rewrite it as
\begin{align}
  \mathcal{L}(W,H,\beta) &=\nonumber \\
  &(1-\alpha)\left(\frac{1}{2np}\lVert X - WH \rVert_F^2 + \frac{\lambda_H}{2nk} \lVert H \rVert_F^2\right) \nonumber \\
  -&\alpha \left(\frac{2}{n_{event}}\ell(W,\beta) - \lambda \{ \xi \lVert \beta \rVert_1 + \tfrac{1-\xi}{2} \lVert \beta \rVert_2^2 \} \right) \nonumber
\end{align}

where $\lVert \cdot \rVert_F$ is the Frobenius norm and
$\beta \in \mathbb{R}^k$ are the Cox coefficients. The Cox log-partial
likelihood $\ell(W,\beta)$ is
\begin{equation}
\ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left[
  Z_i^\top \beta
  -
  \log \left(
    \sum_{j : y_j \ge y_i}
    \exp(Z_j^\top \beta)
  \right)
\right].
\end{equation}
Hyperparameters satisfy $\alpha \in [0,1)$,
$\lambda_H \ge 0$, $\lambda \ge 0$, and $\xi \in [0,1]$.
The constants $1/(2np)$, $2/n_{\text{event}}$, and $1/(2nk)$ are chosen
for numerical convenience and do not affect the minimizer.


\section{Optimization Algorithm}

\subsection{Block coordinate descent scheme}
Algorithm \ref{alg:desurv} summarizes the block coordinate descent (BCD)
scheme used to minimize $\mathcal{L}(W,H,\beta)$. At each outer iteration,
we update $H$, then $W$, then $\beta$ while holding the other blocks
fixed.

\begin{algorithm}[htbp]
\caption{DeSurv block coordinate descent}
\label{alg:desurv}
    \begin{algorithmic}[1]
        \Require $X \in \mathbb{R}_{\geq 0}^{p \times n}$, survival times $y \in \mathbb{R}^n_{> 0}$, event indicators $\delta \in \{0,1\}^{n}$, tolerance $tol$, maximum iterations $maxit$, supervision parameter $\alpha$, rank $k$, penalties $\lambda_H$, $\lambda$, and $\xi$
        \Ensure Fitted DeSurv parameters $(W,H,\beta)$

        \State Initialize $W^{(0)}, H^{(0)}$ with positive entries (e.g., $W^{(0)},H^{(0)} \sim \mathrm{Uniform}(0,\max X)$)
        \State Initialize $\beta^{(0)}$ (e.g., $\beta^{(0)} = \mathbf{1}$)
        \State $loss = \mathcal{L}(W^{(0)},H^{(0)},\beta^{(0)})$
        \State $eps = \infty$, $t = 0$
        \While{$eps \geq tol$ \textbf{and} $t < maxit$}
            \State \textbf{(H-update)}
            \State \hspace{0.5cm} Update $H^{(t+1)}$ from $H^{(t)}$ using the multiplicative rule in Eq. \ref{eqn:Hupdate},
            \State \hspace{0.5cm} holding $W^{(t)}$ and $\beta^{(t)}$ fixed.
            \State \textbf{(W-update)}
            \State \hspace{0.5cm} Update $W^{(t+1)}$ from $W^{(t)}$ using the hybrid multiplicative rule in Eq. \ref{eqn:W_update},
            \State \hspace{0.5cm} holding $H^{(t+1)}$ and $\beta^{(t)}$ fixed, and perform backtracking to ensure
            \State \hspace{0.5cm} $\mathcal{L}$ does not increase.
            \State \textbf{($\beta$-update)}
            \State \hspace{0.5cm} Update $\beta^{(t+1)}$ from $\beta^{(t)}$ using a Newton-like step for the Cox loss
            \State \hspace{0.5cm} (Eq. \ref{eqn:beta_update}), holding $W^{(t+1)}$ and $H^{(t+1)}$ fixed.
            \State $lossNew = \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})$
            \State $eps = |lossNew - loss| / |loss|$
            \State $loss = lossNew$
            \State $t = t + 1$
        \EndWhile
        \State \Return $\hat{W} = W^{(t+1)}, \hat{H} = H^{(t+1)}, \hat{\beta}=\beta^{(t+1)}$
    \end{algorithmic}
\end{algorithm}

\subsection{Update for $H$}
Conditional on $W$, the loss $\mathcal{L}(W,H,\beta)$ reduces to a strictly
convex quadratic function of $H$. We adopt the standard NMF multiplicative
update with $\ell_2$ penalty:
\begin{equation}
\label{eqn:Hupdate}
H \leftarrow
\max\!\left(
  H \odot \frac{W^\top X}{
    W^\top W H + \lambda_H H + \varepsilon_H
  },
  \varepsilon_H
\right),
\end{equation}
where $\odot$ denotes elementwise multiplication, all divisions are
elementwise, and $\varepsilon_H > 0$ is a small floor to prevent entries from
becoming exactly zero. This update is equivalent to a majorization–minimization
step and guarantees a nonincreasing reconstruction term conditional on $W$
[@seung2001algorithms; @pascualmontano2006nonsmooth]. The $\varepsilon_H$-floor
ensures that iterates remain in the interior of the nonnegative orthant, which
simplifies the convergence analysis.


\subsection{Update for $W$}
For $W$, we construct a hybrid multiplicative update that balances the
contributions of the NMF and Cox gradients. Let
\[
\nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H)
=
\frac{1}{np}(W H - X)H^\top,
\]
and let
\[
\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta)
=
\frac{2}{n_{event}}\nabla_W \ell(W,\beta).
\]
where $\nabla_W \ell(W,\beta)$ denotes the Cox gradient with respect to $W$. A closed-form expression for $\nabla_W \ell$ is
\begin{equation}
\nabla_W \ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left(
  x_i
  -\frac{
      \sum_{j : y_j \ge y_i}
        x_j \exp(x_j^\top W \beta)
    }{
      \sum_{j : y_j \ge y_i}
        \exp(x_j^\top W \beta)
    }
\right)\beta^\top,
\end{equation}
where $x_i$ denotes the $i$th column of $X$.

We define the gradient-balancing factor
\begin{equation}
\label{eq:supp-zeta}
\zeta^{(t)}
=
\frac{
  \left\lVert (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W^{(t)},H^{(t+1)}) \right\rVert_F^2
}{
  \left\lVert \alpha \nabla_W \mathcal{L}_{Cox}(W^{(t)},\beta^{(t)}) \right\rVert_F^2
},
\end{equation}
and clip $\zeta^{(t)}$ to avoid extreme values:
$\zeta^{(t)} \leftarrow \min(\zeta^{(t)}, \zeta_{\max})$
with $\zeta_{\max} = 10^6$ in all experiments.

The multiplicative factor for $W$ is then
\begin{equation}
R^{(t)}
=
\frac{
  \frac{(1-\alpha)}{np} X H^{(t+1)\top}
  +\frac{2\alpha}{n_{\mathrm{event}}}\,\zeta^{(t)}\,\nabla_W \ell(W^{(t)},\beta^{(t)})
}{
  \frac{(1-\alpha)}{np} W^{(t)} H^{(t+1)}H^{(t+1)\top}
  +\varepsilon_W
},
\end{equation}
and the proposed update is
\begin{equation}
\label{eqn:W_update}
W^{(t+1)}
=
\max\!\left(
  W^{(t)} \odot R^{(t)},
  \varepsilon_W
\right),
\end{equation}
with a small floor $\varepsilon_W > 0$. When $\alpha=0$, this reduces
to the standard multiplicative update for NMF [@seung2001algorithms];
for $\alpha>0$, the Cox gradient perturbs the update in a direction that
decreases the supervised loss.

Because the combined multiplicative step does not, by itself, guarantee a
nonincrease in the full loss $\mathcal{L}$, we embed it in a backtracking
line search.

\begin{algorithm}[H]
\caption{$W$ update with backtracking}
\label{alg:backtrack}
    \begin{algorithmic}[1]
        \Require Value of $W$ and the previous iteration $t$: $W^{(t)}$, the multiplicative update term at the current iteration $R^{(t+1)}$, the max number of backtracking updates $max\_bt$, the backtracking parameter $\rho \subset (0,1)$
        \Ensure $W^{(t+1)}$
        \State $\theta =1$
        \State b = 1
        \State $flag\_accept = FALSE$
        \While{$b \leq max\_bt$}
            \State $W^{(t+1)}_{cand} = W^{(t)} \odot [(1-\theta) + \theta R^{(t+1)}]$
            \State \textbf{(Column normalization)}
            \State \hspace{0.5cm} Compute $D = \mathrm{diag}\!\big(\lVert W^{(t+1)}_{(:,1)cand}\rVert_2,\dots,\lVert W^{(t+1)}_{(:,k)cand}\rVert_2\big)$
            \State \hspace{0.5cm} and set $W^{(t+1)}_{cand} \leftarrow W^{(t+1)}_{cand} D^{-1}$, $H^{(t+1)}_{cand} \leftarrow D H^{(t+1)}$, and $\beta^{(t)}_{cand} \leftarrow D \beta^{(t)}$
            \If{$\mathcal{L}(W^{(t+1)}_{cand},H^{(t+1)},\beta^{(t)}) \leq \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)})$}
              \State $W^{(t+1)} = W^{(t+1)}_{cand}$
              \State $H^{(t+1)} = H^{(t+1)}_{cand}$
              \State $\beta^{(t)} = \beta^{(t)}_{cand}$
              \State $flag\_accept = TRUE$
              \State break
            \EndIf
            \State $\theta = \theta*\rho$
            \State $b = b + 1$
        \EndWhile
        \If{$flag\_accept = FALSE$}
            \State $W^{(t+1)}=W^{(t)}$
        \EndIf
    \end{algorithmic}
\end{algorithm}

The column normalization preserves both $WH$ and $W\beta$, and therefore
leaves the loss $\mathcal{L}$ invariant up to numerical error. Backtracking
guarantees that the accepted $W$ update does not increase $\mathcal{L}$.

\subsection{Update for $\beta$}

Conditional on $(W,H)$, the loss in $\beta$ reduces to a convex
elastic-net–penalized Cox problem:
\[
\min_{\beta \in \mathbb{R}^k}
\left\{
  -\frac{2\alpha}{n_{\mathrm{event}}}\,\ell(W,\beta)
  +
  \lambda\left(
    \xi \lVert \beta \rVert_1
    +\frac{1-\xi}{2} \lVert \beta \rVert_2^2
  \right)
\right\}.
\]
We solve this subproblem by cyclic coordinate descent following
[@simon2011regularization]. Writing $\ell(\beta) = \ell(W,\beta)$ and
$\tilde{\eta}_i = Z_i^\top \beta$, the update for coordinate $r$ has the
closed form
\begin{equation}
\label{eqn:beta_update}
\hat{\beta}_r
=
\frac{
  S\!\left(
    \dfrac{1}{n}
    \sum_{i=1}^n w(\tilde{\eta})_i
    z_{i,r}
    \left[
      v(\tilde{\eta})_i - \sum_{j \ne r} z_{i,j} \beta_j
    \right],
    \lambda \xi
  \right)
}{
  \frac{1}{n} \sum_{i=1}^n w(\tilde{\eta})_i z_{i,r}^2 + \lambda(1-\xi)
},
\end{equation}
where $S(a,\tau) = \mathrm{sign}(a)\max(|a|-\tau,0)$ is the soft-thresholding
operator, and $w(\tilde{\eta})$, $v(\tilde{\eta})$ are standard local
quadratic approximations of the Cox log-partial likelihood [@simon2011regularization].
We iterate coordinate updates until the subproblem converges to numerical
tolerance, yielding a minimizer in $\beta$ for the current $W$.


\subsection{Normalization of W}

After each accepted $W$ update, we normalize the columns of $W$ as
$W \leftarrow W D^{-1}$, and adjust $H$ and $\beta$ as
\[
H \leftarrow D H, \qquad \beta \leftarrow D \beta,
\]
where $D$ is the diagonal matrix of column $\ell_2$-norms of $W$.
This preserves the reconstruction $WH$ and the linear predictor $W\beta$,
leaving both the NMF and Cox terms in the loss unchanged. Normalization
prevents degeneracy in the scale-nonidentifiable factorization and keeps
columns of $W$ comparable in magnitude and interpretable as gene programs.


\subsection{Derivation of W update from projected coordinate descent}
The W update can be derived directly from the projected coordinate descent update with a specific step size.

Recall that the overall loss function is
\begin{align*}
    \mathcal{L}(W,H,\beta) = \frac{(1-\alpha)}{2np} ||X - WH||^2_F - \frac{2 \alpha}{n_{event}} \ell(W,\beta) + \lambda\left(\xi\|\beta\|_1 + \frac{(1-\xi)}{2}\|\beta\|_2^2\right).
\end{align*}
where $\ell(W,\beta)$ is the log-partial likelihood for the Cox model. Let $\nabla_W \ell$ represent the derivative of $\ell(W,\beta)$ with respect to $W$. Then the derivative of the overall loss with respect to $W$ is

\begin{equation*}
  \frac{\partial L}{\partial W} = \frac{(1-\alpha)}{np}\left( WHH^T - XH^T \right) - \frac{2\alpha}{n_{event}} \nabla_W \ell.
\end{equation*}

Then the gradient descent update rule at iteration $t$ is
\begin{align*}
  W^{(t)} &= W^{(t-1)} - \gamma \left( \frac{\partial L}{\partial W} \right) \\
          &= W^{(t-1)} - \gamma \left( \frac{(1-\alpha)}{np}\left( WHH^T - XH^T \right) - \frac{2\alpha}{n_{event}} \nabla_W \ell \right).
\end{align*}
Let the step size $\gamma$ be defined as in [@seung2001algorithms]:
\begin{equation*}
  \gamma = \frac{np}{(1-\alpha)}\frac{W}{WHH^T}.
\end{equation*}
Then the update becomes
\begin{align*}
  W^{(t)} &= W^{(t-1)} -\frac{np}{(1-\alpha)}\frac{W}{WHH^T} \left( \frac{(1-\alpha)}{np}\left( WHH^T - XH^T \right) - \frac{2\alpha}{n_{event}} \nabla_W \ell \right) \\
          &= \frac{W}{WHH^T} XH^T + \frac{2\alpha np}{n_{event}(1-\alpha)} \nabla_W \ell \\
          &= W \odot \frac{XH^T + \frac{2\alpha np}{n_{event}(1-\alpha)} \nabla_W \ell }{WHH^T} \\
          &= W \odot \frac{\frac{(1-\alpha)}{np} XH^T + \frac{2\alpha}{n_{event}} \nabla_W \ell}{\frac{(1-\alpha)}{np} WHH^T}.
\end{align*}
Finally, projected coordinate descent projects the $W$ update in to the positive space
\begin{equation*}
  W^{(t)} =  \max \left(W \odot \frac{\frac{(1-\alpha)}{np} XH^T + \frac{2\alpha}{n_{event}} \nabla_W \ell}{\frac{(1-\alpha)}{np}WHH^T}, 0\right).
\end{equation*}
Since $W \geq 0$ this is equivalent to
\begin{equation*}
  W^{(t)} =  W \odot \max \left( \frac{\frac{(1-\alpha)}{np} XH^T + \frac{2\alpha}{n_{event}} \nabla_W \ell}{\frac{(1-\alpha)}{np} WHH^T}, 0 \right),
\end{equation*}
which matches the multiplicative form used in the software implementation.


\section{Convergence proof}

We show that, under mild regularity conditions, the block coordinate descent
(BCD) Algorithm \ref{alg:desurv} converges to a stationary point of the DeSurv
loss function
\[
\mathcal{L}(W,H,\beta).
\]
For clarity, we first analyze the algorithm \emph{without} the column
normalization of $W$ inside the loop, and then argue in
Section \ref{subsec:conv_normalization} that the normalization step preserves
stationarity of limit points.

Throughout, let $\theta = (W,H,\beta)$ and denote
\(\mathcal{L}(\theta) = \mathcal{L}(W,H,\beta)\).
The feasible set is
\[
\Theta = \{(W,H,\beta):
  W \in \mathbb{R}_{\ge 0}^{p\times k},\;
  H \in \mathbb{R}_{\ge 0}^{k\times n},\;
  \beta \in \mathbb{R}^k\}.
\]

\textbf{Assumption 1 (Regularity and parameter space).}
We assume:
\begin{enumerate}[label=(\roman*)]
  \item The data $X \in \mathbb{R}_{\ge 0}^{p\times n}$, $y \in \mathbb{R}^n_{>0}$, and $\delta \in \{0,1\}^n$ are fixed and bounded.
  \item The hyperparameters satisfy $\lambda_H > 0$, $\lambda>0$, $\alpha \in [0,1)$, and $\xi \in [0,1]$.
  \item The initial iterate $\theta^{(0)} = (W^{(0)},H^{(0)},\beta^{(0)})$ lies in $\Theta$ and satisfies $W^{(0)},H^{(0)} > 0$ elementwise.
\end{enumerate}

\textbf{Assumption 2 (Block updates).}
At each outer iteration $t$, the three block updates in
Algorithm \ref{alg:desurv} satisfy:
\begin{enumerate}[label=(\roman*)]
  \item \emph{$H$-update.} For fixed $(W^{(t)},\beta^{(t)})$, the update
        $H^{(t)} \mapsto H^{(t+1)}$ is given by the multiplicative rule
        \[
        H \leftarrow H \odot \frac{W^\top X}{W^\top W H + \lambda_H H},
        \]
        applied at $(W^{(t)},H^{(t)})$. This rule preserves nonnegativity and
        yields a nonincreasing value of the reconstruction term conditional on
        $W^{(t)}$ and $\beta^{(t)}$; see e.g.
        (Lee and Seung 2001; Pascual-Montano et al. 2006; Lin 2007).
  \item \emph{$W$-update.} For fixed $(H^{(t+1)},\beta^{(t)})$, the update
        $W^{(t)} \mapsto W^{(t+1)}$ is the hybrid multiplicative step
        followed by backtracking as in Algorithm \ref{alg:backtrack}, producing
        $W^{(t+1)}$ such that
        \[
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)})
          \le
          \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)}).
        \]
        If no candidate satisfies the Armijo-type condition within the allowed
        number of backtracking steps, the algorithm sets
        $W^{(t+1)} := W^{(t)}$.
  \item \emph{$\beta$-update.} For fixed $(W^{(t+1)},H^{(t+1)})$, the update
        $\beta^{(t)} \mapsto \beta^{(t+1)}$ is obtained by running the
        coordinate descent method of Simon et al. (2011) on the convex elastic-net Cox
        subproblem until convergence. Thus
        \[
          \beta^{(t+1)} \in
          \arg\min_{\beta\in\mathbb{R}^k}
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta),
        \]
        and
        \[
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})
          \le
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)}).
        \]
\end{enumerate}

\vspace{0.5em}
\textbf{Lemma 1 (Continuity and bounded level sets).}
Under Assumption 1, $\mathcal{L}:\Theta\to\mathbb{R}$ is continuous, and the
initial sublevel set
\[
  \mathcal{S}_0 := \{\theta \in \Theta:
    \mathcal{L}(\theta) \le \mathcal{L}(\theta^{(0)})\}
\]
is nonempty, closed, and bounded.

\emph{Proof.}
The NMF term $\|X-WH\|_F^2$ is a polynomial in the entries of $W$ and $H$, hence
is continuously differentiable. The penalty $\|H\|_F^2$ is continuous as well.
The Cox partial log-likelihood is a smooth function of the linear predictor
$\eta_i = x_i^\top W \beta$, which is linear in $(W,\beta)$, so this term is
continuously differentiable in $(W,\beta)$. The $\ell_2$ penalty on $\beta$ is
smooth, and the $\ell_1$ penalty is convex and lower semicontinuous. Therefore
$\mathcal{L}$ is continuous on $\Theta$.

The constraints $W,H\ge 0$ define closed convex cones, and
$\beta\in\mathbb{R}^k$ is unconstrained. Because $\lambda_H>0$ and $\lambda>0$,
the terms $\tfrac{\lambda_H}{nk}\|H\|_F^2$ and
$\lambda\tfrac{1-\xi}{2}\|\beta\|_2^2$ dominate the objective as
$\|H\|_F\to\infty$ or $\|\beta\|_2\to\infty$, respectively. Thus the sublevel
set $\mathcal{S}_0$ is bounded in $(H,\beta)$. Since $W\ge 0$ and $X$ is
fixed, the reconstruction term and Cox term being bounded prevent $W$ from
diverging while $\mathcal{L}$ remains below $\mathcal{L}(\theta^{(0)})$, so
$W$ is bounded on $\mathcal{S}_0$. Because $\Theta$ is closed and
$\mathcal{L}$ is continuous, $\mathcal{S}_0$ is closed. Nonemptiness follows
from $\theta^{(0)}\in\mathcal{S}_0$. \hfill$\square$

\vspace{0.5em}
\textbf{Lemma 2 (Monotone descent and existence of limit points).}
Under Assumptions 1 and 2, Algorithm \ref{alg:desurv} (without $W$
normalization) generates a sequence $\{\theta^{(t)}\}_{t\ge 0}\subset\mathcal{S}_0$
such that
\[
  \mathcal{L}(\theta^{(t+1)}) \le \mathcal{L}(\theta^{(t)})\quad \forall\,t,
\]
and $\{\mathcal{L}(\theta^{(t)})\}$ converges to a finite limit $\mathcal{L}^*$.
Moreover, $\{\theta^{(t)}\}$ is bounded and therefore admits at least one limit
point.

\emph{Proof.}
By Assumption 2(i),
\[
  \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)})
  \le
  \mathcal{L}(W^{(t)},H^{(t)},\beta^{(t)}).
\]
By Assumption 2(ii),
\[
  \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)})
  \le
  \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)}).
\]
By Assumption 2(iii),
\[
  \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})
  \le
  \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)}).
\]
Combining these inequalities gives
\[
  \mathcal{L}(\theta^{(t+1)}) \le \mathcal{L}(\theta^{(t)})\quad\forall t,
\]
so $\{\mathcal{L}(\theta^{(t)})\}$ is monotonically nonincreasing. Since
$\mathcal{L}$ is bounded below on $\Theta$, the sequence converges to some
finite $\mathcal{L}^*$.

Each iterate lies in $\mathcal{S}_0$ by construction, and $\mathcal{S}_0$ is
bounded by Lemma 1. Therefore $\{\theta^{(t)}\}$ is bounded and has at least
one limit point by the Bolzano--Weierstrass theorem. \hfill$\square$

\vspace{0.5em}
\textbf{Lemma 3 (Blockwise optimality of limit points).}
Let $\theta^*=(W^*,H^*,\beta^*)$ be any limit point of $\{\theta^{(t)}\}$ under
Assumptions 1–2. Then:
\begin{enumerate}[label=(\roman*)]
  \item $H^*$ satisfies the KKT conditions for
        \[
          \min_{H\ge 0} \mathcal{L}(W^*,H,\beta^*).
        \]
  \item $W^*$ satisfies the KKT conditions for
        \[
          \min_{W\ge 0} \mathcal{L}(W,H^*,\beta^*).
        \]
  \item $\beta^*$ is the unique minimizer of
        \[
          \min_{\beta\in\mathbb{R}^k} \mathcal{L}(W^*,H^*,\beta),
        \]
        and satisfies the KKT conditions for the elastic-net Cox subproblem.
\end{enumerate}

\emph{Proof.}
Let $\{t_j\}$ be a subsequence such that
$\theta^{(t_j)}\to\theta^*=(W^*,H^*,\beta^*)$. By continuity of $\mathcal{L}$,
$\mathcal{L}(\theta^{(t_j)})\to\mathcal{L}^*$.

(i) Conditional on $(W,\beta)$, the $H$-subproblem is
\begin{equation}
  \min_{H \geq 0}\frac{(1-\alpha)}{np}(\lVert X-WH \rVert_F^2 + \frac{\lambda_H}{nk}||H||_F^2) + \text{const in $H$},\nonumber
\end{equation}
a strictly convex quadratic program. The multiplicative update for $H$ is known
to be a descent method whose fixed points coincide with the KKT points of this
constrained subproblem [@seung2001algorithms; @lin2007convergence]. Since the full
objective $\mathcal{L}(\theta^{(t)})$ converges, the per-iteration decrease
in $\mathcal{L}$ due to the $H$-update must vanish along $\{t_j\}$.
If $H^*$ were not KKT for the $H$-subproblem given $(W^*,\beta^*)$, there would
exist a feasible descent direction for $H$ at $(W^*,\beta^*)$, implying that
for sufficiently large $j$ the $H$-update would still produce a strict decrease
in $\mathcal{L}$, contradicting convergence. Hence $H^*$ satisfies the KKT
conditions.

(ii) Conditional on $(H,\beta)$, the $W$-subproblem is differentiable and
convex in $W$ (sum of a quadratic term and a negative Cox partial
log-likelihood in a linear predictor). The hybrid multiplicative step with
backtracking can be viewed as a projected gradient-like update onto the
nonnegative orthant: as shown in the derivation above, the multiplicative
direction is equivalent to a negative gradient step with the adaptive step
size of Seung et al., projected onto the nonnegative orthant. Under mild conditions on the search direction and step
sizes, any limit point of such a projected gradient scheme is a KKT point for
the constrained subproblem; see, for example, [@tseng2001convergence; @grippo2000convergence].
If $W^*$ did not satisfy the KKT conditions, there would be a feasible
descent direction at $(W^*,H^*,\beta^*)$ and a sufficiently small step that
reduces $\mathcal{L}$, which the line search would eventually accept. This
would contradict the fact that $\mathcal{L}(\theta^{(t_j)})\to\mathcal{L}^*$.
Therefore $W^*$ is KKT for the $W$-subproblem.

(iii) For fixed $(W,H)$, the $\beta$-subproblem is the convex elastic-net
penalized Cox objective. The coordinate descent algorithm converges to the unique minimizer. By Assumption 2(iii),$\beta^{(t+1)}$ is taken to be this minimizer at each iteration. Passing to the limit along $\{t_j\}$ shows that $\beta^*$ is the unique minimizer of the
$\beta$-block given $(W^*,H^*)$ and hence satisfies its KKT conditions.
\hfill$\square$

\vspace{0.5em}
\textbf{Theorem 1 (Convergence to a stationary point).}
Under Assumptions 1 and 2, the sequence $\{\theta^{(t)}\}$ produced by
Algorithm \ref{alg:desurv} (without $W$ normalization) satisfies:
\begin{enumerate}[label=(\alph*)]
  \item The objective values $\{\mathcal{L}(\theta^{(t)})\}$ are monotonically
        nonincreasing and converge to a finite limit $\mathcal{L}^*$.
  \item Every limit point $\theta^* = (W^*,H^*,\beta^*)$ of $\{\theta^{(t)}\}$
        is a stationary point of $\mathcal{L}$ on $\Theta$, i.e. it satisfies
        the first-order KKT conditions for
        \[
          \min_{\theta\in\Theta} \mathcal{L}(\theta).
        \]
\end{enumerate}

\emph{Proof.}
Part (a) is Lemma 2. For part (b), Lemma 3 shows that any limit point $\theta^*$
is blockwise optimal: each block is optimal (in the KKT sense) when the others
are held fixed.

The DeSurv loss can be written as
\[
\mathcal{L}(\theta) = f(W,H,\beta) + g(\beta)
                     + I_{\{W\ge 0\}}(W) + I_{\{H\ge 0\}}(H),
\]
where $f$ is continuously differentiable, $g(\beta)=\lambda\xi\|\beta\|_1$ is
a separable convex (possibly nondifferentiable) penalty, and $I_C$ denotes the
indicator of a closed convex set $C$. This is the smooth+nonsmooth composite
form considered in block coordinate descent analyses such as [@tseng2001convergence].
In this setting, any point that is optimal with respect to each block
individually (a block coordinatewise minimizer) is a stationary point for the
full problem: equivalently,
\[
0 \in \nabla_{W,H} f(W^*,H^*,\beta^*)
    + \partial I_{\{W\ge 0\}}(W^*)
    + \partial I_{\{H\ge 0\}}(H^*),
\]
\[
0 \in \nabla_\beta f(W^*,H^*,\beta^*)
    + \partial g(\beta^*),
\]
which are precisely the KKT conditions for
$\min_{\theta\in\Theta}\mathcal{L}(\theta)$.
Hence every limit point of Algorithm \ref{alg:desurv} is stationary.
\hfill$\square$

\subsection{Effect of column normalization of $W$}
\label{subsec:conv_normalization}

The above analysis omits the column-normalization step for $W$ used in
Algorithm \ref{alg:backtrack}. We briefly argue that this normalization does
not affect stationarity of limit points.

Let $D$ be a diagonal matrix with strictly positive diagonal entries. The
transformation
\[
  (W,H,\beta) \mapsto (W',H',\beta') = (W D^{-1},\, D H,\, D\beta)
\]
preserves both the product $WH$ and the linear predictor $W\beta$, and hence
leaves $\mathcal{L}$ invariant. The column-normalization step is exactly such a
transformation, with $D$ chosen from the column norms of $W$.

Let $\{\theta^{(t)}\}$ be the sequence generated by the algorithm without
normalization, and let $\{\tilde{\theta}^{(t)}\}$ be the sequence with
normalization. For each $t$ there exists a diagonal $D^{(t)}$ with strictly
positive entries such that
\[
  \tilde{\theta}^{(t)} =
  (W^{(t)} D^{(t),-1},\, D^{(t)} H^{(t)},\, D^{(t)}\beta^{(t)}),
\]
and $\mathcal{L}(\tilde{\theta}^{(t)}) = \mathcal{L}(\theta^{(t)})$.
Thus limit points of $\{\tilde{\theta}^{(t)}\}$ are obtained from limit points
of $\{\theta^{(t)}\}$ by such invertible diagonal scalings.

Since the KKT conditions are expressed in terms of the gradients with respect
to $WH$ and $W\beta$, and these quantities are invariant under the above
scaling, stationarity is preserved under the transformation. Therefore
Theorem 1 implies that every limit point of the \emph{normalized} algorithm is
also a stationary point of $\mathcal{L}$ (up to this scaling equivalence), which
establishes convergence of the implementation used in practice.

\subsection{Remark (Coxnet implementation)}
Our implementation updates the $\beta$ block using a Coxnet-style coordinate descent that relies on an approximate Hessian. Consequently, the formal optimality proof is established for an idealized version of the $\beta$ update that assumes exact second-order information. Nonetheless, in empirical applications the approximate update is numerically stable and yields a monotone decrease in the objective, consistent with the behavior predicted by the idealized analysis.


\section{Supervision on \(W\) versus \(H\)}

Classical nonnegative matrix factorization (NMF) is non-identifiable: for any
invertible matrix \(R\) with nonnegative entries, the factorization
\((W, H)\) can be transformed to \((W R, R^{-1} H)\) without changing the
reconstruction error, yielding multiple valid solutions
[@donoho2004nmf; @laurberg2008uniqueness; @gillis2014nmf]. Although normalizing
columns of \(W\) or rows of \(H\) resolves the trivial scaling ambiguity,
nonnegative mixing transformations persist, and empirical studies consistently
show that the sample-loading matrix \(H\) is more sensitive to initialization
and local minima than the program matrix \(W\)
[@brunet2004metagenes; @kim2007sparse; @gillis2012accelerated].

In DeSurv, survival supervision enters through the projection $Z = W^\top X$ in the partial log-likelihood, so that the gradient of the DeSurv loss with respect to the programs is
\[
\nabla \mathcal{L} = (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H) - \alpha\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta),
\]
and therefore the update rule for $W$ as in Equation \ref{eqn:W_update} relies on both the reconstruction term and the supervision term. As such, the gene-program definitions are directly shaped by their association with survival. This mechanism amplifies
genes aligned with the hazard gradient and suppresses those that dilute
prognostic structure, ensuring that the learned programs encode survival-relevant
biological variation.

By contrast, if supervision were applied to the sample loadings \(H\), the model
could reduce the Cox loss by redistributing patient-specific coefficients while
leaving the gene-level programs \(W\) nearly unchanged, a behavior documented in
multiple supervised variants of NMF and supervised topic models, where
supervision applied only to the coefficient matrix modifies \(H\) but not the
basis \(W\) [@cai2011graph; @blei2007slda].

Supervising \(W\) also provides a practical advantage for **portability**:
because \(W\) defines gene-level programs, new samples can be embedded by the
closed-form projection \(Z_{\text{new}} = X_{\text{new}}^\top W\). In contrast,
supervising \(H\) would not yield such a mapping; instead, obtaining sample
loadings for external datasets would require solving a nonnegative least-squares
(NNLS) problem for each new sample, introducing optimization noise and reducing
reproducibility.

Together, these considerations motivate supervising through \(W\), which shapes
the gene programs toward prognostic directions, stabilizes solutions across
initializations, and yields biologically interpretable signatures that
generalize across cohorts.


\section{Cross-validation procedure}
Algorithm \ref{alg:cv} describes the cross validation procedure with $F$ folds and $R$ initializations.
We define the c-index using comparable pairs
$\mathcal{P} = \{(i,j) : y_i < y_j,\ \delta_i = 1\}$ and linear predictors
$\hat{\eta}_i$:
\begin{equation}
\label{eqn:cindex}
\hat{c} =
\frac{1}{|\mathcal{P}|}
\sum_{(i,j)\in\mathcal{P}}
\left[
  \mathbbm{1}(\hat{\eta}_i > \hat{\eta}_j)
  + \tfrac{1}{2}\mathbbm{1}(\hat{\eta}_i = \hat{\eta}_j)
\right].
\end{equation}

\begin{algorithm}[H]
\caption{Cross-validation for DeSurv pipeline}
\label{alg:cv}
\begin{algorithmic}[1]
\Require
\Statex Number of folds $F$
\Statex Number of initializations $R$
\Statex Observed data ($X$, $y$, $\delta$)
\Statex Hyperparameters $k$, $\alpha$, $\lambda$, $\xi$, and $\lambda_H$
\Ensure The cross-validated c-index

\State Divide subjects into $F$ folds.
\For{$f = 1$ to $F$}
    \State Split data into training and validation $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(X_{(f)},y_{(f)}, \delta_{(f)})$ sets
    \For{$r = 1$ to $R$}
        \State set.seed$(r)$
        \State Apply Algorithm \ref{alg:desurv} with inputs $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(k, \alpha, \lambda, \xi, \lambda_H)$
        \State Obtain $\hat{W}$ and $\hat{\beta}$ as output from Algorithm \ref{alg:desurv}
        \State Compute the estimated linear predictor:
        $
        \hat{\eta} = X_{(f)}^T\hat{W}\hat{\beta}
        $
        \State Compute c-index, denoted $\hat{c}_{(f)r}$, according to Equation \ref{eqn:cindex}
    \EndFor
    \State \textbf{end loop over $r$}
    \State Compute average c-index across initializations:
    $
    \hat{c}_{(f)} = \frac{1}{R}\sum_{r=1}^R\hat{c}_{(f)r}
    $
\EndFor
\State \textbf{end loop over $f$}
\State Compute final cross-validated c-index:
$
\hat{c} = \frac{1}{F}\sum_{f=1}^F \hat{c}_{(f)}
$
\State \Return Final estimate $\hat{c}$

\end{algorithmic}
\end{algorithm}


\section{Bayesian Optimization}

We treat hyperparameter tuning as a black-box optimization problem over a
bounded domain. Let $\theta$ collect the tuned parameters (at minimum
$\theta = (k,\alpha,\lambda,\xi)$, and optionally discrete choices
such as the signature size $n_{\text{top}}$). For a given $\theta$ we run the
cross-validation procedure in Algorithm \ref{alg:cv}, averaging the c-index
across folds and random initializations to form the objective
\[
\hat{c}(\theta) = \frac{1}{F} \sum_{f=1}^F \hat{c}_{(f)}(\theta).
\]
Because $\hat{c}(\theta)$ is expensive to evaluate and non-differentiable with
respect to $\theta$, we use Bayesian optimization (BO) to adaptively select
candidate configurations.

We place a Gaussian-process (GP) prior on the response surface
$f(\theta) \sim \mathcal{GP}(0, k(\theta,\theta'))$ with a Mat\'ern kernel and
automatic relevance determination length-scales. Continuous parameters are
rescaled to $[0,1]$ and penalties are optimized on a log scale; integer-valued
coordinates (e.g., $k$ or $n_{\text{top}}$) are proposed in the continuous
space and rounded inside the evaluator. After an initial batch of evaluations,
the GP is refit after each new observation and used to guide acquisition.

At iteration $t$, let $\mu_t(\theta)$ and $\sigma_t(\theta)$ denote the GP
posterior mean and standard deviation, and let $f_\star$ be the best observed
score. We select new candidates by maximizing expected improvement (EI),
\[
\operatorname{EI}(\theta) =
\left(\mu_t(\theta) - f_\star - \kappa\right)\Phi(Z(\theta))
  + \sigma_t(\theta)\phi(Z(\theta)),
\quad
Z(\theta) = \frac{\mu_t(\theta) - f_\star - \kappa}{\sigma_t(\theta)},
\]
where $\Phi$ and $\phi$ are the standard normal cdf/pdf and $\kappa \ge 0$
controls the exploration margin. EI is evaluated over a candidate pool sampled
from the bounded search region, and the maximizer is evaluated next. The loop
continues until the iteration budget is exhausted or the improvement plateaus.

For model selection, we take the highest observed $\hat{c}(\theta)$ as the
optimal configuration. When fold-level standard errors are available, we apply
a one-standard-error rule for $k$, selecting the smallest rank whose mean
performance lies within one standard error of the best. The final DeSurv model
is refit at the selected parameters using the consensus-based initialization
described above, and the BO-chosen $n_{\text{top}}$ (when tuned) is used to
truncate each program for downstream analysis.

The parameter $n_{\text{top}}$ controls which genes enter the survival linear predictor. After factorization, each sample's score for factor $j$ is computed as $\widetilde{W}_j^\top x$, where $\widetilde{W}_j$ retains only the $n_{\text{top}}$ genes with the largest factor-specificity scores $s_{ij}$ (SI Appendix, Section~\ref{sec:gene-program-correlation}); genes outside this set still shape the factorization but do not enter the linear predictor. BO selects $n_{\text{top}}$ jointly with $k$, $\alpha$, $\lambda$, and $\xi$ by optimizing cross-validated C-index. When $\alpha = 0$ (no survival supervision), gene selection and survival outcomes are decoupled, so varying $n_{\text{top}}$ has negligible effect on precision in the null simulation.

\section{PDAC Datasets}

We analyzed PDAC cohorts from TCGA-PAAD and CPTAC-PDAC as training data
[@raphael2017integrated;@ellis2013clinical]. External validation used
independent cohorts from Dijk, Moffitt, PACA-AU (array and RNA-seq), and Puleo
[@dijk2020unsupervised;@moffitt2015virtual;@zhao2018gene;@puleo2018stratification].
Training models were fit on the combined TCGA and CPTAC expression matrices
after harmonizing gene identifiers; validation datasets were processed with the
same gene set and transformations before prediction and clustering.

Dataset-specific inclusion and preprocessing steps were as follows. For
TCGA-PAAD, we excluded samples lacking tumor grade and mapped features to gene
symbols. For CPTAC, we retained samples annotated as PDAC by histology. For
Dijk, all 90 samples passed quality-control filters and were retained. For
Moffitt, only primary tumor specimens were kept. For PACA-AU array and RNA-seq,
we retained primary PDAC tumors and
collapsed duplicated gene symbols by median; the RNA-seq cohort received a
``\_seq'' suffix on sample IDs to avoid collisions with the array cohort. Puleo
required no additional dataset-specific filtering beyond survival QC.

Across cohorts, samples without valid survival outcomes (nonpositive time,
missing time, or missing event indicator) were removed. Expression matrices
were analyzed on a log scale: RNA-seq cohorts as log$_2$(TPM + 1); microarray cohorts in platform-normalized log-scale intensities. For each training cohort (TCGA-PAAD and CPTAC) separately, genes were ranked
by mean expression and by variance independently; the 3,000 genes with the
lowest sum of these two ranks were retained, selecting genes that are both
highly expressed and highly variable.
The intersection of these two gene sets yielded 1,970 genes used for model
training. Validation datasets were restricted to this same 1,970-gene set.
Within-sample rank transformation was applied to all datasets to mitigate
platform effects, with validation data transformed identically to training.

Table \ref{tab:cohorts} summarizes the cohorts, platforms, sample sizes after
quality control, and patient-level metadata used in this study. Overall survival
time and event indicators were available for all cohorts. PurIST
[@rashid2020purity] and DeCAF [@peng2024determination] molecular
classifications, applied to bulk expression data as described in those
references, were used as covariates in the covariate-adjusted external
validation analysis. Gene expression data and molecular
classifications for the validation cohorts were originally compiled for the
DeCAF study [@peng2024determination]; training data were obtained directly
from GDC.

\begin{table}[ht]
\centering
\caption{Summary of PDAC cohorts used in this study. $N$ denotes the number of
samples after all quality-control filters (histology, survival validity,
dataset-specific exclusions). Patient-level metadata: OS = overall survival
(time and event indicator); PurIST = tumor subtype classification
(Rashid et al. 2020); DeCAF = CAF subtype classification
(Peng et al. 2020). Expression data and molecular classifications
for the validation cohorts were originally compiled for the DeCAF study
(Peng et al. 2020).}
\label{tab:cohorts}
\footnotesize
\begin{tabular}{llrllll}
\toprule
Cohort & Platform & $N$ & Role & Accession & Metadata & Reference \\
\midrule
TCGA-PAAD        & RNA-seq    & 144 & Training   & GDC (TCGA-PAAD) & OS, PurIST, DeCAF & Raphael et al. (2017) \\
CPTAC            & RNA-seq    & 129 & Training   & GDC (CPTAC-3)    & OS, PurIST, DeCAF & Ellis et al. (2013) \\
\addlinespace
Dijk             & RNA-seq    &  90 & Validation & E-MTAB-6830      & OS, PurIST, DeCAF & Dijk et al. (2020) \\
Moffitt          & Microarray & 123 & Validation & GSE71729         & OS, PurIST, DeCAF & Moffitt et al. (2015) \\
PACA-AU (array)  & Microarray &  63 & Validation & GSE36924         & OS, PurIST, DeCAF & Zhao et al. (2018) \\
PACA-AU (RNA-seq) & RNA-seq   &  52 & Validation & EGAS00001000154  & OS, PurIST, DeCAF & Zhao et al. (2018) \\
Puleo            & Microarray & 288 & Validation & E-MTAB-6134      & OS, PurIST, DeCAF & Puleo et al. (2018) \\
\addlinespace
\multicolumn{2}{l}{Training total} & 273 & & & & \\
\multicolumn{2}{l}{Validation total} & 616 & & & & \\
\bottomrule
\end{tabular}
\end{table}

\section{Survival Analysis}

Survival analyses used overall survival time with event indicators as provided
by each cohort. For fitted DeSurv models, sample-level program scores were
computed as $Z = \widetilde{W}^\top X$ and standardized using the training mean and
standard deviation before constructing the linear predictor
$\hat{\eta} = Z \hat{\beta}$. Prognostic performance was summarized using the
concordance index as defined above. For visualization and group-level testing,
we generated Kaplan--Meier curves and log-rank tests for risk groups defined by
median splits of $\hat{\eta}$ within each dataset to avoid cross-cohort scaling
differences. When multiple datasets were pooled, Cox models were stratified
by dataset to allow cohort-specific baseline hazards. All p-values reported for
survival differences between groups correspond to two-sided log-rank tests.

\section{Gene Program Correlation Analysis}
\label{sec:gene-program-correlation}

To assess the biological identity of learned factors, we evaluated the
correspondence between factor-specific gene rankings and established PDAC gene
programs. For each factor column $j$ of the gene program matrix $W$, we
identified the top 50 genes by factor specificity score. Each column of $W$
was first scaled by its column maximum to produce $W^*_{ij} = W_{ij} /
\max_i W_{ij}$, placing all factors on a common $[0,1]$ scale. The
specificity score for gene $i$ in factor $j$ was then defined as
$$s_{ij} = W^*_{ij} - \max_{j' \neq j} W^*_{ij'}.$$
Genes were ranked in descending order of $s_{ij}$, and the top 50 per factor
were retained. The union of these gene sets across all factors was then
intersected with the genes present in any reference program to form the set of
genes used in the correlation analysis.

For each factor $j$ and each reference gene program $G_k$, we computed the
Spearman rank correlation between the continuous factor loadings $W_{\cdot j}$
(restricted to the intersection gene set) and a binary indicator vector
$\mathbf{1}[\text{gene} \in G_k]$. P-values were derived from the asymptotic
approximation to the Spearman test statistic and adjusted for multiple
comparisons using the Benjamini--Hochberg procedure applied within each factor
column independently. Gene programs were included in the visualization only if
at least one factor yielded a Spearman correlation exceeding 0.2. Asterisks
($*$) indicate adjusted p-value $< 0.1$.

Reference gene programs included Classical and Basal-like tumor cell signatures
from Moffitt et al.\ [@moffitt2015virtual]; iCAF and myCAF
cancer-associated fibroblast signatures from Elyada et al.\
[@elyada2019cross]; promoting (proCAF) and restraining (restCAF) cancer-associated fibroblast subtypes
from the DeCAF study [@peng2024determination]; proCAF and restCAF subtypes from SCISSORS [@leary2023s]; tumor,
stromal, and immune compartment signatures from DECODER [@peng2019novo];
and Classical and Basal-like subtype classifiers from PurIST
[@rashid2020purity]. Programs from Bailey et al.\ [@Bailey2016],
periductal subtypes, and MSI subtypes were excluded from visualization due to
redundancy with other included programs or low overall correlation across all
factors.

\section{Variance and Survival Contribution per Factor}

For each factor $j \in \{1, \ldots, k\}$, we quantified two quantities to assess
the alignment between expression variance and prognostic relevance. The
conditional variance explained by factor $j$ was quantified as the semi-partial
$R^2$:
$$V_j = \frac{\mathrm{RSS}_{-j} - \mathrm{RSS}_{\text{full}}}{\|X\|_F^2},$$
where $\mathrm{RSS}_{\text{full}} = \|X - WH\|_F^2$ is the residual sum of
squares from the full $k$-factor reconstruction and
$\mathrm{RSS}_{-j} = \|X - W_{-j}H_{-j}\|_F^2$ is the residual from the
$(k-1)$-factor reconstruction omitting factor $j$. This measures the incremental
fraction of total variance accounted for by factor $j$ given all other factors.

The survival contribution of factor $j$ was quantified as the Type III partial
likelihood increment: the reduction in Cox partial log-likelihood when factor $j$
is omitted from the full $k$-factor model,
$$\Delta\ell_j = \ell(\beta_1, \ldots, \beta_k) - \ell(\beta_1, \ldots,
\beta_{j-1}, \beta_{j+1}, \ldots, \beta_k),$$
where $\ell(\cdot)$ denotes the Cox partial log-likelihood evaluated at the
maximum partial likelihood estimates. Larger $\Delta\ell_j$ indicates a greater
marginal contribution of factor $j$ to survival prediction given all other
factors. These two quantities are plotted against each other in
main text, Fig.\ 3C.

\section{Factor Correspondence Analysis}

To assess which biological programs are preserved, altered, or suppressed
under survival supervision, we quantified the pairwise correspondence between
NMF and DeSurv factor loadings. For each pair of factors
$(j_{\text{NMF}},\, j_{\text{DeSurv}})$, we computed the Spearman rank
correlation between the corresponding columns of the NMF gene program matrix
$W_{\text{NMF}}$ and the DeSurv gene program matrix $W_{\text{DeSurv}}$,
restricted to the top 50 genes per factor (by factor-specificity score $s_{ij}$;
SI Appendix, Section~\ref{sec:gene-program-correlation}) present in both models.
Spearman correlation was used to capture monotone correspondence in gene loading
ranks while remaining robust to scale differences between the two methods.
Results are displayed as a heatmap with NMF factors as rows and DeSurv factors
as columns (main text, Fig.\ 3D).

\clearpage

\newpage

\part{Supplementary Results}

<!-- PNAS CONVENTION: All supplementary figures should be labeled "Fig. S1", "Fig. S2", etc.
     not "Figure 1", "Figure 2". Tables should be "Table S1", etc.
     Internal scenario code names (R0_easy, R00_null, R_mixed) should be replaced with
     descriptive labels (primary, null, mixed) consistent with main text terminology. -->
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Determine figure output path (respects DESURV_NTOP subfolder)
.ntop_env <- Sys.getenv("DESURV_NTOP", "")
.ntop_lo  <- Sys.getenv("DESURV_NTOP_LOWER", "")
.ntop_hi  <- Sys.getenv("DESURV_NTOP_UPPER", "")
.fig_sub  <- if (nzchar(.ntop_env)) {
  paste0("ntop_", .ntop_env)
} else if (nzchar(.ntop_lo) && nzchar(.ntop_hi)) {
  paste0("ntop_bo_", .ntop_lo, "_", .ntop_hi)
} else {
  ""
}
.param_rule_env <- Sys.getenv("DESURV_PARAM_RULE", "best")
if (identical(.param_rule_env, "1se") && nzchar(.fig_sub)) {
  .fig_sub <- paste0(.fig_sub, "_1se")
}
.fig_path <- if (nzchar(.fig_sub)) {
  file.path("figures", .fig_sub, "")
} else {
  "figures/"
}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo = FALSE,
                      fig.path = .fig_path,
                      dev = "png",
                      dpi = 300)
source("paper/load_precomputed.R")
library(dplyr)
library(ggplot2)
library(cowplot)
if (!exists("set_fig_font", inherits = TRUE)) {
  set_fig_font <- function(plot_obj, size = 10) {
    if (inherits(plot_obj, "ggplot")) {
      plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
    } else {
      plot_obj
    }
  }
}
```

```{r ksens-data, echo=FALSE}
#
# -- Load K-sensitivity RDS outputs --------------------------------------------
# Generated by:
#   Rscript inst/cv_grid_validation_analysis_amber_optimal.R  -> adj_p_270, hcor_270, master_rows_270
#   Rscript inst/cv_grid_validation_analysis_allgenes.R       -> adj_p_all,  hcor_all,  master_rows_all
#   Rscript inst/k3_k7_nesting_heatmap.R                     -> k3_k7_cor_mat, k3_k7_summary
#   Rscript inst/create_ksens_rds.R                          -> lam300_summary, production_summary (bootstrap)
RDS_DIR <- "results/cv_grid"

adj_p_270      <- readRDS(file.path(RDS_DIR, "adj_p_270_matrix.rds"))
adj_p_all      <- readRDS(file.path(RDS_DIR, "adj_p_all_matrix.rds"))
hcor_270       <- readRDS(file.path(RDS_DIR, "hcor_270_matrix.rds"))
hcor_all       <- readRDS(file.path(RDS_DIR, "hcor_all_matrix.rds"))
master_270     <- readRDS(file.path(RDS_DIR, "master_rows_270.rds"))
master_all     <- readRDS(file.path(RDS_DIR, "master_rows_all.rds"))
k3k7_sum       <- readRDS(file.path(RDS_DIR, "k3_k7_summary.rds"))
lam300         <- readRDS(file.path(RDS_DIR, "lam300_summary.rds"))
prod_sum       <- readRDS(file.path(RDS_DIR, "production_summary.rds"))

# Labels for table formatting
alpha_labs <- c("$\\alpha$=0.00", "$\\alpha$=0.25", "$\\alpha$=0.35", "$\\alpha$=0.55", "$\\alpha$=0.75", "$\\alpha$=0.85", "$\\alpha$=0.95")
k_labs     <- c("K=2", "K=3", "K=5", "K=7", "K=9")

# -- Derived scalars from P-value matrices -------------------------------------
n_sig_270        <- sum(adj_p_270 < 0.05)
n_sig_all        <- sum(adj_p_all < 0.05)
k3_alpha55_p_270 <- adj_p_270["K=3", "a=0.55"]
k3_alpha55_p_all <- adj_p_all["K=3", "a=0.55"]
k5_alpha55_p_270 <- adj_p_270["K=5", "a=0.55"]
k5_alpha25_p_all <- adj_p_all["K=5", "a=0.25"]
k7_alpha75_p_270 <- adj_p_270["K=7", "a=0.75"]
n_k2_sig_270     <- sum(adj_p_270["K=2", ] < 0.05)
n_k3_sig_270     <- sum(adj_p_270["K=3", ] < 0.05)
n_k5_sig_270     <- sum(adj_p_270["K=5", ] < 0.05)

# -- H-cor values (from hcor matrices) ----------------------------------------
hcor_k3_a55_270  <- hcor_270["K=3", "a=0.55"]   # K=3 alpha=0.55, ntop=270
hcor_k5_a55_270  <- hcor_270["K=5", "a=0.55"]   # K=5 alpha=0.55, ntop=270
hcor_k5_a25_all  <- hcor_all["K=5", "a=0.25"]   # K=5 alpha=0.25, ntop=NULL

# -- K3xK7 nesting correlations (from k3_k7_summary) --------------------------
k3f1_best_r    <- k3k7_sum$k3f1_best_r
k3f1_best_k7   <- k3k7_sum$k3f1_best_k7
k3f2_best_r    <- k3k7_sum$k3f2_best_r
k3f2_best_k7   <- k3k7_sum$k3f2_best_k7
k3f3_range_min <- k3k7_sum$k3f3_range_min
k3f3_range_max <- k3k7_sum$k3f3_range_max
k7f4_range_min <- k3k7_sum$k7f4_range_min
k7f4_range_max <- k3k7_sum$k7f4_range_max

# -- Overfitting HR and LRT P-value ranges (high alpha, all-genes analysis) ----
high_alpha_rows  <- master_all[master_all$alpha >= 0.75, ]
hr_overfit_min   <- round(min(high_alpha_rows$hr_t_adj, na.rm = TRUE), 2)
hr_overfit_max   <- round(max(high_alpha_rows$hr_t_adj, na.rm = TRUE), 2)
lrt_overfit_min  <- min(high_alpha_rows$lrt_p_t, na.rm = TRUE)   # ~ 10^-60

# -- lambda=0.300 analysis scalars (from lam300_summary) -----------------------
n_sig_all_prior  <- lam300$n_sig
k3_p_all_prior   <- lam300$k3_alpha55_p
lam300_best_k    <- lam300$best_k
lam300_best_a    <- lam300$best_alpha
lam300_best_p    <- lam300$best_adj_p

# -- Production model adjusted P-value and validation counts -------------------
production_adj_p <- prod_sum$adj_p
n_val_total      <- prod_sum$n_val_total
n_val_cohorts    <- prod_sum$n_val_cohorts
n_genes_total    <- prod_sum$n_genes_total

# -- Grid dimensions -----------------------------------------------------------
n_combinations   <- length(k_labs) * length(alpha_labs)
```

# The DeSurv model shows consistent convergence and stability across restarts

Across datasets and initialization schemes, the DeSurv algorithm consistently converged to numerically stable solutions within the designated iteration budget. Warm-start strategies, in which solutions at $\alpha=0$ initialized supervised runs, substantially reduced variability across restarts and improved reproducibility. Fig. \ref{fig:fig-converge} shows representative loss trajectories demonstrating monotone decreases until convergence across restarts. Compared with naive random initialization, warm-starts produced tighter distributions of cross-validated C-index and partial likelihood, confirming stability of the optimization procedure.

```{r fig-converge, fig.width = 5, fig.height= 3, fig.cap="Convergence trajectories of the DeSurv objective across five randomly selected multi-start initializations on the PDAC training cohort (TCGA + CPTAC). Each line shows the relative loss (loss at iteration $t$ divided by the initial loss for that restart) as a function of iteration number, restricted to the first 5{,}000 iterations for readability. The total objective decreases monotonically and stabilizes well within the iteration budget across all initializations, confirming convergence of the alternating block-update scheme.", fig.env='figure*', fig.pos='htbp', out.height= "3in", out.width='5in'}
load_result("desurv_seed_fits_tcgacptac")
set.seed(147)
seeds = sample(1:length(desurv_seed_fits_tcgacptac$fits),5,replace = FALSE)
iters <- list()
for (i in seeds) {
  loss_vec <- desurv_seed_fits_tcgacptac$fits[[i]]$lossit
  iters[[i]] <- data.frame(
    lossit = loss_vec,
    slossit = desurv_seed_fits_tcgacptac$fits[[i]]$slossit,
    nlossit = desurv_seed_fits_tcgacptac$fits[[i]]$nlossit,
    iter = seq_along(loss_vec),
    init = i
  )
}

lossit <- dplyr::bind_rows(iters) |> dplyr::filter(iter < 5000)

seed_labels <- setNames(paste0("Init ", seq_along(seeds)), as.character(seeds))

lossit = lossit %>%
  group_by(init) %>%
  arrange(iter) %>%
  mutate(
    rel_delta_loss = lossit / dplyr::first(lossit),
    rel_delta_sloss = slossit / dplyr::first(slossit),
    rel_delta_nloss = nlossit / dplyr::first(nlossit),
    init_label = seed_labels[as.character(init)]
  ) %>%
  filter(!is.na(rel_delta_loss))

ggplot2::ggplot(lossit, ggplot2::aes(y = rel_delta_loss, x = iter, color = init_label)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::labs(x = "Iteration", y = "Loss normalized to initial value", color = "Initialization")

```


# Simulation results under null and mixed scenarios

To characterize the operating characteristics of DeSurv across different signal regimes, we evaluated performance under three simulation scenarios. In the primary scenario (main text Fig. 3), prognostic programs explained low variance relative to outcome-neutral background signals, and survival depended on 150 factor-specific marker genes ($\beta_1 = 2$, $\beta_2 = \beta_3 = 0$). In the null scenario, survival times were generated independently of gene expression ($\beta = 0$), so no prognostic structure existed. In the mixed scenario, survival depended on 300 genes per lethal factor, of which 50% were factor-specific markers and 50% were background genes with loadings shared across all factors. All scenarios used $G = 3{,}000$ genes, $N = 200$ samples, $K = 3$ true factors, and identical gamma distribution parameters for the gene loading matrix $W$ (see Materials and Methods). Performance was evaluated using cross-validated concordance index (C-index), precision of recovered prognostic gene sets, and the distribution of selected ranks across simulation replicates.

Under the null scenario (Fig. \ref{fig:sim-null-mixed}A--B), DeSurv showed no advantage over unsupervised NMF ($\alpha = 0$). Both methods yielded C-index values near 0.5, consistent with the absence of survival signal. The distribution of selected ranks was concentrated at low values for both methods, with NMF more strongly favoring $k = 2$. These results confirm that DeSurv does not hallucinate prognostic structure when none exists: Bayesian optimization selected low values of the supervision parameter $\alpha$, effectively recovering standard NMF as a special case. This specificity control is important because it establishes that the advantages observed in the primary scenario reflect genuine signal recovery rather than overfitting to noise in the survival labels.

Under the mixed scenario (Fig. \ref{fig:sim-null-mixed}C--F), four mechanistic observations characterize DeSurv's advantage. First, when precision is evaluated against the factor-specific marker genes alone (150 genes; Fig. \ref{fig:sim-null-mixed}D left), DeSurv substantially outperforms NMF---matching or exceeding primary-scenario levels---even though overall survival-gene precision (300-gene survival set including background genes; Fig. \ref{fig:sim-null-mixed}D right) is more modestly improved. This dissociation reveals that background-gene recovery is similar between methods, and DeSurv's edge comes specifically from concentrating marker programs. Second, the learned factor most aligned with the true marker program receives a substantially larger Cox coefficient ($|\beta|$) under DeSurv than under NMF (Fig. \ref{fig:sim-null-mixed}E), confirming that survival supervision preferentially amplifies the marker-specific signal. Third, C-index values are modestly higher for DeSurv (Fig. \ref{fig:sim-null-mixed}C), consistent with the moderate overall improvement when background-gene dilution attenuates the survival signal. Together, these results demonstrate that DeSurv's advantage in the mixed scenario is mechanistic: the method specifically recovers marker programs and assigns them elevated survival weights, rather than incidental capture of background-gene prognosis.

Together, the three scenarios provide a dose-response relationship: DeSurv's advantage is greatest when variance and prognosis diverge (primary), moderate when they partially overlap (mixed), and absent when no survival signal exists (null). The mechanistic basis for the attenuation in the mixed scenario is now clear: DeSurv continues to specifically recover marker programs (high marker precision, elevated $|\beta|$), but the overall survival-gene precision improvement is dampened because background-gene dilution in the 300-gene survival signal reduces the discrimination advantage. This gradient is consistent with the predictions of sufficient dimension reduction theory and offers practical guidance for users: DeSurv is most beneficial in settings where the dominant transcriptional axes are expected to be prognostically neutral, as is common in cancers with high tumor purity variation or dominant tissue-composition signals.

```{r fig-sim-null-mixed, fig.width = 6.5, fig.height = 7, fig.cap = "DeSurv performance under null and mixed simulation scenarios. (A--B) Null scenario ($\\beta = 0$): both methods yield C-index $\\approx 0.5$ (A) and comparable rank distributions (B), confirming DeSurv does not hallucinate prognostic structure; precision is omitted as no true survival genes exist. (C--F) Mixed scenario: survival depends on factor-specific marker genes (50\\%) and shared background genes (50\\%). (C) C-index is modestly higher for DeSurv. (D) Precision by gene-set definition: marker-only precision (150 genes; left facet) substantially favors DeSurv; full 300-gene survival-set precision (right facet) shows a more modest improvement. (E) The matched factor receives a substantially larger Cox coefficient $|\\beta|$ under DeSurv, confirming preferential amplification of marker-specific signal. (F) Selected-rank distributions. DeSurv (blue); NMF (red, $\\alpha = 0$); both Bayesian-optimized. \\label{fig:sim-null-mixed}", fig.env='figure*', fig.pos='t', out.height = "6.5in", out.width='\\textwidth'}
library(cowplot)
library(ggplot2)

load_result("sim_figs_by_scenario")

scenario_ids <- sapply(sim_figs_by_scenario, function(x) x$scenario_id)
analysis_ids <- sapply(sim_figs_by_scenario, function(x) x$analysis_id)

null_idx <- which(scenario_ids == "R00_null" & analysis_ids == "bo_tune_ntop")
mixed_idx <- which(scenario_ids == "R_mixed" & analysis_ids == "bo_tune_ntop")

null_plots <- sim_figs_by_scenario[[null_idx]]
mixed_plots <- sim_figs_by_scenario[[mixed_idx]]

null_plots <- lapply(null_plots, set_fig_font, size = 10)
mixed_plots <- lapply(mixed_plots, set_fig_font, size = 10)

# Null row: C-index and k-hist only (precision undefined when beta = 0)
null_cindex <- null_plots$cindex_box +
  labs(title = NULL) +
  scale_y_continuous(limits = c(0.3, 1))
null_cindex$data$scenario_id <- "Null scenario (\u03b2 = 0)"

null_khist <- null_plots$k_hist + labs(title = NULL)
null_khist$data$scenario_id <- "Null scenario (\u03b2 = 0)"

null_row <- plot_grid(
  null_cindex, null_khist,
  ncol = 2, labels = c("A", "B"), rel_widths = c(1, 1.2)
)

# Mixed row 1: C-index and precision breakdown (2-facet panel)
mixed_cindex <- mixed_plots$cindex_box +
  labs(title = NULL) +
  scale_y_continuous(limits = c(0.3, 1)) +
  ggplot2::theme(
    plot.clip = "off",
    plot.margin = ggplot2::margin(t = 5, r = 20, b = 5, l = 5, unit = "pt")
  )
mixed_cindex$data$scenario_id <- "Mixed scenario"

mixed_prec_breakdown <- if (!is.null(mixed_plots$precision_breakdown)) {
  mixed_plots$precision_breakdown + labs(title = NULL)
} else {
  # Fallback: use existing precision_box if new panel not available
  mixed_plots$precision_box + labs(title = NULL)
}

mixed_row1 <- plot_grid(
  mixed_cindex, mixed_prec_breakdown,
  ncol = 2, labels = c("C", "D"), rel_widths = c(1, 2)
)

# Mixed row 2: matched beta and k-hist
mixed_beta <- if (!is.null(mixed_plots$matched_beta_box)) {
  mixed_plots$matched_beta_box + labs(title = NULL)
} else {
  NULL
}

mixed_khist <- mixed_plots$k_hist + labs(title = NULL)
mixed_khist$data$scenario_id <- "Mixed scenario"

if (!is.null(mixed_beta)) {
  mixed_row2 <- plot_grid(
    mixed_beta, mixed_khist,
    ncol = 2, labels = c("E", "F"), rel_widths = c(1, 1.2)
  )
  plot_grid(null_row, mixed_row1, mixed_row2, nrow = 3, rel_heights = c(1, 1, 1))
} else {
  mixed_row2 <- plot_grid(
    mixed_khist,
    ncol = 1, labels = c("E")
  )
  plot_grid(null_row, mixed_row1, mixed_row2, nrow = 3, rel_heights = c(1, 1, 0.7))
}
```


# Cross-validated C-index across factorization rank

To evaluate how the number of latent factors $k$ affects prognostic performance, we performed an exhaustive grid search over $k \in \{2, 3, \ldots, 12\}$ and supervision strength $\alpha \in \{0, 0.05, \ldots, 0.95\}$ using 5-fold cross-validation on the TCGA+CPTAC training cohort. For each $k$, the best-performing $\alpha$ was selected by maximizing the mean CV C-index. The unsupervised NMF baseline ($\alpha = 0$) was evaluated at the same $k$ values for comparison. Results are shown for two gene-selection settings: $n_{\text{top}} = 270$ top genes per factor and $n_{\text{top}} = \text{ALL}$ (all genes retained), both at $\lambda = 0.349$ and $\xi = 0.056$.

Fig. \ref{fig:fig-cindex-by-k} shows the resulting C-index curves for both training (CV) and external validation (pooled across held-out cohorts), with shaded ribbons indicating $\pm 1$ standard error. In training, DeSurv consistently outperforms NMF across all $k$, confirming that survival supervision improves in-sample concordance. In external validation, DeSurv at $k = 3$ achieves concordance comparable to NMF at higher ranks (e.g., $k = 7$; Table \ref{tab:tab-cindex-comparison}), demonstrating that survival supervision concentrates prognostic information into fewer factors rather than requiring additional rank to capture it.

```{r fig-cindex-by-k, fig.width = 10, fig.height = 10, fig.cap="Training (CV) and external validation C-index as a function of factorization rank $k$ for DeSurv (blue) and standard NMF (red, $\\alpha = 0$). Shaded ribbons show $\\pm 1$ SE. For each $k$, the best $\\alpha$ was selected by maximizing the mean CV C-index. Validation C-index is pooled across all external cohorts. (A) Results using $n_{\\text{top}} = 270$ top genes per factor ($\\lambda = 0.349$, $\\xi = 0.056$). (B) Results retaining all genes ($n_{\\text{top}} = \\text{ALL}$; $\\lambda = 0.349$, $\\xi = 0.056$).", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
# This figure is produced by the cv_grid pipeline (separate store);
# read directly from the file output.
# Use magick to render the PDF as a raster (avoids LaTeX path issues)
# NOTE: DL flagged duplicate legend across panels A and B. The legend is baked
# into the pre-rendered PDF; consolidating it requires regenerating the source
# figure in inst/cv_grid_validation_analysis_amber_optimal.R on HPC.
img <- magick::image_read("figures/cv_grid/cv_cindex_by_k_primary.pdf", density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

# Standard NMF rank selection heuristics

Standard unsupervised rank selection criteria --- reconstruction residuals, cophenetic correlation coefficient, and mean silhouette width --- yielded inconsistent guidance for selecting the factorization rank $k$ in the PDAC training cohort (TCGA + CPTAC). Reconstruction residuals decreased smoothly without a clear elbow (Fig. \ref{fig:nmf-diagnostics}A). Cophenetic correlation fluctuated without a distinct transition point (Fig. \ref{fig:nmf-diagnostics}B). Mean silhouette width favored very low ranks that conflicted with other criteria (Fig. \ref{fig:nmf-diagnostics}C). These contradictions motivated the survival-supervised rank selection approach described in the main text.

```{r fig-nmf-diagnostics, fig.width = 6, fig.height = 3.5, fig.cap = "Standard unsupervised NMF rank selection heuristics for the PDAC training cohort (TCGA + CPTAC). (A) Reconstruction residuals as a function of factorization rank $k$. (B) Cophenetic correlation coefficient as a function of $k$. (C) Mean silhouette width across multiple distance metrics as a function of $k$. These criteria yield inconsistent guidance, motivating the survival-supervised model selection approach (main text Fig. 2D). \\label{fig:nmf-diagnostics}", fig.env='figure*', fig.pos='t', out.height= "3.5in", out.width='\\textwidth'}
library(cowplot)
library(ggplot2)

load_result("fig_residuals_tcgacptac")
load_result("fig_cophenetic_tcgacptac")
load_result("fig_silhouette_tcgacptac")
load_result("fit_std_tcgacptac")

fig_residuals_tcgacptac  <- set_fig_font(fig_residuals_tcgacptac, size = 10)
fig_cophenetic_tcgacptac <- set_fig_font(fig_cophenetic_tcgacptac, size = 10)
fig_silhouette_tcgacptac <- set_fig_font(fig_silhouette_tcgacptac, size = 10)

# Shared legend from fit_std plot
legend_plot <- plot(fit_std_tcgacptac) +
  ggplot2::theme(legend.position = "bottom")
legend_grob <- ggplotGrob(legend_plot)
legend <- legend_grob$grobs[
  sapply(legend_grob$grobs, function(x) x$name) == "guide-box"
][[1]]

panels <- plot_grid(
  fig_residuals_tcgacptac  + theme(legend.position = "none"),
  fig_cophenetic_tcgacptac + theme(legend.position = "none"),
  fig_silhouette_tcgacptac + theme(legend.position = "none"),
  ncol = 3, labels = c("A", "B", "C")
)

plot_grid(panels, legend, nrow = 2, rel_heights = c(1, 0.15))
```

# Cutpoint selection and validation of dichotomized risk groups

To convert the continuous DeSurv linear predictor into a clinically interpretable binary risk stratification, we selected an optimal z-score cutpoint via cross-validation on the TCGA+CPTAC training cohort. For each candidate cutpoint in $\{-2.0, -1.8, \ldots, 2.0\}$, we computed the mean absolute log-rank z-statistic across held-out folds and selected the cutpoint maximizing this metric (Fig. \ref{fig:fig-cutpoint-km}A). We then applied this cutpoint to external validation cohorts by standardizing each patient's linear predictor using the training mean and standard deviation. Fig. \ref{fig:fig-cutpoint-km}B shows the resulting Kaplan--Meier curves for the pooled external validation analysis, with a stratified log-rank test accounting for dataset of origin. Panels C--F display per-cohort validation KM curves for Dijk, Moffitt, PACA-AU, and Puleo, respectively. Across all cohorts, the DeSurv-derived high-risk group showed consistently shorter survival, confirming that the training-derived cutpoint generalizes to independent datasets.

```{r fig-cutpoint-km, fig.width = 10, fig.height = 11, fig.cap = "Cutpoint selection and external validation of dichotomized risk groups. (A) Cross-validated mean absolute log-rank z-statistic as a function of z-score cutpoint; red dashed line indicates the selected optimum. (B) Pooled validation Kaplan--Meier curves (stratified log-rank test). (C--D) Per-cohort validation KM curves: (C) Dijk, (D) Moffitt. (E--F) Per-cohort validation KM curves: (E) PACA-AU, (F) Puleo.", fig.env='figure*', fig.pos='t', out.height = "8in", out.width='\\textwidth', echo=FALSE}
# Production model cutpoint and KM validation PDFs (direct paths)
fig_dir <- normalizePath("figures")
km_dir  <- file.path(fig_dir, "km_dichot")

read_pdf_as_grob <- function(path) {
  img <- magick::image_read(path, density = 300)
  cowplot::ggdraw() + cowplot::draw_image(img)
}

p_cutpoint <- read_pdf_as_grob(file.path(fig_dir, "cutpoint_curve_logrank_tcgacptac.pdf"))
p_pooled   <- read_pdf_as_grob(file.path(km_dir, "km_val_pooled_logrank_tcgacptac.pdf"))
p_dijk     <- read_pdf_as_grob(file.path(km_dir, "km_val_Dijk_logrank_tcgacptac.pdf"))
p_moffitt  <- read_pdf_as_grob(file.path(km_dir, "km_val_Moffitt_GEO_array_logrank_tcgacptac.pdf"))
p_paca     <- read_pdf_as_grob(file.path(km_dir, "km_val_PACA_AU_logrank_tcgacptac.pdf"))
p_puleo    <- read_pdf_as_grob(file.path(km_dir, "km_val_Puleo_array_logrank_tcgacptac.pdf"))

row1 <- plot_grid(
  p_cutpoint, p_pooled,
  ncol = 2, labels = c("A", "B"), label_size = 14
)

row2 <- plot_grid(
  p_dijk, p_moffitt,
  ncol = 2, labels = c("C", "D"), label_size = 14
)

row3 <- plot_grid(
  p_paca, p_puleo,
  ncol = 2, labels = c("E", "F"), label_size = 14
)

plot_grid(row1, row2, row3, nrow = 3, rel_heights = c(1, 1, 1))
```

# Overlap of DeSurv risk groups with known molecular subtypes

To assess whether the DeSurv-derived risk stratification captures biologically meaningful transcriptional programs, we examined the overlap between dichotomized risk groups (High vs Low) and two established PDAC molecular classifiers: PurIST (Basal-like vs Classical) and DeCAF (proCAF vs restCAF). Fig. \ref{fig:fig-subtype-overlap} shows the composition of each risk group across all pooled external validation cohorts, with Fisher's exact test P-values quantifying the association. The DeSurv high-risk group was significantly enriched for both Basal-like (PurIST) and proCAF (DeCAF) subtypes, consistent with the known poor prognosis of these molecular classes. By contrast, standard NMF risk groups at the matched rank ($k = 3$) separated Basal-like from Classical tumors (PurIST) but showed no significant enrichment for proCAF versus restCAF (DeCAF), indicating that survival-unsupervised factorization does not recover the tumor--stroma coupling captured by DeSurv's D1. This concordance confirms that DeSurv's survival-driven factorization recovers clinically relevant biology without requiring subtype labels during training.

```{r fig-subtype-overlap, fig.width = 8, fig.height = 10, fig.cap = "Subtype composition of DeSurv (top) and standard NMF $k=3$ (bottom) risk groups in pooled external validation cohorts. Stacked bar charts show the proportion of PurIST (left) and DeCAF (right) subtypes within Low and High risk groups defined by the log-rank-optimized cutpoint. Numbers indicate sample counts; Fisher's exact test P-values are shown above each panel. DeSurv risk groups are enriched for both Basal-like (PurIST) and proCAF (DeCAF) poor-prognosis subtypes; NMF risk groups separate Basal-like from Classical tumors but show no significant CAF subtype enrichment.", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
# Composite DeSurv (top) and NMF k=3 (bottom) subtype overlap figures
img_desurv <- magick::image_read(
  "figures/km_dichot/subtype_overlap_pooled_logrank_tcgacptac.pdf", density = 300)
img_nmf <- magick::image_read(
  "figures/km_dichot/subtype_overlap_pooled_logrank_std_tcgacptac.pdf", density = 300)
img_combined <- magick::image_append(c(img_desurv, img_nmf), stack = TRUE)
cowplot::ggdraw() + cowplot::draw_image(magick::image_trim(img_combined))
```

<!-- ## Analysis of scRNA-seq {.unnumbered} -->
<!-- Next, we verify our findings at the cellular level using the Elyada scRNA-seq data [@elyada2019cross]. We found that our DeSurv factor 1 signature was expressed primarily in iCAF and B cells, factor 2 was expressed in Acinar and Classical PDAC 2, and factor 3 was expressed primarily in Basal-like PDAC cells (Figure \ref{fig:fig-sc}B-E). This is consistent with our ORA analysis which found that factor 1 was mostly immune and factor 3 was basal-like, and with the overlap we saw with published gene lists that identified factor 1 as iCAF/classical/immune and factor 3 as basal-like. Interestingly, we do not see a large overlap of factor 1 with the classical PDAC cell types despite some overlap with classical gene lists. Need more here... -->

<!-- ```{r fig-sc, fig.cap="A. Cell type clusters from the Elyada scRNA-seq data. B-D. VAM scores for the DeSurv factor signatures in each cell shown on the UMAP of cells in the Elyada-sc data. E. Heatmap of average VAM scores by cell type and factor signature in the Elyada-sc data.", fig.env='figure*', fig.pos='t', out.height= "8in", out.width='\\textwidth'} -->

<!-- tar_load(fig_sc_plot_tcgacptac) -->
<!-- fig_sc_plot_tcgacptac -->
<!-- ``` -->

# Standard NMF factor structure at independently selected rank

Throughout the main text, both DeSurv and standard NMF were fit at the same rank ($k = 3$), enabling a controlled comparison of the effect of survival supervision on factor structure. When each method selects its own rank independently, standard NMF chooses $k = 5$ via residual elbow detection or $k = 7$ via Bayesian optimization with $\alpha = 0$, while supervised DeSurv BO selects $k = 3$. We examine here the gene program structure at elbow-selected $k = 5$ for both DeSurv (Fig. \ref{fig:fig-desurv-k5-heatmap}) and standard NMF (Fig. \ref{fig:fig-nmf-k5-heatmap}), and the NMF $k = 7$ structure to illustrate additional fragmentation when supervision is absent (Fig. \ref{fig:fig-nmf-k7-heatmap}).

At elbow-selected $k = 5$, supervised and unsupervised factorizations yield substantially different gene-program structures. DeSurv at $k = 5$ (Fig. \ref{fig:fig-desurv-k5-heatmap}) splits the ECM/activated-stroma (proCAF-associated) signal into two factors with opposing tumor correlations --- one positively correlated with Classical tumor programs, one negatively correlated --- and isolates restCAF as a separate factor; exocrine and Basal-like programs are not represented as dominant factors in the supervised solution. Standard NMF at $k = 5$ (Fig. \ref{fig:fig-nmf-k5-heatmap}), by contrast, recovers the three core programs from its $k = 3$ solution and adds two factors capturing Exocrine-like and iCAF variance --- transcriptional variance not aligned with survival outcomes. These biologically redundant NMF factors explain why supervised BO selects $k = 3$ in the production model: the additional NMF factors at $k = 5$ contribute no independent prognostic content beyond the three core programs (Fig. \ref{fig:varsurvival-k5}), even though the overall higher-rank model still validates externally on per-cohort C-index (Table \ref{tab:tab-cindex-comparison}).

```{r fig-desurv-k5-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene program overlap heatmap for DeSurv at elbow-selected rank $k = 5$. Each cell shows the number of top genes from a given factor (rows) overlapping established PDAC gene programs (columns). DeSurv at $k = 5$ splits the ECM/activated-stroma (proCAF-associated) signal into two factors with opposing Classical-tumor correlations and isolates restCAF as a separate factor; exocrine and Basal-like programs are not represented as dominant factors. \\label{fig:fig-desurv-k5-heatmap}", fig.env='figure', fig.pos='t'}
library(cowplot)
load_result("fig_gene_overlap_heatmap_desurv_elbowk_tcgacptac")

cowplot::plot_grid(
  fig_gene_overlap_heatmap_desurv_elbowk_tcgacptac$plot,
  cowplot::ggdraw(fig_gene_overlap_heatmap_desurv_elbowk_tcgacptac$legend),
  ncol = 2, rel_widths = c(4, 1)
)
```

```{r fig-nmf-k5-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene program overlap heatmap for standard NMF at elbow-selected rank $k = 5$. Each cell shows the number of top genes from a given factor (rows) overlapping established PDAC gene programs (columns). The three core programs from $k = 3$ are recovered; two additional factors overlap Exocrine-like and iCAF signatures, capturing transcriptional variance not aligned with survival outcomes. \\label{fig:fig-nmf-k5-heatmap}", fig.env='figure', fig.pos='t'}
library(cowplot)
load_result("fig_gene_overlap_heatmap_std_elbowk_tcgacptac")

cowplot::plot_grid(
  fig_gene_overlap_heatmap_std_elbowk_tcgacptac$plot,
  cowplot::ggdraw(fig_gene_overlap_heatmap_std_elbowk_tcgacptac$legend),
  ncol = 2, rel_widths = c(4, 1)
)
```

At BO-selected $k = 7$ under $\alpha = 0$ (Fig. \ref{fig:fig-nmf-k7-heatmap}), standard NMF shows increased fragmentation relative to $k = 5$: multiple factors partially overlap the same PDAC gene signatures, and no new biologically coherent program emerges beyond those present at $k = 3$.

```{r fig-nmf-k7-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene program overlap heatmap for standard NMF at $k = 7$ (BO-selected rank with $\\alpha = 0$). Compared with $k = 3$ and $k = 5$, the 7-factor solution shows increased fragmentation: multiple factors partially capture the same PDAC gene signatures and no new coherent biological program emerges. \\label{fig:fig-nmf-k7-heatmap}", fig.env='figure', fig.pos='t'}
library(cowplot)
load_result("fig_gene_overlap_heatmap_desurv_alpha0_tcgacptac")
cowplot::plot_grid(
  fig_gene_overlap_heatmap_desurv_alpha0_tcgacptac$plot,
  cowplot::ggdraw(fig_gene_overlap_heatmap_desurv_alpha0_tcgacptac$legend),
  ncol = 2, rel_widths = c(4, 1)
)
```

```{r tab-cindex-comparison, echo=FALSE}
library(dplyr)
library(knitr)
load_result("val_cindex_desurv_tcgacptac")
load_result("val_cindex_std_desurvk_tcgacptac")
load_result("val_cindex_std_elbowk_tcgacptac")
load_result("val_cindex_desurv_alpha0_tcgacptac")

fmt_cindex <- function(tbl, col_name) {
  tbl %>%
    dplyr::select(dataset, cindex) %>%
    dplyr::rename(!!col_name := cindex)
}

cindex_wide <- fmt_cindex(val_cindex_desurv_tcgacptac, "DeSurv (K=3)") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_desurvk_tcgacptac, "Std NMF (K=3)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_elbowk_tcgacptac, "Std NMF (K=5, elbow)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_desurv_alpha0_tcgacptac, "Std NMF (K=7, BO $\\alpha$=0)"), by = "dataset") %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))) %>%
  dplyr::rename(Cohort = dataset)

knitr::kable(
  cindex_wide,
  caption = "External validation C-index by method and cohort. DeSurv at BO-selected rank ($k = 3$) is compared with standard NMF at matched rank ($k = 3$), elbow-selected rank ($k = 5$), and BO-selected rank ($k = 7$, $\\alpha = 0$). Higher values indicate better discrimination. \\label{tab:tab-cindex-comparison}",
  label = "tab-cindex-comparison",
  booktabs = TRUE,
  linesep = "",
  escape = FALSE
)
```

Standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ both substantially improve over NMF at $k = 3$ across all cohorts (Table \ref{tab:tab-cindex-comparison}), confirming that additional unsupervised factors capture prognostically relevant variation missed at the matched rank. At $k = 7$, NMF matches or exceeds DeSurv's per-cohort C-index in most cohorts, demonstrating that unsupervised factorization can approach supervised concordance given enough factors. However, this comes at the cost of a less parsimonious and less interpretable factorization: the $k = 7$ NMF solution shows increased fragmentation (Fig. \ref{fig:fig-nmf-k7-heatmap}). DeSurv's advantage thus combines biological concordance and methodological complementarity: its risk groups identify the same poor-prognosis subtypes as established classifiers (Basal-like, proCAF; Fig. \ref{fig:fig-subtype-overlap}), while its linear predictor retains significant prognostic value after adjustment for those classifiers (Table \ref{tab:tab-hr-adjusted}).

Table \ref{tab:tab-hr-adjusted} reports pooled validation hazard ratios (per SD of the linear predictor) for DeSurv at $k = 3$ and the BO-selected NMF model at $k = 7$ ($\alpha = 0$), both unadjusted and after adjustment for PurIST and DeCAF molecular classifiers. Both methods retain significant prognostic value after adjustment, with DeSurv's larger HR drop reflecting biological concordance with the established subtypes (per Fig. \ref{fig:fig-subtype-overlap}) and NMF $k = 7$'s smaller drop reflecting residual content from its additional, less interpretable factors. The NMF $k = 7$ model selects its rank via DeSurv's BO infrastructure (with $\alpha = 0$, no factorization-level supervision); DeSurv uses survival at both levels, concentrating the prognostic signal into three factors rather than seven.

```{r tab-hr-adjusted, echo=FALSE, results='asis'}
library(survival)
library(knitr)
library(kableExtra)

# -- Load validation latent targets --------------------------------------------
load_result("val_latent_desurv_tcgacptac")
load_result("val_latent_desurv_alpha0_tcgacptac")
load_result("data_val_filtered_tcgacptac")

# -- Extract pooled risk_score + survival + PurIST/DeCAF -----------------------
extract_pool_df_with_covariates <- function(val_latent, data_val) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
    df <- data.frame(
      sample_id = rownames(surv),
      time = surv$time, event = surv$event,
      dataset = entry$dataset, risk_score = entry$risk_score,
      stringsAsFactors = FALSE
    )
    # Merge PurIST/DeCAF from validation data
    ds_match <- Filter(function(d) d$sampInfo$dataset[1] == entry$dataset, data_val)
    if (length(ds_match) > 0) {
      si <- ds_match[[1]]$sampInfo
      si$sample_id <- rownames(si)
      covars <- si[, intersect(c("sample_id", "PurIST", "DeCAF"), names(si)), drop = FALSE]
      if ("DeCAF" %in% names(covars)) {
        covars$DeCAF[covars$DeCAF == "permCAF"] <- "proCAF"
      }
      df <- merge(df, covars, by = "sample_id", all.x = TRUE)
    }
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time) & is.finite(df$risk_score), ]
}

# -- Fit unadjusted + PurIST/DeCAF-adjusted stratified Cox --------------------
pooled_cox_unadj_adj <- function(df) {
  rs_sd <- sd(df$risk_score, na.rm = TRUE)
  if (is.na(rs_sd) || rs_sd < 1e-12) rs_sd <- 1
  df$risk_z <- df$risk_score / rs_sd
  # Unadjusted
  cox_u <- summary(coxph(Surv(time, event) ~ risk_z + strata(dataset), data = df))
  # Adjusted for PurIST + DeCAF
  df_adj <- df[complete.cases(df[, c("PurIST", "DeCAF")]), ]
  cox_a <- summary(coxph(Surv(time, event) ~ risk_z + PurIST + DeCAF + strata(dataset),
                          data = df_adj))
  fmt <- function(cx) {
    coef_row <- cx$coefficients["risk_z", , drop = FALSE]
    ci_row   <- cx$conf.int["risk_z", , drop = FALSE]
    list(
      hr    = round(ci_row[, "exp(coef)"], 2),
      ci_lo = round(ci_row[, "lower .95"], 2),
      ci_hi = round(ci_row[, "upper .95"], 2),
      p     = coef_row[, "Pr(>|z|)"]
    )
  }
  list(unadj = fmt(cox_u), adj = fmt(cox_a))
}

fmt_p <- function(p) {
  if (p < 0.001) return("$< 0.001$")
  paste0("$", formatC(p, format = "f", digits = 3), "$")
}
fmt_hr <- function(s) {
  paste0(sprintf("%.2f", s$hr), " (",
         sprintf("%.2f", s$ci_lo), "--",
         sprintf("%.2f", s$ci_hi), "), ", fmt_p(s$p))
}

# -- Compute -------------------------------------------------------------------
pool_desurv <- extract_pool_df_with_covariates(
  val_latent_desurv_tcgacptac, data_val_filtered_tcgacptac)
stats_desurv <- pooled_cox_unadj_adj(pool_desurv)

pool_nmf7 <- extract_pool_df_with_covariates(
  val_latent_desurv_alpha0_tcgacptac, data_val_filtered_tcgacptac)
stats_nmf7 <- pooled_cox_unadj_adj(pool_nmf7)

tab_df <- data.frame(
  Method = c("DeSurv ($k = 3$)", "Std NMF ($k = 7$, BO $\\alpha = 0$)"),
  Unadjusted = c(fmt_hr(stats_desurv$unadj), fmt_hr(stats_nmf7$unadj)),
  Adjusted   = c(fmt_hr(stats_desurv$adj), fmt_hr(stats_nmf7$adj)),
  stringsAsFactors = FALSE
)
colnames(tab_df) <- c("Method", "Unadjusted HR per SD (95\\% CI), $P$",
                       "Adjusted HR per SD (95\\% CI), $P$")

kable(tab_df, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = "Pooled external validation hazard ratios (per SD of linear predictor) for DeSurv at $k = 3$ and standard NMF at BO-selected $k = 7$ ($\\alpha = 0$). Adjusted models include PurIST and DeCAF molecular classifiers as covariates, with stratification by dataset.") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

The variance--survival decomposition at $k = 5$ (Fig. \ref{fig:varsurvival-k5}) confirms that elbow selection does not recover additional prognostic structure. At $k = 5$, standard NMF continues to distribute survival contribution negligibly across all factors; the two factors new relative to $k = 3$ (N4 and N5) contribute minimally, confirming that the additional rank captures background transcriptional variance rather than independent prognostic programs.

```{r fig-varsurvival-k5, fig.width = 4.5, fig.height = 3.5, fig.cap = "Fraction of expression variance explained versus Type III partial likelihood contribution for each standard NMF factor at elbow-selected $k = 5$. Each point represents one factor, labeled N1--N5. Survival contribution is quantified as the reduction in Cox partial log-likelihood when each factor is removed from the full model. The two factors new relative to $k = 3$ (N4, N5) contribute negligibly to survival, confirming that elbow selection recovers no additional prognostic structure. \\label{fig:varsurvival-k5}", fig.env='figure', fig.pos='t'}
library(survival)
library(dplyr)
library(ggrepel)
load_result("fit_std_elbowk_tcgacptac")
fit_std_elbowk=fit_std_elbowk_tcgacptac
load_result("tar_data_filtered_elbowk_tcgacptac")
tar_data_filtered_elbowk=tar_data_filtered_elbowk_tcgacptac

compute_variance_explained_inline <- function(X, W, H) {
  X <- as.matrix(X); W <- as.matrix(W); H <- as.matrix(H)
  total_ss   <- sum(X^2, na.rm = TRUE)
  X_hat_full <- W %*% H
  rss_full   <- sum((X - X_hat_full)^2, na.rm = TRUE)
  tibble::tibble(
    factor = seq_len(ncol(W)),
    variance_explained = purrr::map_dbl(seq_len(ncol(W)), function(j) {
      X_hat_mj <- W[, -j, drop = FALSE] %*% H[-j, , drop = FALSE]
      rss_mj   <- sum((X - X_hat_mj)^2, na.rm = TRUE)
      (rss_mj - rss_full) / total_ss
    })
  )
}

compute_survival_explained_inline <- function(X, W, time, event) {
  keep     <- intersect(rownames(X), rownames(W))
  XtW      <- t(X[keep, ]) %*% W[keep, ]
  full_fit <- coxph(Surv(time, event) ~ XtW)
  ll_full  <- full_fit$loglik[2]
  tibble::tibble(
    factor = seq_len(ncol(XtW)),
    delta_loglik = purrr::map_dbl(seq_len(ncol(XtW)), function(j) {
      XtW_mj  <- XtW[, -j, drop = FALSE]
      red_fit <- if (ncol(XtW_mj) == 0) {
        coxph(Surv(time, event) ~ 1)
      } else {
        coxph(Surv(time, event) ~ XtW_mj)
      }
      ll_full - red_fit$loglik[2]
    })
  )
}

X     <- tar_data_filtered_elbowk$ex
time  <- tar_data_filtered_elbowk$sampInfo$time
event <- tar_data_filtered_elbowk$sampInfo$event

df_nmf <- dplyr::left_join(
  compute_variance_explained_inline(X, fit_std_elbowk$W, fit_std_elbowk$H),
  compute_survival_explained_inline(X, fit_std_elbowk$W, time, event),
  by = "factor"
) |> dplyr::mutate(method = "NMF", factor_label = paste0("N", factor))

ggplot2::ggplot(df_nmf, ggplot2::aes(x = variance_explained, y = delta_loglik,
                                      label = factor_label)) +
  ggplot2::geom_point(size = 4, color = "red") +
  ggrepel::geom_text_repel(size = 4, max.overlaps = Inf, box.padding = 0.4,
                            point.padding = 0.3, segment.size = 0.3) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(
    x = "Conditional variance explained\n(semi-partial R\u00b2)",
    y = expression(atop(Delta ~ "partial log-likelihood", "(full vs. k-1 factor model)"))
  ) +
  ggplot2::theme_classic(base_size = 10)
```

\section{Factor-Specific Gene Lists}

Tables \ref{tab:gene-list-270-D1}--\ref{tab:gene-list-270-D3} list the top 270 genes for each of the three DeSurv factors, ranked by the factor specificity score $s_{ij}$ defined in SI Appendix, Section \ref{sec:gene-program-correlation}. The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\text{top}} = 270$); these are the genes used to compute sample-level factor scores for survival prediction. Factor D1 is enriched for Classical tumor identity genes, Factor D2 for stromal and immune-related genes, and Factor D3 for Basal-like tumor and proliferation-associated genes.

```{r tab-gene-list-270, echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)
library(tidyr)
library(dplyr)

genes <- read.csv("paper/gene_lists_top270_k3.csv", stringsAsFactors = FALSE)

# Keep only DeSurv factors
desurv_genes <- genes %>%
  dplyr::filter(factor %in% c("D1", "D2", "D3"))

# For each factor, split 270 genes into 6 columns of 45
build_factor_table <- function(df, fct) {
  g <- df %>% dplyr::filter(factor == fct) %>% dplyr::arrange(rank) %>% dplyr::pull(gene)
  data.frame(
    a = g[1:45], b = g[46:90], c = g[91:135],
    d = g[136:180], e = g[181:225], f = g[226:270],
    stringsAsFactors = FALSE
  )
}

factor_labels <- c("D1", "D2", "D3")
factor_captions <- c(
  "Factor D1 (Classical tumor)",
  "Factor D2 (stromal/immune)",
  "Factor D3 (Basal-like tumor)"
)

for (i in seq_along(factor_labels)) {
  tbl <- build_factor_table(desurv_genes, factor_labels[i])
  cat("")
  kable(tbl, format = "latex", booktabs = TRUE, linesep = "",
        col.names = c("1--45", "46--90", "91--135", "136--180", "181--225", "226--270"),
        caption = paste0("Top 270 genes for DeSurv factor ", factor_labels[i],
          " (", factor_captions[i], "), ranked by factor specificity score $s_{ij}$.",
          if (i == 1) paste0(" The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\\text{top}}$).") else "",
          " \\label{tab:gene-list-270-", factor_labels[i], "}"),
        escape = FALSE) %>%
    kable_styling(font_size = 8, latex_options = c("HOLD_position")) %>%
    add_header_above(setNames(6, factor_captions[i]), bold = TRUE) %>%
    cat()
  cat("")
}
```
