---
title: "Supplementary Information"
subtitle: "Survival-supervised deconvolution identifies a reproducible tumor--stroma prognostic axis in pancreatic cancer"
author: Amber M. Young, Alisa Yurovsky, Xianlu Laura Peng, Jen Jen Yeh, Didong Li, Naim U. Rashid
header-includes:
- \providecommand{\pandocbounded}[1]{#1}
- \usepackage{booktabs}
- \usepackage{multirow}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{xr-hyper}
- \externaldocument{paper}
- \usepackage{amsmath}
- \usepackage{amsfonts}
- \usepackage{amssymb}
- \usepackage{bbm}
- \usepackage{algorithm}
- \usepackage{algpseudocode}
- \usepackage{setspace}
- \usepackage{enumitem}
- \usepackage[nolists,nomarkers]{endfloat}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator\diag{diag}
- \renewcommand{\algorithmicrequire}{\textbf{Input:}}
- \renewcommand{\algorithmicensure}{\textbf{Output:}}
- \renewcommand{\thefigure}{S\arabic{figure}}
- \renewcommand{\thetable}{S\arabic{table}}
- \renewcommand{\thealgorithm}{S\arabic{algorithm}}
bibliography: references_30102025.bib
csl: nature.csl
link-citations: true
reference-section-title: "SI References"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    keep_tex: true
  html_document: default
---
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\newpage

\part{Supplementary Methods}

\section{Model details}
Let $X \in \mathbb{R}_{\ge 0}^{p \times n}$ denote the nonnegative gene
expression matrix with $p$ genes and $n$ subjects. DeSurv approximates
$X \approx WH$ with $W \in \mathbb{R}_{\ge 0}^{p \times k}$ and
$H \in \mathbb{R}_{\ge 0}^{k \times n}$, and links the shared program matrix
$W$ to survival via an elastic-net–penalized Cox model. Supervision is placed on the program matrix $W$, the basis that transfers to new cohorts, whereas the sample loadings $H$ serve only to reconstruct the training expression matrix; new samples are therefore scored by projection onto the learned programs, $Z = W^\top X$, rather than by re-estimating loadings.

Let $y \in \mathbb{R}^n_{> 0}$ be the observed survival times,
$\delta \in \{0,1\}^n$ the event indicators, and
$Z = W^\top X \in \mathbb{R}^{k \times n}$ the sample-wise factor scores. We write $Z_i$ for the $i$th column of $Z$, and
$n_{\text{event}} = \sum_{i=1}^n \delta_i$. For convenience we recall the
DeSurv loss from Equation 1 in the main text and rewrite it as
\begin{align}
  \mathcal{L}(W,H,\beta) &=\nonumber \\
  &(1-\alpha)\left(\frac{1}{2\lVert X \rVert_F^2}\lVert X - WH \rVert_F^2 + \frac{\lambda_H}{2nk} \lVert H \rVert_F^2\right) \nonumber \\
  -&\alpha \left(\frac{2}{n_{event}}\ell(W,\beta) - \lambda \{ \xi \lVert \beta \rVert_1 + \tfrac{1-\xi}{2} \lVert \beta \rVert_2^2 \} \right) \nonumber
\end{align}

where $\lVert \cdot \rVert_F$ is the Frobenius norm and
$\beta \in \mathbb{R}^k$ are the Cox coefficients. The Cox log-partial
likelihood $\ell(W,\beta)$ is
\begin{equation}
\ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left[
  Z_i^\top \beta
  -
  \log \left(
    \sum_{j : y_j \ge y_i}
    \exp(Z_j^\top \beta)
  \right)
\right].
\end{equation}
The inner sum runs over the risk set $\{j : y_j \ge y_i\}$ of subjects still at risk at the observed time $y_i$; tied event times are handled by the Breslow approximation in the \texttt{glmnet} Coxnet routine used for the $\beta$ update (below).
Hyperparameters satisfy $\alpha \in [0,1)$,
$\lambda_H \ge 0$, $\lambda \ge 0$, and $\xi \in [0,1]$.
The constants $1/(2\lVert X \rVert_F^2)$, $2/n_{\text{event}}$, and $1/(2nk)$ place the reconstruction, loadings-penalty, and survival terms on comparable numerical scales for stable optimization.


\section{Optimization Algorithm}

\subsection{Block coordinate descent scheme}
Optimization proceeds by block coordinate descent (BCD; Algorithm \ref{alg:desurv}), which at each outer iteration updates $H$, then $W$, then $\beta$ while holding the other blocks fixed.

\begin{algorithm}[htbp]
\caption{DeSurv block coordinate descent}
\label{alg:desurv}
    \begin{algorithmic}[1]
        \Require $X \in \mathbb{R}_{\geq 0}^{p \times n}$, survival times $y \in \mathbb{R}^n_{> 0}$, event indicators $\delta \in \{0,1\}^{n}$, tolerance $tol$, maximum iterations $maxit$, supervision parameter $\alpha$, rank $k$, penalties $\lambda_H$, $\lambda$, and $\xi$
        \Ensure Fitted DeSurv parameters $(W,H,\beta)$

        \State Initialize $W^{(0)}, H^{(0)}$ with positive entries (e.g., $W^{(0)},H^{(0)} \sim \mathrm{Uniform}(0,\max X)$)
        \State Initialize $\beta^{(0)}$ (e.g., $\beta^{(0)} = \mathbf{1}$)
        \State $loss = \mathcal{L}(W^{(0)},H^{(0)},\beta^{(0)})$
        \State $eps = \infty$, $t = 0$
        \While{$eps \geq tol$ \textbf{and} $t < maxit$}
            \State \textbf{(H-update)}
            \State \hspace{0.5cm} Update $H^{(t+1)}$ from $H^{(t)}$ using the multiplicative rule in Eq. \ref{eqn:Hupdate},
            \State \hspace{0.5cm} holding $W^{(t)}$ and $\beta^{(t)}$ fixed.
            \State \textbf{(W-update)}
            \State \hspace{0.5cm} Update $W^{(t+1)}$ from $W^{(t)}$ using the hybrid multiplicative rule in Eq. \ref{eqn:W_update},
            \State \hspace{0.5cm} holding $H^{(t+1)}$ and $\beta^{(t)}$ fixed, and perform backtracking to ensure
            \State \hspace{0.5cm} $\mathcal{L}$ does not increase.
            \State \textbf{($\beta$-update)}
            \State \hspace{0.5cm} Update $\beta^{(t+1)}$ from $\beta^{(t)}$ using a Newton-like step for the Cox loss
            \State \hspace{0.5cm} (Eq. \ref{eqn:beta_update}), holding $W^{(t+1)}$ and $H^{(t+1)}$ fixed.
            \State $lossNew = \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})$
            \State $eps = |lossNew - loss| / |loss|$
            \State $loss = lossNew$
            \State $t = t + 1$
        \EndWhile
        \State \Return $\hat{W} = W^{(t)}, \hat{H} = H^{(t)}, \hat{\beta}=\beta^{(t)}$
    \end{algorithmic}
\end{algorithm}

\subsection{Update for $H$}
Conditional on $W$, the loss $\mathcal{L}(W,H,\beta)$ reduces to a strictly
convex quadratic function of $H$. We adopt the standard NMF multiplicative
update with $\ell_2$ penalty:
\begin{equation}
\label{eqn:Hupdate}
H \leftarrow
\max\!\left(
  H \odot \frac{W^\top X}{
    W^\top W H + \dfrac{\lVert X \rVert_F^2}{nk}\,\lambda_H H + \varepsilon_H
  },
  \varepsilon_H
\right),
\end{equation}
where $\odot$ denotes elementwise multiplication, all divisions are
elementwise, and $\varepsilon_H > 0$ is a small floor to prevent entries from
becoming exactly zero. This update is equivalent to a majorization–minimization
step and guarantees a nonincreasing reconstruction term conditional on $W$
[@seung2001algorithms; @pascualmontano2006nonsmooth]. The $\varepsilon_H$-floor
ensures that iterates remain in the interior of the nonnegative orthant, which
simplifies the convergence analysis.


\subsection{Update for $W$}
For $W$, we construct a hybrid multiplicative update that balances the
contributions of the NMF and Cox gradients. Let
\[
\nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H)
=
\frac{1}{\lVert X \rVert_F^2}(W H - X)H^\top,
\]
and let
\[
\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta)
=
\frac{2}{n_{event}}\nabla_W \ell(W,\beta).
\]
where $\nabla_W \ell(W,\beta)$ denotes the Cox gradient with respect to $W$. A closed-form expression for $\nabla_W \ell$ is
\begin{equation}
\nabla_W \ell(W,\beta)
=
\sum_{i=1}^n
\delta_i
\left(
  x_i
  -\frac{
      \sum_{j : y_j \ge y_i}
        x_j \exp(x_j^\top W \beta)
    }{
      \sum_{j : y_j \ge y_i}
        \exp(x_j^\top W \beta)
    }
\right)\beta^\top,
\end{equation}
where $x_i$ denotes the $i$th column of $X$.

We define a gradient-scaling heuristic
\begin{equation}
\label{eq:supp-zeta}
\zeta^{(t)}
=
\frac{
  \left\lVert (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W^{(t)},H^{(t+1)}) \right\rVert_F^2
}{
  \left\lVert \alpha \nabla_W \mathcal{L}_{Cox}(W^{(t)},\beta^{(t)}) \right\rVert_F^2
},
\end{equation}
and clip $\zeta^{(t)}$ to avoid extreme values:
$\zeta^{(t)} \leftarrow \min(\zeta^{(t)}, \zeta_{\max})$
with $\zeta_{\max} = 10^6$ in all experiments.

The multiplicative factor for $W$ is then
\begin{equation}
R^{(t)}
=
\frac{
  \frac{(1-\alpha)}{\lVert X \rVert_F^2} X H^{(t+1)\top}
  +\frac{2\alpha}{n_{\mathrm{event}}}\,\zeta^{(t)}\,\nabla_W \ell(W^{(t)},\beta^{(t)})
}{
  \frac{(1-\alpha)}{\lVert X \rVert_F^2} W^{(t)} H^{(t+1)}H^{(t+1)\top}
  +\varepsilon_W
},
\end{equation}
and the proposed update is
\begin{equation}
\label{eqn:W_update}
W^{(t+1)}
=
\max\!\left(
  W^{(t)} \odot R^{(t)},
  \varepsilon_W
\right),
\end{equation}
with a small floor $\varepsilon_W > 0$. When $\alpha=0$, this reduces
to the standard multiplicative update for NMF [@seung2001algorithms];
for $\alpha>0$, the Cox gradient perturbs the update in a direction that
decreases the supervised loss.

Because the combined multiplicative step does not, by itself, guarantee a
nonincrease in the full loss $\mathcal{L}$, we embed it in a backtracking
line search.

\begin{algorithm}[H]
\caption{$W$ update with backtracking}
\label{alg:backtrack}
    \begin{algorithmic}[1]
        \Require Value of $W$ and the previous iteration $t$: $W^{(t)}$, the multiplicative update term at the current iteration $R^{(t)}$, the max number of backtracking updates $max\_bt$, the backtracking parameter $\rho \in (0,1)$
        \Ensure $W^{(t+1)}$
        \State $\theta =1$
        \State b = 1
        \State $flag\_accept = FALSE$
        \While{$b \leq max\_bt$}
            \State $W^{(t+1)}_{cand} = \max\!\left( W^{(t)} \odot [(1-\theta) + \theta R^{(t)}],\ \varepsilon_W \right)$
            \State \textbf{(Column normalization)}
            \State \hspace{0.5cm} Compute $D = \mathrm{diag}\!\big(\lVert W^{(t+1)}_{(:,1)cand}\rVert_2,\dots,\lVert W^{(t+1)}_{(:,k)cand}\rVert_2\big)$
            \State \hspace{0.5cm} and set $W^{(t+1)}_{cand} \leftarrow W^{(t+1)}_{cand} D^{-1}$, $H^{(t+1)}_{cand} \leftarrow D H^{(t+1)}$, and $\beta^{(t)}_{cand} \leftarrow D \beta^{(t)}$
            \If{$\mathcal{L}(W^{(t+1)}_{cand},H^{(t+1)},\beta^{(t)}) \leq \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)})$}
              \State $W^{(t+1)} = W^{(t+1)}_{cand}$
              \State $H^{(t+1)} = H^{(t+1)}_{cand}$
              \State $\beta^{(t)} = \beta^{(t)}_{cand}$
              \State $flag\_accept = TRUE$
              \State break
            \EndIf
            \State $\theta = \theta*\rho$
            \State $b = b + 1$
        \EndWhile
        \If{$flag\_accept = FALSE$}
            \State $W^{(t+1)}=W^{(t)}$
        \EndIf
    \end{algorithmic}
\end{algorithm}

The column normalization preserves both $WH$ and $W\beta$, and therefore
leaves the loss $\mathcal{L}$ invariant up to numerical error. Backtracking
guarantees that the accepted $W$ update does not increase $\mathcal{L}$.

\subsection{Update for $\beta$}

Conditional on $(W,H)$, the loss in $\beta$ reduces to a convex
elastic-net–penalized Cox problem; the common factor $\alpha > 0$ multiplies both
the log-partial-likelihood and the penalty in the joint loss (Equation 1) and is
dropped here, as it does not affect the minimizer:
\[
\min_{\beta \in \mathbb{R}^k}
\left\{
  -\frac{2}{n_{\mathrm{event}}}\,\ell(W,\beta)
  +
  \lambda\left(
    \xi \lVert \beta \rVert_1
    +\frac{1-\xi}{2} \lVert \beta \rVert_2^2
  \right)
\right\}.
\]
We solve this subproblem by cyclic coordinate descent following
[@simon2011regularization]. Writing $\ell(\beta) = \ell(W,\beta)$ and
$\tilde{\eta}_i = Z_i^\top \beta$, the update for coordinate $r$ has the
closed form
\begin{equation}
\label{eqn:beta_update}
\hat{\beta}_r
=
\frac{
  S\!\left(
    \dfrac{1}{n}
    \sum_{i=1}^n w(\tilde{\eta})_i
    z_{i,r}
    \left[
      v(\tilde{\eta})_i - \sum_{j \ne r} z_{i,j} \beta_j
    \right],
    \lambda \xi
  \right)
}{
  \frac{1}{n} \sum_{i=1}^n w(\tilde{\eta})_i z_{i,r}^2 + \lambda(1-\xi)
},
\end{equation}
where $S(a,\tau) = \mathrm{sign}(a)\max(|a|-\tau,0)$ is the soft-thresholding
operator, and $w(\tilde{\eta})$, $v(\tilde{\eta})$ are standard local
quadratic approximations of the Cox log-partial likelihood [@simon2011regularization].
We iterate coordinate updates until the subproblem converges to numerical
tolerance, yielding a minimizer in $\beta$ for the current $W$.


\subsection{Normalization of W}

After each accepted $W$ update, we normalize the columns of $W$ as
$W \leftarrow W D^{-1}$, and adjust $H$ and $\beta$ as
\[
H \leftarrow D H, \qquad \beta \leftarrow D \beta,
\]
where $D$ is the diagonal matrix of column $\ell_2$-norms of $W$.
This preserves the reconstruction $WH$ and the linear predictor $W\beta$,
leaving both the NMF and Cox terms in the loss unchanged. Normalization
prevents degeneracy in the scale-nonidentifiable factorization and keeps
columns of $W$ comparable in magnitude and interpretable as gene programs.


\subsection{Derivation of W update from projected coordinate descent}
The W update can be derived directly from the projected coordinate descent update with a specific step size.

Recall that the overall loss function is
\begin{align*}
    \mathcal{L}(W,H,\beta) = \frac{(1-\alpha)}{2\|X\|_F^2} ||X - WH||^2_F - \alpha\left(\frac{2}{n_{event}} \ell(W,\beta) - \lambda\left(\xi\|\beta\|_1 + \frac{(1-\xi)}{2}\|\beta\|_2^2\right)\right).
\end{align*}
where $\ell(W,\beta)$ is the log-partial likelihood for the Cox model. Let $\nabla_W \ell$ represent the derivative of $\ell(W,\beta)$ with respect to $W$. Then the derivative of the overall loss with respect to $W$ is

\begin{equation*}
  \frac{\partial L}{\partial W} = \frac{(1-\alpha)}{\lVert X \rVert_F^2}\left( WHH^T - XH^T \right) - \frac{2\alpha}{n_{event}} \nabla_W \ell.
\end{equation*}

Then the gradient descent update rule at iteration $t$ is
\begin{align*}
  W^{(t)} &= W^{(t-1)} - \gamma \left( \frac{\partial L}{\partial W} \right) \\
          &= W^{(t-1)} - \gamma \left( \frac{(1-\alpha)}{\lVert X \rVert_F^2}\left( WHH^T - XH^T \right) - \frac{2\alpha}{n_{event}} \nabla_W \ell \right).
\end{align*}
Let the step size $\gamma$ be defined as in [@seung2001algorithms]:
\begin{equation*}
  \gamma = \frac{\lVert X \rVert_F^2}{(1-\alpha)}\frac{W}{WHH^T}.
\end{equation*}
Then the update becomes
\begin{align*}
  W^{(t)} &= W^{(t-1)} -\frac{\lVert X \rVert_F^2}{(1-\alpha)}\frac{W}{WHH^T} \left( \frac{(1-\alpha)}{\lVert X \rVert_F^2}\left( WHH^T - XH^T \right) - \frac{2\alpha}{n_{event}} \nabla_W \ell \right) \\
          &= \frac{W}{WHH^T} XH^T + \frac{2\alpha \lVert X \rVert_F^2}{n_{event}(1-\alpha)} \nabla_W \ell \\
          &= W \odot \frac{XH^T + \frac{2\alpha \lVert X \rVert_F^2}{n_{event}(1-\alpha)} \nabla_W \ell }{WHH^T} \\
          &= W \odot \frac{\frac{(1-\alpha)}{\lVert X \rVert_F^2} XH^T + \frac{2\alpha}{n_{event}} \nabla_W \ell}{\frac{(1-\alpha)}{\lVert X \rVert_F^2} WHH^T}.
\end{align*}
Finally, projected coordinate descent projects the $W$ update in to the positive space
\begin{equation*}
  W^{(t)} =  \max \left(W \odot \frac{\frac{(1-\alpha)}{\lVert X \rVert_F^2} XH^T + \frac{2\alpha}{n_{event}} \nabla_W \ell}{\frac{(1-\alpha)}{\lVert X \rVert_F^2}WHH^T}, 0\right).
\end{equation*}
Since $W \geq 0$ this is equivalent to
\begin{equation*}
  W^{(t)} =  W \odot \max \left( \frac{\frac{(1-\alpha)}{\lVert X \rVert_F^2} XH^T + \frac{2\alpha}{n_{event}} \nabla_W \ell}{\frac{(1-\alpha)}{\lVert X \rVert_F^2} WHH^T}, 0 \right),
\end{equation*}
which matches the multiplicative form used in the software implementation.


\section{Convergence proof}

We show that, under mild regularity conditions, the block coordinate descent
(BCD) Algorithm \ref{alg:desurv} converges to a stationary point of the DeSurv
loss function
\[
\mathcal{L}(W,H,\beta).
\]
For clarity, we first analyze the algorithm \emph{without} the column
normalization of $W$ inside the loop, and then argue in
Section \ref{subsec:conv_normalization} that the normalization step preserves
stationarity of limit points.

Throughout, let $\theta = (W,H,\beta)$ and denote
\(\mathcal{L}(\theta) = \mathcal{L}(W,H,\beta)\).
The feasible set is
\[
\Theta = \{(W,H,\beta):
  W \in \mathbb{R}_{\ge 0}^{p\times k},\;
  H \in \mathbb{R}_{\ge 0}^{k\times n},\;
  \beta \in \mathbb{R}^k\}.
\]

\textbf{Assumption 1 (Regularity and parameter space).}
We assume:
\begin{enumerate}[label=(\roman*)]
  \item The data $X \in \mathbb{R}_{\ge 0}^{p\times n}$, $y \in \mathbb{R}^n_{>0}$, and $\delta \in \{0,1\}^n$ are fixed and bounded.
  \item The hyperparameters satisfy $\lambda_H > 0$, $\lambda>0$, $\alpha \in [0,1)$, and $\xi \in [0,1]$.
  \item The initial iterate $\theta^{(0)} = (W^{(0)},H^{(0)},\beta^{(0)})$ lies in $\Theta$ and satisfies $W^{(0)},H^{(0)} > 0$ elementwise.
\end{enumerate}

\textbf{Assumption 2 (Block updates).}
At each outer iteration $t$, the three block updates in
Algorithm \ref{alg:desurv} satisfy:
\begin{enumerate}[label=(\roman*)]
  \item \emph{$H$-update.} For fixed $(W^{(t)},\beta^{(t)})$, the update
        $H^{(t)} \mapsto H^{(t+1)}$ is given by the multiplicative rule
        \[
        H \leftarrow H \odot \frac{W^\top X}{W^\top W H + \lambda_H H},
        \]
        applied at $(W^{(t)},H^{(t)})$. This rule preserves nonnegativity and
        yields a nonincreasing value of the reconstruction term conditional on
        $W^{(t)}$ and $\beta^{(t)}$; see e.g.
        (Lee and Seung 2001; Pascual-Montano et al. 2006; Lin 2007).
  \item \emph{$W$-update.} For fixed $(H^{(t+1)},\beta^{(t)})$, the update
        $W^{(t)} \mapsto W^{(t+1)}$ is the hybrid multiplicative step
        followed by backtracking as in Algorithm \ref{alg:backtrack}, producing
        $W^{(t+1)}$ such that
        \[
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)})
          \le
          \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)}).
        \]
        If no candidate satisfies the Armijo-type condition within the allowed
        number of backtracking steps, the algorithm sets
        $W^{(t+1)} := W^{(t)}$.
  \item \emph{$\beta$-update.} For fixed $(W^{(t+1)},H^{(t+1)})$, the update
        $\beta^{(t)} \mapsto \beta^{(t+1)}$ is obtained by running the
        coordinate descent method of Simon et al. (2011) on the convex elastic-net Cox
        subproblem until convergence. Thus
        \[
          \beta^{(t+1)} \in
          \arg\min_{\beta\in\mathbb{R}^k}
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta),
        \]
        and
        \[
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})
          \le
          \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)}).
        \]
\end{enumerate}

\vspace{0.5em}
\textbf{Lemma 1 (Continuity and bounded level sets).}
Under Assumption 1, $\mathcal{L}:\Theta\to\mathbb{R}$ is continuous, and the
initial sublevel set
\[
  \mathcal{S}_0 := \{\theta \in \Theta:
    \mathcal{L}(\theta) \le \mathcal{L}(\theta^{(0)})\}
\]
is nonempty, closed, and bounded.

\emph{Proof.}
The NMF term $\|X-WH\|_F^2$ is a polynomial in the entries of $W$ and $H$, hence
is continuously differentiable. The penalty $\|H\|_F^2$ is continuous as well.
The Cox partial log-likelihood is a smooth function of the linear predictor
$\eta_i = x_i^\top W \beta$, which is linear in $(W,\beta)$, so this term is
continuously differentiable in $(W,\beta)$. The $\ell_2$ penalty on $\beta$ is
smooth, and the $\ell_1$ penalty is convex and lower semicontinuous. Therefore
$\mathcal{L}$ is continuous on $\Theta$.

The constraints $W,H\ge 0$ define closed convex cones, and
$\beta\in\mathbb{R}^k$ is unconstrained. Because $\lambda_H>0$ and $\lambda>0$,
the terms $\tfrac{\lambda_H}{nk}\|H\|_F^2$ and
$\lambda\tfrac{1-\xi}{2}\|\beta\|_2^2$ dominate the objective as
$\|H\|_F\to\infty$ or $\|\beta\|_2\to\infty$, respectively. Thus the sublevel
set $\mathcal{S}_0$ is bounded in $(H,\beta)$. Since $W\ge 0$ and $X$ is
fixed, the reconstruction term and Cox term being bounded prevent $W$ from
diverging while $\mathcal{L}$ remains below $\mathcal{L}(\theta^{(0)})$, so
$W$ is bounded on $\mathcal{S}_0$. Because $\Theta$ is closed and
$\mathcal{L}$ is continuous, $\mathcal{S}_0$ is closed. Nonemptiness follows
from $\theta^{(0)}\in\mathcal{S}_0$. \hfill$\square$

\vspace{0.5em}
\textbf{Lemma 2 (Monotone descent and existence of limit points).}
Under Assumptions 1 and 2, Algorithm \ref{alg:desurv} (without $W$
normalization) generates a sequence $\{\theta^{(t)}\}_{t\ge 0}\subset\mathcal{S}_0$
such that
\[
  \mathcal{L}(\theta^{(t+1)}) \le \mathcal{L}(\theta^{(t)})\quad \forall\,t,
\]
and $\{\mathcal{L}(\theta^{(t)})\}$ converges to a finite limit $\mathcal{L}^*$.
Moreover, $\{\theta^{(t)}\}$ is bounded and therefore admits at least one limit
point.

\emph{Proof.}
By Assumption 2(i),
\[
  \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)})
  \le
  \mathcal{L}(W^{(t)},H^{(t)},\beta^{(t)}).
\]
By Assumption 2(ii),
\[
  \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)})
  \le
  \mathcal{L}(W^{(t)},H^{(t+1)},\beta^{(t)}).
\]
By Assumption 2(iii),
\[
  \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t+1)})
  \le
  \mathcal{L}(W^{(t+1)},H^{(t+1)},\beta^{(t)}).
\]
Combining these inequalities gives
\[
  \mathcal{L}(\theta^{(t+1)}) \le \mathcal{L}(\theta^{(t)})\quad\forall t,
\]
so $\{\mathcal{L}(\theta^{(t)})\}$ is monotonically nonincreasing. Since
$\mathcal{L}$ is bounded below on $\Theta$, the sequence converges to some
finite $\mathcal{L}^*$.

Each iterate lies in $\mathcal{S}_0$ by construction, and $\mathcal{S}_0$ is
bounded by Lemma 1. Therefore $\{\theta^{(t)}\}$ is bounded and has at least
one limit point by the Bolzano--Weierstrass theorem. \hfill$\square$

\vspace{0.5em}
\textbf{Lemma 3 (Blockwise optimality of limit points).}
Let $\theta^*=(W^*,H^*,\beta^*)$ be any limit point of $\{\theta^{(t)}\}$ under
Assumptions 1–2. Then:
\begin{enumerate}[label=(\roman*)]
  \item $H^*$ satisfies the KKT conditions for
        \[
          \min_{H\ge 0} \mathcal{L}(W^*,H,\beta^*).
        \]
  \item $W^*$ satisfies the KKT conditions for
        \[
          \min_{W\ge 0} \mathcal{L}(W,H^*,\beta^*).
        \]
  \item $\beta^*$ is the unique minimizer of
        \[
          \min_{\beta\in\mathbb{R}^k} \mathcal{L}(W^*,H^*,\beta),
        \]
        and satisfies the KKT conditions for the elastic-net Cox subproblem.
\end{enumerate}

\emph{Proof.}
Let $\{t_j\}$ be a subsequence such that
$\theta^{(t_j)}\to\theta^*=(W^*,H^*,\beta^*)$. By continuity of $\mathcal{L}$,
$\mathcal{L}(\theta^{(t_j)})\to\mathcal{L}^*$.

(i) Conditional on $(W,\beta)$, the $H$-subproblem is
\begin{equation}
  \min_{H \geq 0}\frac{(1-\alpha)}{\lVert X \rVert_F^2}(\lVert X-WH \rVert_F^2 + \frac{\lambda_H}{nk}||H||_F^2) + \text{const in $H$},\nonumber
\end{equation}
a strictly convex quadratic program. The multiplicative update for $H$ is known
to be a descent method whose fixed points coincide with the KKT points of this
constrained subproblem [@seung2001algorithms; @lin2007convergence]. Since the full
objective $\mathcal{L}(\theta^{(t)})$ converges, the per-iteration decrease
in $\mathcal{L}$ due to the $H$-update must vanish along $\{t_j\}$.
If $H^*$ were not KKT for the $H$-subproblem given $(W^*,\beta^*)$, there would
exist a feasible descent direction for $H$ at $(W^*,\beta^*)$, implying that
for sufficiently large $j$ the $H$-update would still produce a strict decrease
in $\mathcal{L}$, contradicting convergence. Hence $H^*$ satisfies the KKT
conditions.

(ii) Conditional on $(H,\beta)$, the $W$-subproblem is differentiable and
convex in $W$ (sum of a quadratic term and a negative Cox partial
log-likelihood in a linear predictor). The hybrid multiplicative step with
backtracking can be viewed as a projected gradient-like update onto the
nonnegative orthant: as shown in the derivation above, the multiplicative
direction is equivalent to a negative gradient step with the adaptive step
size of Seung et al., projected onto the nonnegative orthant. Under mild conditions on the search direction and step
sizes, any limit point of such a projected gradient scheme is a KKT point for
the constrained subproblem; see, for example, [@tseng2001convergence; @grippo2000convergence].
If $W^*$ did not satisfy the KKT conditions, there would be a feasible
descent direction at $(W^*,H^*,\beta^*)$ and a sufficiently small step that
reduces $\mathcal{L}$, which the line search would eventually accept. This
would contradict the fact that $\mathcal{L}(\theta^{(t_j)})\to\mathcal{L}^*$.
Therefore $W^*$ is KKT for the $W$-subproblem.

(iii) For fixed $(W,H)$, the $\beta$-subproblem is the convex elastic-net
penalized Cox objective. The coordinate descent algorithm converges to the unique minimizer. By Assumption 2(iii),$\beta^{(t+1)}$ is taken to be this minimizer at each iteration. Passing to the limit along $\{t_j\}$ shows that $\beta^*$ is the unique minimizer of the
$\beta$-block given $(W^*,H^*)$ and hence satisfies its KKT conditions.
\hfill$\square$

\vspace{0.5em}
\textbf{Theorem 1 (Convergence to a stationary point).}
Under Assumptions 1 and 2, the sequence $\{\theta^{(t)}\}$ produced by
Algorithm \ref{alg:desurv} (without $W$ normalization) satisfies:
\begin{enumerate}[label=(\alph*)]
  \item The objective values $\{\mathcal{L}(\theta^{(t)})\}$ are monotonically
        nonincreasing and converge to a finite limit $\mathcal{L}^*$.
  \item Every limit point $\theta^* = (W^*,H^*,\beta^*)$ of $\{\theta^{(t)}\}$
        is a stationary point of $\mathcal{L}$ on $\Theta$, i.e. it satisfies
        the first-order KKT conditions for
        \[
          \min_{\theta\in\Theta} \mathcal{L}(\theta).
        \]
\end{enumerate}

\emph{Proof.}
Part (a) is Lemma 2. For part (b), Lemma 3 shows that any limit point $\theta^*$
is blockwise optimal: each block is optimal (in the KKT sense) when the others
are held fixed.

The DeSurv loss can be written as
\[
\mathcal{L}(\theta) = f(W,H,\beta) + g(\beta)
                     + I_{\{W\ge 0\}}(W) + I_{\{H\ge 0\}}(H),
\]
where $f$ is continuously differentiable, $g(\beta)=\lambda\xi\|\beta\|_1$ is
a separable convex (possibly nondifferentiable) penalty, and $I_C$ denotes the
indicator of a closed convex set $C$. This is the smooth+nonsmooth composite
form considered in block coordinate descent analyses such as [@tseng2001convergence].
In this setting, any point that is optimal with respect to each block
individually (a block coordinatewise minimizer) is a stationary point for the
full problem: equivalently,
\[
0 \in \nabla_{W,H} f(W^*,H^*,\beta^*)
    + \partial I_{\{W\ge 0\}}(W^*)
    + \partial I_{\{H\ge 0\}}(H^*),
\]
\[
0 \in \nabla_\beta f(W^*,H^*,\beta^*)
    + \partial g(\beta^*),
\]
which are precisely the KKT conditions for
$\min_{\theta\in\Theta}\mathcal{L}(\theta)$.
Hence every limit point of Algorithm \ref{alg:desurv} is stationary.
\hfill$\square$

\subsection{Effect of column normalization of $W$}
\label{subsec:conv_normalization}

The above analysis omits the column-normalization step for $W$ used in
Algorithm \ref{alg:backtrack}. We briefly argue that this normalization does
not affect stationarity of limit points.

Let $D$ be a diagonal matrix with strictly positive diagonal entries. The
transformation
\[
  (W,H,\beta) \mapsto (W',H',\beta') = (W D^{-1},\, D H,\, D\beta)
\]
preserves both the product $WH$ and the linear predictor $W\beta$, and hence
leaves $\mathcal{L}$ invariant. The column-normalization step is exactly such a
transformation, with $D$ chosen from the column norms of $W$.

Let $\{\theta^{(t)}\}$ be the sequence generated by the algorithm without
normalization, and let $\{\tilde{\theta}^{(t)}\}$ be the sequence with
normalization. For each $t$ there exists a diagonal $D^{(t)}$ with strictly
positive entries such that
\[
  \tilde{\theta}^{(t)} =
  (W^{(t)} D^{(t),-1},\, D^{(t)} H^{(t)},\, D^{(t)}\beta^{(t)}),
\]
and $\mathcal{L}(\tilde{\theta}^{(t)}) = \mathcal{L}(\theta^{(t)})$.
Thus limit points of $\{\tilde{\theta}^{(t)}\}$ are obtained from limit points
of $\{\theta^{(t)}\}$ by such invertible diagonal scalings.

Since the KKT conditions are expressed in terms of the gradients with respect
to $WH$ and $W\beta$, and these quantities are invariant under the above
scaling, stationarity is preserved under the transformation. Therefore
Theorem 1 implies that every limit point of the \emph{normalized} algorithm is
also a stationary point of $\mathcal{L}$ (up to this scaling equivalence), which
establishes convergence of the implementation used in practice.

\subsection{Remark (Coxnet implementation)}
Our implementation updates the $\beta$ block using a Coxnet-style coordinate descent that relies on an approximate Hessian. Consequently, the formal optimality proof is established for an idealized version of the $\beta$ update that assumes exact second-order information. Nonetheless, in empirical applications the approximate update is numerically stable and yields a monotone decrease in the objective, consistent with the behavior predicted by the idealized analysis.


\section{Supervision on \(W\) versus \(H\)}

Classical nonnegative matrix factorization (NMF) is non-identifiable: for any
invertible matrix \(R\) with \(W R \ge 0\) and \(R^{-1} H \ge 0\), the factorization
\((W, H)\) can be transformed to \((W R, R^{-1} H)\) without changing the
reconstruction error, yielding multiple valid solutions
[@donoho2004nmf; @laurberg2008uniqueness; @gillis2014nmf]. Although normalizing
columns of \(W\) or rows of \(H\) resolves the trivial scaling ambiguity,
nonnegative mixing transformations persist, and empirical studies consistently
show that the sample-loading matrix \(H\) is more sensitive to initialization
and local minima than the program matrix \(W\)
[@brunet2004metagenes; @kim2007sparse; @gillis2012accelerated].

In DeSurv, survival supervision enters through the projection $Z = W^\top X$ in the partial log-likelihood, so that the gradient of the DeSurv loss with respect to the programs is
\[
\nabla \mathcal{L} = (1-\alpha) \nabla_W \mathcal{L}_{\mathrm{NMF}}(W,H) - \alpha\nabla_W \mathcal{L}_{\mathrm{Cox}}(W,\beta),
\]
and therefore the update rule for $W$ as in Equation \ref{eqn:W_update} relies on both the reconstruction term and the supervision term. As such, the gene-program definitions are directly shaped by their association with survival. This mechanism amplifies
genes aligned with the hazard gradient and suppresses those that dilute
prognostic structure, favoring programs enriched for survival-relevant variation.

By contrast, if supervision were applied to the sample loadings \(H\), the model
could reduce the Cox loss by redistributing patient-specific coefficients while
leaving the gene-level programs \(W\) nearly unchanged, a behavior documented in
multiple supervised variants of NMF and supervised topic models, where
supervision applied only to the coefficient matrix modifies \(H\) but not the
basis \(W\) [@cai2011graph; @blei2007slda].

Supervising \(W\) also provides a practical advantage for **portability**:
because \(W\) defines gene-level programs, new samples can be embedded by the
closed-form projection \(Z_{\text{new}} = W^\top X_{\text{new}}\). In contrast,
supervising \(H\) would not yield such a mapping; instead, obtaining sample
loadings for external datasets would require solving a nonnegative least-squares
(NNLS) problem for each new sample, introducing optimization noise and reducing
reproducibility.

Together, these considerations motivate supervising through \(W\), which shapes
the gene programs toward prognostic directions, stabilizes solutions across
initializations, and yields biologically interpretable signatures that
generalize across cohorts.


\section{Cross-validation procedure}
Algorithm \ref{alg:cv} describes the cross validation procedure with $F$ folds and $R$ initializations.
We define the c-index using comparable pairs
$\mathcal{P} = \{(i,j) : y_i < y_j,\ \delta_i = 1\}$ and linear predictors
$\hat{\eta}_i$:
\begin{equation}
\label{eqn:cindex}
\hat{c} =
\frac{1}{|\mathcal{P}|}
\sum_{(i,j)\in\mathcal{P}}
\left[
  \mathbbm{1}(\hat{\eta}_i > \hat{\eta}_j)
  + \tfrac{1}{2}\mathbbm{1}(\hat{\eta}_i = \hat{\eta}_j)
\right].
\end{equation}

\begin{algorithm}[H]
\caption{Cross-validation for DeSurv pipeline}
\label{alg:cv}
\begin{algorithmic}[1]
\Require
\Statex Number of folds $F$
\Statex Number of initializations $R$
\Statex Observed data ($X$, $y$, $\delta$)
\Statex Hyperparameters $k$, $\alpha$, $\lambda$, $\xi$, $\lambda_H$, and number of top genes per factor $n_{top}$
\Ensure The cross-validated c-index

\State Divide subjects into $F$ folds.
\For{$f = 1$ to $F$}
    \State Split data into training and validation $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(X_{(f)},y_{(f)}, \delta_{(f)})$ sets
    \For{$r = 1$ to $R$}
        \State set.seed$(r)$
        \State Apply Algorithm \ref{alg:desurv} with inputs $(X_{(-f)},y_{(-f)}, \delta_{(-f)})$, and $(k, \alpha, \lambda, \xi, \lambda_H)$
        \State Obtain $\hat{W}$ and $\hat{\beta}$ as output from Algorithm \ref{alg:desurv}
        \State Restrict $\hat{W}$ to the top $n_{top}$ genes per factor (by factor-specificity score) and compute the estimated linear predictor on this support:
        $
        \hat{\eta} = X_{(f)}^T\hat{W}\hat{\beta}
        $
        \State Compute c-index, denoted $\hat{c}_{(f)r}$, according to Equation \ref{eqn:cindex}
    \EndFor
    \State \textbf{end loop over $r$}
    \State Compute average c-index across initializations:
    $
    \hat{c}_{(f)} = \frac{1}{R}\sum_{r=1}^R\hat{c}_{(f)r}
    $
\EndFor
\State \textbf{end loop over $f$}
\State Compute final cross-validated c-index:
$
\hat{c} = \frac{1}{F}\sum_{f=1}^F \hat{c}_{(f)}
$
\State \Return Final estimate $\hat{c}$

\end{algorithmic}
\end{algorithm}


\section{Bayesian Optimization}

We treat hyperparameter tuning as a black-box optimization problem over a
bounded domain. Let $\theta$ collect the tuned parameters (at minimum
$\theta = (k,\alpha,\lambda,\xi)$, and optionally discrete choices
such as the signature size $n_{\text{top}}$). For a given $\theta$ we run the
cross-validation procedure in Algorithm \ref{alg:cv}, averaging the c-index
across folds and random initializations to form the objective
\[
\hat{c}(\theta) = \frac{1}{F} \sum_{f=1}^F \hat{c}_{(f)}(\theta).
\]
Because $\hat{c}(\theta)$ is expensive to evaluate and non-differentiable with
respect to $\theta$, we use Bayesian optimization (BO) to adaptively select
candidate configurations.

We place a Gaussian-process (GP) prior on the response surface
$f(\theta) \sim \mathcal{GP}(0, k(\theta,\theta'))$ with a Mat\'ern kernel and
automatic relevance determination length-scales. Continuous parameters are
rescaled to $[0,1]$ and penalties are optimized on a log scale; integer-valued
coordinates (e.g., $k$ or $n_{\text{top}}$) are proposed in the continuous
space and rounded inside the evaluator. After an initial batch of evaluations,
the GP is refit after each new observation and used to guide acquisition.

At iteration $t$, let $\mu_t(\theta)$ and $\sigma_t(\theta)$ denote the GP
posterior mean and standard deviation, and let $f_\star$ be the best observed
score. We select new candidates by maximizing expected improvement (EI),
\[
\operatorname{EI}(\theta) =
\left(\mu_t(\theta) - f_\star - \kappa\right)\Phi(Z(\theta))
  + \sigma_t(\theta)\phi(Z(\theta)),
\quad
Z(\theta) = \frac{\mu_t(\theta) - f_\star - \kappa}{\sigma_t(\theta)},
\]
where $\Phi$ and $\phi$ are the standard normal cdf/pdf and $\kappa \ge 0$
controls the exploration margin. EI is evaluated over a candidate pool sampled
from the bounded search region, and the maximizer is evaluated next. The loop
continues until the iteration budget is exhausted or the improvement plateaus.

For model selection, we take the highest observed $\hat{c}(\theta)$ as the
optimal configuration. When fold-level standard errors are available, we apply
a one-standard-error rule for $k$, selecting the smallest rank whose mean
performance lies within one standard error of the best. The final DeSurv model
is refit at the selected parameters using the consensus-based initialization
described above, and the BO-chosen $n_{\text{top}}$ (when tuned) is used to
truncate each program for downstream analysis.

The parameter $n_{\text{top}}$ controls which genes enter the survival linear predictor. After factorization, each sample's score for factor $j$ is computed as $\widetilde{W}_j^\top x$, where $\widetilde{W}_j$ retains only the $n_{\text{top}}$ genes with the largest factor-specificity scores $s_{ij}$ (Supplementary Information, Section \ref{sec:gene-program-correlation}); genes outside this set still shape the factorization but do not enter the linear predictor. BO selects $n_{\text{top}}$ jointly with $k$, $\alpha$, $\lambda$, and $\xi$ by optimizing cross-validated C-index. When $\alpha = 0$ (no survival supervision), gene selection and survival outcomes are decoupled, so varying $n_{\text{top}}$ has negligible effect on precision in the null simulation.

Supplementary Fig. \ref{fig:fig-si-bo-tuning} illustrates the tuned response surface for the two most influential coordinates in the PDAC training cohort: the cross-validated C-index over factorization rank $k$ and supervision strength $\alpha$. Performance is maximized at intermediate supervision and moderate rank and degrades toward the unsupervised corner ($\alpha = 0$), consistent with a benefit of survival supervision; this is an empirical model-selection sensitivity surface, not a simulation result.

```{r fig-si-bo-tuning, fig.cap="Bayesian-optimization tuning surface for DeSurv in the PDAC training cohort (TCGA + CPTAC). Gaussian-process--predicted mean cross-validated C-index over factorization rank ($k$) and supervision strength ($\\alpha$; $\\alpha = 0$ corresponds to standard NMF). The response surface is smooth and single-peaked, with the best cross-validated performance at intermediate supervision and moderate rank; discrimination degrades toward the unsupervised corner. This is empirical model-selection tuning on the training data, not a simulation result.", fig.env='figure*', fig.pos='t', out.height= '3.6in', out.width='4.5in', fig.align='center'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_bo_tuning_surface_tcgacptac.pdf", density = 300))

```

\subsection{Consensus initialization for the final model}

The final DeSurv model at the selected hyperparameters was initialized from a consensus of multiple independent runs to reduce sensitivity to random starts. We first fit several DeSurv models to the training data from independent random nonnegative initializations at the selected $(k,\alpha,\lambda,\xi,n_{\text{top}})$. For each run, the top $n_{\text{top}}$ genes of every learned program were extracted by factor-specificity score, and a gene-by-gene consensus matrix was accumulated across runs, recording how often each pair of genes appeared together among the top genes of the same program. Genes appearing in the top sets of fewer than $m$ runs (default $m = 2$) were treated as inactive and excluded from clustering. The consensus matrix over the remaining genes was partitioned into $k$ groups by agglomerative hierarchical clustering (average linkage), and each cluster defined one column of the initial gene-program matrix $W_0$; inactive genes and otherwise-empty entries were set to a small positive floor ($\sqrt{\epsilon_{\text{mach}}}$) to preserve strict positivity. The initial loadings $H_0$ were obtained by nonnegative least squares given $W_0$ and the training expression matrix, and the initial Cox coefficients $\beta_0$ were set to zero. This consensus seed used only training-cohort expression and was applied solely to initialize the single final refit after hyperparameter selection; it used no validation-cohort data and therefore introduced no information leakage.

\section{PDAC Datasets}

We analyzed PDAC cohorts from TCGA-PAAD and CPTAC-PDAC as training data
[@raphael2017integrated;@cao2021proteogenomic]. External validation used
independent cohorts from Dijk, Moffitt, PACA-AU (array and RNA-seq), and Puleo
[@dijk2020unsupervised;@moffitt2015virtual;@Bailey2016;@puleo2018stratification].
Training models were fit on the combined TCGA and CPTAC expression matrices
after harmonizing gene identifiers; validation datasets were processed with the
same gene set and transformations before projection, scoring and downstream validation analyses.

Dataset-specific inclusion and preprocessing steps were as follows. For
TCGA-PAAD, we excluded samples lacking tumor grade and mapped features to gene
symbols. For CPTAC, we retained samples annotated as PDAC by histology. For
Dijk, all 90 samples passed quality-control filters and were retained. For
Moffitt, only primary tumor specimens were kept. For PACA-AU array and RNA-seq,
we retained primary PDAC tumors and
collapsed duplicated gene symbols by median; the RNA-seq cohort received a
``\_seq'' suffix on sample IDs to avoid collisions with the array cohort. Puleo
required no additional dataset-specific filtering beyond survival QC.

Across cohorts, samples without valid survival outcomes (nonpositive time,
missing time, or missing event indicator) were removed. Expression matrices
were analyzed on a log scale: RNA-seq cohorts as log$_2$(TPM + 1); microarray cohorts in platform-normalized log-scale intensities. For each training cohort (TCGA-PAAD and CPTAC) separately, genes were ranked
by mean expression and by variance independently; the 3,000 genes with the
lowest sum of these two ranks were retained, selecting genes that are both
highly expressed and highly variable.
The intersection of these two gene sets yielded 1,970 genes used for model
training. Validation datasets were restricted to this same 1,970-gene set.
Within-sample rank transformation was applied to all datasets to mitigate
platform effects, with validation data transformed identically to training.

Table \ref{tab:cohorts} summarizes the cohorts, platforms, sample sizes after
quality control, and patient-level metadata used in this study. Overall survival
time and event indicators were available for all cohorts. PurIST
[@rashid2020purity] and DeCAF [@peng2024determination] molecular
classifications, applied to bulk expression data as described in those
references, were used as covariates in the covariate-adjusted external
validation analysis. Gene expression data and molecular
classifications for the validation cohorts were originally compiled for the
DeCAF study [@peng2024determination]; training data were obtained directly
from GDC.

```{r cohorts-events, echo=FALSE, message=FALSE, warning=FALSE}
.coh <- local({
  tr <- read_result("tar_data_filtered_tcgacptac")$sampInfo
  vl <- read_result("val_latent_desurv_tcgacptac")
  ev <- list(TCGA_PAAD = tr$event[tr$dataset == "TCGA_PAAD"],
             CPTAC     = tr$event[tr$dataset == "CPTAC"])
  for (x in vl) ev[[x$dataset]] <- x$survival$event
  ev$train_tot <- tr$event
  ev$val_tot   <- unlist(lapply(vl, function(x) x$survival$event))
  N    <- sapply(ev, length)
  E    <- sapply(ev, function(z) sum(z == 1, na.rm = TRUE))
  rate <- round(100 * E / N)
  cell <- sprintf("%d (%d\\%%)", E, rate); names(cell) <- names(ev)
  vr <- rate[c("Dijk","Moffitt_GEO_array","PACA_AU_array","PACA_AU_seq","Puleo_array")]
  list(N = N, cell = cell,
       train_lo = min(rate[c("TCGA_PAAD","CPTAC")]), train_hi = max(rate[c("TCGA_PAAD","CPTAC")]),
       val_lo = min(vr), val_hi = max(vr), val_pooled = unname(rate["val_tot"]))
})
```

\begin{table}[ht]
\centering
\caption{Summary of PDAC cohorts used in this study. $N$ denotes the number of
samples after all quality-control filters (histology, survival validity,
dataset-specific exclusions). Patient-level metadata: OS = overall survival
(time and event indicator); PurIST = tumor subtype classification
(Rashid et al. 2020); DeCAF = CAF subtype classification
(Peng et al. 2026). Expression data and molecular classifications
for the validation cohorts were originally compiled for the DeCAF study
(Peng et al. 2026). Events denotes the number of overall-survival events (deaths) with the corresponding event rate; outcome frequency was lower and more homogeneous in the training cohorts (`r .coh$train_lo`--`r .coh$train_hi`\%) than across the validation cohorts (`r .coh$val_lo`--`r .coh$val_hi`\%, pooled `r .coh$val_pooled`\%), a case-mix difference relevant to the interpretation of external-validation performance.}
\label{tab:cohorts}
\footnotesize
\begin{tabular}{llrrllll}
\toprule
Cohort & Platform & $N$ & Events & Role & Accession & Metadata & Reference \\
\midrule
TCGA-PAAD        & RNA-seq    & `r .coh$N[["TCGA_PAAD"]]` & `r .coh$cell[["TCGA_PAAD"]]`  & Training   & GDC (TCGA-PAAD) & OS, PurIST, DeCAF & Raphael et al. (2017) \\
CPTAC            & RNA-seq    & `r .coh$N[["CPTAC"]]` & `r .coh$cell[["CPTAC"]]`  & Training   & GDC (CPTAC-3)    & OS, PurIST, DeCAF & Cao et al. (2021) \\
\addlinespace
Dijk             & RNA-seq    &  `r .coh$N[["Dijk"]]` & `r .coh$cell[["Dijk"]]`  & Validation & E-MTAB-6830      & OS, PurIST, DeCAF & Dijk et al. (2020) \\
Moffitt          & Microarray & `r .coh$N[["Moffitt_GEO_array"]]` & `r .coh$cell[["Moffitt_GEO_array"]]`  & Validation & GSE71729         & OS, PurIST, DeCAF & Moffitt et al. (2015) \\
PACA-AU (array)  & Microarray &  `r .coh$N[["PACA_AU_array"]]` & `r .coh$cell[["PACA_AU_array"]]`  & Validation & GSE36924         & OS, PurIST, DeCAF & Bailey et al. (2016) \\
PACA-AU (RNA-seq) & RNA-seq   &  `r .coh$N[["PACA_AU_seq"]]` & `r .coh$cell[["PACA_AU_seq"]]`  & Validation & EGAS00001000154  & OS, PurIST, DeCAF & Bailey et al. (2016) \\
Puleo            & Microarray & `r .coh$N[["Puleo_array"]]` & `r .coh$cell[["Puleo_array"]]` & Validation & E-MTAB-6134      & OS, PurIST, DeCAF & Puleo et al. (2018) \\
\addlinespace
\multicolumn{2}{l}{Training total} & `r .coh$N[["train_tot"]]` & `r .coh$cell[["train_tot"]]` & & & & \\
\multicolumn{2}{l}{Validation total} & `r .coh$N[["val_tot"]]` & `r .coh$cell[["val_tot"]]` & & & & \\
\bottomrule
\end{tabular}
\end{table}

\section{Survival Analysis}

Survival analyses used overall survival time with event indicators as provided
by each cohort. For fitted DeSurv models, sample-level program scores were
computed as $Z = \widetilde{W}^\top X$ and standardized using the training mean and
standard deviation before constructing the linear predictor
$\hat{\eta} = Z \hat{\beta}$. Prognostic performance was summarized using the
concordance index as defined above. For visualization and group-level testing,
we generated Kaplan--Meier curves and log-rank tests for risk groups defined by
a single cutpoint selected by cross-validation in the TCGA and CPTAC training
cohort, applied unchanged to the pooled validation samples after standardizing
each linear predictor by the training mean and standard deviation. When multiple datasets were pooled, Cox models were stratified
by dataset to allow cohort-specific baseline hazards. P-values for Kaplan--Meier group comparisons correspond to two-sided log-rank tests; continuous-score associations were evaluated using Cox models stratified by dataset.

\section{Gene Program Correlation Analysis}
\label{sec:gene-program-correlation}

To assess the biological identity of learned factors, we evaluated the
correspondence between factor-specific gene rankings and established PDAC gene
programs. For each factor column $j$ of the gene program matrix $W$, we
identified the top 50 genes by factor specificity score. Each column of $W$
was first scaled by its column maximum to produce $W^*_{ij} = W_{ij} /
\max_i W_{ij}$, placing all factors on a common $[0,1]$ scale. The
specificity score for gene $i$ in factor $j$ was then defined as
$$s_{ij} = W^*_{ij} - \max_{j' \neq j} W^*_{ij'}.$$
Genes were ranked in descending order of $s_{ij}$, and the top 50 per factor
were retained. The union of these gene sets across all factors was then
intersected with the genes present in any reference program to form the set of
genes used in the correlation analysis.

For each factor $j$ and each reference gene program $G_k$, we computed the
Spearman rank correlation between the continuous factor loadings $W_{\cdot j}$
(restricted to the intersection gene set) and a binary indicator vector
$\mathbf{1}[\text{gene} \in G_k]$. P-values were derived from the asymptotic
approximation to the Spearman test statistic and adjusted for multiple
comparisons using the Benjamini--Hochberg procedure applied within each factor
column independently. Gene programs were included in the visualization only if
at least one factor yielded a Spearman correlation exceeding 0.2. Asterisks
($*$) indicate adjusted p-value $< 0.1$.

Reference gene programs displayed spanned classical and basal-like tumor
signatures and activated/immune stromal programs from Moffitt et al.\
[@moffitt2015virtual]; classical and exocrine subtypes from Collisson et al.\
[@collisson2011subtypes]; classical-tumor, basal-like-tumor, exocrine, and
activated-stroma compartment signatures from DECODER [@peng2019novo]; the pure
classical, immune classical, desmoplastic, and basal-like tumor
microenvironment programs from Puleo et al.\ [@puleo2018stratification]; the ECM
and immune programs from Maurer et al.\ [@Maurer2019]; the iCAF
cancer-associated fibroblast signature from Elyada et al.\ [@elyada2019cross];
the proCAF, restCAF, and panCAF fibroblast subtypes from SCISSORS
[@leary2023s]; the proCAF and restCAF programs from the DeCAF study
[@peng2024determination]; the classical A subtype from Chan-Seng-Yue et al.\
[@chansengyue2020transcription]; and the ADEX (aberrantly differentiated
endocrine exocrine) program from Bailey et al.\ [@Bailey2016]. Bailey's ADEX
program is retained because it is unique, whereas redundant Bailey programs,
periductal subtypes, and MSI subtypes were dropped as exact duplicates of, or
low-correlation relative to, other retained programs.

\section{Reconstruction and Survival Contribution per Factor}

NMF does not provide an orthogonal variance decomposition; it seeks a low-rank
reconstruction of the expression matrix $X \approx WH$. We therefore measure
each factor's contribution to reconstruction rather than to ``variance
explained.'' For a subset of factors $S \subseteq \{1, \ldots, k\}$, define the
reconstruction value
$$v(S) = \|X\|_F^2 - \Big\| X - \sum_{j \in S} W_{\cdot j} H_{j \cdot} \Big\|_F^2,$$
the reduction in residual error relative to the null (zero-factor) model, where
$W_{\cdot j} H_{j \cdot}$ is the rank-one outer-product contribution of factor $j$ (the outer product of column $j$ of $W$ with row $j$ of $H$). The
reconstruction Shapley value of factor $j$,
$$\phi_j = \sum_{S \subseteq \{1,\ldots,k\} \setminus \{j\}}
\frac{|S|!\,(k - |S| - 1)!}{k!}\,\big(v(S \cup \{j\}) - v(S)\big),$$
is its average marginal reconstruction gain over all orderings of the factors.
By the Shapley efficiency axiom, $\sum_j \phi_j = v(\{1,\ldots,k\})$, the total
reconstruction gain of the full model, so the normalized shares
$\phi_j / \sum_l \phi_l$ form an additive decomposition of reconstruction that
sums to one; individual shares may be signed in principle but were all positive
in the reported PDAC models. This is a genuine additive decomposition, in contrast to per-factor
variance fractions, which for non-orthogonal factorizations such as NMF do not
partition (cross-terms between rank-one contributions are non-zero). The Shapley
shares are computed exactly by enumerating the $2^k$ subsets (trivial at
$k = 3$ or $k = 5$).

The survival contribution of factor $j$ was quantified as the Type III partial
likelihood increment: the reduction in Cox partial log-likelihood when factor $j$
is omitted from the full $k$-factor model,
$$\Delta\ell_j = \ell_{\text{full}}(\hat{\beta}_{\text{full}}) - \ell_{-j}(\hat{\beta}_{-j}),$$
where $\ell_{\text{full}}(\hat{\beta}_{\text{full}})$ is the maximized unpenalized
Cox partial log-likelihood of the full $k$-factor model and
$\ell_{-j}(\hat{\beta}_{-j})$ is that of the reduced model refit on the remaining
$k - 1$ factors after omitting factor $j$, each evaluated at its own maximum
partial likelihood estimate. Larger $\Delta\ell_j$ indicates a greater
marginal contribution of factor $j$ to survival prediction given all other
factors. These two quantities are plotted against each other in
main text, Fig.\ 2C. In PDAC at $k = 3$, the reconstruction shares are
$41.8\%/18.6\%/39.7\%$ for NMF (N1/N2/N3) and $18.8\%/25.2\%/56.0\%$ for DeSurv
(D1/D2/D3), while the survival contribution $\Delta\ell$ concentrates in DeSurv
D1 ($\Delta\ell = 66.4$) with all NMF factors negligible --- the factors that
contribute most to reconstruction are not the most prognostic.

\section{Factor Correspondence Analysis}

To assess which biological programs are preserved, altered, or suppressed
under survival supervision, we quantified the pairwise correspondence between
NMF and DeSurv factor loadings. For each pair of factors
$(j_{\text{NMF}},\, j_{\text{DeSurv}})$, we computed the Spearman rank
correlation between the corresponding columns of the NMF gene program matrix
$W_{\text{NMF}}$ and the DeSurv gene program matrix $W_{\text{DeSurv}}$ over all
genes. We use the full loading profile rather than a top-gene subset, which
would impose an arbitrary threshold and overstate apparent preservation.
Spearman correlation was used to capture monotone correspondence in gene loading
ranks while remaining robust to scale differences between the two methods.
Results are displayed as a heatmap with NMF factors as rows and DeSurv factors
as columns (main text, Fig.\ 2D).

\subsection{Cross-method projection \texorpdfstring{$R^2$}{R2}: novelty versus inheritance}

The correlation heatmap summarizes pairwise similarity; to make a direction-aware
statement of how much of each factor is reconstructable from the \emph{other}
method's basis, we regressed each factor's gene-loading vector onto the full set
of the other method's loading vectors and report the ordinary least-squares
$R^2$ (Table \ref{tab:projection-r2}). A high $R^2$ means the factor is inherited
(reconstructable from the other basis); a low $R^2$ means it is novel.

```{r tab-projection-r2, echo=FALSE, results='asis'}
source("R/reconstruction_helpers.R")
.fit_d  <- read_result("tar_fit_desurv_tcgacptac")
.fit_n  <- read_result("fit_std_desurvk_tcgacptac")
.r2_N <- projection_r2(.fit_n$W, .fit_d$W)   # NMF factor -> DeSurv basis
.r2_D <- projection_r2(.fit_d$W, .fit_n$W)   # DeSurv factor -> NMF basis
.pr <- data.frame(
  Factor = c("NMF N1 (tumor)", "NMF N2 (exocrine)", "NMF N3 (microenviron.)",
             "DeSurv D1 (classical/restCAF)", "DeSurv D2 (proCAF)", "DeSurv D3 (basal-like)"),
  R2 = sprintf("%.2f", c(.r2_N, .r2_D)),
  Interpretation = c("preserved", "dissolved", "preserved",
                     "not reconstructable", "inherited ($\\approx$N3)", "inherited ($\\approx$N1)"),
  stringsAsFactors = FALSE)
colnames(.pr) <- c("Factor", "$R^2$ from other basis", "Interpretation")
knitr::kable(.pr, format = "latex", escape = FALSE, booktabs = TRUE,
  caption = sprintf("Cross-method projection $R^2$: how reconstructable each factor's gene-loading vector is from the other method's full basis (OLS $R^2$). DeSurv retains the bulk of the unsupervised structure (NMF N1 and N3 carry over at $R^2 \\ge %.2f$) and reorganizes it: the survival-aligned D1 program is essentially unreconstructable from the NMF basis ($R^2 = %.2f$), i.e. not well reconstructed as a linear combination of the NMF factors, while the exocrine NMF factor N2 is poorly reconstructed from the DeSurv basis ($R^2 = %.2f$). \\label{tab:projection-r2}", min(.r2_N[1], .r2_N[3]), .r2_D[1], .r2_N[2])) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

The headline is DeSurv D1: $R^2 = `r sprintf("%.2f", .r2_D[1])`$ from the NMF basis, indicating that
its gene-loading vector is not well reconstructed as a linear combination of the
matched-rank NMF factors. The biologically dominant tumor (N1) and stromal (N3)
programs, by contrast, are largely reconstructable from the DeSurv basis
($R^2 = `r sprintf("%.2f", .r2_N[1])`$ and $`r sprintf("%.2f", .r2_N[3])`$), so most of the unsupervised structure is
retained; the principal reorganization is that the DeSurv basis resolves tumor
signal into basal- and classical-aligned programs while reconstructing the
exocrine factor (N2) less well ($R^2 = `r sprintf("%.2f", .r2_N[2])`$).

\subsection{Overall reconstruction \texorpdfstring{$R^2$}{R2}: the price of supervision}

```{r tab-overall-r2, echo=FALSE, results='asis'}
.X <- as.matrix(read_result("tar_data_filtered_tcgacptac")$ex)
.or <- data.frame(
  Model = c("NMF ($k = 3$)", "DeSurv ($k = 3$)"),
  R2 = sprintf("%.3f", c(overall_recon_r2(.fit_n$W, .fit_n$H, .X),
                         overall_recon_r2(.fit_d$W, .fit_d$H, .X))),
  Resid = sprintf("%.3f", c(1 - overall_recon_r2(.fit_n$W, .fit_n$H, .X),
                            1 - overall_recon_r2(.fit_d$W, .fit_d$H, .X))),
  stringsAsFactors = FALSE)
colnames(.or) <- c("Model", "Reconstruction $R^2$", "Residual fraction")
knitr::kable(.or, format = "latex", escape = FALSE, booktabs = TRUE,
  caption = "Whole-matrix reconstruction $R^2$ for unsupervised NMF versus DeSurv at $k = 3$. DeSurv sacrifices only $\\approx 1.5$ percentage points of reconstruction $R^2$ to gain its prognostic structure. \\label{tab:overall-r2}") |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

DeSurv gives up only about $1.5$ percentage points of reconstruction $R^2$
relative to unsupervised NMF (Table \ref{tab:overall-r2}) --- very little
reconstruction fidelity is sacrificed in exchange for the prognostic
reorganization documented above.

\subsection{The D1 axis is recovered by survival supervision, not by DeSurv specifically}

```{r suprec-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("supervised_recovery_stats")
.sr <- supervised_recovery_stats
.srt <- .sr$table
.srcor <- function(m) abs(.srt$score_cor_D1[.srt$method == m])
```

If the D1 tumor--stroma axis were an artifact of DeSurv's particular
optimizer, independent survival-supervised methods should not recover it. We
therefore fit three other supervised dimension-reduction methods on the same
training cohort ($n = `r .sr$meta$n`$, `r .sr$meta$n_events` events) --- supervised
principal components (Cox-screened genes followed by PCA), one-component Cox
partial least squares against martingale residuals, and a sparse Cox model
(lasso-penalized, `r .sr$meta$n_sparse_cox_genes` selected genes) --- and compared
their leading survival-aligned direction to the DeSurv D1 score. All three
recovered D1 closely (per-patient $|r| = `r sprintf("%.2f", .srcor("Supervised PCA"))`$
for supervised PCA, $`r sprintf("%.2f", .srcor("Cox-PLS"))`$ for Cox-PLS, and
$`r sprintf("%.2f", .srcor("Sparse Cox"))`$ for sparse Cox; mean
$`r sprintf("%.2f", .sr$sup_mean_score_cor)`$), and each aligned specifically with
D1 rather than the stromal D2 or basal-like D3 programs
(Fig. \ref{fig:suprec}). By contrast, the leading \emph{unsupervised} direction
(the first principal component, a variance-dominant unsupervised direction) was nearly orthogonal to D1
($|r| = `r sprintf("%.2f", .srcor("Unsupervised PCA"))`$) and instead tracked the
high-variance D2/D3 stroma--basal axis. This D1 axis is thus a property of
survival supervision shared across methods --- consistent with its being
unreconstructable from the unsupervised NMF basis (Table \ref{tab:projection-r2})
and with standard NMF's failure to enrich for the proCAF stroma at matched rank
--- and not an idiosyncrasy of the DeSurv algorithm.

```{r fig-suprec, echo=FALSE, fig.width = 8, fig.height = 3.6, fig.cap = "Independent survival-supervised methods recover the D1 axis; an unsupervised method does not. (A) Absolute correlation (across training-cohort patients) of each method's leading direction with the DeSurv D1 score, in the training cohort: supervised principal components, Cox partial least squares, and sparse Cox all recover D1 (green), whereas the leading unsupervised principal component does not (grey). (B) Sample-level agreement with D1 for Cox-PLS (supervised, green) versus the first unsupervised principal component (grey). \\label{fig:suprec}", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
knitr::include_graphics(file.path("..", "figures", "fig_supervised_recovery.pdf"), error = FALSE)
```

```{r suptuned-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_vs_supervised_tuned")
.stv    <- desurv_vs_supervised_tuned
.stvp   <- function(m) sprintf("%.2f", .stv$prediction[m, "POOLED"])
.stva   <- function(m, f) sprintf("%.2f", abs(.stv$axis_decomposition[m, f]))
.stvhit <- function(m) unname(.stv$supervised_attribution_across_seeds[[m]]["BasalTumor"])
.stvg   <- function(m, q) unname(.stv$genes_per_seed[[m]][q])
.stvc   <- function(m) sprintf("%.2f", .stv$stromal_carriage[m])
.stvns  <- .stv$meta$nseed
```

\subsection{Robustness to tuned supervised comparators and compartment attribution}

The recovery above establishes the D1 axis as a genuine survival direction shared across methods; here we ask what DeSurv adds beyond recovering it, using canonical, independently tuned comparators in place of the simple single-component versions above: supervised principal components (\texttt{superpc}; Bair--Tibshirani, with cross-validated screening threshold and component number) and penalized (lasso) Cox (\texttt{glmnet}, a single $\lambda_{\min}$ model). Cox partial least squares was not re-run in this pass, as a validated multi-component implementation was unavailable in our environment; the two canonical methods are sufficient. Every method began from the same 1{,}970-gene training-only expression/variance universe and performed its own feature selection; DeSurv was scored with its Bayesian-optimization-selected top-270-per-factor programs; all scoring used the frozen, rank-transformed transfer with no validation refitting.

Prediction parity was preserved: the pooled frozen-transfer C-index was $`r .stvp("DeSurv full LP")`$ for the DeSurv linear predictor, $`r .stvp("Supervised PCA")`$ for supervised PCA, and $`r .stvp("Sparse Cox")`$ for penalized Cox. Consistent with the score-level recovery above, both tuned methods aligned with the D1 direction (training $|r| = `r .stva("Sparse Cox","D1")`$ and $`r .stva("Supervised PCA","D1")`$) and with the DeSurv linear predictor ($|r| = `r .stva("Sparse Cox","fullLP")`$ and $`r .stva("Supervised PCA","fullLP")`$), but were essentially orthogonal to the activated-stroma/proCAF program D2 ($|r| = `r .stva("Sparse Cox","D2")`$ and $`r .stva("Supervised PCA","D2")`$): a single supervised risk axis captures the dominant tumour-axis signal but not the separate stromal program.

The compartment attribution made this explicit. Because D1 runs from a classical-tumour, restraining-stroma protective pole to a basal-like high-risk pole, a sparse supervised risk score aligned with D1 is naturally composed of basal-like tumour markers. Accordingly, across `r .stvns` cross-validation seeds the genes selected by both methods were reproducibly enriched (hypergeometric $P < 0.01$) for the basal-like tumour program (in `r .stvhit("Sparse Cox")`/`r .stvns` seeds for penalized Cox and `r .stvhit("Supervised PCA")`/`r .stvns` for supervised PCA) and in no seed for the activated-stroma/proCAF, restraining-stroma, immune, or exocrine compartments. The supervised risk scores did nonetheless retain some stromal signal: each correlated modestly with an external activated-stroma/proCAF signature ($|r| = `r .stvc("Sparse Cox")`$ for penalized Cox and $`r .stvc("Supervised PCA")`$ for supervised PCA), but well below DeSurv's D2 program ($|r| = `r .stvc("DeSurv D2 (proCAF factor)")`$), which tracks that signature directly. The supervised risk scores retained modest stromal signal but did not resolve it as a distinct program. The supervised feature selection was also unstable: the number of genes selected by supervised PCA ranged from `r .stvg("Supervised PCA","min")` to `r .stvg("Supervised PCA","max")` across seeds while its discrimination stayed near-constant, whereas DeSurv's programs are fixed. DeSurv instead placed the dominant survival-associated direction within a three-program representation containing separately measurable classical/restCAF-like, proCAF-like and basal-like programs. Thus, the distinction is representational: DeSurv preserves stromal context as an explicit program even when that program does not contribute independently to the fitted linear predictor.

```{r brier-stats, echo=FALSE, message=FALSE, warning=FALSE}
load_result("desurv_brier_stats")
.br <- desurv_brier_stats; .brp <- .br$pooled; .bripa <- range(.br$per_cohort$ipa)
```

\subsection*{Proper-scoring sensitivity of the external validation}

The main-text external validation reports discrimination with Harrell's C-index
under a Cox proportional-hazards model. Because Harrell's C can be optimistic when
hazards are non-proportional [@birolo2025beyondcox; @hao2021dnnsurv], we additionally
evaluated the DeSurv linear predictor---its direction frozen from training---with
a proper score, the integrated Brier score (IBS), in each validation cohort,
recalibrating only its scalar slope and baseline hazard
within each cohort and scoring over that cohort's follow-up. For a single
time-invariant risk score under proportional hazards the monotone score-to-survival
map preserves the concordance ordering, so Antolini's time-dependent concordance
coincides with Harrell's C; the additive quantity is therefore the proper score, not
a second concordance. Across the `r nrow(.br$per_cohort)` cohorts the DeSurv predictor achieved an
event-weighted integrated Brier score of `r sprintf("%.3f", .brp["ibs"])` versus
`r sprintf("%.3f", .brp["ibs_null"])` for a Kaplan--Meier reference, an index of
prediction accuracy (IPA $= 1 - \mathrm{IBS}/\mathrm{IBS}_{\text{null}}$) of
`r sprintf("%.1f", 100*.brp["ipa"])`\% (per-cohort range
`r sprintf("%.1f", 100*.bripa[1])`--`r sprintf("%.1f", 100*.bripa[2])`\%). Because each
cohort's IBS is integrated over its own event-time horizon, this pooled value is a
descriptive event-weighted summary rather than a single common-horizon proper score,
and the per-cohort IPA range is the more directly interpretable quantity. This
modest proper-score improvement over the null is consistent with the moderate
discrimination (C $= `r sprintf("%.2f", .brp["cindex"])`$) and with DeSurv's
contribution being a separately resolved tumour- and stroma-associated decomposition
rather than higher predictive accuracy.

\subsection*{Proportional-hazards diagnostic}

```{r si-ph-diagnostic, echo=FALSE, message=FALSE, warning=FALSE}
.phdf <- do.call(rbind, lapply(val_latent_desurv_tcgacptac, function(x)
  data.frame(risk_score = x$risk_score, time = x$survival$time,
             event = x$survival$event, dataset = x$dataset)))
.phdf$risk_z <- .phdf$risk_score / sd(.phdf$risk_score)
.phz <- survival::cox.zph(
  survival::coxph(survival::Surv(time, event) ~ risk_z + strata(dataset), data = .phdf))
.ph_chisq <- .phz$table["GLOBAL", "chisq"]
.ph_df    <- .phz$table["GLOBAL", "df"]
.ph_p     <- .phz$table["GLOBAL", "p"]
```

Because the pooled hazard ratio assumes proportional hazards, we tested the scaled Schoenfeld residuals of the stratified validation model ($\mathrm{Surv}(t,\delta) \sim \text{risk\_z} + \mathrm{strata}(\text{dataset})$) with \texttt{survival::cox.zph}. The proportional-hazards assumption was not supported for the standardized DeSurv linear predictor ($\chi^2 = `r sprintf("%.1f", .ph_chisq)`$, $\mathrm{df} = `r .ph_df`$, $P = `r sprintf("%.1e", .ph_p)`$), with the score's effect attenuating over follow-up rather than reversing. The reported pooled hazard ratio should therefore be read as a time-averaged summary of a consistently prognostic score rather than a constant multiplicative effect; this time-varying behavior is precisely why we additionally report the horizon-integrated proper-scoring (IBS/IPA) sensitivity above, which integrates prediction accuracy over each cohort's event-time horizon (to the 80th percentile of event times) rather than through a single hazard-ratio summary.

\subsection*{Independence of discovery and validation}

To make the separation between discovery, adjustment, and validation unambiguous: DeSurv
is trained with survival supervision only; it never uses PurIST, DeCAF, or any molecular
subtype label during factorization or hyperparameter selection. The PurIST and DeCAF
subtype calls were fixed before the DeSurv analysis (loaded as precomputed per-cohort
annotations) and enter only as covariates in the adjusted Cox models, not as inputs to
DeSurv. All rank and hyperparameter selection (including the one-standard-error rule) was
performed by cross-validation entirely within the TCGA and CPTAC training data;
no external validation cohort was used for training, cutpoint selection, or tuning.
Validation-cohort expression enters only at scoring, a fixed-basis projection
($Z = \widetilde{W}^\top X_{\text{new}}$) that uses no validation outcome, and
validation-cohort survival outcomes were used only for the final, out-of-sample
performance evaluation.

\subsection*{Adjustment for clinical covariates and tumor purity}

```{r clinical-adj, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
load_result("desurv_clinical_adjustment")
.ca  <- desurv_clinical_adjustment
.caf <- function(m) if (is.null(m)) "---" else sprintf("%.2f (%.2f--%.2f), %s ($n$=%d, %d ev)", m$hr, m$lo, m$hi,
  if (m$p < 0.001) "$<0.001$" else paste0("$", sprintf("%.3f", m$p), "$"), m$n, m$events)
.ord  <- c("PACA_AU_array","PACA_AU_seq","Puleo_array","Dijk","Moffitt_GEO_array","TCGA_PAAD","CPTAC")
.disp <- c(PACA_AU_array="PACA-AU array", PACA_AU_seq="PACA-AU RNA-seq", Puleo_array="Puleo",
           Dijk="Dijk", Moffitt_GEO_array="Moffitt", TCGA_PAAD="TCGA (train)", CPTAC="CPTAC (train)")
.carows <- lapply(.ord, function(k){ r <- .ca[[k]]
  covs <- if (!is.null(r$clinical_adjusted)) paste(r$clinical_adjusted$covars, collapse=", ") else "---"
  data.frame(Cohort=.disp[[k]], Covariates=covs,
             Unadjusted=.caf(r$unadjusted), `Clinical-adjusted`=.caf(r$clinical_adjusted),
             `Purity-adjusted`=.caf(r$purity_adjusted), check.names=FALSE, stringsAsFactors=FALSE) })
.catab <- do.call(rbind, .carows)
knitr::kable(.catab, format="latex", escape=FALSE, booktabs=TRUE,
  caption="Adjusted hazard ratios (per SD of the DeSurv linear predictor) for overall survival, before and after adjustment for standard clinical covariates and, where annotated, tumor purity, fitted separately in each cohort by complete-case analysis (per-cell $n$ and events shown). The Covariates column lists the clinical variables entered (stage, grade, age, sex, nodal status, or resection margin, as available). The DeSurv score remains independently prognostic after clinical adjustment in the fully-annotated external PACA-AU cohorts and both training cohorts; where tumor-purity annotation was available, it also remained prognostic after purity adjustment in the adequately-powered cohorts (both PACA-AU cohorts and TCGA), consistent with the association not being driven solely by purity or composition. The smaller Dijk and Moffitt cohorts were individually underpowered (non-significant even unadjusted; the Moffitt purity-adjusted model, $n = 29$, likewise). Training-cohort (TCGA, CPTAC) hazard ratios are in-sample. \\label{tab:clinical-adjustment}") |>
  kableExtra::kable_styling(latex_options=c("HOLD_position","scale_down"), font_size=8)
```

\clearpage

\newpage

\part{Supplementary Results}

<!-- Supplementary numbering convention: all SI figures labeled "Fig. S1", "Fig. S2", etc.
     (not "Figure 1"). Tables: "Table S1", etc. Internal scenario code names
     (R0_easy, R00_null, R_mixed) are replaced with descriptive labels (prognostic,
     null, mixed) consistent with main-text terminology. -->
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo    = FALSE,
                      fig.path = "figures/",
                      dev     = "png",
                      dpi     = 300)
source("paper/load_precomputed.R")
library(dplyr)
library(ggplot2)
library(cowplot)
if (!exists("set_fig_font", inherits = TRUE)) {
  set_fig_font <- function(plot_obj, size = 10) {
    if (inherits(plot_obj, "ggplot")) {
      plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
    } else {
      plot_obj
    }
  }
}
```


# The DeSurv model shows consistent convergence and stability across restarts

Supplementary Theorem 1 establishes convergence to a stationary point under $\lambda_H > 0$. The reported analyses used $\lambda_H = 0$, so the theorem does not formally apply to the fitted models; convergence was therefore assessed empirically across multiple initializations and was consistently observed (Fig.\ \ref{fig:fig-converge}). The software implementation also supports an optional ridge penalty on the gene programs ($\lambda_W$), which was likewise set to zero in all reported analyses. In the PDAC training cohort, the DeSurv objective converged to numerically stable solutions within the designated iteration budget across random multi-start initializations. Fig. \ref{fig:fig-converge} shows representative loss trajectories demonstrating monotone decreases until convergence across restarts.


```{r fig-converge, fig.width = 5, fig.height= 3, fig.cap="Convergence trajectories of the DeSurv objective across five randomly selected multi-start initializations on the PDAC training cohort (TCGA + CPTAC). Each line shows the relative loss (loss at iteration $t$ divided by the initial loss for that restart) as a function of iteration number, restricted to the first 5{,}000 iterations for readability. The total objective decreases monotonically and stabilizes well within the iteration budget across all initializations, confirming convergence of the alternating block-update scheme.", fig.env='figure*', fig.pos='t', out.height= "3in", out.width='5in', fig.align='center'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_converge_tcgacptac.pdf", density = 300))

```


# Simulation results under null and mixed scenarios

To characterize the operating characteristics of DeSurv across different signal regimes, we evaluated performance under three simulation scenarios. In the prognostic scenario (main text Fig. \ref{fig:sim}), prognostic programs explained low variance relative to outcome-neutral background signals, and survival depended on 150 factor-specific marker genes ($\beta_1 = 2$, $\beta_2 = \beta_3 = 0$). In the null scenario, survival times were generated independently of gene expression ($\beta = 0$), so no prognostic structure existed. In the mixed scenario, survival depended on 300 genes per lethal factor, of which 50% were factor-specific markers and 50% were background genes with loadings shared across all factors. All scenarios used $G = 3{,}000$ genes, $N = 200$ samples, $K = 3$ true factors, and identical gamma distribution parameters for the gene loading matrix $W$ (see Materials and Methods). Performance was evaluated using cross-validated concordance index (C-index), precision of recovered prognostic gene sets, and the distribution of selected ranks across simulation replicates.

```{r si-null-alpha, echo=FALSE}
load_result("sim_figs_by_scenario")
sids <- sapply(sim_figs_by_scenario, `[[`, "scenario_id")
aids <- sapply(sim_figs_by_scenario, `[[`, "analysis_id")
null_i <- which(sids == "R00_null" & aids == "bo_tune_ntop")
null_alpha <- with(sim_figs_by_scenario[[null_i]]$k_hist$data, alpha[method == "DeSurv"])
null_alpha_n <- length(null_alpha); null_alpha_median <- median(null_alpha)
null_alpha_q1 <- quantile(null_alpha, .25); null_alpha_q3 <- quantile(null_alpha, .75)
null_alpha_min <- min(null_alpha); null_alpha_max <- max(null_alpha)
```

Under the null scenario (Fig. \ref{fig:sim-null-mixed}A--B), DeSurv showed no advantage over unsupervised NMF ($\alpha = 0$). Both methods yielded C-index values near 0.5, consistent with the absence of survival signal. The distribution of selected ranks was concentrated at low values for both methods, with NMF more strongly favoring $k = 2$. These results confirm that DeSurv does not recover spurious prognostic structure when none exists. Because the survival objective is uninformative under the null, the cross-validated concordance is essentially flat in $\alpha$, so the Bayesian-optimized supervision strength is unidentified and disperses across its admissible range rather than concentrating near zero (median $\alpha = `r round(null_alpha_median, 2)`$, IQR $`r round(null_alpha_q1, 2)`$--$`r round(null_alpha_q3, 2)`$; range $`r round(null_alpha_min, 2)`$--$`r round(null_alpha_max, 2)`$ across the `r null_alpha_n` replicates). The specificity guarantee therefore comes not from the optimizer driving $\alpha$ to zero, but from the flat concordance surface itself: across the full range of selected $\alpha$, DeSurv attains C-index $\approx 0.5$, matching unsupervised NMF and confirming that the advantages observed in the prognostic scenario reflect genuine signal recovery rather than overfitting to noise in the survival labels.

Under the mixed scenario (Fig. \ref{fig:sim-null-mixed}C--F), three mechanistic observations characterize DeSurv's advantage. First, when precision is evaluated against the factor-specific marker genes alone (150 genes; Fig. \ref{fig:sim-null-mixed}D left), DeSurv substantially outperforms NMF---matching or exceeding prognostic-scenario levels---even though overall survival-gene precision (300-gene survival set including background genes; Fig. \ref{fig:sim-null-mixed}D right) is more modestly improved. This dissociation reveals that background-gene recovery is similar between methods, and DeSurv's edge comes specifically from concentrating marker programs. Second, the learned factor most aligned with the true marker program receives a substantially larger Cox coefficient ($|\beta|$) under DeSurv than under NMF (Fig. \ref{fig:sim-null-mixed}E), confirming that survival supervision preferentially amplifies the marker-specific signal. Third, C-index values are modestly higher for DeSurv (Fig. \ref{fig:sim-null-mixed}C), consistent with the moderate overall improvement when background-gene dilution attenuates the survival signal. Together, these results demonstrate that DeSurv's advantage in the mixed scenario is mechanistic: the method specifically recovers marker programs and assigns them elevated survival weights, rather than incidental capture of background-gene prognosis.

Together, the three scenarios provide a graded pattern: DeSurv's advantage is greatest when variance and prognosis diverge (prognostic), moderate when they partially overlap (mixed), and absent when no survival signal exists (null). The mechanistic basis for the attenuation in the mixed scenario is now clear: DeSurv continues to specifically recover marker programs (high marker precision, elevated $|\beta|$), but the overall survival-gene precision improvement is dampened because background-gene dilution in the 300-gene survival signal reduces the discrimination advantage. This gradient offers practical guidance for users: DeSurv is most beneficial in settings where the dominant transcriptional axes are expected to be prognostically neutral, as is common in cancers with high tumor purity variation or dominant tissue-composition signals.

```{r fig-sim-null-mixed, fig.width = 6.5, fig.height = 7, fig.cap = "DeSurv performance under null and mixed simulation scenarios. (A--B) Null scenario ($\\beta = 0$): both methods yield C-index $\\approx 0.5$ (A) and comparable rank distributions (B), confirming DeSurv does not recover spurious prognostic structure; precision is omitted as no true survival genes exist. (C--F) Mixed scenario: survival depends on factor-specific marker genes (50\\%) and shared background genes (50\\%). (C) C-index is modestly higher for DeSurv. (D) Precision by gene-set definition: marker-only precision (150 genes; left facet) substantially favors DeSurv; full 300-gene survival-set precision (right facet) shows a more modest improvement. (E) The matched factor receives a substantially larger Cox coefficient $|\\beta|$ under DeSurv, confirming preferential amplification of marker-specific signal. (F) Selected-rank distributions. DeSurv (blue); NMF (red, $\\alpha = 0$); both Bayesian-optimized. \\label{fig:sim-null-mixed}", fig.env='figure*', fig.pos='t', out.height = "6.5in", out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_sim_null_mixed_tcgacptac.pdf", density = 300))
```


# Cross-validated C-index across factorization rank

To evaluate how the number of latent factors $k$ affects prognostic performance, as a sensitivity analysis we performed an exhaustive grid search over $k \in \{2, 3, \ldots, 12\}$ and supervision strength $\alpha \in \{0, 0.05, \ldots, 0.95\}$ using 5-fold cross-validation on the TCGA+CPTAC training cohort. For each $k$, the best-performing $\alpha$ was selected by maximizing the mean CV C-index. The unsupervised NMF baseline ($\alpha = 0$) was evaluated at the same $k$ values for comparison. Results are shown for two gene-selection settings: $n_{\text{top}} = 270$ top genes per factor and $n_{\text{top}} = \text{ALL}$ (all genes retained), both at $\lambda = 0.349$ and $\xi = 0.056$. External validation C-index is shown descriptively; validation outcomes were not used to select $k$, $\alpha$, $\lambda$, $\xi$, or $n_{\text{top}}$.

Fig. \ref{fig:fig-cindex-by-k} shows the resulting C-index curves for both training (CV) and external validation (pooled across held-out cohorts), with shaded ribbons indicating $\pm 1$ standard error. In training, DeSurv consistently outperforms NMF across all $k$, confirming that survival supervision improves in-sample concordance. In external validation, DeSurv at $k = 3$ achieves concordance comparable to NMF at higher ranks (e.g., $k = 7$; Table \ref{tab:tab-cindex-comparison}), demonstrating that survival supervision concentrates prognostic information into fewer factors rather than requiring additional rank to capture it.

```{r fig-cindex-by-k, fig.width = 10, fig.height = 10, fig.cap="Training (CV) and external validation C-index as a function of factorization rank $k$ for DeSurv (blue) and standard NMF (red, $\\alpha = 0$). Shaded ribbons show $\\pm 1$ SE. For each $k$, the best $\\alpha$ was selected by maximizing the mean CV C-index. Validation C-index is pooled across all external cohorts. (A) Results using $n_{\\text{top}} = 270$ top genes per factor ($\\lambda = 0.349$, $\\xi = 0.056$). (B) Results retaining all genes ($n_{\\text{top}} = \\text{ALL}$; $\\lambda = 0.349$, $\\xi = 0.056$).", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
# This figure is produced by the cv_grid pipeline (separate store);
# read directly from the file output.
# Use magick to render the PDF as a raster (avoids LaTeX path issues)
# NOTE: DL flagged duplicate legend across panels A and B. The legend is baked
# into the pre-rendered PDF; consolidating it requires regenerating the source
# figure in inst/cv_grid_validation_analysis_amber_optimal.R on HPC.
img <- magick::image_read("figures/cv_cindex_by_k_primary.pdf", density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

# Standard NMF rank selection heuristics

Standard unsupervised rank selection criteria (reconstruction residuals, cophenetic correlation coefficient, and mean silhouette width) yielded inconsistent guidance for selecting the factorization rank $k$ in the PDAC training cohort (TCGA + CPTAC). Reconstruction residuals decreased smoothly without a clear elbow (Fig. \ref{fig:nmf-diagnostics}A). Cophenetic correlation fluctuated without a distinct transition point (Fig. \ref{fig:nmf-diagnostics}B). Mean silhouette width favored very low ranks that conflicted with other criteria (Fig. \ref{fig:nmf-diagnostics}C). These contradictions motivated the survival-supervised rank selection approach described in the main text.

```{r fig-nmf-diagnostics, fig.width = 6, fig.height = 3.5, fig.cap = "Standard unsupervised NMF rank selection heuristics for the PDAC training cohort (TCGA + CPTAC). (A) Reconstruction residuals as a function of factorization rank $k$. (B) Cophenetic correlation coefficient as a function of $k$. (C) Mean silhouette width across multiple distance metrics as a function of $k$. These criteria yield inconsistent guidance, motivating the survival-supervised model-selection approach described in the main text. \\label{fig:nmf-diagnostics}", fig.env='figure*', fig.pos='t', out.height= "3.5in", out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_diagnostics_tcgacptac.pdf", density = 300))
```

# Cutpoint selection and validation of dichotomized risk groups

To convert the continuous DeSurv linear predictor into a clinically interpretable binary risk stratification, we selected an optimal z-score cutpoint via cross-validation on the TCGA+CPTAC training cohort. For each candidate cutpoint in $\{-2.0, -1.8, \ldots, 2.0\}$, we computed the mean absolute log-rank z-statistic across held-out folds and selected the cutpoint maximizing this metric (Fig. \ref{fig:fig-cutpoint-km}A). We then applied this cutpoint to external validation cohorts by standardizing each patient's linear predictor using the training mean and standard deviation. Fig. \ref{fig:fig-cutpoint-km}B shows the resulting Kaplan--Meier curves for the pooled external validation analysis, with a two-sided log-rank test on the pooled samples; the hazard ratio is from a Cox model on the pooled samples. Panels C--F display per-cohort validation KM curves for Dijk, Moffitt, PACA-AU (array and RNA-seq subsets pooled), and Puleo, respectively. Across all cohorts, the DeSurv-derived high-risk group showed directionally shorter survival across cohorts, confirming that the training-derived cutpoint generalizes to independent datasets.

```{r fig-cutpoint-km, fig.width = 10, fig.height = 11, fig.cap = "Cutpoint selection and external validation of dichotomized risk groups. (A) Cross-validated mean absolute log-rank z-statistic as a function of z-score cutpoint; red dashed line indicates the selected optimum. (B) Pooled validation Kaplan--Meier curves (log-rank test). (C--D) Per-cohort validation KM curves: (C) Dijk, (D) Moffitt. (E--F) Per-cohort validation KM curves: (E) PACA-AU (array and RNA-seq subsets pooled), (F) Puleo.", fig.env='figure*', fig.pos='t', out.height = "8in", out.width='\\textwidth', echo=FALSE}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_cutpoint_km_tcgacptac.pdf", density = 300))
```

# Overlap of DeSurv risk groups with known molecular subtypes

To assess whether the DeSurv-derived risk stratification captures biologically meaningful transcriptional programs, we examined the overlap between dichotomized risk groups (High vs Low) and two established PDAC molecular classifiers: PurIST (basal-like vs classical) and DeCAF (proCAF vs restCAF). Fig. \ref{fig:fig-subtype-overlap} shows the composition of each risk group across all pooled external validation cohorts, with Fisher's exact test P-values quantifying the association. The DeSurv high-risk group was significantly enriched for both basal-like (PurIST) and proCAF (DeCAF) subtypes, consistent with the established higher-risk biology of basal-like tumors and proCAF-enriched stroma. By contrast, standard NMF risk groups at the matched rank ($k = 3$) separated basal-like from classical tumors (PurIST) but showed no significant enrichment for proCAF versus restCAF (DeCAF), indicating that survival-unsupervised factorization does not recover the tumor--stroma axis resolved by the DeSurv linear predictor. This concordance supports that DeSurv's survival-driven factorization recovers clinically relevant biology without requiring subtype labels during training.

```{r fig-subtype-overlap, fig.width = 8, fig.height = 10, fig.cap = "Subtype composition of DeSurv (top) and standard NMF $k=3$ (bottom) risk groups in pooled external validation cohorts. Stacked bar charts show the proportion of PurIST (left) and DeCAF (right) subtypes within Low and High risk groups, among samples with available classifier calls, defined by the training-derived cross-validated cutpoint. Numbers indicate sample counts; Fisher's exact test P-values are shown above each panel. DeSurv risk groups are enriched for both basal-like (PurIST) and proCAF (DeCAF) poor-prognosis subtypes; NMF risk groups separate basal-like from classical tumors but show no significant CAF subtype enrichment.", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_subtype_overlap_tcgacptac.pdf", density = 300))
```

<!-- ## Analysis of scRNA-seq {.unnumbered} -->
<!-- Next, we verify our findings at the cellular level using the Elyada scRNA-seq data [@elyada2019cross]. We found that our DeSurv factor 1 signature was expressed primarily in iCAF and B cells, factor 2 was expressed in Acinar and Classical PDAC 2, and factor 3 was expressed primarily in basal-like PDAC cells (Figure \ref{fig:fig-sc}B-E). This is consistent with our ORA analysis which found that factor 1 was mostly immune and factor 3 was basal-like, and with the overlap we saw with published gene lists that identified factor 1 as iCAF/classical/immune and factor 3 as basal-like. Interestingly, we do not see a large overlap of factor 1 with the classical PDAC cell types despite some overlap with classical gene lists. Need more here... -->

<!-- ```{r fig-sc, fig.cap="A. Cell type clusters from the Elyada scRNA-seq data. B-D. VAM scores for the DeSurv factor signatures in each cell shown on the UMAP of cells in the Elyada-sc data. E. Heatmap of average VAM scores by cell type and factor signature in the Elyada-sc data.", fig.env='figure*', fig.pos='t', out.height= "8in", out.width='\\textwidth'} -->

<!-- tar_load(fig_sc_plot_tcgacptac) -->
<!-- fig_sc_plot_tcgacptac -->
<!-- ``` -->

# Standard NMF factor structure at independently selected rank

Throughout the main text, both DeSurv and standard NMF were fit at the same rank ($k = 3$), enabling a controlled comparison of the effect of survival supervision on factor structure. When each method selects its own rank independently, a residual-elbow heuristic suggested $k = 5$, whereas Bayesian optimization of the unsupervised ($\alpha=0$) baseline selected $k = 7$, while supervised DeSurv BO selects $k = 3$. We examine here the gene program structure at $k = 5$ (a matched-rank sensitivity fit) for both DeSurv (Fig. \ref{fig:fig-desurv-k5-heatmap}) and standard NMF (Fig. \ref{fig:fig-nmf-k5-heatmap}), and the NMF $k = 7$ structure to illustrate additional fragmentation when supervision is absent (Fig. \ref{fig:fig-nmf-k7-heatmap}).

At elbow-selected $k = 5$, supervised and unsupervised factorizations yield substantially different gene-program structures. DeSurv at $k = 5$ (Fig. \ref{fig:fig-desurv-k5-heatmap}) splits the ECM/activated-stroma (proCAF-associated) signal into two factors with opposing tumor correlations (one positively correlated with classical tumor programs, one negatively correlated) and isolates restCAF as a separate factor; exocrine and basal-like programs are redistributed across additional factors at $k = 5$ rather than forming the compact three-program organization selected by BO. Standard NMF at $k = 5$ (Fig. \ref{fig:fig-nmf-k5-heatmap}), by contrast, recovers the three core programs from its $k = 3$ solution and adds two factors capturing Exocrine-like and restCAF variance, transcriptional variance not aligned with survival outcomes. These biologically redundant NMF factors explain why supervised BO selects $k = 3$ in the production model: the additional NMF factors at $k = 5$ contribute no independent prognostic content beyond the three core programs (Fig. \ref{fig:varsurvival-k5}), even though the overall higher-rank model still validates externally on per-cohort C-index (Table \ref{tab:tab-cindex-comparison}).

```{r fig-desurv-k5-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene program overlap heatmap for DeSurv at elbow-selected rank $k = 5$. Each cell shows the number of top genes from a given factor (rows) overlapping established PDAC gene programs (columns). DeSurv at $k = 5$ splits the ECM/activated-stroma (proCAF-associated) signal into two factors with opposing classical-tumor correlations and isolates restCAF as a separate factor; exocrine and basal-like programs are not represented as dominant factors. \\label{fig:fig-desurv-k5-heatmap}", fig.env='figure', fig.pos='t'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_desurv_k5_heatmap_tcgacptac.pdf", density = 300))
```

```{r fig-nmf-k5-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene program overlap heatmap for standard NMF at elbow-selected rank $k = 5$. Each cell shows the number of top genes from a given factor (rows) overlapping established PDAC gene programs (columns). The three core programs from $k = 3$ are recovered; two additional factors overlap Exocrine-like and restCAF signatures, capturing transcriptional variance not aligned with survival outcomes. \\label{fig:fig-nmf-k5-heatmap}", fig.env='figure', fig.pos='t'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_k5_heatmap_tcgacptac.pdf", density = 300))
```

At BO-selected $k = 7$ under $\alpha = 0$ (Fig. \ref{fig:fig-nmf-k7-heatmap}), standard NMF shows increased fragmentation relative to $k = 5$: multiple factors partially overlap the same PDAC gene signatures, and no additional factor maps cleanly onto a distinct established PDAC program; the added factors fragment existing signatures.

```{r fig-nmf-k7-heatmap, fig.width = 6, fig.height = 5, fig.cap = "Gene program overlap heatmap for standard NMF at $k = 7$ (BO-selected rank with $\\alpha = 0$). Compared with $k = 3$ and $k = 5$, the 7-factor solution shows increased fragmentation: multiple factors partially capture the same PDAC gene signatures and no new coherent biological program emerges. \\label{fig:fig-nmf-k7-heatmap}", fig.env='figure', fig.pos='t'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_nmf_k7_heatmap_tcgacptac.pdf", density = 300))
```

```{r tab-cindex-comparison, echo=FALSE, results='asis'}
library(dplyr)
library(knitr)
library(kableExtra)
load_result("val_cindex_desurv_tcgacptac")
load_result("val_cindex_std_desurvk_tcgacptac")
load_result("val_cindex_std_elbowk_tcgacptac")
load_result("val_cindex_desurv_alpha0_tcgacptac")

fmt_cindex <- function(tbl, col_name) {
  tbl %>%
    dplyr::select(dataset, cindex) %>%
    dplyr::rename(!!col_name := cindex)
}

cindex_wide <- fmt_cindex(val_cindex_desurv_tcgacptac, "DeSurv (K=3)") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_desurvk_tcgacptac, "Std NMF (K=3)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_std_elbowk_tcgacptac, "Std NMF (K=5, elbow)"), by = "dataset") %>%
  dplyr::left_join(fmt_cindex(val_cindex_desurv_alpha0_tcgacptac, "Std NMF (K=7, BO $\\alpha$=0)"), by = "dataset") %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))) %>%
  dplyr::rename(Cohort = dataset) %>%
  dplyr::mutate(Cohort = gsub("_", "\\_", Cohort, fixed = TRUE))

# Explicit \label is the canonical referenced name (main paper + SI \ref it);
# push knitr's auto chunk-label to a throwaway name so it neither collides with
# the explicit one nor is silently relied upon in its place.
knitr::opts_current$set(label = "tab-cindex-comparison-auto")
kable(cindex_wide, format = "latex", booktabs = TRUE, linesep = "", escape = FALSE,
      caption = "External validation C-index by method and cohort. DeSurv at BO-selected rank ($k = 3$) is compared with standard NMF at matched rank ($k = 3$), elbow-selected rank ($k = 5$), and BO-selected rank ($k = 7$, $\\alpha = 0$). Higher values indicate better discrimination. \\label{tab:tab-cindex-comparison}") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

Standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ both substantially improve over NMF at $k = 3$ across all cohorts (Table \ref{tab:tab-cindex-comparison}), confirming that additional unsupervised factors capture prognostically relevant variation missed at the matched rank. This external C-index gain reflects a more fragmented multivariate representation of the existing prognostic signal rather than independent Type III survival content in the added factors (Fig. \ref{fig:varsurvival-k5}): the added factors are individually negligible in the per-factor decomposition even as the higher-rank predictor validates externally. At $k = 7$, NMF matches or exceeds DeSurv's per-cohort C-index in most cohorts, demonstrating that unsupervised factorization can approach supervised concordance given enough factors. However, this comes at the cost of a less parsimonious and less interpretable factorization: the $k = 7$ NMF solution shows increased fragmentation (Fig. \ref{fig:fig-nmf-k7-heatmap}). DeSurv's advantage thus combines biological concordance and methodological complementarity: its risk groups identify the same poor-prognosis subtypes as established classifiers (basal-like, proCAF; Fig. \ref{fig:fig-subtype-overlap}), while its linear predictor retains significant prognostic value after adjustment for those classifiers (main-text Table \ref{tab:hr-adjusted}).

Table \ref{tab:tab-bic-percohort} provides a complementary perspective that isolates the prognostic information carried by each method's factor decomposition from the transfer quality of its trained linear predictor. Whereas main-text Table \ref{tab:hr-adjusted} freezes the training-derived $\beta$ coefficients into a single risk-score covariate, Table \ref{tab:tab-bic-percohort} re-estimates $\beta$ on each validation cohort using the individual factor scores (F1, F2, F3 for DeSurv; F1--F7 for NMF) as separate covariates and reports the Bayesian Information Criterion (BIC) of the resulting Cox proportional hazards fits. BIC uses $k \log n_{\text{events}}$ as the complexity penalty (Volinsky and Raftery, \textit{Biometrics} 52, 203--223, 1996; Kass and Raftery, \textit{JASA} 90, 773--795, 1995). DeSurv's parsimonious 3-factor decomposition is BIC-preferred over NMF's 7-factor decomposition in three of five validation cohorts (Dijk, Moffitt, PACA-AU array), with the remaining two cohorts essentially tied: NMF's four additional factors do not provide sufficient log-likelihood improvement to overcome their complexity penalty. This comparison is descriptive and was not used for model selection or the primary external-validation claims, because the Cox coefficients are re-estimated using validation-cohort outcomes.

```{r tab-bic-percohort, echo=FALSE, results='asis'}
library(survival); library(knitr); library(kableExtra)

extract_factor_df <- function(val_latent) {
  pool_dfs <- lapply(val_latent, function(entry) {
    surv <- entry$survival
    if (is.null(surv) || !all(c("time","event") %in% names(surv))) return(NULL)
    if (is.null(entry$latent)) return(NULL)
    k <- ncol(entry$latent)
    df <- data.frame(time = surv$time, event = surv$event, dataset = entry$dataset)
    for (j in seq_len(k)) df[[paste0("F",j)]] <- entry$latent[, j]
    df
  })
  df <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
  df[is.finite(df$time), ]
}

pool_d_bic <- extract_factor_df(val_latent_desurv_tcgacptac)
pool_n_bic <- extract_factor_df(val_latent_desurv_alpha0_tcgacptac)

fit_bic <- function(df, fmla) {
  fit <- coxph(fmla, data = df)
  list(BIC = BIC(fit), n = nrow(df), events = sum(df$event))
}

display_name <- function(ds) {
  nm <- c("Dijk" = "Dijk", "Moffitt_GEO_array" = "Moffitt GEO array",
          "PACA_AU_array" = "PACA-AU array", "PACA_AU_seq" = "PACA-AU RNA-seq",
          "Puleo_array" = "Puleo array")
  if (ds %in% names(nm)) nm[[ds]] else ds
}

rows <- list()
for (ds in unique(pool_d_bic$dataset)) {
  d <- pool_d_bic[pool_d_bic$dataset == ds, ]
  n <- pool_n_bic[pool_n_bic$dataset == ds, ]
  if (sum(d$event) < 5) next
  r_d <- fit_bic(d, Surv(time, event) ~ F1 + F2 + F3)
  r_n <- fit_bic(n, Surv(time, event) ~ F1 + F2 + F3 + F4 + F5 + F6 + F7)
  rows[[ds]] <- data.frame(
    Cohort = display_name(ds), n = r_d$n, events = r_d$events,
    BICd = sprintf("%.1f", r_d$BIC),
    BICn = sprintf("%.1f", r_n$BIC),
    dBIC = sprintf("%+.1f", r_n$BIC - r_d$BIC),
    stringsAsFactors = FALSE)
}

tbl <- dplyr::bind_rows(rows)
colnames(tbl) <- c("Cohort", "$n$", "events",
                    "DeSurv ($k = 3$) BIC", "NMF ($k = 7$) BIC", "$\\Delta$ BIC")

# Explicit \label is the canonical referenced name; auto chunk-label pushed aside.
knitr::opts_current$set(label = "tab-bic-percohort-auto")
kable(tbl, format = "latex", booktabs = TRUE, linesep = "", escape = FALSE,
      align = c("l","r","r","r","r","r"),
      caption = "Per-cohort Bayesian Information Criterion (BIC) for Cox proportional hazards models using individual factor scores as predictors of overall survival in each external validation cohort. $\\beta$ coefficients are re-fit on each validation cohort separately. DeSurv uses three factor scores (F1, F2, F3); NMF at BO-selected $k = 7$ uses seven (F1--F7). BIC penalises complexity as $k \\log n_{\\text{events}}$. Positive $\\Delta$BIC (NMF $-$ DeSurv) indicates DeSurv's parsimonious 3-factor decomposition is preferred. DeSurv is BIC-preferred in three of five validation cohorts; the remaining two are essentially tied. \\label{tab:tab-bic-percohort}") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 9)
```

The reconstruction--survival decomposition at $k = 5$ (Fig. \ref{fig:varsurvival-k5}) confirms that elbow selection does not recover additional prognostic structure. At $k = 5$, standard NMF continues to distribute survival contribution negligibly across all factors; the two factors new relative to $k = 3$ (N4 and N5) contribute minimally, confirming that the additional rank captures background transcriptional variance rather than independent prognostic programs.

```{r fig-varsurvival-k5, fig.width = 4.5, fig.height = 3.5, fig.cap = "Per-factor contribution to reconstruction (normalized reconstruction Shapley share) versus Type III partial likelihood contribution for each standard NMF factor at elbow-selected $k = 5$. Each point represents one factor, labeled N1--N5. Survival contribution is quantified as the reduction in Cox partial log-likelihood when each factor is removed from the full model. The two factors new relative to $k = 3$ (N4, N5) contribute negligibly to survival, confirming that elbow selection recovers no additional prognostic structure. \\label{fig:varsurvival-k5}", fig.env='figure', fig.pos='t'}
cowplot::ggdraw() + cowplot::draw_image(
  magick::image_read("figures/si_fig_varsurvival_k5_tcgacptac.pdf", density = 300))
```

\section{Factor-Specific Gene Lists}

Tables \ref{tab:gene-list-270-D1}--\ref{tab:gene-list-270-D3} list the top 270 genes for each of the three DeSurv factors, ranked by the factor specificity score $s_{ij}$ defined in Supplementary Information, Section \ref{sec:gene-program-correlation}. The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\text{top}} = 270$); these are the genes used to compute sample-level factor scores for survival prediction. Factor D1 is enriched for classical tumor and restCAF-like stromal genes, Factor D2 for proCAF-associated stromal genes, and Factor D3 for basal-like tumor genes.

```{r tab-gene-list-270, echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)
library(tidyr)
library(dplyr)

genes <- read.csv("paper/gene_lists_top270_k3.csv", stringsAsFactors = FALSE)

# Keep only DeSurv factors
desurv_genes <- genes %>%
  dplyr::filter(factor %in% c("D1", "D2", "D3"))

# For each factor, split 270 genes into 6 columns of 45
build_factor_table <- function(df, fct) {
  g <- df %>% dplyr::filter(factor == fct) %>% dplyr::arrange(rank) %>% dplyr::pull(gene)
  data.frame(
    a = g[1:45], b = g[46:90], c = g[91:135],
    d = g[136:180], e = g[181:225], f = g[226:270],
    stringsAsFactors = FALSE
  )
}

factor_labels <- c("D1", "D2", "D3")
factor_captions <- c(
  "Factor D1 (classical tumor + restCAF-like stroma)",
  "Factor D2 (proCAF stroma)",
  "Factor D3 (basal-like tumor)"
)

for (i in seq_along(factor_labels)) {
  tbl <- build_factor_table(desurv_genes, factor_labels[i])
  # Make knitr's auto-generated table label unique per iteration so the three
  # loop tables don't all collide on tab:tab-gene-list-270 (multiply-defined).
  # The referenced per-factor labels (tab:gene-list-270-D<i>) are set in the caption.
  knitr::opts_current$set(label = paste0("tab-gene-list-270-", factor_labels[i]))
  cat("")
  kable(tbl, format = "latex", booktabs = TRUE, linesep = "",
        col.names = c("1--45", "46--90", "91--135", "136--180", "181--225", "226--270"),
        caption = paste0("Top 270 genes for DeSurv factor ", factor_labels[i],
          " (", factor_captions[i], "), ranked by factor specificity score $s_{ij}$.",
          if (i == 1) paste0(" The 270-gene cutoff corresponds to the Bayesian optimization--selected projection dimension ($n_{\\text{top}}$).") else "",
          " \\label{tab:gene-list-270-", factor_labels[i], "}"),
        escape = FALSE) %>%
    kable_styling(font_size = 8, latex_options = c("HOLD_position")) %>%
    add_header_above(setNames(6, factor_captions[i]), bold = TRUE) %>%
    cat()
  cat("")
}
```

\section{Translational analysis in independent bulk treated PDAC cohorts (Rash-Accept and Linehan)}
\label{si:treated}

```{r treated-setup, echo=FALSE, message=FALSE, warning=FALSE}
load_result("treated_cohort_stats")
.tcs  <- treated_cohort_stats
.pfmt <- function(p) if (is.na(p)) "---" else if (p < 0.001) "$<0.001$" else sprintf("%.3f", p)
.os   <- function(coh, prog, mod) { r <- .tcs$os[.tcs$os$cohort==coh & .tcs$os$program==prog & .tcs$os$model==mod, ]
  if (!nrow(r)) return("---"); sprintf("$%.2f$ ($%.2f$--$%.2f$), %s", r$hr, r$lo, r$hi, .pfmt(r$p)) }
.ax   <- function(coh, clf) { r <- .tcs$axis[.tcs$axis$cohort==coh & .tcs$axis$classifier==clf, ]
  if (!nrow(r)) return("---"); sprintf("$%.2f$ ($%.2f$--$%.2f$), %s", r$hr, r$lo, r$hi, .pfmt(r$p)) }
.prev <- function(coh) .tcs$prevalence[.tcs$prevalence$cohort==coh, ]
```

The primary analysis evaluated DeSurv in treatment-naive, non-metastatic PDAC, the setting in which the gene-program matrix $W$ and Cox coefficients $\beta$ were trained. As an exploratory translational test we asked whether the fixed treatment-naive programs retained interpretable prognostic associations in treated PDAC, and specifically whether they remain prognostic after adjustment, where appropriate, for the established PurIST (tumor-intrinsic) and DeCAF (stromal) molecular classifiers --- the same dual adjustment applied to the treatment-naive validation cohorts in the main paper (Table \ref{tab:hr-adjusted}). We used three bulk-tissue treated cohorts spanning two disease settings: the two metastatic gemcitabine-backbone trials Rash ($n = `r .prev("Rash")$n`$, gemcitabine + erlotinib) and Accept ($n = `r .prev("Accept")$n`$, gemcitabine $\pm$ afatinib) --- together the Rash-Accept series ($n = `r .prev("Rash")$n + .prev("Accept")$n`$, bulk RNA-seq, first-line chemotherapy) --- and Linehan ($n = `r .prev("Linehan")$n`$ borderline-resectable / locally advanced PDAC, bulk biopsy; FOLFIRINOX $\pm$ the CCR2 inhibitor PF-04136309). All are bulk tissue, so PurIST and DeCAF are both computable (unlike microdissected cohorts, where the stromal classifier cannot be applied), although adjusted estimates are interpreted cautiously given limited sample size and confounding. Because the cohorts differ in stage and regimen, they are analysed per cohort rather than pooled (below; Fig. \ref{fig:treated}). The elastic-net penalty shrank the trained survival coefficient for the proCAF program $D_2$ to exactly zero in the treatment-naive model (an estimated selection outcome, not an imposed constraint), so $D_2$ contributes nothing to the trained risk score; any association of $D_2$ in the treated cohorts therefore reflects a marginal re-evaluation of the projected program in a new clinical context rather than a property of the trained predictor.

```{r fig-treated, fig.cap = "DeSurv programs show context-dependent prognostic associations in treated PDAC. Trained DeSurv programs were projected without retraining onto three bulk treated cohorts: Linehan, borderline-resectable/locally advanced disease ($n=66$); Rash, metastatic disease ($n=85$); and Accept, metastatic disease ($n=70$). (A) Per-cohort overall-survival hazard ratios per SD for each program; D1 and D3 are shown marginally, D2 adjusted for PurIST and DeCAF. (B) Overall-survival hazard ratios for the classical-versus-basal contrast (PurIST, Moffitt and Bailey classifiers) in less-advanced versus metastatic disease. (C) The proCAF program D2 before and after treatment in paired Linehan biopsies (paired Wilcoxon test). (D) The D1 score versus GATA6 RNA in situ hybridization in COMPASS samples (Spearman correlation). Error bars denote 95\\% confidence intervals. \\label{fig:treated}", fig.env='figure*', fig.pos='t', fig.width=7.2, fig.height=5, out.width='\\textwidth'}
cowplot::ggdraw() + cowplot::draw_image(magick::image_read("figures/fig_treated_context.pdf", density = 300))
```

\subsection{Per-cohort overall survival (no pooling)}

Factor scores were computed by projection using the same trained, top-gene-truncated $\widetilde{W}$ ($n_{\text{top}} = `r .tcs$ntop`$ per factor, union across factors) and within-sample rank convention as the main-paper validation cohorts ($Z_{\text{new}} = \widetilde{W}^\top X_{\text{new}}$), z-standardized within cohort; results were near-identical under a full-gene projection (Spearman $r \geq 0.86$; see Robustness below). Because the three treated cohorts differ in disease setting and regimen (Linehan borderline-resectable/locally-advanced on FOLFIRINOX~$\pm$~CCR2 inhibitor; Rash and Accept metastatic on gemcitabine backbones), and the basal--classical program $D_1$ is heterogeneous across them, we report each cohort separately rather than as a pooled estimate. $D_1$ is reported \emph{marginally}: because $D_1$ tracks the basal--classical contrast and is collinear with the PurIST classifier (it separates PurIST calls at AUC $`r sprintf("%.2f", min(.tcs$d1_auc, na.rm=TRUE))`$--$`r sprintf("%.2f", max(.tcs$d1_auc, na.rm=TRUE))`$), adjusting $D_1$ for PurIST conditions the same axis on itself and over-adjusts; the marginal estimate is the appropriate measure of its prognostic value. $D_2$ (a stromal program, not strongly collinear with the classifiers) is reported both marginal and PurIST + DeCAF-adjusted.

\begin{table}[H]
\centering
\caption{Per-cohort overall-survival hazard ratios per SD in the three bulk treated PDAC cohorts. $D_1$ marginal; $D_2$ marginal and PurIST + DeCAF-adjusted; $D_3$ marginal. Sample sizes: Linehan $n=`r .prev("Linehan")$n`$ (`r .prev("Linehan")$events` events), Rash $n=`r .prev("Rash")$n`$ (`r .prev("Rash")$events` events), Accept $n=`r .prev("Accept")$n`$ (`r .prev("Accept")$events` events); basal-like prevalence Linehan `r .prev("Linehan")$pct_basal`\% (`r .prev("Linehan")$n_basal` cases), Rash `r .prev("Rash")$pct_basal`\%, Accept `r .prev("Accept")$pct_basal`\%. \label{tab:treated-os}}
\begin{tabular}{lcccc}
\toprule
Cohort (setting) & $D_1$ (marginal) & $D_2$ (marginal) & $D_2$ (+PurIST+DeCAF) & $D_3$ (marginal) \\
\midrule
Linehan (BR/LA)     & `r .os("Linehan","D1","marginal")` & `r .os("Linehan","D2","marginal")` & `r .os("Linehan","D2","adjusted")` & `r .os("Linehan","D3","marginal")` \\
Rash (metastatic)   & `r .os("Rash","D1","marginal")`    & `r .os("Rash","D2","marginal")`    & `r .os("Rash","D2","adjusted")`    & `r .os("Rash","D3","marginal")` \\
Accept (metastatic) & `r .os("Accept","D1","marginal")`  & `r .os("Accept","D2","marginal")`  & `r .os("Accept","D2","adjusted")`  & `r .os("Accept","D3","marginal")` \\
\bottomrule
\end{tabular}
\end{table}

The proCAF stromal program $D_2$ was protective in all three cohorts (both marginally and after adjustment, with significance reached in Linehan), with a consistent direction across the independent settings and chemotherapy backbones; it was significant in Linehan and non-significant but directionally concordant in the metastatic cohorts (Accept and Rash). Because $D_2$ is stromal and not strongly collinear with the classifiers, its association was not an artifact of adjustment: in Linehan it was significant both before and after adjustment, and across all three cohorts the adjusted estimate strengthened rather than attenuated relative to the marginal one, indicating that the $D_2$ association was not explained by PurIST or DeCAF adjustment. The basal--classical program $D_1$ was prognostic in the less-advanced Linehan cohort (marginal HR `r sub(",.*","",.os("Linehan","D1","marginal"))`) but not in the metastatic cohorts --- a setting-dependence we show below is a property of the basal--classical axis itself (shared by PurIST, Moffitt, and Bailey), not a failure of $D_1$. $D_3$ (basal-like) was not associated with overall survival in any treated cohort, consistent with the basal--classical contrast residing in $D_1$.

\subsection{The basal--classical axis is setting-dependent (classifier triangulation)}

That $D_1$ attenuates in metastatic disease reflects the basal--classical prognostic axis itself rather than $D_1$ specifically: every established basal--classical classifier shows the same pattern (Table \ref{tab:axis}). In the less-advanced Linehan cohort the axis is strongly prognostic (PurIST classical-vs-basal HR `r sub(",.*","",.ax("Linehan","PurIST"))`), whereas in the metastatic cohorts PurIST, Moffitt, and Bailey are all non-significant. Basal-like prevalence in the metastatic cohorts is the expected `r {m <- c(.prev("Rash")$pct_basal, .prev("Accept")$pct_basal); sprintf("%g--%g", min(m), max(m))}`\% (not range-restricted; the Linehan estimate rests on only `r .prev("Linehan")$n_basal` basal-like cases and is correspondingly imprecise), and the same axis is prognostic in treatment-naive validation cohorts, supporting disease-setting dependence rather than reduced power alone; larger real-world cohorts have likewise reported adverse outcomes for the basal subtype in metastatic disease [@singh2024basal], although stage, treatment and cohort remain confounded.

\begin{table}[H]
\centering
\caption{Classical-vs-basal overall-survival hazard ratios for established subtype classifiers, by cohort. PurIST is strongly prognostic in the less-advanced Linehan cohort; all three classifiers attenuate in metastatic disease (Rash, Accept). \label{tab:axis}}
\begin{tabular}{lccc}
\toprule
Classifier & Linehan (BR/LA) & Rash (metastatic) & Accept (metastatic) \\
\midrule
PurIST  & `r .ax("Linehan","PurIST")` & `r .ax("Rash","PurIST")`  & `r .ax("Accept","PurIST")` \\
Moffitt & ---                          & `r .ax("Rash","Moffitt")` & `r .ax("Accept","Moffitt")` \\
Bailey (squamous) & ---                & `r .ax("Rash","Bailey")`  & `r .ax("Accept","Bailey")` \\
\bottomrule
\end{tabular}
\end{table}

\subsection{D2 is associated with progression-free survival and increases after treatment}

In the metastatic Rash-Accept cohort, $D_2$ showed a nominal association with longer progression-free survival (not significant after Benjamini--Hochberg correction) (HR per SD `r {r<-.tcs$pfs[.tcs$pfs$program=="D2",]; sprintf("$%.2f$ ($%.2f$--$%.2f$), %s", r$hr, r$lo, r$hi, .pfmt(r$p))}`); $D_3$ was associated with shorter progression-free survival (significant after Benjamini--Hochberg correction) (`r {r<-.tcs$pfs[.tcs$pfs$program=="D3",]; sprintf("HR $%.2f$ ($%.2f$--$%.2f$), %s", r$hr, r$lo, r$hi, .pfmt(r$p))}`) --- a single-cohort, single-endpoint observation consistent with the aggressive biology of basal-like tumors. In the `r .tcs$paired$n` Linehan patients with paired pre- and post-treatment biopsies, $D_2$ increased significantly from pre- to post-treatment (paired Wilcoxon `r .pfmt(.tcs$paired$wilcox_p)`; mean $\Delta = `r sprintf("%+.2f", .tcs$paired$mean_delta)`$ SD), consistent with treatment-associated remodeling of the proCAF program.

\subsection{D2 molecular sub-components (exploratory)}

Partitioning $D_2$'s projection additively into gene-group contributions and testing each against overall survival (cohort-stratified; sub-scores are correlated, so the joint attribution is interpreted cautiously) showed the protective association was strongest for the neural-reactive/axon-guidance component (Table \ref{tab:decomp}): the neural-reactive component is protective both alone and in the joint model that controls for collinearity, whereas the extracellular-matrix-core component is protective alone but absorbed jointly, and the CD163-associated macrophage-like component is non-protective alone and, after joint adjustment, significantly adverse (`r {m<-.tcs$decomp[.tcs$decomp$component=="Macrophage" & .tcs$decomp$model=="joint",]; sprintf("HR %.2f, $P = %.3f$", m$hr, m$p)}`), consistent with reports that cancer-associated-fibroblast--driven macrophage reprogramming is tumor-promoting [@hussain2024cafmacrophage; @raymant2024macrophagefibroblast].

```{r treated-decomp, echo=FALSE, results='asis'}
dc <- .tcs$decomp
rowfmt <- function(comp){ a<-dc[dc$component==comp & dc$model=="alone",]; j<-dc[dc$component==comp & dc$model=="joint",]
  sprintf("%s & $%.2f$ ($%.2f$--$%.2f$), %s & $%.2f$ ($%.2f$--$%.2f$), %s \\\\",
          comp, a$hr,a$lo,a$hi,.pfmt(a$p), j$hr,j$lo,j$hi,.pfmt(j$p)) }
cat("\\begin{table}[H]\\centering\n",
    "\\caption{Exploratory additive decomposition of the $D_2$ projection into gene-group sub-scores; per-SD overall-survival HR (cohort-stratified, PurIST + DeCAF-adjusted), each sub-score alone and all jointly. \\label{tab:decomp}}\n",
    "\\begin{tabular}{lcc}\n\\toprule\n",
    "$D_2$ sub-component & Alone HR/SD (95\\% CI), $P$ & Joint HR/SD (95\\% CI), $P$ \\\\\n\\midrule\n",
    rowfmt("Neural"), "\n", rowfmt("ECMcore"), "\n", rowfmt("Macrophage"), "\n",
    "\\bottomrule\n\\end{tabular}\n\\end{table}\n", sep="")
```

\subsection{Orthogonal validation of program identity}

Independently of survival, the $D_1$ program score correlated with GATA6 RNA in situ hybridization in the COMPASS cohort (Spearman $\rho = `r sprintf("%.2f", .tcs$gata6$rho)`$, `r .pfmt(.tcs$gata6$p)`, $n = `r .tcs$gata6$n`$), corroborating the classical/GATA6 identity of $D_1$ beyond the signature-level correlations in the main-text factor--signature heatmap.

\subsection{Robustness to projection convention}

The treated-cohort scores were computed with the top-gene $\widetilde{W}$ used in primary validation. Recomputing them with the full trained $W$ (all `r nrow(readRDS("results/tar_fit_desurv_tcgacptac.rds")$W)` genes) gives near-identical per-sample scores (pooled Spearman correlation $D_1$ $`r sprintf("%.2f", .tcs$concordance["D1"])`$, $D_2$ $`r sprintf("%.2f", .tcs$concordance["D2"])`$, $D_3$ $`r sprintf("%.2f", .tcs$concordance["D3"])`$) and left the qualitative per-cohort conclusions unchanged, so the treated findings do not depend on the projection convention. We report results per cohort and do not pool: the protective direction of $D_2$ is consistent across all three cohorts, which is the relevant cross-cohort evidence in the absence of a common-effect assumption. The exploratory $D_2$ sub-component decomposition is the exception: it uses a cohort-stratified model (pooled across cohorts with cohort as a stratum) to improve stability and is interpreted descriptively.

\subsection{Caveats}

This analysis is exploratory and prognostic, not predictive (treatment is confounded with cohort and stage, and biopsies are predominantly pre-treatment). (i) The cohorts are analysed per cohort and not pooled, because their settings differ (metastatic Rash/Accept vs borderline-resectable/locally-advanced Linehan) and $D_1$ is heterogeneous across them; the consistent protective direction of $D_2$ across all three cohorts is the cross-cohort evidence. (ii) $D_1$ is reported marginally because it is collinear with the PurIST classifier, so adjusting it for PurIST over-adjusts the same basal--classical axis. (iii) The treated cohorts are modestly sized (Linehan $n = `r .prev("Linehan")$n`$, `r .prev("Linehan")$events` events; the metastatic Rash and Accept cohorts $n = `r .prev("Rash")$n`$ and $`r .prev("Accept")$n`$) so per-cohort $D_2$ estimates are individually under-powered, although the protective direction is consistent across all three. (iv) No analysis plan was preregistered. (v) The different behavior of $D_2$ in treated versus treatment-naive settings is the central context-dependent observation (Discussion); the sub-component decomposition and the established restraining-CAF literature provide candidate explanations, but mechanism is not established here. Robust treated-cohort validation will require larger, harmonized bulk-tissue series, ideally in the neoadjuvant setting where treated and resectable disease coincide.
